---
title: "Hour 1 — EDA: Analyzing horror movie profits in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 037 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`boxplot_jitter` · `coord_flip_horizontal` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_histogram_density` · `geom_smooth_loess` · `grouped_summary_ggplot` · `interactive_plotly` · `log_scale` · `lubridate_dates` · `pivot_reshape`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(tidyverse)
library(lubridate)
library(scales)
library(plotly)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_02}
movie_profit_raw <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018-10-23/movie_profit.csv")
```
```{r chunk_03}
# Data cleaning
movie_profit <- movie_profit_raw |>
  select(-X1) |>
  mutate(release_date = as.Date(parse_date_time(release_date, "%m!/%d/%Y"))) |>
  filter(release_date < "2018-01-01") |>
  arrange(desc(row_number())) |>
  distinct(movie, release_date, .keep_all = TRUE) |>
  mutate(distributor = fct_lump(distributor, 5)) |>
  filter(worldwide_gross > 0) |>
  mutate(profit_ratio = worldwide_gross / production_budget,
         decade = 10 * floor(year(release_date) / 10))
```
```{r chunk_04}
movie_profit |>
  count(distributor, sort = TRUE)

movie_profit |>
  ggplot(aes(production_budget)) +
  geom_histogram() +
  scale_x_log10(labels = dollar_format())

movie_profit |>
  ggplot(aes(distributor, production_budget)) +
  geom_boxplot() +
  scale_y_log10(labels = dollar_format()) +
  coord_flip()
```
```{r chunk_05}
movie_profit |>
  ggplot(aes(distributor, worldwide_gross)) +
  geom_boxplot() +
  scale_y_log10(labels = dollar_format()) +
  coord_flip()
```

## What genres make the most money?
```{r chunk_06}
movie_profit |>
  mutate(genre = fct_reorder(genre, production_budget)) |>
  filter(!is.na(distributor)) |>
  ggplot(aes(genre, production_budget)) +
  geom_boxplot() +
  scale_y_log10(labels = dollar_format()) +
  coord_flip() +
  facet_wrap(~ distributor)
```
```{r chunk_07}
movie_profit |>
  mutate(genre = fct_reorder(genre, worldwide_gross)) |>
  filter(!is.na(distributor)) |>
  ggplot(aes(genre, worldwide_gross)) +
  geom_boxplot() +
  scale_y_log10(labels = dollar_format()) +
  coord_flip() +
  facet_wrap(~ distributor)
```
```{r chunk_08}
movie_profit |>
  arrange(desc(profit_ratio)) |>
  head(20) |>
  mutate(movie = fct_reorder(movie, profit_ratio)) |>
  ggplot(aes(movie, profit_ratio, fill = genre)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = function(x) paste0(x, "X")) +
  labs(x = "",
       y = "Ratio of worldwide gross to production budget",
       title = "What movies have most outgrossed their budget?",
       fill = "Genre")
```

## What are the most common genres over time?
```{r chunk_09}
movie_profit |>
  count(decade, genre) |>
  group_by(decade) |>
  mutate(percent = n / sum(n)) |>
  ggplot(aes(decade, percent, color = genre)) +
  geom_line() +
  scale_y_continuous(labels = percent_format())
```
```{r chunk_10}
movie_profit |>
  filter(!is.na(distributor)) |>
  count(distributor, genre) |>
  ggplot(aes(genre, n, fill = genre)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ distributor, scales = "free_x") +
  coord_flip()
```

## Which genres have the biggest payoff?
```{r chunk_11}
movie_profit |>
  group_by(genre) |>
  summarize(median_profit_ratio = median(profit_ratio)) |>
  arrange(desc(median_profit_ratio)) |>
  mutate(genre = fct_reorder(genre, median_profit_ratio)) |>
  ggplot(aes(genre, median_profit_ratio)) +
  geom_col() +
  scale_y_continuous(labels = function(x) paste0(x, "X")) +
  coord_flip()
```
```{r chunk_12}
movie_profit |>
  group_by(genre, year = year(release_date)) |>
  summarize(median_profit_ratio = median(profit_ratio),
            movies = n()) |>
  ungroup() |>
  filter(year >= 2000) |>
  arrange(desc(median_profit_ratio)) |>
  mutate(genre = fct_reorder(genre, median_profit_ratio)) |>
  ggplot(aes(year, median_profit_ratio, color = genre)) +
  geom_line() +
  scale_y_continuous(labels = function(x) paste0(x, "X"))
```
```{r horror_movies}
horror_movies <- movie_profit |>
  filter(genre == "Horror") |>
  arrange(desc(profit_ratio))
```
```{r chunk_13}
horror_movies |>
  head(20) |>
  mutate(movie = paste0(movie, " (", year(release_date), ")"),
         movie = fct_reorder(movie, profit_ratio)) |>
  ggplot(aes(movie, profit_ratio, fill = distributor)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = function(x) paste0(x, "X")) +
  labs(x = "",
       y = "Ratio of worldwide gross to production budget",
       title = "What horror movies have most outgrossed their budget?")
```
```{r chunk_14}
horror_movies |>
  filter(release_date >= "1990-01-01",
         profit_ratio >= .01) |>
  ggplot(aes(release_date, profit_ratio)) +
  geom_point() +
  geom_smooth(method = "lm") +
  geom_text(aes(label = movie), vjust = 1, hjust = 1, check_overlap = TRUE) +
  scale_y_log10(labels = function(x) paste0(x, "X"), breaks = c(.1, 1, 10, 100))
```
```{r chunk_15}
g <- movie_profit |>
  filter(release_date >= "1990-01-01",
         profit_ratio >= .01) |>
  ggplot(aes(release_date, profit_ratio, label = movie)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_y_log10(labels = function(x) paste0(x, "X"), breaks = c(.1, 1, 10, 100)) +
  facet_wrap(~ genre)

library(plotly)

ggplotly(g)
```
```{r chunk_16}
movie_profit |>
  group_by(genre, distributor, decade) |>
  summarize(median_profit_ratio = median(profit_ratio),
            movies = n()) |>
  ungroup() |>
  filter(decade >= 1990,
         !is.na(distributor)) |>
  arrange(movies)
  mutate(genre = fct_reorder(genre, median_profit_ratio)) |>
  ggplot(aes(decade, median_profit_ratio, color = genre)) +
  geom_line() +
  facet_wrap(~ distributor) +
  scale_y_continuous(labels = function(x) paste0(x, "X"))
```

## What are typical budgets over time?
```{r chunk_17}
movie_profit |>
  group_by(decade) |>
  summarize_at(vars(production_budget:worldwide_gross), median, na.rm = TRUE) |>
  gather(metric, value, -decade) |>
  ggplot(aes(decade, value, color = metric)) +
  geom_line() +
  scale_y_continuous(labels = dollar_format())
```

---

**→ [Hour 2: Modeling](h2_modeling_horror_movie_profits.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
