---
title: "Hour 1 — EDA: Analyzing X-Men comics in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 079 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`error_bar_coef_plot` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `grouped_summary_ggplot` · `label_repulsion_ggrepel` · `log_scale` · `multi_table_join` · `spline_regression` · `str_detect_filter` · `theme_set_global` · `tidytuesdayr_load`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(ggrepel)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
character_visualization <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-06-30/character_visualization.csv", show_col_types = FALSE)
characters <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-06-30/characters.csv", show_col_types = FALSE)
comic_bechdel <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-06-30/comic_bechdel.csv", show_col_types = FALSE)
covers <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-06-30/covers.csv", show_col_types = FALSE)
issue_collaborators <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-06-30/issue_collaborators.csv", show_col_types = FALSE)
locations <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-06-30/locations.csv", show_col_types = FALSE)
xmen_bechdel <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-06-30/xmen_bechdel.csv", show_col_types = FALSE)
glimpse(character_visualization)
glimpse(characters)
```

## Explore
```{r chunk_02}
character_visualization <- xmen$character_visualization |>
  separate(character, c("superhero", "secret_identity"), sep = " = ", fill = "right")
```
```{r chunk_03}
character_visualization |>
  group_by(superhero)
```
```{r chunk_04}
by_character |>
  left_join(characters, by = c(superhero = "name")) |>

characters |>
  filter(str_detect(name, "Angel")) |>
  distinct(name)
```
```{r chunk_05}
library(ggrepel)

by_character <- character_visualization |>
  group_by(superhero) |>
  summarize(across(speech:depicted,
                   list(total = sum,
                        issues = ~ sum(. > 0),
                        avg = ~ mean(.[depicted > 0]))))

by_character |>
  arrange(desc(speech_issues)) |>
  filter(speech_issues > 0) |>
  mutate(superhero = fct_reorder(superhero, speech_issues)) |>
  ggplot(aes(speech_issues, superhero)) +
  geom_col() +
  labs(title = "Which X-Men speak in the most issues?",
       x = "# of issues",
       y = "")

by_character |>
  ggplot(aes(depicted_issues, speech_avg)) +
  geom_point() +
  geom_text_repel(aes(label = superhero)) +
  labs(x = "# of issues in which character appears",
       y = "Average lines per issue depicted") +
  expand_limits(y = 0)

by_character |>
  ggplot(aes(depicted_issues, thought_avg)) +
  geom_point() +
  geom_text_repel(aes(label = superhero)) +
  scale_x_log10() +
  labs(x = "# of issues in which character appears",
       y = "Average thoughts per issue") +
  expand_limits(y = 0)

by_character |>
  ggplot(aes(depicted_issues, speech_avg / thought_avg)) +
  geom_point() +
  geom_text_repel(aes(label = superhero)) +
  expand_limits(y = 1) +
  scale_y_log10(breaks = c(1, 3, 10, 30),
                labels = c("Same", "3X", "10X", "30X")) +
  labs(x = "# of issues depicted",
       y = "Speech to thought ratio (log scale)",
       title = "Which X-Men think, and which speak?")
```
```{r chunk_06}
by_character_costume <- character_visualization |>
  group_by(superhero, costume) |>
  summarize(across(speech:depicted,
                   list(total = sum,
                        issues = ~ sum(. > 0),
                        avg = ~ mean(.[depicted > 0])))) |>
  ungroup() |>
  mutate(speech_thought_ratio = speech_avg / thought_avg)

by_character_costume |>
  mutate(superhero = fct_reorder(superhero, speech_total)) |>
  filter(speech_avg > 0) |>
  ggplot(aes(speech_total,
             superhero,
             color = costume,
             size = depicted_issues)) +
  geom_point() +
  labs(x = "# of lines (in/out of costume)")

costume_ratios <- by_character_costume |>
  filter(speech_avg > 0) |>
  group_by(superhero) |>
  summarize(costume_ratio = speech_total[1] / speech_total[2]) |>
  inner_join(by_character, by = "superhero")

costume_ratios |>
  ggplot(aes(speech_total,
             costume_ratio)) +
  geom_point() +
  geom_text_repel(aes(label = superhero)) +
  scale_y_log10() +
  labs(x = "Total # of lines",
       y = "Lines in costume / lines out of costume")

costume_ratios |>
  mutate(superhero = fct_reorder(superhero, costume_ratio)) |>
  ggplot(aes(costume_ratio, y = superhero)) +
  geom_errorbarh(aes(xmin = 1, xmax = costume_ratio), height = 0) +
  geom_point(aes(size = speech_total, color = costume_ratio > 1)) +
  scale_x_log10() +
  scale_color_discrete(guide = FALSE) +
  labs(size ="# of lines",
       x = "Lines in costume / lines out of costume",
       y = "",
       title = "Which X-Men tend to speak in costume/out of costume?")
```
```{r chunk_07}
xmen$locations |>
  group_by(location) |>
  summarize(total = n(),
            issues = n_distinct(issue)) |>
  arrange(desc(issues))
```
```{r chunk_08}
xmen$comic_bechdel |>
  bind_rows(xmen$xmen_bechdel |> mutate(writer = "Chris Claremont")) |>
  filter(!is.na(pass_bechdel),
         !is.na(writer)) |>
  group_by(writer = fct_lump(writer, 5)) |>
  summarize(n_issues = n(),
            pct_bechdel = mean(pass_bechdel == "yes")) |>
  arrange(desc(n_issues))
```
```{r chunk_09}
xmen$xmen_bechdel |>
  filter(!is.na(pass_bechdel)) |>
  group_by(issue_group = 20 * (issue %/% 20)) |>
  summarize(pct_bechdel = mean(pass_bechdel == "yes"),
            n = n()) |>
  ggplot(aes(issue_group, pct_bechdel)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent)
```
```{r chunk_10}
by_issue_character <- character_visualization |>
  group_by(issue_group = 20 * (issue %/% 20), superhero) |>
  summarize(across(speech:depicted, sum)) |>
  ungroup()

by_issue_character |>
  filter(fct_lump(superhero, 9, w = speech) != "Other") |>
  mutate(superhero = fct_reorder(superhero, -speech, sum)) |>
  ggplot(aes(issue_group, speech)) +
  geom_col() +
  facet_wrap(~ superhero)
```

---

**→ [Hour 2: Modeling](h2_modeling_x_men_comics.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
