---
title: "Hour 1 — EDA: Analyzing Maryland bridges with R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 041 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`coord_flip_horizontal` · `error_bar_coef_plot` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_histogram_density` · `grouped_summary_ggplot` · `linear_model_broom` · `log_scale` · `logistic_regression` · `scales_axis_labels` · `spline_regression`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
:::

```{r setup}
library(here)
library(tidyverse)
library(scales)
library(broom)
library(splines)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
library(scales)
theme_set(theme_light())

maryland_bridges <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018-11-27/baltimore_bridges.csv") |>
  select(-vehicles) |>
  mutate(inspection_yr = inspection_yr + 2000,
         decade = 10 * (yr_built %/% 10),
         responsibility = fct_lump(responsibility, 4),
         county = str_to_title(county))
```
```{r chunk_02}
maryland_bridges |>
  filter(yr_built >= 1900) |>
  count(decade = 10 * (yr_built %/% 10), sort = TRUE) |>
  ggplot(aes(decade, n)) +
  geom_line() +
  expand_limits(y = 0) +
  labs(y = "# of bridges in Baltimore built this decade")
```
```{r chunk_03}
maryland_bridges |>
  filter(yr_built >= 1900) |>
  group_by(decade) |>
  summarize(pct_good = mean(bridge_condition == "Good"),
            total = n()) |>
  ggplot(aes(decade, pct_good)) +
  geom_line() +
  scale_y_continuous(labels = percent_format()) +
  expand_limits(y = 0)
```
```{r chunk_04}
maryland_bridges |>
  replace_na(list(responsibility = "Other")) |>
  count(responsibility = fct_lump(responsibility, 4), sort = TRUE) |>
  mutate(responsibility = fct_reorder(responsibility, n)) |>
  ggplot(aes(responsibility, n)) +
  geom_col() +
  coord_flip()
```
```{r chunk_05}
maryland_bridges |>
  filter(yr_built >= 1900) |>
  group_by(responsibility = fct_lump(responsibility, 4),
           decade) |>
  summarize(pct_good = mean(bridge_condition == "Good"),
            total = n()) |>
  filter(responsibility != "Other") |>
  ggplot(aes(decade, pct_good, color = responsibility)) +
  geom_line() +
  scale_y_continuous(labels = percent_format()) +
  expand_limits(y = 0) +
  labs(y = "% of bridges rated 'Good'")
```
```{r chunk_06}
maryland_bridges |>
  ggplot(aes(avg_daily_traffic)) +
  geom_histogram() +
  scale_x_log10(labels = comma_format())
```
```{r chunk_07}
maryland_bridges |>
  filter(yr_built >= 1990) |>
  group_by(traffic_category = cut(avg_daily_traffic, c(0, 1000, 10000, Inf),
                                  labels = c("<1000", "1000-10,000", "10,000+"))) |>
  summarize(pct_good = mean(bridge_condition == "Good"),
            total = n())
```
```{r chunk_08}
maryland_bridges |>
  ggplot(aes(long, lat, color = avg_daily_traffic)) +
  borders("state", regions = "Maryland") +
  geom_point() +
  scale_color_gradient2(low = "blue",
                        high = "red",
                        midpoint = log10(median(maryland_bridges$avg_daily_traffic)),
                        trans = "log10",
                        labels = comma_format()) +
  coord_map() +
  theme_void()
```
```{r chunk_09}
maryland_bridges |>
  ggplot(aes(long, lat, color = bridge_condition)) +
  borders("state", regions = "Maryland") +
  geom_point(size = 1) +
  coord_map() +
  theme_void()
```
```{r chunk_10}
maryland_bridges |>
  filter(yr_built >= 1900) |>
  ggplot(aes(long, lat, color = county)) +
  borders("state", regions = "Maryland") +
  geom_point(size = 1) +
  coord_map() +
  theme_void()
```
```{r chunk_11}
maryland_bridges |>
  filter(yr_built >= 1900) |>
  group_by(county, decade) |>
  summarize(pct_good = mean(bridge_condition == "Good"),
            total = n()) |>
  arrange(county, decade) |>
  ggplot(aes(decade, pct_good, color = county)) +
  geom_line() +
  scale_y_continuous(labels = percent_format()) +
  expand_limits(y = 0) +
  labs(y = "% of bridges rated 'Good'")
```

## Effect of county, responsibility, traffic and time on bridge condition
```{r chunk_12}
# fit a logistic model
bridges <- maryland_bridges |>
  filter(yr_built >= 1900)

library(broom)
library(splines)

simple_model <- bridges |>
  mutate(good = bridge_condition == "Good") |>
  glm(good ~ ns(yr_built, 4), data = ., family = "binomial")

model <- bridges |>
  mutate(good = bridge_condition == "Good") |>
  glm(good ~ ns(yr_built, 4) + responsibility + county, data = ., family = "binomial")

augment(simple_model, bridges, type.predict = "response") |>
  ggplot(aes(yr_built, .fitted)) +
  geom_line() +
  expand_limits(y = 0) +
  scale_y_continuous(labels = percent_format()) +
  labs(y = "Predicted probability a bridge is rated 'Good'")

augment(model, bridges, type.predict = "response") |>
  ggplot(aes(yr_built, .fitted, color = responsibility)) +
  geom_line() +
  expand_limits(y = 0) +
  facet_wrap(~ county) +
  scale_y_continuous(labels = percent_format()) +
  labs(y = "Predicted probability a bridge is rated 'Good'")
```
```{r chunk_13}
model |>
  tidy(conf.int = TRUE) |>
  filter(str_detect(term, "responsibility|county")) |>
  mutate(term = reorder(term, estimate)) |>
  ggplot(aes(estimate, term)) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +
  geom_vline(xintercept = 0, color = "red", lty = 2)
```

---

**→ [Hour 2: Modeling](h2_modeling_maryland_bridges.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
