---
title: "Hour 2 — Modeling: Analyzing Maryland bridges with R"
subtitle: "Going deeper with statistical models | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 041 · Hour 2 · [← Hour 1 EDA](h1_eda_maryland_bridges.qmd)
:::


::: {.callout-note collapse="true"}
## Atomic Concepts — Hour 2
`coord_flip_horizontal` · `error_bar_coef_plot` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_histogram_density` · `grouped_summary_ggplot` · `linear_model_broom` · `log_scale` · `logistic_regression`
:::

## Setup

```{r setup}
library(tidyverse)
library(scales)
library(broom)
library(here)
theme_set(theme_light(base_size = 13))
```

## Reload data

```{r load-data}
# Re-load from Hour 1 — same local paths
source_qmd <- here::here("output/episodes/maryland_bridges/h1_eda_maryland_bridges.qmd")
# Data files are re-read directly below if needed
```

## Model 1 — Linear regression baseline

```{r lm-setup}
# df <- read_csv(...)
glimpse(df)
```

```{r lm-fit}
# Identify your outcome and predictors first — adjust formula below
# fit <- lm(outcome ~ predictor1 + predictor2, data = df)
# tidy(fit) |> filter(term != "(Intercept)") |>
#   mutate(term = fct_reorder(term, estimate)) |>
#   ggplot(aes(estimate, term, xmin = estimate - 2*std.error,
#              xmax = estimate + 2*std.error)) +
#   geom_point() + geom_errorbarh(height = 0.2) +
#   geom_vline(xintercept = 0, lty = 2)
cat("Linear model: adjust formula to your outcome variable\n")
```

## Model 2 — Variable importance

```{r importance}
# augment(fit) |>
#   ggplot(aes(.fitted, outcome)) +
#   geom_point(alpha = 0.4) +
#   geom_abline(color = "red", lty = 2) +
#   labs(title = "Predicted vs actual", x = "Fitted", y = "Observed")
cat("Residual diagnostics: wire up after lm fit above\n")
```


---

**→ [Hour 3: Communicate](h3_communicate_maryland_bridges.qmd)**
