---
title: "Hour 1 — EDA: Analyzing Himalayan climbers in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 035 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`boxplot_jitter` · `case_when_conditional_col` · `empirical_bayes` · `error_bar_coef_plot` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_histogram_density` · `grouped_summary_ggplot` · `linear_model_broom` · `log_scale` · `logistic_regression` · `multi_table_join`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(ebbr)
library(broom)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
expeditions <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-09-22/expeditions.csv", show_col_types = FALSE)
members <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-09-22/members.csv", show_col_types = FALSE)
peaks <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-09-22/peaks.csv", show_col_types = FALSE)
glimpse(expeditions)
glimpse(members)
```

## Load the weekly Data
```{r chunk_01}
expeditions |>
  count(termination_reason, sort = TRUE)

expeditions |>
  filter(!is.na(days_to_highpoint), !is.na(peak_name)) |>
  filter(success == "Success") |>
  mutate(peak_name = fct_lump(peak_name, 10),
         peak_name = fct_reorder(peak_name, days_to_highpoint)) |>
  ggplot(aes(days_to_highpoint, peak_name)) +
  geom_boxplot() +
  labs(x = "Days from basecamp to highpoint",
       y = "",
       title = "How long does it take to get to the high point?",
       subtitle = "Successful climbs only")
```
```{r chunk_02}
summarize_expeditions <- function(tbl) {
  tbl |>
    summarize(n_climbs = n(),
              pct_success = mean(success == "Success"),
              across(members:hired_staff_deaths, sum),
              first_climb = min(year)) |>
    mutate(pct_death = member_deaths / members,
         pct_hired_staff_deaths = hired_staff_deaths / hired_staff)
}

peaks_summarized <- expeditions |>
  group_by(peak_id, peak_name) |>
  summarize_expeditions() |>
  ungroup() |>
  arrange(desc(n_climbs)) |>
  inner_join(peaks |> select(peak_id, height_meters), by = "peak_id")
```
```{r chunk_03}
# devtools::install_github("dgrtwo/ebbr")
library(ebbr)

peaks_eb <- peaks_summarized |>
  filter(members >= 20) |>
  arrange(desc(pct_death)) |>
  add_ebb_estimate(member_deaths, members)

peaks_eb |>
  ggplot(aes(pct_death, .fitted)) +
  geom_point(aes(size = members, color = members)) +
  geom_abline(color = "red") +
  scale_x_continuous(labels = percent) +
  scale_y_continuous(labels = percent) +
  scale_color_continuous(trans = "log10") +
  labs(x = "Death rate (raw)",
       y = "Death rate (empirical Bayes adjusted)")

peaks_eb |>
  filter(members >= 200) |>
  arrange(desc(.fitted)) |>
  mutate(peak_name = fct_reorder(peak_name, .fitted)) |>
  ggplot(aes(.fitted, peak_name)) +
  geom_point(aes(size = members)) +
  geom_errorbarh(aes(xmin = .low, xmax = .high)) +
  expand_limits(x = 0) +
  scale_x_continuous(labels = percent) +
  labs(x = "Death rate (empirical Bayes adjusted + 95% credible interval)",
       y = "",
       title = "How deadly is each peak in the Himalayas?",
       subtitle = "Only peaks that at least 200 climbers have attempted")
```
```{r chunk_04}
# No relationship between height and death rate
peaks_eb |>
  filter(members >= 100) |>
  ggplot(aes(height_meters, .fitted)) +
  geom_point(aes(size = members))
```
```{r chunk_05}
expeditions |>
  filter(peak_name == "Everest") |>
  ggplot(aes(days_to_highpoint, fill = success)) +
  geom_density(alpha = .5)

expeditions |>
  filter(peak_name == "Everest") |>
  filter(success == "Success") |>
  arrange(days_to_highpoint)

expeditions |>
  filter(success == "Success") |>
  ggplot(aes(year)) +
  geom_histogram()

everest_by_decade <- expeditions |>
  filter(peak_name == "Everest") |>
  mutate(decade = pmax(10 * (year %/% 10), 1970)) |>
  group_by(decade) |>
  summarize_expeditions()

everest_by_decade |>
  ggplot(aes(decade, pct_death)) +
  geom_line(aes(color = "All climbers")) +
  geom_line(aes(y = pct_hired_staff_deaths, color = "Hired staff")) +
  geom_point(aes(color = "All climbers", size = members)) +
  geom_point(aes(y = pct_hired_staff_deaths, size = hired_staff, color = "Hired staff")) +
  scale_x_continuous(breaks = seq(1970, 2010, 10),
                     labels = c("< 1980", seq(1980, 2010, 10))) +
  scale_y_continuous(labels = percent) +
  expand_limits(y = 0) +
  labs(x = "Decade",
       y = "Death rate",
       title = "Everest has been getting less deadly over time",
       subtitle = "Though trends have recently reversed for hired staff",
       size = "# of climbers",
       color = "")
```

## Examine death probability per member
```{r chunk_06}
members <- tt$members

library(broom)

everest <- members |>
  filter(peak_name == "Everest")

everest |>
  group_by(age = 10 * (age %/% 10)) |>
  summarize(n_climbers = n(),
            pct_death = mean(died))

everest |>
  group_by(hired) |>
  summarize(n_climbers = n(),
            pct_death = mean(died))

model <- everest |>
  mutate(leader = expedition_role == "Leader") |>
  glm(died ~ year + age + sex + leader + hired + oxygen_used, data = ., family = "binomial")

tidied <- model |>
  tidy(conf.int = TRUE, exponentiate = TRUE)

tidied |>
  filter(term != "(Intercept)") |>
  mutate(term = reorder(term, estimate)) |>
  ggplot(aes(estimate, term)) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high))
```

---

**→ [Hour 2: Modeling](h2_modeling_himalayan_climbers.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
