---
title: "Hour 1 — EDA: Analyzing African-American history in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 002 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`boxplot_jitter` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_histogram_density` · `grouped_summary_ggplot` · `network_graph_ggraph` · `pivot_reshape` · `reorder_within_facets` · `spline_regression` · `theme_set_global` · `tidytuesdayr_load`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- text is multilingual without language detection
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- text is long-form English and benefits from word-level decomposition
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(grid)
library(ggraph)
library(ggwordcloud)
library(tidytext)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
african_names <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-06-16/african_names.csv", show_col_types = FALSE)
blackpast <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-06-16/blackpast.csv", show_col_types = FALSE)
census <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-06-16/census.csv", show_col_types = FALSE)
slave_routes <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-06-16/slave_routes.csv", show_col_types = FALSE)
glimpse(african_names)
glimpse(blackpast)
```

## Explore
```{r chunk_02}
slave_routes <- slave_routes

slave_routes |>
  summarize(sum(n_slaves_arrived, na.rm = TRUE))

slave_routes |>
  count(ship_name, sort = TRUE)

slave_routes |>
  ggplot(aes(n_slaves_arrived)) +
  geom_histogram()
```
```{r chunk_03}
slave_routes |>
  filter(!is.na(ship_name), !is.na(n_slaves_arrived)) |>
  filter(fct_lump(ship_name, 12) != "Other") |>
  mutate(ship_name = fct_reorder(ship_name, n_slaves_arrived)) |>
  ggplot(aes(n_slaves_arrived, ship_name)) +
  geom_boxplot()
```
```{r chunk_04}
slave_routes |>
  count(port_arrival, sort = TRUE)

slave_routes |>
  count(port_origin, sort = TRUE)

slave_routes |>
  ggplot(aes(year_arrival)) +
  geom_histogram() +
  geom_vline(xintercept = c(1776, 1865), lty = 2) +
  labs(title = "# of slave ships over time")
```
```{r chunk_05}
slave_routes |>
  filter(!is.na(port_origin)) |>
  mutate(port_origin = str_trunc(port_origin, 25),
         port_origin = fct_lump(port_origin, 24),
         port_origin = fct_reorder(port_origin, year_arrival)) |>
  count(port_origin,
        decade = 10 * (year_arrival %/% 10)) |>
  ggplot(aes(decade, n)) +
  geom_line() +
  facet_wrap(~ port_origin) +
  geom_vline(xintercept = c(1807), lty = 2) +
  theme(strip.text = element_text(size = 6))

slave_routes |>
  filter(port_origin == "New York") |>
  arrange(desc(year_arrival)) |>
```
```{r chunk_06}
slave_routes |>
  filter(fct_lump(ship_name, 12) != "Other") |>
  count(ship_name, port_origin, port_arrival, sort = TRUE) |>
  group_by(ship_name) |>
  mutate(percent = n / sum(n)) |>
  top_n(1, percent) |>
  arrange(desc(percent))
```
```{r chunk_07}
library(grid)
library(ggraph)

set.seed(2020)

slave_routes |>
  count(port_origin, port_arrival, sort = TRUE) |>
  filter(port_origin != port_arrival) |>
  head(40) |>
  ggraph(layout = "fr") +
  geom_edge_link(aes(width = n),
                 arrow = arrow(type = "closed",
                               length = unit(.1, "inch"))) +
  geom_node_point() +
  geom_node_text(aes(label = name), repel = TRUE) +
  scale_edge_width_continuous(range = c(.5, 4)) +
  labs(title = "Common slave routes",
       edge_width = "# of ships")
```
```{r chunk_08}
slave_routes |>
  group_by(port_origin) |>
  summarize(n_ships = n(),
            total_recorded = sum(n_slaves_arrived, na.rm = TRUE),
            pct_missing = mean(is.na(n_slaves_arrived)),
            estimated_total = mean(n_slaves_arrived, na.rm = TRUE) * n()) |>
  arrange(desc(n_ships))
```
```{r chunk_09}
census <- census

census_gathered <- census |>
  mutate(other = total - white - black) |>
  arrange(year) |>
  gather(racial_category, population, white, black_free, black_slaves, other)

census_gathered |>
  filter(region == "USA Total") |>
  ggplot(aes(year, population, fill = racial_category)) +
  geom_col() +
  scale_y_continuous(labels = comma) +
  labs(x = "Year",
       y = "Census Population",
       fill = "Racial category",
       title = "Census racial makeup of US, 1790-1870",
       subtitle = "No 'other' category existed before 1860")
```
```{r chunk_10}
census_gathered |>
  filter(region != "USA Total", !is.na(division)) |>
  mutate(division = fct_reorder(division, -population, sum)) |>
  ggplot(aes(year, population, fill = racial_category)) +
  geom_col() +
  scale_y_continuous(labels = comma) +
  facet_wrap(~ division) +
  labs(x = "Year",
       y = "Census Population",
       fill = "Racial category",
       title = "Census racial makeup of US, 1790-1870",
       subtitle = "No 'other' category existed before 1860")
```
```{r chunk_11}
census_gathered |>
  filter(region != "USA Total", !is.na(division)) |>
  mutate(division = fct_reorder(division, -population, sum)) |>
  group_by(division, year) |>
  mutate(percent = population / sum(population)) |>
  ggplot(aes(year, percent, fill = racial_category)) +
  geom_col() +
  scale_y_continuous(labels = percent) +
  facet_wrap(~ division) +
  labs(x = "Year",
       y = "% of Census Population",
       fill = "Racial category",
       title = "Census racial makeup of US, 1790-1870",
       subtitle = "No 'other' category existed before 1860")
```
```{r chunk_12}
library(ggwordcloud)

name_counts <- african_names |>
  count(name, gender, sort = TRUE)

wordcloud(name_counts$name, name_counts$n)

name_counts |>
  head(100) |>
  ggplot(aes(label = name, size = n, color = gender)) +
  geom_text_wordcloud()
```
```{r chunk_13}
library(tidytext)

african_names |>
  filter(!is.na(gender)) |>
  mutate(gender = fct_recode(gender, Man = "Boy", Woman = "Girl")) |>
  count(name, gender, sort = TRUE) |>
  group_by(gender) |>
  top_n(20, n) |>
  ungroup() |>
  mutate(name = reorder_within(name, n, gender)) |>
  ggplot(aes(n, name)) +
  geom_col() +
  scale_y_reordered() +
  facet_wrap(~ gender, scales = "free")
```
```{r chunk_14}
african_names |>
  count(country_origin, sort = TRUE)
```

---

**→ [Hour 2: Modeling](h2_modeling_african_american_history.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
