---
title: "Hour 2 — Modeling: Analyzing car fuel efficiency in R"
subtitle: "Go deeper with statistics and models | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode: `car_fuel_efficiency` | [All episodes](https://thedatashrink.com.au/tidytuesday) | [YouTube](#) | [Instagram](#)
:::


::: {.callout-note collapse="true"}
## Atomic Concepts (DataX Index)
These reusable concepts are demonstrated in this episode. See `CONCEPT_INDEX.md`.

- `cross_validation_tidymodels`
- `grouped_summary_ggplot`
- `pivot_reshape`
- `time_series_trend`
:::


::: {.callout-warning collapse="true"}
## FAILS WHEN (DataX Stage 5)
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN (DataX Stage 5)
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- time dimension is regular and spans multiple periods with meaningful change
:::


## Episode Context

- **Analysis type:** `Modeling` — this shapes the modeling approach below
- **Pattern:** `load_inspect_reshape_trend_model_evaluate_visualize`
- **Hour 2 goal:** Go deeper than David went in the original screencast

## Hour 2 Checklist
- [ ] Engineer features from EDA findings
- [ ] Split with initial_split(), create v-fold CV
- [ ] Fit baseline model (linear/logistic)
- [ ] Tune LASSO / random forest / xgboost
- [ ] Compare models with collect_metrics()
- [ ] Interpret with vip / SHAP values
- [ ] Residual analysis + calibration plot


## Feature Engineering

```{r}
#| label: features
#| eval: false
# Build features from EDA findings
# Use recipes::recipe() to encode categoricals and impute
```

## Train / Test Split

```{r}
#| label: split
#| eval: false
library(tidymodels)
set.seed(42)
split <- initial_split(data, prop = 0.8, strata = outcome)
train <- training(split)
test  <- testing(split)
folds <- vfold_cv(train, v = 5, strata = outcome)
```

## Baseline Model

```{r}
#| label: baseline
#| eval: false
# Fit a simple linear/logistic model as the baseline
# Compare to null model AUC / RMSE
```

## Tuned Model (LASSO / XGBoost)

```{r}
#| label: tuned-model
#| eval: false
# Use tune_grid() or tune_bayes() to find optimal hyperparameters
```

## Model Interpretation

```{r}
#| label: vip
#| eval: false
library(vip)
# Variable importance plot
# SHAP values if tree model
```


## Model Story

> What does the model reveal that raw EDA couldn't?
> One paragraph for the YouTube "deep dive" section.

*[Write model narrative here]*

## Cross-Reference

- [← Hour 1 EDA](h1_eda_car_fuel_efficiency.qmd)
- [Hour 3 Communicate →](h3_communicate_car_fuel_efficiency.qmd)
