---
title: "Hour 1 — EDA: Analyzing bird collisions with bootstrapping in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 010 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`bootstrap_ci` · `confidence_ribbon` · `coord_flip_horizontal` · `crossing_simulation_grid` · `faceted_comparison` · `fct_reorder_stat` · `geom_histogram_density` · `grouped_summary_ggplot` · `log_scale` · `lubridate_dates` · `multi_table_join` · `pivot_reshape`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(lubridate)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## R Markdown
```{r chunk_01}
library(tidyverse)
theme_set(theme_light())

mp_light <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-04-30/mp_light.csv")

bird_collisions <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-04-30/bird_collisions.csv") |>
  left_join(mp_light, by = "date")
```
```{r chunk_02}
bird_collisions |>
  ggplot(aes(date, fill = locality)) +
  geom_histogram()
```
```{r chunk_03}
bird_collisions |>
  gather(category, value, -date, -light_score) |>
  count(category, value, light_score_missing = is.na(light_score)) |>
  group_by(category) |>
  top_n(16, n) |>
  ungroup() |>
  mutate(value = fct_reorder(value, n, sum),
         category = fct_reorder(category, n, length)) |>
  ggplot(aes(value, n, fill = light_score_missing)) +
  geom_col() +
  facet_wrap(~ category, scales = "free_y") +
  coord_flip() +
  labs(x = "# of collisions",
       y = "",
       fill = "Light score missing",
       title = "Category breakdowns of collisions")
```
```{r chunk_04}
bird_collisions |>
  filter(!is.na(light_score)) |>
  count(date, locality) |>
  ggplot(aes(n, color = locality)) +
  geom_density() +
  scale_x_log10() +
  labs(x = "# of collisions per night")

bird_collisions |>
  filter(!is.na(light_score)) |>
  distinct(date, light_score) |>
  ggplot(aes(light_score)) +
  geom_histogram()

geom_mean <- function(x) {
  exp(mean(log(x + 1)) - 1)
}

by_day_mp <- bird_collisions |>
  filter(!is.na(light_score)) |>
  group_by(date, locality) |>
  summarize(collisions = n()) |>
  ungroup() |>
  complete(date, locality, fill = list(collisions = 0)) |>
  right_join(mp_light |> crossing(locality = c("CHI", "MP")), by = c("date", "locality")) |>
  filter(date <= "2016-11-13") |>
  replace_na(list(collisions = 0)) |>
  mutate(locality = ifelse(locality == "CHI", "Greater Chicago", "McCormick Place"))

bootstrap_cis <- by_day_mp |>
  bootstraps(times = 1000) |>
  unnest(map(splits, as.data.frame)) |>
  group_by(light_score, locality, id) |>
  summarize(avg_collisions = geom_mean(collisions)) |>
  summarize(bootstrap_low = quantile(avg_collisions, .025),
            bootstrap_high = quantile(avg_collisions, .975))

by_day_mp |>
  group_by(light_score, locality) |>
  summarize(avg_collisions = geom_mean(collisions),
            nights = n()) |>
  ggplot(aes(light_score, color = locality)) +
  geom_line(aes(y = avg_collisions)) +
  geom_ribbon(aes(ymin = bootstrap_low, ymax = bootstrap_high),
              data = bootstrap_cis,
              alpha = .25) +
  expand_limits(y = 0) +
  labs(x = "Light score at McCormick place (higher means more lights on)",
       y = "Geometric mean of the number of collisions",
       title = "Brighter lights at McCormick place correlate with more bird collisions there, and not with Chicago overall",
       subtitle = "Ribbon shows 95% bootstrapped percentile confidence interval",
       color = "")
```

## Look at confounders
```{r chunk_05}
library(lubridate)

bird_collisions |>
  filter(date >= "2005-01-01") |>
  count(month = month(date, label = TRUE), #floor_date(date, "month"),
        locality) |>
  ggplot(aes(month, n, color = locality, group = locality)) +
  geom_line()
```

---

**→ [Hour 2: Modeling](h2_modeling_bird_collisions_bootstrapping.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
