---
title: "Hour 1 — EDA: Analyzing water access points in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 076 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`animated_transitions` · `countrycode_standardize` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_area_stacked` · `grouped_summary_ggplot` · `lubridate_dates` · `purrr_map_iterate` · `spline_regression` · `theme_set_global` · `tidytuesdayr_load`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(lubridate)
library(ggthemes)
library(gganimate)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Load the weekly Data
```{r Load}
library(lubridate)

tt <- tt_load("2021-05-04")

water <- tt$water |>
  mutate(report_date = mdy(report_date)) |>
  rename(lat = lat_deg,
         lon = lon_deg,
         country = country_name) |>
  separate(water_tech, c("water_tech", "brand"), sep = " - ",
           fill = "right") |>
  mutate(install_year = ifelse(install_year > 2021, NA_real_, install_year)) |>
  filter(!country %in% c("Peru", "Dominican Republic", "Timor-Leste"),
         !is.na(country)) |>
  filter(between(lat, -35, 37),
         between(lon, -40, 60))
```
```{r chunk_01}
water |>
  count(status_id)

water |>
  count(water_tech, sort = TRUE)

water |>
  count(water_source, sort = TRUE)

water |>
  count(water_source, water_tech, sort = TRUE)
```
```{r chunk_02}
water |>
  filter(install_year > 2021) |>

water |>
  filter(install_year > 1980) |>
  count(install_year) |>
  ggplot(aes(install_year, n)) +
  geom_col()

water |>
  count(installer, sort = TRUE) |>
```
```{r chunk_03}
water |>
  count(status_id, status, sort = TRUE)
```

## Maps
```{r chunk_04}
library(ggthemes)

water |>
  group_by(country) |>
  summarize(lat = mean(lat),
            lon = mean(lon)) |>
  ggplot(aes(lon, lat)) +
  geom_point() +
  geom_text(aes(label = country), vjust = 1, hjust = 1)

countries <- unique(water$country)

africa_map_data <- map_data("world") |>
  as_tibble() |>
  mutate(continent = countrycode(region, "country.name", "continent")) |>
  filter(continent == "Africa")

water |>
  sample_n(10000) |>
  ggplot(aes(lon, lat)) +
  geom_polygon(aes(long, lat, group = group),
               color = "gray",
               fill = "white",
               data = africa_map_data,
               size = .25) +
  geom_point(size = .1, alpha = .25) +
  theme_map()

water |>
  count(country, sort = TRUE)

water |>
  filter(country == "Uganda") |>
  sample_n(10000) |>
  ggplot(aes(lon, lat)) +
  geom_polygon(aes(long, lat, group = group),
               color = "gray",
               fill = "white",
               data = africa_map_data,
               size = .25) +
  geom_point(size = .1, alpha = .25) +
  theme_map()
```
```{r chunk_05}
water_uganda <- water |>
  filter(country == "Uganda",
         between(lat, -2, 4),
         between(lon, 29, 40))

water_uganda |>
  #sample_n(20000) |>
  ggplot(aes(lon, lat, color = status_id)) +
  borders("world", regions = "Uganda") +
  geom_point(size = .1, alpha = .25) +
  theme_map() +
  scale_color_discrete(guide = guide_legend(override.aes = list(size = 2, alpha = 1)))
```
```{r chunk_06}
bbox <- c(left = 29.2, bottom = -2, right = 35, top = 4.2)

uganda_map <- get_stamenmap(bbox, zoom = 8)
```
```{r chunk_07}
water_uganda_lumped <- water_uganda |>
  mutate(water_source = fct_lump(water_source, 5)) |>
  replace_na(list(water_source = "Other")) |>
  mutate(water_source = fct_reorder(water_source, water_source, length, .desc = TRUE))

ggmap(uganda_map) +
  geom_point(aes(lon, lat),
             data = water_uganda_lumped, size = .1, alpha = .1) +
  facet_wrap(~ water_source)

water_uganda |>
  count(water_source, sort = TRUE)

water_uganda |>
  count(pay, sort = T)
```
```{r chunk_08}
water_uganda_lumped |>
  mutate(report_year = year(report_date)) |>
  count(report_year, water_source) |>
  complete(report_year, water_source, fill = list(n = 0)) |>
  group_by(report_year) |>
  mutate(year_total = sum(n)) |>
  filter(year_total >= 500) |>
  ggplot(aes(report_year, n / year_total, fill = water_source)) +
  geom_area()

water_uganda_lumped |>
  count(water_tech, sort = TRUE)

water_uganda_lumped |>
  mutate(water_tech = fct_lump(water_tech, 5)) |>
  mutate(report_year = year(report_date)) |>
  count(report_year, water_tech) |>
  complete(report_year, water_tech, fill = list(n = 0)) |>
  group_by(report_year) |>
  mutate(year_total = sum(n)) |>
  filter(year_total >= 500) |>
  ggplot(aes(report_year, n, fill = water_tech)) +
  geom_area()
```
```{r chunk_09}
water_uganda |>
  ggplot(aes(report_date, install_year)) +
  geom_point()
```
```{r chunk_10}
ggmap(uganda_map) +
  geom_point(aes(lon, lat, color = install_year),
             data = water_uganda |> sample_n(20000),
             size = .2) +
  scale_color_gradient2(low = "red", high = "brown",
                        midpoint = 1990)

library(gganimate)

water_uganda |>
  filter(!is.na(install_year)) |>
  sample_n(10000) |>
  mutate(install_year = pmax(1990, install_year)) |>
  mutate(year = map(install_year, ~ seq(., 2021))) |>
  unnest(year) |>
  ggplot(aes(lon, lat)) +
  borders("world", regions = "Uganda") +
  geom_point(size = .1, alpha = .25) +
  theme_map() +
  transition_manual(year) +
  labs(title = "Water sources in Uganda in year: { current_frame }")
```
```{r chunk_11}
point_data <- water_uganda |>
  filter(!is.na(install_year)) |>
  mutate(install_year = pmax(1990, install_year)) |>
  mutate(year = map(install_year, ~ seq(., 2021))) |>
  unnest(year)

ggmap(uganda_map) +
  geom_point(aes(lon, lat), data = point_data, size = .1, alpha = .25) +
  transition_manual(year) +
  labs(title = "Water sources in Uganda in year: { current_frame }")
```
```{r chunk_12}
get_map(c(lon = -95.3632715, lat = 29.7632836),
        source = "osm")
```
```{r chunk_13}
water_uganda |>
```

---

**→ [Hour 2: Modeling](h2_modeling_water_access_points.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
