---
title: "Hour 1 — EDA: Analyzing car fuel efficiency in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 014 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`boxplot_jitter` · `coord_flip_horizontal` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_smooth_loess` · `grouped_summary_ggplot` · `linear_model_broom` · `purrr_map_iterate` · `spline_regression` · `theme_set_global` · `time_series_trend`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(broom)
library(splines)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_02}
big_epa_cars <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-10-15/big_epa_cars.csv") |>
  mutate(uses_electricity = ifelse(highwayE > 0, "Uses Electricity", "Doesn't Use Electricity"))
```
```{r chunk_03}
big_epa_cars_alphabetical <- big_epa_cars |>
  select(sort(colnames(big_epa_cars)))
```
```{r chunk_04}
big_epa_cars |>
  ggplot(aes(highway08, city08)) +
  geom_point() +
  geom_abline(color = "red") +
  facet_wrap(~ uses_electricity, scales = "free") +
  expand_limits(x = 0, y = 0) +
  labs(x = "Highway MPG",
       y = "City MPG",
       title = "How does fuel efficiency differ between city + highway?")
```
```{r chunk_05}
big_epa_cars |>
  select(city08, highway08, make, model, cylinders, displ, drive, engId, eng_dscr)
```
```{r chunk_06}
big_epa_cars |>
  filter(cityE == 0) |>
  mutate(VClass = fct_lump(VClass, 8),
         VClass = fct_reorder(VClass, city08)) |>
  ggplot(aes(VClass, city08)) +
  geom_boxplot() +
  coord_flip()

big_epa_cars |>
  filter(cityE == 0) |>
  mutate(drive = fct_reorder(drive, city08)) |>
  ggplot(aes(drive, city08)) +
  geom_boxplot() +
  coord_flip()

big_epa_cars |>
  filter(cityE == 0) |>
  ggplot(aes(cylinders, city08, group = cylinders)) +
  geom_boxplot()

big_epa_cars |>
  filter(cityE == 0) |>
  ggplot(aes(displ, city08)) +
  geom_point() +
  expand_limits(x = 0, y = 0)
```

## Goal: Predict city fuel efficiency for single-fuel, non-electric cars
```{r chunk_07}
# Cross validation holdout set
non_electric_cars <- big_epa_cars |>
  filter(cityA08 == 0,
         cityE == 0) |>
  sample_frac(1)

training_set <- non_electric_cars |>
  filter(row_number() %% 5 != 0)
```
```{r chunk_08}
library(broom)

training_set |>
  ggplot(aes(displ, city08)) +
  geom_point() +
  geom_smooth(method = "lm")

library(splines)
augmented_data <- lm(city08 ~ ns(displ, 2), data = training_set) |>
  augment(data = training_set)

augmented_data |>
  ggplot(aes(displ, city08)) +
  geom_point() +
  geom_line(aes(y = .fitted), color = "red", size = 2)

models <- tibble(df = 1:10) |>
  mutate(lm_model = map(df, ~ lm(city08 ~ ns(displ, df = .), data = training_set)))

augmented_unnested <- models |>
  mutate(augmented = map(lm_model, augment, data = training_set)) |>
  unnest(augmented)

augmented_unnested |>
  ggplot(aes(displ, city08)) +
  geom_point(data = training_set) +
  geom_line(aes(y = .fitted, color = factor(df)), size = 2) +
  labs(x = "Engine volume (L)",
       y = "City MPG",
       color = "# of degrees of freedom") +
  expand_limits(x = 0, y = 0)

augmented_unnested |>
  ggplot(aes(displ, .resid)) +
  geom_point() +
  facet_wrap(~ df)

glanced_models <- models |>
  rename(spline_df = df) |>
  mutate(glanced = map(lm_model, glance, data = training_set)) |>
  unnest(glanced)

glanced_models |>
  ggplot(aes(spline_df, adj.r.squared)) +
  geom_line()
```
```{r chunk_09}
lm(city08 ~ ns(displ, 4), data = training_set) |>
  anova() |>
  tidy() |>
  mutate(pct_variation = sumsq / sum(sumsq))

ggplot(training_set, aes(cylinders, displ, group = cylinders)) +
  geom_boxplot()
```
```{r chunk_10}
training_set |>
  ggplot(aes(year, city08)) +
  geom_point() +
  geom_smooth(method = "loess")

efficiency_time <- training_set |>
  mutate(VClass = fct_lump(VClass, 6),
         guzzler = !is.na(guzzler)) |>
  group_by(year = 2 * floor(year / 2), VClass) |>
  summarize_at(vars(city08, cylinders, displ, guzzler), mean)

efficiency_time |>
  ggplot(aes(year, city08, color = VClass)) +
  geom_line() +
  expand_limits(y = 0)

efficiency_time |>
  ggplot(aes(year, displ, color = VClass)) +
  geom_line() +
  expand_limits(y = 0)

efficiency_time |>
  ggplot(aes(year, guzzler, color = VClass)) +
  geom_line() +
  expand_limits(y = 0)
```

---

**→ [Hour 2: Modeling](h2_modeling_car_fuel_efficiency.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
