---
title: "Hour 1 — EDA: Analyzing the Kenya census in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 065 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`choropleth_sf` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `grouped_summary_ggplot` · `heatmap_tile` · `janitor_clean` · `log_scale` · `multi_table_join` · `pivot_reshape` · `spline_regression` · `theme_set_global`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- region name spelling differs between data and shapefile — silent dropped rows
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- geographic column can be matched to a standard ISO/FIPS shapefile without fuzzy joins
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(rKenyaCensus)
library(sf)
library(ggthemes)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
crops <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-01-19/crops.csv", show_col_types = FALSE)
gender <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-01-19/gender.csv", show_col_types = FALSE)
households <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-01-19/households.csv", show_col_types = FALSE)
glimpse(crops)
glimpse(gender)
```

## Load the weekly Data
```{r chunk_01}
gender |>
  filter(county != "Total") |>
  mutate(county = fct_reorder(county, total)) |>
  ggplot(aes(total, county)) +
  geom_col() +
  scale_x_continuous(labels = comma)

gender |>
  filter(county != "Total") |>
  gather(gender, population, male, female, intersex) |>
  mutate(gender = str_to_title(gender)) |>
  mutate(county = fct_reorder(county, total, sum)) |>
  ggplot(aes(population, county, fill = gender)) +
  geom_col() +
  scale_x_continuous(labels = comma)

gender |>
  filter(county != "Total") |>
  mutate(pct_male = male / total) |>
  arrange(desc(pct_male)) |>
  ggplot(aes(total, pct_male)) +
  geom_point() +
  geom_text(aes(label = county),
            vjust = 1, hjust = 1) +
  geom_hline(color = "red", yintercept = .5) +
  scale_y_continuous(labels = percent)
```
```{r chunk_02}
households |>
  filter(county != "Kenya") |>
  arrange(desc(average_household_size)) |>
  ggplot(aes(population, average_household_size)) +
  geom_point() +
  geom_text(aes(label = county),
            vjust = 1, hjust = 1) +
  scale_x_log10(labels = comma) +
  expand_limits(y = 0)
  # geom_hline(color = "red", yintercept = .5) +
  scale_y_continuous(labels = percent)
```
```{r chunk_03}
crop_counties <- tt$crops |>
  gather(crop, households, -SubCounty, -Farming) |>
  filter(!is.na(households)) |>
  mutate(county = str_to_title(SubCounty)) |>
  filter(county != "Kenya") |>
  select(-SubCounty) |>
  inner_join(households, by = "county") |>
  mutate(county = fct_reorder(county, households, sum),
         crop = fct_reorder(crop, households, sum)) |>
  complete(crop, county, fill = list(number_of_households = 0))

crop_counties |>
  ggplot(aes(households, county, fill = crop)) +
  geom_col() +
  labs(x = "# of households farming this crop",
       y = "",
       fill = "Crop")

crop_counties |>
  complete(crop, county, fill = list(households = 0)) |>
  ggplot(aes(crop, county, fill = households)) +
  geom_tile() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Crop",
       y = "",
       fill = "# of households")

crop_counties |>
  mutate(pct_households = households / number_of_households) |>
  arrange(desc(pct_households)) |>
  ggplot(aes(number_of_households, pct_households)) +
  geom_point() +
  geom_text(aes(label = county), vjust = 1, hjust = 1,
            check_overlap = TRUE) +
  scale_y_continuous(labels = percent) +
  labs(y = "% of households growing this crop") +
  facet_wrap(~ crop)
```

## The package
```{r chunk_04}
library(rKenyaCensus)
library(sf)
library(ggthemes)

kenya_sf <- rKenyaCensus::KenyaCounties_SHP |>
  st_as_sf() |>
  st_simplify(dTolerance = 200) |>
  mutate(county = str_to_title(County)) |>
  left_join(crop_counties |>
              filter(crop == "Avocado"), by = "county")

kenya_sf |>
  ggplot(aes(fill = number_of_households)) +
  geom_sf() +
  theme_map() +
  labs(fill = "Households growing avocado")
```
```{r chunk_05}
graph_by_county <- function(tbl) {
  tbl |>
    tibble() |>
    gather(category, population, -County) |>
    filter(!is.na(population)) |>
    mutate(category = respace(category)) |>
    mutate(county = str_to_title(County)) |>
    filter(county != "Kenya") |>
    mutate(category = fct_reorder(category, population, sum),
           county = fct_reorder(county, population, sum)) |>
    ggplot(aes(population, county, fill = category)) +
    geom_col()
}

V4_T2.30 |>
  select(-Total) |>
  graph_by_county()

V4_T2.19 |>
  filter(AdminArea == "County") |>
  gather(category, value, MainsElectricity:NotStated) |>
  filter(!is.na(value)) |>
  mutate(value = ConventionalHouseholds / 100 * value) |>
  group_by(County, category = fct_lump(category, 5, w = value)) |>
  summarize(value = sum(value)) |>
  spread(category, value) |>
  graph_by_county()

V4_T2.33 |>
  ungroup() |>
  filter(AdminArea == "County") |>
  mutate(County = str_to_title(County),
         County = fct_reorder(County, UoI_Total_Perc)) |>
  ggplot(aes(UoI_Total_Perc / 100, County)) +
  geom_col() +
  scale_x_continuous(labels = percent)
```

---

**→ [Hour 2: Modeling](h2_modeling_kenya_census.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
