---
title: "Hour 1 — EDA: Analyzing penguins in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 048 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`boxplot_jitter` · `error_bar_coef_plot` · `faceted_comparison` · `fct_lump_infrequent` · `geom_histogram_density` · `grouped_summary_ggplot` · `pivot_reshape` · `spline_regression` · `theme_set_global` · `tidymodels_cv` · `tidytuesdayr_load` · `unnest_list_cols`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(tidymodels)
library(modeldata)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
penguins <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-07-28/penguins.csv", show_col_types = FALSE)
penguins_raw <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-07-28/penguins_raw.csv", show_col_types = FALSE)
glimpse(penguins)
glimpse(penguins_raw)
```

## Load the weekly Data
```{r chunk_01}
penguins <- tt$penguins
```
```{r chunk_02}
penguins |>
  count(species, sort = TRUE)

penguins |>
  count(island, sort = TRUE)

penguins_pivoted <- penguins |>
  pivot_longer(cols = bill_length_mm:body_mass_g,
               names_to = "metric",
               values_to = "value")

penguins_pivoted |>
  ggplot(aes(value, fill = species)) +
  geom_histogram(bins = 20) +
  facet_wrap(~ metric, scales = "free_x")

penguins_pivoted |>
  ggplot(aes(value, fill = species)) +
  geom_density(alpha = .5) +
  facet_wrap(~ metric, scales = "free")

penguins_pivoted |>
  ggplot(aes(species, value)) +
  geom_boxplot() +
  facet_wrap(~ metric, scales = "free_y")
```
```{r chunk_03}
penguins |>
  ggplot(aes(year, fill = species)) +
  geom_bar()

penguins |>
  ggplot(aes(island, fill = species)) +
  geom_bar()
```

## Models to predict species
```{r chunk_04}
library(tidymodels)

set.seed(2020)
split <- penguins |>
  filter(!is.na(bill_length_mm)) |>
  mutate(species = fct_lump(species, 1)) |>
  initial_split()

training_data <- training(split)

splits <- training_data |>
  rsample::vfold_cv(v = 10)

my_metrics <- metric_set(accuracy, kap, roc_auc)
control <- control_resamples(extract = extract_model,
                             save_pred = TRUE)
logistic_spec <- logistic_reg(mode = "classification") |>
  set_engine("glm")

logistic_model <- logistic_spec |>
  fit_resamples(species ~ bill_length_mm,
                resamples = splits,
                metrics = my_metrics)

logistic_model_extended <- logistic_spec |>
  fit_resamples(species ~ bill_length_mm + bill_depth_mm +
                  flipper_length_mm + body_mass_g,
                resamples = splits,
                metrics = my_metrics)

nn <- nearest_neighbor(mode = "classification",
                 neighbors = 10) |>
  set_engine("kknn") |>
  fit_resamples(species ~ bill_length_mm + bill_depth_mm +
                  flipper_length_mm + body_mass_g,
                resamples = splits,
                metrics = my_metrics)

dt <- decision_tree(mode = "classification") |>
  set_engine("rpart") |>
  fit_resamples(species ~ bill_length_mm + bill_depth_mm +
                  flipper_length_mm + body_mass_g,
                resamples = splits,
                control = control,
                metrics = my_metrics)

# Look at one decision tree
dt$.extracts[[1]]$.extracts[[1]]

bind_rows(
  collect_metrics(logistic_model) |> mutate(model = "Logistic: bill length"),
  collect_metrics(logistic_model_extended) |> mutate(model = "Logistic: 4 predictors"),
  collect_metrics(nn) |> mutate(model = "KNN: 10 neighbors"),
  collect_metrics(dt) |> mutate(model = "Decision Tree")
) |>
  ggplot(aes(mean, .metric, color = model)) +
  geom_point() +
  geom_errorbarh(aes(xmin = mean - std_err,
                     xmax = mean + std_err)) +
  labs(title = "Cross validated accuracy metrics across models",
       x = "Estimated metric (+/- standard error)",
       y = "")
```
```{r chunk_05}
knn_trained <- nearest_neighbor(mode = "classification",
                 neighbors = 10) |>
  set_engine("kknn") |>
  fit(species ~ bill_length_mm + bill_depth_mm +
                  flipper_length_mm + body_mass_g,
      data = training(split))

predictions <- bind_cols(predict(knn_trained, new_data = testing(split)),
                         predict(knn_trained, new_data = testing(split), type = "prob")) |>
  bind_cols(testing(split))

metric_set(kap, accuracy)(predictions, species, estimate = .pred_class)
metric_set(roc_auc)(predictions, species, .pred_Adelie)

predictions |>
  ggplot(aes(species, .pred_Adelie)) +
  geom_boxplot()
```
```{r chunk_06}
split_multiclass <- penguins |>
  filter(!is.na(bill_length_mm)) |>
  mutate(species = factor(species)) |>
  initial_split()

splits_multiclass <- training(split_multiclass) |>
  vfold_cv(10)

multiclass <- multinom_reg(penalty = 0) |>
  set_engine("glmnet") |>
  fit_resamples(species ~ bill_length_mm + bill_depth_mm +
                  flipper_length_mm + body_mass_g,
                resamples = splits_multiclass,
                control = control,
                metrics = my_metrics)

multiclass |>
  unnest(.predictions) |>
  pivot_longer(cols = c(.pred_Adelie, .pred_Chinstrap, .pred_Gentoo),
               names_to = "predicted_class",
               values_to = "prob") |>
  ggplot(aes(predicted_class, prob)) +
  geom_boxplot() +
  facet_wrap(~ species)
```
```{r chunk_07}
predict(model, new_data = testing(split)) |>
  bind_cols(testing(split)) |>
  metrics(species, .pred_class)
```
```{r chunk_08}
library(modeldata)
data("lending_club")
lr_mod <- logistic_reg()

using_formula <-
  lr_mod |>
  set_engine("glm") |>
  fit(Class ~ funded_amnt + int_rate, data = lending_club)

using_formula
```

---

**→ [Hour 2: Modeling](h2_modeling_penguins.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
