---
title: "Hour 1 — EDA: Analyzing Netflix titles in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 044 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `boxplot_jitter` · `error_bar_coef_plot` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_area_stacked` · `geom_histogram_density` · `grouped_summary_ggplot` · `lasso_glmnet` · `linear_model_broom` · `log_odds_ratio`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- text is multilingual without language detection
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- text is long-form English and benefits from word-level decomposition
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(lubridate)
library(tidytext)
library(snakecase)
library(tidylo)
library(widyr)
library(tidygraph)
library(ggraph)
library(glmnet)
library(broom)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Load the weekly Data
```{r Load}
library(lubridate)

tt <- tt_load("2021-04-20")

netflix_titles <- tt$netflix_titles |>
  separate(duration, c("duration", "duration_units"), sep = " ", convert = TRUE) |>
  mutate(date_added = mdy(date_added),
         year_added = year(date_added)) |>
  mutate(mature = rating %in% c("TV-MA", "R", "NC-17"))
```
```{r chunk_01}
netflix_titles |>
  ggplot(aes(release_year, fill = type)) +
  geom_histogram(binwidth = 5) +
  facet_wrap(~ type, ncol = 1, scales = "free_y")

netflix_titles |>
  #count(year = 2 * (release_year %/% 2), type) |>
  count(year = release_year, type) |>
  group_by(type) |>
  mutate(percent = n / sum(n)) |>
  ggplot(aes(year, percent, color = type)) +
  geom_line()
```
```{r chunk_02}
netflix_titles |>
  filter(type == "Movie") |>
  mutate(decade = 10 * (release_year %/% 10)) |>
  ggplot(aes(decade, duration, group = decade)) +
  geom_boxplot()
```
```{r chunk_03}
summarize_titles <- function(tbl) {
  tbl |>
    summarize(n = n(),
              median_duration = median(duration),
              median_year = median(release_year)) |>
    arrange(desc(n))
}
```
```{r chunk_04}
netflix_titles |>
  count(rating, sort = TRUE)

netflix_titles |>
  separate_rows(listed_in, sep = ", ") |>
  group_by(type, genre = listed_in) |>
  summarize_titles() |>
  filter(type == "Movie") |>
  filter(genre != "Movies") |>
  mutate(genre = fct_reorder(genre, median_duration)) |>
  ggplot(aes(median_duration, genre)) +
  geom_col()
```
```{r chunk_05}
head(netflix_titles$date_added)
```
```{r chunk_06}
netflix_titles |>
  filter(!is.na(date_added)) |>
  arrange(date_added) |>
  select(type, title, date_added)

netflix_titles |>
  filter(!is.na(date_added)) |>
  count(year_added, type) |>
  ggplot(aes(year_added, n, fill = type)) +
  geom_area()

netflix_titles |>
  filter(year_added >= 2015) |>
  filter(!is.na(date_added), !is.na(rating)) |>
  mutate(rating = fct_lump(rating, 5)) |>
  ungroup() |>
  count(type, year_added, rating) |>
  group_by(type, year_added) |>
  mutate(percent = n / sum(n)) |>
  ggplot(aes(year_added, percent, fill = rating)) +
  geom_area() +
  facet_wrap(~ type)
```
```{r chunk_07}
netflix_titles |>
  filter(!is.na(country)) |>
  count(country = fct_lump(country, 20),
        type,
        sort = TRUE) |>
  mutate(country = fct_reorder(country, n)) |>
  ggplot(aes(n, country, fill = type)) +
  geom_col()

netflix_titles |>
  filter(!is.na(country)) |>
  filter(type == "Movie") |>
  group_by(country) |>
  summarize_titles()
```
```{r chunk_08}
netflix_titles |>
  filter(rating %in% c("R", "TV-MA")) |>
  count(country, sort = TRUE)

netflix_titles |>
  filter(!is.na(rating), !is.na(country)) |>
  group_by(type, country = fct_lump(country, 9)) |>
  summarize(n_mature = sum(rating %in% c("R", "TV-MA", "NC-17")),
            n = n(),
            .groups = "drop") |>
  mutate(pct_mature = n_mature / n,
         conf_low = qbeta(.025, n_mature + .5, n - n_mature + .5),
         conf_high = qbeta(.975, n_mature + .5, n - n_mature + .5)) |>
  ggplot(aes(pct_mature, country, color = type)) +
  geom_point(aes(size = n)) +
  geom_errorbar(aes(xmin = conf_low, xmax = conf_high), width = .1) +
  scale_x_continuous(labels = percent) +
  expand_limits(x = 0) +
  labs(x = "% of titles that are R/TV-MA")
```
```{r chunk_09}
library(tidytext)
library(snakecase)
library(tidylo)

words_unnested <- netflix_titles |>
  unnest_tokens(word, description) |>
  anti_join(stop_words, by = "word")

words_unnested |>
  count(type, word, sort = TRUE) |>
  mutate(type = to_snake_case(type)) |>
  spread(type, n, fill = 0) |>
  mutate(total = movie + tv_show) |>
  arrange(desc(total)) |>
  head(100) |>
  ggplot(aes(movie, tv_show)) +
  geom_point() +
  geom_text(aes(label = word) , vjust = 1, hjust = 1) +
  scale_x_log10() +
  scale_y_log10()

words_unnested |>
  count(type, word) |>
  bind_log_odds(type, word, n) |>
  arrange(desc(log_odds_weighted)) |>
  group_by(type) |>
  top_n(10, log_odds_weighted) |>
  ungroup() |>
  mutate(word = fct_reorder(word, log_odds_weighted)) |>
  ggplot(aes(log_odds_weighted, word)) +
  geom_col() +
  facet_wrap(~ type, scales = "free_y")
```
```{r chunk_10}
library(widyr)
library(tidygraph)
library(ggraph)

set.seed(2021)

words_unnested |>
  distinct(type, title, word) |>
  add_count(word, name = "word_total") |>
  filter(word_total >= 40) |>
  pairwise_cor(word, title, sort = TRUE) |>
  filter(correlation >= .1) |>
  igraph::graph_from_data_frame() |>
  ggraph(layout = "fr") +
  geom_edge_link(aes(alpha = correlation)) +
  geom_node_point() +
  geom_node_text(aes(label = name),
                 repel = TRUE) +
  theme(legend.position = "none")
```
```{r chunk_11}
word_genre_log_odds <- words_unnested |>
  distinct(type, title, word, genre = listed_in) |>
  add_count(word, name = "word_total") |>
  filter(word_total >= 25) |>
  separate_rows(genre, sep = ", ") |>
  filter(fct_lump(genre, 9) != "Other") |>
  count(genre, word) |>
  bind_log_odds(genre, word, n)
```
```{r chunk_12}
word_genre_log_odds |>
  group_by(genre) |>
  top_n(10, log_odds_weighted) |>
  ungroup() |>
  mutate(word = reorder_within(word, log_odds_weighted, genre)) |>
  ggplot(aes(log_odds_weighted, word, fill = genre)) +
  geom_col() +
  facet_wrap(~ genre, scales = "free_y") +
  scale_y_reordered() +
  theme(legend.position = "none") +
  labs(x = "Log-odds of word's specificity to genre",
       y = "")
```

## Lasso regression
```{r chunk_13}
word_ratings <- words_unnested |>
  count(type, title, rating, word) |>
  filter(!is.na(rating)) |>
  add_count(word, name = "word_total") |>
  filter(word_total >= 30)
```
```{r chunk_14}
library(glmnet)
library(broom)

other_features <- netflix_titles |>
  select(title, director, cast, genre = listed_in, country) |>
  gather(feature_type, feature, director, cast, genre, country) |>
  filter(!is.na(feature)) |>
  separate_rows(feature, sep = ", ") |>
  mutate(feature_type = str_to_title(feature_type)) |>
  unite(feature, feature_type, feature, sep = ": ") |>
  add_count(feature, name = "feature_count") |>
  filter(feature_count >= 10)

feature_matrix <- word_ratings |>
  mutate(feature = paste("Description:", word)) |>
  bind_rows(other_features) |>
  cast_sparse(title, feature)

y <- netflix_titles$mature[match(rownames(feature_matrix), netflix_titles$title)]

mod <- cv.glmnet(feature_matrix, y, family = "binomial")

plot(mod)

mod$glmnet.fit |>
  tidy() |>
  separate(term, c("feature_type", "feature"), sep = ": ") |>
  filter(lambda == mod$lambda.1se) |>
  top_n(60, abs(estimate)) |>
  mutate(feature = fct_reorder(feature, estimate)) |>
  ggplot(aes(estimate, feature, fill = feature_type)) +
  geom_col() +
  labs(x = "Coefficient: does this make the title more likely to be TV-MA/R?",
       y = "",
       fill = "Feature Type")
```

---

**→ [Hour 2: Modeling](h2_modeling_netflix_titles.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
