---
title: "Hour 2 — Modeling: Analyzing Netflix titles in R"
subtitle: "Going deeper with statistical models | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 044 · Hour 2 · [← Hour 1 EDA](h1_eda_netflix_titles.qmd)
:::


::: {.callout-note collapse="true"}
## Atomic Concepts — Hour 2
`add_count_inline` · `boxplot_jitter` · `error_bar_coef_plot` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_area_stacked` · `geom_histogram_density` · `grouped_summary_ggplot` · `lasso_glmnet`
:::

## Setup

```{r setup}
library(tidyverse)
library(scales)
library(broom)
library(here)
theme_set(theme_light(base_size = 13))
```

## Reload data

```{r load-data}
# Re-load from Hour 1 — same local paths
source_qmd <- here::here("output/episodes/netflix_titles/h1_eda_netflix_titles.qmd")
# Data files are re-read directly below if needed
```

## Model 1 — Sentiment analysis

```{r text-setup}
library(tidytext)
# df <- read_csv(...)
```

```{r sentiment}
afinn <- get_sentiments("afinn")

# Adjust column names to match your data
# df |> unnest_tokens(word, text_column) |> inner_join(afinn, by = "word") |>
#   group_by(group_col) |> summarise(mean_sentiment = mean(value), .groups = "drop") |>
#   mutate(group_col = fct_reorder(group_col, mean_sentiment)) |>
#   ggplot(aes(mean_sentiment, group_col)) + geom_col()
cat("Sentiment analysis: customise the column names above to your dataset\n")
```

## Model 2 — TF-IDF: most distinctive terms by group

```{r tfidf}
# df |> unnest_tokens(word, text_column) |>
#   anti_join(stop_words, by = "word") |>
#   count(group_col, word) |>
#   bind_tf_idf(word, group_col, n) |>
#   group_by(group_col) |>
#   slice_max(tf_idf, n = 8, with_ties = FALSE) |>
#   mutate(word = reorder_within(word, tf_idf, group_col)) |>
#   ggplot(aes(tf_idf, word, fill = group_col)) +
#   geom_col(show.legend = FALSE) +
#   facet_wrap(~group_col, scales = "free_y") +
#   scale_y_reordered()
cat("TF-IDF: customise column names above\n")
```


## Model 1 — Linear regression baseline

```{r lm-setup}
# df <- read_csv(...)
glimpse(df)
```

```{r lm-fit}
# Identify your outcome and predictors first — adjust formula below
# fit <- lm(outcome ~ predictor1 + predictor2, data = df)
# tidy(fit) |> filter(term != "(Intercept)") |>
#   mutate(term = fct_reorder(term, estimate)) |>
#   ggplot(aes(estimate, term, xmin = estimate - 2*std.error,
#              xmax = estimate + 2*std.error)) +
#   geom_point() + geom_errorbarh(height = 0.2) +
#   geom_vline(xintercept = 0, lty = 2)
cat("Linear model: adjust formula to your outcome variable\n")
```

## Model 2 — Variable importance

```{r importance}
# augment(fit) |>
#   ggplot(aes(.fitted, outcome)) +
#   geom_point(alpha = 0.4) +
#   geom_abline(color = "red", lty = 2) +
#   labs(title = "Predicted vs actual", x = "Fitted", y = "Observed")
cat("Residual diagnostics: wire up after lm fit above\n")
```


## Model 1 — Time trend

```{r time-setup}
library(lubridate)
# df <- read_csv(...)
```

```{r trend}
# df |>
#   mutate(date_col = ymd(date_col)) |>
#   group_by(floor_date(date_col, "month")) |>
#   summarise(n = n(), .groups = "drop") |>
#   ggplot(aes(`floor_date(date_col, "month")`, n)) +
#   geom_line() + geom_smooth(method = "loess")
cat("Time trend: adjust date_col to your date column\n")
```

## Model 2 — Seasonality / year-over-year

```{r seasonality}
# df |>
#   mutate(year = year(date_col), month = month(date_col, label = TRUE)) |>
#   group_by(year, month) |>
#   summarise(n = n(), .groups = "drop") |>
#   ggplot(aes(month, n, group = year, color = factor(year))) +
#   geom_line()
cat("Seasonality: adjust date_col above\n")
```


---

**→ [Hour 3: Communicate](h3_communicate_netflix_titles.qmd)**
