---
title: "KER-Guided Analysis: The Bechdel Test"
subtitle: "A from-scratch Quarto walkthrough built from the Knowledge Extraction Register"
author: "Andrei Stoian"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show the analysis code"
    number-sections: true
    fig-width: 9
    fig-height: 5.5
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup

library(tidyverse)
library(scales)

theme_set(theme_light(base_size = 13))

simple_stop_words <- c(
  "the", "and", "for", "that", "with", "this", "from", "their", "have", "into",
  "about", "after", "before", "while", "where", "when", "what", "which", "will",
  "would", "there", "these", "those", "they", "them", "then", "than", "just",
  "like", "been", "being", "over", "under", "between", "through", "during",
  "against", "because", "around", "also", "only", "very", "more", "most", "some",
  "such", "much", "many", "each", "other", "another", "into", "onto", "upon",
  "movie", "film", "story", "life", "lives", "young", "man", "men", "woman",
  "women", "boy", "girl", "family", "friends", "father", "mother", "son",
  "daughter", "new", "old", "one", "two", "three", "four", "five", "who",
  "whose", "your", "our", "his", "her", "hers", "him", "she", "himself",
  "herself", "its", "it's", "are", "was", "were", "has", "had", "did", "does",
  "not", "but", "out", "off", "all", "any", "can", "may", "get", "gets", "got",
  "set", "made", "make", "takes", "take"
)

raw_bechdel_path <- "tidytuesday/data/2021/2021-03-09/raw_bechdel.csv"
movies_path <- "tidytuesday/data/2021/2021-03-09/movies.csv"

raw_bechdel <- readr::read_csv(raw_bechdel_path, show_col_types = FALSE)
movies <- readr::read_csv(movies_path, show_col_types = FALSE) %>%
  mutate(
    across(contains("gross"), readr::parse_number),
    binary = factor(binary, levels = c("FAIL", "PASS"))
  )
```

## Why this dataset

This walkthrough uses the Bechdel dataset because it is a strong fit for the workflow our KER already surfaced:

-   `load -> inspect -> reshape -> compare -> visualize`
-   plus a light predictive extension
-   plus optional text analysis from plot summaries

From the KER, the matching episode pattern is:

-   `analysis_type`: `Predictive`
-   `pattern_signature`: `load_inspect_reshape_join_tokenize_trend_model_evaluate_visualize`

That gives us a useful starting point without forcing us to copy the original screencast.

::: callout-note
## Thought Process

The KER suggests that this is not just a "make one chart" dataset. It supports three layers:

1.  trend over time
2.  commercial and genre comparison
3.  light modeling and plot-text comparison

So the plan is to build the analysis in that same order.
:::

## Step 1: Inspect the shape of the data

We have two related files:

-   `raw_bechdel.csv`: longer historical coverage with the raw test outcome
-   `movies.csv`: richer movie metadata, but only through 2013

```{r}
#| label: inspect-shape

tibble(
  dataset = c("raw_bechdel", "movies"),
  rows = c(nrow(raw_bechdel), nrow(movies)),
  columns = c(ncol(raw_bechdel), ncol(movies))
)
```

```{r}
#| label: inspect-columns

tibble(
  movies_columns = names(movies)
) %>%
  slice_head(n = 15)
```

::: callout-note
## Thought Process

Before choosing visuals, we want to answer two boring-but-critical questions:

-   Which file is best for long-run trend?
-   Which file is best for richer comparisons?

The answer is:

-   use `raw_bechdel` for the historical pass-rate timeline
-   use `movies` for genre, revenue, ratings, and plot text
:::

## Step 2: Build a clean analytical base

The enriched `movies` table needs only a little cleaning for this analysis:

-   gross columns parsed as numeric
-   decade bucket
-   a simplified pass flag
-   a 1990+ subset for modern comparisons

```{r}
#| label: clean-data

movies_clean <- movies %>%
  mutate(
    decade = 10 * (year %/% 10),
    pass_flag = binary == "PASS",
    log_intgross_2013 = log10(intgross_2013),
    log_budget_2013 = log10(budget_2013)
  )

movies_modern <- movies_clean %>%
  filter(year >= 1990)

glimpse(movies_modern)
```

## Step 3: Start with the broadest question

How has the pass rate changed over time?

```{r}
#| label: trend-raw

raw_bechdel_trend <- raw_bechdel %>%
  mutate(
    pass_flag = rating == 3,
    decade = 10 * (year %/% 10)
  ) %>%
  group_by(decade) %>%
  summarise(
    n_movies = n(),
    pct_pass = mean(pass_flag, na.rm = TRUE),
    .groups = "drop"
  )
raw_bechdel_trend |> 
ggplot(aes(decade, pct_pass)) +
  geom_line(linewidth = 1, color = "#1f4e79") +
  geom_point(aes(size = n_movies), color = "#d55e00", alpha = 0.9) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_size_continuous(range = c(2.5, 8)) +
  expand_limits(y = 0) +
  labs(
    title = "Bechdel pass rates have generally improved over time",
    subtitle = "Point size shows how many movies are represented in each decade",
    x = "Decade",
    y = "Share of movies that pass",
    size = "Movies"
  )
```

::: callout-note
## Interpretation Guide

This is the classic KER move: start broad before getting specific.

The long-run trend helps us decide whether later comparisons should control for time. Because the pass rate changes across decades, it makes sense to focus many downstream comparisons on more modern films.
:::

## Step 4: Compare representation and commercial outcomes

One natural next question is whether passing the Bechdel test travels with commercial success.

```{r}
#| label: revenue-summary

movies_modern %>%
  group_by(binary) %>%
  summarise(
    n_movies = n(),
    median_budget = median(budget_2013, na.rm = TRUE),
    median_intgross = median(intgross_2013, na.rm = TRUE),
    median_imdb_rating = median(imdb_rating, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    across(c(median_budget, median_intgross), dollar)
  )
```

```{r}
#| label: gross-density

ggplot(
  movies_modern %>% filter(!is.na(intgross_2013)),
  aes(intgross_2013, fill = binary)
) +
  geom_density(alpha = 0.45) +
  scale_x_log10(labels = dollar_format()) +
  labs(
    title = "Pass and fail movies overlap heavily in gross revenue",
    subtitle = "The question is less about one clean split and more about differences in distribution",
    x = "International gross (2013 dollars, log scale)",
    y = "Density",
    fill = "Bechdel result"
  )
```

```{r}
#| label: top-grossing

movies_modern %>%
  filter(!is.na(intgross_2013)) %>%
  slice_max(intgross_2013, n = 20, with_ties = FALSE) %>%
  mutate(title = fct_reorder(title, intgross_2013)) %>%
  ggplot(aes(intgross_2013, title, fill = binary)) +
  geom_col() +
  scale_x_continuous(labels = dollar_format(scale_cut = cut_short_scale())) +
  labs(
    title = "Many of the biggest-grossing films still fail the test",
    subtitle = "A simple ranking is often more persuasive than a model at this stage",
    x = "International gross (2013 dollars)",
    y = NULL,
    fill = "Bechdel result"
  )
```

::: callout-note
## Thought Process

The KER pattern says to compare distributions, then rankings.

-   distribution plots tell us whether the two groups separate cleanly
-   ranked plots tell us what the high-profile exceptions look like

Using both gives a better guide-through than either one alone.
:::

## Step 5: Bring genre into the story

Genres are stored as comma-separated values, so we need to reshape them before comparing pass rates.

```{r}
#| label: genre-data

movie_genres <- movies_modern %>%
  filter(!is.na(genre)) %>%
  separate_rows(genre, sep = ",\\s*") %>%
  filter(genre != "")

genre_summary <- movie_genres %>%
  group_by(genre) %>%
  summarise(
    n_movies = n(),
    pct_pass = mean(pass_flag, na.rm = TRUE),
    median_intgross = median(intgross_2013, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_movies >= 25)

genre_summary %>%
  arrange(desc(pct_pass)) %>%
  slice_head(n = 12)
```

```{r}
#| label: genre-bubble

ggplot(genre_summary, aes(median_intgross, pct_pass)) +
  geom_point(aes(size = n_movies), color = "#2c7fb8", alpha = 0.8) +
  geom_text(aes(label = genre), check_overlap = TRUE, nudge_y = 0.015, size = 3.4) +
  scale_x_log10(labels = dollar_format()) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  expand_limits(x = 1, y = 0) +
  labs(
    title = "Genre matters more than a single overall pass/fail average",
    subtitle = "Some genres pair stronger representation with stronger median grosses",
    x = "Median international gross (2013 dollars, log scale)",
    y = "Share of movies that pass",
    size = "Movies"
  )
```

## Step 6: Use plot summaries for a text signal

The KER flagged this episode as a good candidate for tokenization. We can compare the words that show up more often in plot summaries for passing versus failing films.

```{r}
#| label: tokenize-plots

movie_words <- movies_modern %>%
  filter(!is.na(plot)) %>%
  select(imdb, binary, plot) %>%
  mutate(
    plot = str_to_lower(plot),
    plot = str_replace_all(plot, "[^a-z\\s]", " ")
  ) %>%
  mutate(word = str_split(plot, "\\s+")) %>%
  unnest(word) %>%
  filter(
    word != "",
    !word %in% simple_stop_words,
    str_detect(word, "^[a-z]+$"),
    str_length(word) >= 3
  )

word_rates <- movie_words %>%
  count(binary, word, sort = TRUE) %>%
  group_by(binary) %>%
  mutate(total_words = sum(n)) %>%
  ungroup() %>%
  select(binary, word, n, total_words) %>%
  pivot_wider(
    names_from = binary,
    values_from = c(n, total_words),
    values_fill = 0
  ) %>%
  mutate(
    total_n = n_FAIL + n_PASS,
    pass_rate = (n_PASS + 1) / (total_words_PASS + 1),
    fail_rate = (n_FAIL + 1) / (total_words_FAIL + 1),
    log_ratio = log(pass_rate / fail_rate)
  ) %>%
  filter(total_n >= 40)

word_extremes <- bind_rows(
  word_rates %>% slice_max(log_ratio, n = 12, with_ties = FALSE),
  word_rates %>% slice_min(log_ratio, n = 12, with_ties = FALSE)
) %>%
  mutate(direction = if_else(log_ratio > 0, "More common in PASS plots", "More common in FAIL plots")) %>%
  mutate(word = fct_reorder(word, log_ratio))
```

```{r}
#| label: word-plot

ggplot(word_extremes, aes(log_ratio, word, fill = direction)) +
  geom_col() +
  labs(
    title = "Plot summaries also carry a representation signal",
    subtitle = "Positive values mean the word appears more often in PASS movie plots",
    x = "Log ratio of word use in PASS vs FAIL plots",
    y = NULL,
    fill = NULL
  )
```

::: callout-note
## Interpretation Guide

This is not a causal claim. These words do not *cause* a movie to pass.

They are just a compact way to show that the kinds of stories associated with passing and failing films are not randomly distributed.
:::

## Step 7: Add a light model

The KER also suggested a predictive extension. We will keep it intentionally modest: a logistic regression predicting whether a movie passes the Bechdel test using year, ratings, and broad commercial signals.

```{r}
#| label: model-data

model_data <- movies_modern %>%
  mutate(
    genre_action = str_detect(coalesce(genre, ""), "Action"),
    genre_comedy = str_detect(coalesce(genre, ""), "Comedy"),
    genre_drama = str_detect(coalesce(genre, ""), "Drama"),
    genre_romance = str_detect(coalesce(genre, ""), "Romance"),
    genre_horror = str_detect(coalesce(genre, ""), "Horror"),
    genre_thriller = str_detect(coalesce(genre, ""), "Thriller"),
    log_budget_2013 = log1p(budget_2013),
    log_intgross_2013 = log1p(intgross_2013)
  ) %>%
  filter(
    !is.na(pass_flag),
    !is.na(year),
    !is.na(imdb_rating),
    !is.na(metascore),
    !is.na(log_budget_2013),
    !is.na(log_intgross_2013)
  )

bechdel_model <- glm(
  pass_flag ~ year + imdb_rating + metascore + log_budget_2013 + log_intgross_2013 +
    genre_action + genre_comedy + genre_drama + genre_romance + genre_horror + genre_thriller,
  data = model_data,
  family = binomial()
)

model_coefficients <- summary(bechdel_model)$coefficients %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  as_tibble() %>%
  filter(term != "(Intercept)") %>%
  transmute(
    term,
    estimate = Estimate,
    std_error = `Std. Error`,
    z_value = `z value`,
    p_value = `Pr(>|z|)`
  ) %>%
  arrange(desc(abs(estimate)))

model_coefficients
```

```{r}
#| label: model-coefs

model_coefficients %>%
  slice_head(n = 12) %>%
  mutate(term = fct_reorder(term, estimate)) %>%
  ggplot(aes(estimate, term, fill = estimate > 0)) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "The model is a guide, not the goal",
    subtitle = "Positive coefficients align with a higher probability of passing",
    x = "Log-odds coefficient",
    y = NULL
  )
```

## What we learned

This KER-guided build points to a few useful takeaways:

-   pass rates improve over time, so time is part of the story
-   commercial success and representation are related in a nuanced way, not a clean split
-   genre composition matters a lot
-   plot summaries carry a measurable representation signal
-   a light model can organize the relationships, but the explanatory visuals do most of the teaching

## Why this matches the KER

This document followed the KER-derived workflow quite closely:

1.  **Context** Pick the question and define the unit of analysis.
2.  **Inspect** Decide which table is best for which question.
3.  **Transform** Parse numerics, create decade buckets, and unnest genres.
4.  **Compare** Look at time, revenue, and genre-level differences.
5.  **Tokenize** Use plot summaries for a lightweight language comparison.
6.  **Model** Add a simple predictive layer after the descriptive work is clear.
7.  **Reflect** Summarize reusable moves for the next dataset.

## Reuse pattern

If we wanted to port this exact approach to another dataset, the reusable workflow would be:

-   start with a broad trend chart
-   separate descriptive comparison from modeling
-   reshape multi-value text fields into long form
-   create one "story chart", one "business chart", one "text signal chart", and one "model chart"
-   keep the narrative explicit so the analysis reads like a guided walkthrough, not a code dump
