---
title: "Hour 1 — EDA: Analyzing wealth and income in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 077 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`boxplot_jitter` · `crossing_simulation_grid` · `fct_lump_infrequent` · `fct_relevel_manual` · `fct_reorder_stat` · `geom_histogram_density` · `grouped_summary_ggplot` · `linear_model_broom` · `log_scale` · `lubridate_dates` · `parse_number_dirty` · `skim_inspect`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(broom)
library(tidymodels)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Load the weekly Data
```{r Load}
tt <- tt_load("2021-05-18")

survey <- tt$survey |>
  mutate(timestamp = mdy_hms(timestamp),
         age_category = fct_relevel(fct_reorder(how_old_are_you, parse_number(how_old_are_you)), "under 18"),
         experience_overall = str_replace(overall_years_of_professional_experience, " - ", "-"),
         experience_overall = fct_reorder(experience_overall, parse_number(experience_overall)),
         experience_in_field = str_replace(years_of_experience_in_field, " - ", "-"),
         experience_in_field = fct_reorder(experience_in_field, parse_number(experience_in_field))) |>
  mutate(gender = fct_collapse(coalesce(gender, "Other or prefer not to answer"), "Other or prefer not to answer" = c("Other or prefer not to answer", "Prefer not to answer")),
         race = fct_lump(coalesce(race, "Other"), 4))

survey |>
  count(years_of_experience_in_field)

survey |>
  count(industry, sort = TRUE)

survey |>
  count(job_title, sort = TRUE)

survey |>
  count(currency, sort = TRUE)
```
```{r chunk_01}
survey_usd <- survey |>
  filter(currency == "USD") |>
  filter(annual_salary >= 5000,
         annual_salary <= 2e6) |>
  mutate(state = str_remove(state, ", .*"))
```
```{r chunk_02}
survey_usd |>
  ggplot(aes(annual_salary)) +
  geom_histogram() +
  scale_x_log10(labels = dollar_format()) +
  labs(x = "Annual")

summarize_salary <- function(tbl) {
  tbl |>
    summarize(n = n(),
              median_salary = median(annual_salary)) |>
    arrange(desc(n))
}

plot_categorical <- function(tbl, column, n_levels = 9, reorder = TRUE) {
  lumped_tbl <- tbl |>
    filter(!is.na({{ column }})) |>
    mutate({{ column }} := fct_lump({{ column }}, n_levels))
  
  if (reorder) {
    lumped_tbl <- lumped_tbl |>
      mutate({{ column }} := fct_reorder({{ column }}, annual_salary))
  }

  lumped_tbl |>
    group_by({{ column }}) |>
    summarize_salary() |>
    ggplot(aes(median_salary, {{ column }})) +
    geom_col() +
    scale_x_continuous(labels = dollar_format()) +
    labs(x = "Median salary")
}

survey_usd |>
  plot_categorical(state)

survey_usd |>
  plot_categorical(industry)

survey_usd |>
  plot_categorical(job_title, n_levels = 10)

survey_usd |>
  plot_categorical(experience_overall, reorder = FALSE)

survey_usd |>
  plot_categorical(experience_in_field, reorder = FALSE)

survey_usd |>
  plot_categorical(age_category, reorder = FALSE)

survey_usd |>
  plot_categorical(gender)

survey_usd |>
  plot_categorical(race, n_levels = 4)
```

## ANOVA
```{r chunk_03}
survey_usd |>
  filter(!is.na(experience_overall)) |>
  ggplot(aes(annual_salary, experience_overall)) +
  geom_boxplot() +
  scale_x_log10()

library(broom)

lm(log2(annual_salary) ~ experience_overall, data = survey_usd)
lm(log2(annual_salary) ~ experience_in_field, data = survey_usd) |>
  summary()

survey_usd |>
  mutate(job_title = fct_lump(job_title, 20)) |>
  lm(log2(annual_salary) ~ job_title, data = .) |>
  summary()

survey_usd |>
  mutate(job_title = fct_lump(job_title, 10),
         state = fct_lump(state, 10),
         industry = fct_lump(industry, 10)) |>
  lm(log2(annual_salary) ~ job_title + state + city + experience_in_field + gender + race + industry, data = .) |>
  anova() |>
  tidy() |>
  mutate(pct_variation = sumsq / sum(sumsq)) |>
  arrange(desc(pct_variation))
```
```{r chunk_04}
survey_usd |>
  count(overall_years_of_professional_experience, sort = TRUE)
```

## Machine Learning
```{r chunk_05}
library(tidymodels)

set.seed(2021)

survey_usd_split <- initial_split(survey_usd)

survey_usd_training <- training(survey_usd_split)
survey_usd_testing <- testing(survey_usd_split)

rec <- survey_usd_training |>
  recipe(annual_salary ~ job_title + state + city + experience_in_field + gender + race + industry +
           highest_level_of_education_completed) |>
  step_novel(all_nominal()) |>
  step_unknown(job_title, industry, state, city, highest_level_of_education_completed) |>
  step_log(annual_salary, base = 2) |>
  step_other(job_title, industry, state, city, threshold = tune())
```
```{r chunk_06}
training_cv <- vfold_cv(survey_usd_training)

threshold_grid <- crossing(threshold = c(.001, .003, .01, .03, .1))

linear_model_cv_tune_threshold <- linear_reg() |>
  set_engine("lm") |>
  tune_grid(rec, training_cv, grid = threshold_grid)

linear_model_cv_tune_threshold |>
  autoplot() +
  scale_x_log10()
```
```{r chunk_07}
rec_with_threshold <- rec |>
  finalize_recipe(list(threshold = .001))
```
```{r chunk_08}
doParallel::registerDoParallel(cores = 4)

rf_grid <- crossing(mtry = c(2, 3, 4, 5),
                    trees = c(30, 100, 200, 300))

training_cv5 <- vfold_cv(survey_usd_training, v = 5)

linear_model_cv <- linear_reg() |>
  set_engine("lm") |>
  fit_resamples(rec_with_threshold, training_cv5)

random_forest_tune <- rand_forest(mode = "regression", mtry = tune(), trees = tune()) |>
  set_engine("ranger") |>
  tune_grid(rec_with_threshold, training_cv5, grid = rf_grid,
            control = control_grid(verbose = TRUE))

linear_model_metrics <- linear_model_cv |>
  collect_metrics()

linear_model_metrics

random_forest_tune |>
  collect_metrics() |>
  filter(.metric == "rsq") |>
  ggplot(aes(trees, mean, color = factor(mtry))) +
  geom_line() +
  geom_hline(yintercept = linear_model_metrics$mean[2], lty = 2) +
  labs(y = "R Squared",
       color = "# of splits",
       x = "# of trees")
```

---

**→ [Hour 2: Modeling](h2_modeling_wealth_income.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
