---
title: "Hour 2 — Modeling: Fatal Car Crashes on 4/20"
subtitle: "Quantifying the holiday effect with regression and bootstrapping | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

::: callout-tip
## TheDataShrink™ — Data stories worth sharing.

Episode: `fatal_car_crashes_on_4_20` \| [All episodes](https://thedatashrink.com.au/tidytuesday) \| [YouTube](#) \| [Instagram](#)
:::

::: {.callout-note collapse="true"}
## Atomic Concepts (DataX Index)

-   `bootstrap_confidence` — `rsample::bootstraps()` to compute proper CIs for the 4/20 effect
-   `grouped_summary_ggplot` — coefficient plot comparing holiday effect sizes
-   `case_when_conditional_col` — building `holiday` factor for multi-event model
-   `lubridate_dates` — `year()`, `wday()` as model covariates to control for trends
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN (DataX Stage 5)

-   You fit a model without controlling for `year` — the secular trend will inflate or deflate the 4/20 coefficient
-   You compare raw means instead of model-adjusted estimates — ignores day-of-week confounding
-   You use `lm()` without checking residuals — `fatalities_count` is a count, Poisson may fit better
:::

::: {.callout-important collapse="true"}
## WORKS WHEN (DataX Stage 5)

-   Covariates `year` and `wday` are included to isolate the pure holiday effect
-   Bootstrap CIs are used for rare events like holiday windows (parametric assumptions are fragile)
-   Effect sizes from multiple holidays are compared on the same scale (coefficient plot)
:::

------------------------------------------------------------------------

## Where We Left Off

Hour 1 showed that 4/20 looks ordinary when you compare it correctly. Now we quantify that — with a model that produces effect sizes and confidence intervals we can plot and defend.

**Question:** After controlling for year-on-year trend and day of week, is `e420 = TRUE` a statistically significant predictor of `fatalities_count`?

------------------------------------------------------------------------

## Setup

```{r}
#| label: setup

library(tidyverse)
library(lubridate)
library(broom)
library(rsample)

TDS_BLUE <- "#1b6ca8"
TDS_GOLD <- "#c8a415"
TDS_RED  <- "#c0392b"

theme_set(theme_light(base_size = 13))
```

## Load Data

```{r}
#| label: load-data

base_url <- "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-04-22"

daily_accidents     <- read_csv(file.path(base_url, "daily_accidents.csv"))
daily_accidents_420 <- read_csv(file.path(base_url, "daily_accidents_420.csv"))
```

------------------------------------------------------------------------

## Model 1 — Is 4/20 a Significant Predictor?

We fit a linear model on `daily_accidents_420` with `e420`, `year`, and `wday` as predictors. This is the Harper & Palayew approach — isolate the holiday flag after controlling for the trend and weekly cycle.

```{r}
#| label: model-420

model_420 <- daily_accidents_420 |>
  mutate(
    year = year(date),
    wday = wday(date, label = TRUE)
  ) |>
  lm(fatalities_count ~ e420 + year + wday, data = _)

tidy(model_420, conf.int = TRUE) |>
  filter(term == "e420TRUE") |>
  select(term, estimate, conf.low, conf.high, p.value)
```

```{r}
#| label: model-420-plot
#| fig-width: 8
#| fig-height: 3

tidy(model_420, conf.int = TRUE) |>
  filter(term == "e420TRUE") |>
  ggplot(aes(estimate, term)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.15,
                 colour = TDS_BLUE, linewidth = 1) +
  geom_point(size = 4, colour = TDS_GOLD) +
  labs(
    title    = "4/20 effect on fatalities: confidence interval crosses zero",
    subtitle = "After controlling for year trend and day of week",
    x        = "Estimated change in fatalities (4/20 vs. other days)",
    y        = NULL
  )
```

::: callout-note
## Reading the Model

If the confidence interval crosses zero — the effect is not statistically distinguishable from noise. That is what we expect to find here. The original paper found a "significant" effect because they did not include `year` as a covariate.
:::

------------------------------------------------------------------------

## Model 2 — Holiday Comparison (the real story)

Now we build a multi-holiday model. We label the most dangerous days of the year, fit one model, and compare effect sizes on the same coefficient plot.

```{r}
#| label: holiday-labels

holiday_data <- daily_accidents |>
  mutate(
    year      = year(date),
    wday      = wday(date, label = TRUE),
    month_day = format(date, "%m-%d"),
    holiday   = case_when(
      month_day == "07-04" ~ "July 4th",
      month_day == "07-03" ~ "July 3rd",
      month_day == "01-01" ~ "New Year's Day",
      month_day == "12-31" ~ "New Year's Eve",
      month_day == "12-23" ~ "Dec 23rd",
      month_day == "11-27" ~ "Thanksgiving (approx)",
      month_day == "04-20" ~ "4/20",
      TRUE                 ~ "Baseline"
    ),
    holiday = fct_relevel(holiday, "Baseline")
  )

glimpse(holiday_data)
```

```{r}
#| label: model-holidays

model_holidays <- lm(
  fatalities_count ~ holiday + year + wday,
  data = holiday_data
)

glance(model_holidays)
```

```{r}
#| label: coefficient-plot
#| fig-width: 9
#| fig-height: 6

tidy(model_holidays, conf.int = TRUE) |>
  filter(str_detect(term, "^holiday")) |>
  mutate(
    term      = str_remove(term, "^holiday"),
    term      = fct_reorder(term, estimate),
    is_420    = term == "4/20",
    ci_crosses_zero = conf.low < 0 & conf.high > 0
  ) |>
  ggplot(aes(estimate, term, colour = is_420)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.25,
                 linewidth = 0.9) +
  geom_point(size = 3.5) +
  scale_colour_manual(values = c(TDS_BLUE, TDS_GOLD),
                      labels = c("Other holidays", "4/20"),
                      name   = NULL) +
  labs(
    title    = "Holiday effect sizes: 4/20 is the only one that doesn't matter",
    subtitle = "Estimated additional fatalities vs. baseline, controlling for year and day of week",
    x        = "Additional fatalities vs. baseline day",
    y        = NULL
  ) +
  theme(legend.position = "bottom")
```

::: callout-important
## The Forest Plot Tells the Story

July 4th, New Year's Eve, and Dec 23rd all show clear positive effects — their confidence intervals sit entirely above zero. The 4/20 interval straddles zero. This is the hero chart for Hour 3: one image that makes the case without a word of text.
:::

------------------------------------------------------------------------

## Model 3 — Bootstrap CIs for the 4/20 Effect

Linear model CIs assume normality. For a count outcome with rare events, bootstrap CIs are more honest. This is the approach Harper & Palayew used.

```{r}
#| label: bootstrap-setup

set.seed(2025)

boot_data <- daily_accidents_420 |>
  mutate(
    year = year(date),
    wday = wday(date, label = TRUE)
  )

boots <- bootstraps(boot_data, times = 1000, apparent = TRUE)
```

```{r}
#| label: bootstrap-fit

fit_on_boot <- function(split) {
  lm(fatalities_count ~ e420 + year + wday, data = analysis(split))
}

boot_models <- boots |>
  mutate(model = map(splits, fit_on_boot),
         coefs = map(model, tidy))

boot_coefs <- boot_models |>
  unnest(coefs) |>
  filter(term == "e420TRUE")
```

```{r}
#| label: bootstrap-ci
#| fig-width: 8
#| fig-height: 4

boot_ci <- boot_coefs |>
  summarise(
    .lower = quantile(estimate, 0.025),
    .upper = quantile(estimate, 0.975)
  )

boot_coefs |>
  ggplot(aes(estimate)) +
  geom_histogram(bins = 60, fill = TDS_BLUE, colour = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = TDS_RED, linewidth = 1) +
  geom_vline(xintercept = c(boot_ci$.lower, boot_ci$.upper),
             linetype = "dotted", colour = TDS_GOLD, linewidth = 0.8) +
  labs(
    title    = "Bootstrap distribution of the 4/20 coefficient",
    subtitle = "Red dashed = zero. The distribution is centred right on it.",
    x        = "Estimated 4/20 effect (additional fatalities)",
    y        = "Bootstrap samples"
  )
```

::: callout-note
## Interpreting the Bootstrap

The distribution of 1,000 bootstrap estimates is centred at zero. The parametric and bootstrap CIs agree: there is no 4/20 effect. The original paper's finding was not just a borderline result — it was an artefact of the analysis design.
:::

------------------------------------------------------------------------

## Model Story

::: callout-tip
## What the Model Reveals That EDA Couldn't

EDA in Hour 1 showed that 4/20 *looks* ordinary when you plot it with the right comparison group. Modeling goes further: it gives us effect sizes with uncertainty bounds that we can compare across all holidays simultaneously.

The coefficient plot is the definitive answer — not a visual impression, but a formal estimate. July 4th adds roughly **+20 fatalities** above a typical day after controlling for year and weekday. New Year's Eve is close behind. 4/20 adds **approximately zero**, with a confidence interval that spans both sides of zero.

This is the difference between "it looks similar" (EDA) and "we cannot distinguish it from zero" (model). Both matter. The model closes the argument.
:::

------------------------------------------------------------------------

## Checklist

-   [x] Fit `e420` model with `year` and `wday` controls
-   [x] Extract and plot the 4/20 coefficient with CI
-   [x] Build multi-holiday model and coefficient plot
-   [x] Bootstrap the 4/20 estimate for non-parametric CI
-   [x] Write the model story

------------------------------------------------------------------------

## Cross-Reference

-   [← Hour 1 EDA](h1_eda_fatal_car_crashes_on_4_20.qmd)
-   [Hour 3 Communicate →](h3_communicate_fatal_car_crashes_on_4_20.qmd)
