---
title: "Hour 1 — EDA: Analyzing employment and earnings in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 025 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`case_when_conditional_col` · `error_bar_coef_plot` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `glue_strings` · `grouped_summary_ggplot` · `label_repulsion_ggrepel` · `pivot_reshape` · `spline_regression` · `theme_set_global` · `tidytuesdayr_load`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(glue)
library(ggrepel)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
bls-2015 <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-02-23/bls-2015.xlsx", show_col_types = FALSE)
bls-2016 <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-02-23/bls-2016.xlsx", show_col_types = FALSE)
bls-2017 <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-02-23/bls-2017.xlsx", show_col_types = FALSE)
bls-2018 <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-02-23/bls-2018.xlsx", show_col_types = FALSE)
bls-2019 <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-02-23/bls-2019.xlsx", show_col_types = FALSE)
bls-2020 <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-02-23/bls-2020.xlsx", show_col_types = FALSE)
earn <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-02-23/earn.csv", show_col_types = FALSE)
employed <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-02-23/employed.csv", show_col_types = FALSE)
~$bls-2020 <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-02-23/~$bls-2020.xlsx", show_col_types = FALSE)
glimpse(bls-2015)
glimpse(bls-2016)
```

## Load the weekly Data
```{r chunk_01}
employed_cleaned <- employed |>
  filter(!is.na(employ_n)) |>
  mutate(industry = fct_lump(industry, 15, w = employ_n),
         industry = fct_reorder(industry, employ_n, sum))

employed_cleaned |>
  filter(dimension == "Total") |>
  ggplot(aes(year, employ_n, fill = industry)) +
  geom_col() +
  scale_y_continuous(labels = comma) +
  labs(y = "# employed in industry",
       x = "Year")

employed_cleaned |>
  filter(dimension == "Gender") |>
  group_by(industry, year, race_gender) |>
  summarize(employ_n = sum(employ_n)) |>
  ggplot(aes(year, employ_n, color = race_gender)) +
  geom_line() +
  facet_wrap(~ industry, scales = "free_y") +
  expand_limits(y = 0) +
  scale_y_continuous(labels = comma) +
  labs(y = "# employed in industry",
       x = "Year",
       color = "Gender")

employed_cleaned |>
  filter(year == 2020,
         industry == "Wholesale and retail trade") |>
```
```{r chunk_02}
comparison <- employed_cleaned |>
  filter(year %in% c(2019, 2020)) |>
  mutate(major_occupation = paste(industry, major_occupation, sep = " - "),
         minor_occupation = paste(major_occupation, minor_occupation, sep = " - ")) |>
  gather(level, occupation, industry, major_occupation, minor_occupation) |>
  group_by(dimension, race_gender, level, occupation, year) |>
  summarize(employ_n = sum(employ_n)) |>
  ungroup() |>
  arrange(year) |>
  group_by(dimension, level, occupation, race_gender) |>
  summarize(ratio = last(employ_n) / first(employ_n),
            change = ratio - 1,
            employed_2019 = first(employ_n)) |>
  group_by(dimension, level, occupation) |>
  mutate(total_2019 = sum(employed_2019)) |>
  ungroup()

comparison |>

comparison |>
  filter(dimension == "Total", level == "industry") |>
  mutate(occupation = fct_reorder(occupation, change)) |>
  ggplot(aes(change, occupation)) +
  geom_col() +
  scale_x_continuous(labels = percent) +
  labs(title = "What industries suffered the most in 2020?",
       x = "Shift in # employed from 2019 to 2020 (BLS)",
       y = "")

library(glue)

compare_lollipop <- function(tbl) {
  tbl |>
    mutate(occupation = glue("{ occupation } ({ comma(total_2019 / 1000) }K)"),
           occupation = fct_reorder(occupation, change)) |>
    ggplot(aes(change, occupation)) +
    geom_errorbarh(aes(xmin = 0, xmax = change, color = race_gender),
                   height = 0,
                   position = position_dodge(width = .7)) +
    geom_point(aes(size = employed_2019, color = race_gender),
               position = position_dodge(width = .7)) +
    geom_vline(lty = 2, xintercept = 0) +
    scale_x_continuous(labels = percent) +
    scale_color_discrete(guide = guide_legend(reverse = TRUE)) +
    scale_size_continuous(labels = comma, guide = FALSE) +
    labs(x = "Shift in # employed from 2019 to 2020 (BLS)",
         y = "",
         color = "",
         size = "# employed 2019")
}

comparison |>
  filter(dimension == "Gender", level == "industry") |>
  compare_lollipop()

comparison |>
  filter(dimension == "Race", level == "industry") |>
  compare_lollipop() +
  labs(title = "What industries suffered the most in 2020?",
       subtitle = "Size of point represents # employed in industry in 2019")

comparison |>
  filter(dimension == "Race", level == "major_occupation") |>
  separate(occupation, c("industry", "occupation"), sep = " - ") |>
  filter(industry == "Construction",
         employed_2019 >= 10000) |>
  compare_lollipop()

comparison |>
  filter(dimension == "Gender", level == "major_occupation") |>
  separate(occupation, c("industry", "occupation"), sep = " - ") |>
  filter(industry == "Construction",
         employed_2019 >= 10000) |>
  compare_lollipop()

comparison |>
  filter(dimension == "Race",
         level == "major_occupation") |>
  separate(occupation, c("industry", "occupation"), sep = " - ") |>
  filter(industry == "Leisure and hospitality",
         employed_2019 >= 10000) |>
  compare_lollipop() +
  labs(title = "Within 'Leisure and hospitality', what occupations suffered?")

compare_2019_2020 |>
  filter(dimension == "Gender") |>
  mutate(industry = fct_reorder(industry, change)) |>
  ggplot(aes(change, industry, fill = race_gender)) +
  geom_col(position = "dodge") +
  scale_x_continuous(labels = percent) +
  scale_fill_discrete(guide = guide_legend(reverse = TRUE)) +
  labs(title = "What industries suffered the most in 2020?",
       x = "Shift in # employed from 2019 to 2020 (BLS)",
       y = "",
       fill = "Gender")

compare_2019_2020 |>
  filter(dimension == "Race") |>
  mutate(industry = fct_reorder(industry, change)) |>
  ggplot(aes(change, industry, fill = race_gender)) +
  geom_col(position = "dodge") +
  scale_x_continuous(labels = percent) +
  scale_fill_discrete(guide = guide_legend(reverse = TRUE)) +
  labs(title = "What industries suffered the most in 2020?",
       x = "Shift in # employed from 2019 to 2020 (BLS)",
       y = "",
       fill = "Race")

compare_2019_2020 |>
  filter(dimension == "Race")

compare_2019_2020 |>
  filter(dimension == "Gender") |>
  mutate(industry = fct_reorder(industry, change)) |>
  ggplot(aes(change, industry)) +
  geom_errorbarh(aes(xmin = 0, xmax = change, color = race_gender),
                 height = 0,
                 position = position_dodge(width = .7)) +
  geom_point(aes(size = employed_2019, color = race_gender),
             position = position_dodge(width = .7)) +
  geom_vline(lty = 2, xintercept = 0) +
  scale_x_continuous(labels = percent) +
  scale_color_discrete(guide = guide_legend(reverse = TRUE)) +
  scale_size_continuous(labels = comma, guide = FALSE) +
  labs(title = "What industries suffered the most in 2020?",
       subtitle = "Size of point represents # employed in industry in 2019",
       x = "Shift in # employed from 2019 to 2020 (BLS)",
       y = "",
       color = "Gender",
       size = "# employed 2019")
```
```{r chunk_03}
library(ggrepel)

compare_2019_2020 |>
  filter(dimension == "Total") |>
  mutate(industry = fct_reorder(industry, change)) |>
  ggplot(aes(employed_2019, change)) +
  geom_point() +
  geom_text_repel(aes(label = industry)) +
  geom_hline(lty = 2, yintercept = 0) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = percent)
```
```{r chunk_04}
employed_cleaned |>
  filter(year %in% c(2019, 2020)) |>
  filter(industry == "Construction") |>
  group_by(dimension, race_gender, major_occupation, year) |>
  summarize(employ_n = sum(employ_n)) |>
  arrange(year) |>
  summarize(ratio = last(employ_n) / first(employ_n),
            change = ratio - 1,
            employed_2019 = first(employ_n)) |>
  ungroup()
```

---

**→ [Hour 2: Modeling](h2_modeling_employment_earnings.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
