---
title: "Hour 1 — EDA: Analyzing the Big Mac index in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 062 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `animated_transitions` · `faceted_comparison` · `fct_reorder_stat` · `geom_smooth_loess` · `grouped_summary_ggplot` · `label_repulsion_ggrepel` · `log_scale` · `pairwise_token_correlation` · `pivot_reshape` · `purrr_map_iterate` · `spline_regression`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(ggrepel)
library(gganimate)
library(widyr)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
big-mac <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-12-22/big-mac.csv", show_col_types = FALSE)
glimpse(big-mac)
```

## Load the weekly Data
```{r chunk_01}
big_mac |>
  mutate(country = fct_reorder(country, local_price, function(.) last(.) / first(.))) |>
  ggplot(aes(date, local_price, color = country)) +
  geom_line() +
  expand_limits(y = 0) +
  facet_wrap(~ country, scales = "free_y") +
  theme(legend.position = "none") +
  labs(x = "Time",
       y = "Price of Big Mac in local currency")
```
```{r chunk_02}
big_mac |>
  group_by(country) |>
  summarize(big_mac_inflation = last(local_price) / first(local_price)) |>
  arrange(desc(big_mac_inflation)) |>
  mutate(country = fct_reorder(country, big_mac_inflation)) |>
  ggplot(aes(big_mac_inflation, country)) +
  geom_col() +
  geom_text(aes(label = paste0(round(big_mac_inflation, 1), "X")), hjust = 0) +
  scale_x_log10(breaks = c(1, 3, 10, 30, 100)) +
  expand_limits(x = 130) +
  labs(x = "Price of Big Mac in 2020 / Price of Big Mac in 2000",
       y = "")
```
```{r chunk_03}
big_mac |>
  group_by(date) |>
  mutate(usd_price = local_price[iso_a3 == "USA"],
         us_gdp = gdp_dollar[iso_a3 == "USA"]) |>
  ungroup() |>
  mutate(big_mac_ex = local_price / usd_price) |>
  select(date, iso_a3, country, local_price, dollar_ex, usd_price, big_mac_ex, usd_raw, gdp_dollar, us_gdp, usd_adjusted)
```
```{r chunk_04}
big_mac |>
  group_by(date) |>
  mutate(usd_price = local_price[iso_a3 == "USA"],
         us_gdp = gdp_dollar[iso_a3 == "USA"]) |>
  ungroup() |>
  filter(country == "Argentina",
         !is.na(gdp_dollar)) |>
  mutate(price_from_usd = usd_price * dollar_ex) |>
  ggplot(aes(date, local_price)) +
  geom_line(aes(color = "Price (in local currency)")) +
  geom_line(aes(y = price_from_usd, color = "Price from USD"))
```
```{r chunk_05}
big_mac |>
  filter(country != "United States") |>
  select(date, country, local_price, dollar_ex, usd_raw, gdp_dollar, usd_adjusted) |>
  filter(!is.na(gdp_dollar)) |>
  mutate(country = fct_reorder(country, usd_raw)) |>
  ggplot(aes(date, usd_adjusted)) +
  geom_line() +
  geom_hline(color = "red", lty = 2, yintercept = 0) +
  expand_limits(y = 0) +
  facet_wrap(~ country) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y = "GDP-adjusted Big Max Index relative to USD",
       x = "")

library(ggrepel)

big_mac |>
  filter(date == max(date)) |>
  ggplot(aes(gdp_dollar, usd_raw)) +
  geom_point() +
  geom_smooth(method = "lm") +
  geom_text_repel(aes(label = country)) +
  labs(x = "GDP per capita (dollars)",
       y = "Raw Big Max Index relative to USD")
```
```{r chunk_06}
big_mac |>
  filter(date == max(date)) |>
  ggplot(aes(gdp_dollar, usd_adjusted)) +
  geom_point() +
  geom_smooth(method = "lm") +
  geom_text_repel(aes(label = country)) +
  labs(x = "GDP per capita (dollars)",
       y = "Adjusted Big Max Index relative to USD")
```
```{r chunk_07}
library(gganimate)

big_mac |>
  filter(!is.na(gdp_dollar)) |>
  ggplot(aes(gdp_dollar, usd_adjusted)) +
  geom_point() +
  geom_smooth(method = "lm") +
  geom_text_repel(aes(label = country)) +
  transition_time(date) +
  labs(x = "GDP per capita (dollars)",
       y = "Adjusted Big Max Index relative to USD",
       title = "{frame_time }")
```
```{r chunk_08}
big_mac |>
  filter(date == max(date)) |>
  mutate(country = fct_reorder(country, usd_adjusted)) |>
  ggplot(aes(usd_adjusted, country)) +
  geom_col() +
  labs(x = "Big Mac Index relative to USD (GDP-adjusted)",
       y = "")
```
```{r chunk_09}
big_mac |>
  select(date, country, local_price, dollar_ex, usd_raw, gdp_dollar,
         ends_with("adjusted")) |>
  pivot_longer(ends_with("adjusted"), names_to = "base_currency",
               values_to = "adjusted") |>
  mutate(base_currency = str_to_upper(str_remove(base_currency, "_adjusted"))) |>
  filter(!is.na(gdp_dollar)) |>
  mutate(country = fct_reorder(country, adjusted)) |>
  ggplot(aes(date, adjusted, color = base_currency)) +
  geom_line() +
  geom_hline(color = "red", lty = 2, yintercept = 0) +
  expand_limits(y = 0) +
  facet_wrap(~ country) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y = "GDP-adjusted Big Max Index relative to each currency",
       x = "",
       color = "Base Currency")
```
```{r chunk_10}
library(widyr)

big_mac |>
  pairwise_cor(country, date, local_price, sort = TRUE) |>
  filter(item1 == "United States")
```

## Readme
```{r Readme}
tt
```

## Glimpse Data
```{r Glimpse}
tt |> 
  map(glimpse)
```

## Save Image
```{r chunk_11}
# This will save your most recent plot
ggsave(
  filename = "My TidyTuesday Plot.png",
  device = "png")
```

---

**→ [Hour 2: Modeling](h2_modeling_big_mac_index.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
