---
title: "Hour 1 — EDA: Analyzing data on women in the workplace in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 022 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`coord_flip_horizontal` · `fct_reorder_stat` · `grouped_summary_ggplot` · `interactive_plotly` · `scales_axis_labels` · `spline_regression` · `theme_set_global` · `time_series_trend`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(tidyverse)
library(scales)
library(plotly)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
library(scales)
theme_set(theme_light())

jobs_gender <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-03-05/jobs_gender.csv")
earnings_female <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-03-05/earnings_female.csv") 
employed_gender <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-03-05/employed_gender.csv")
```
```{r chunk_02}
summarize_jobs_gender <- function(tbl) {
  tbl |>
    summarize(total_earnings = sum(total_earnings * total_workers) / sum(total_workers),
              total_earnings_male = sum(total_earnings_male * workers_male, na.rm = TRUE) /
                sum(workers_male[!is.na(total_earnings_male)]),
              total_earnings_female = sum(total_earnings_female * workers_female, na.rm = TRUE) /
                sum(workers_female[!is.na(total_earnings_female)]),
              total_workers = sum(total_workers),
              workers_male = sum(workers_male),
              workers_female = sum(workers_female)) |>
    mutate(wage_percent_of_male = total_earnings_female / total_earnings_male)
}

by_year_major_category <- jobs_gender |>
  filter(complete.cases(.)) |>
  group_by(year, major_category) |>
  summarize_jobs_gender()

by_year_major_category |>
  mutate(major_category = fct_reorder(major_category, -total_earnings)) |>
  ggplot(aes(year, total_earnings, color = major_category)) +
  geom_line() +
  expand_limits(y = 0)

by_year_major_category |>
  mutate(major_category = fct_reorder(major_category, -wage_percent_of_male)) |>
  ggplot(aes(year, wage_percent_of_male, color = major_category)) +
  geom_line()
```
```{r chunk_03}
by_minor_category_2016 <- jobs_gender |>
  filter(year == 2016) |>
  group_by(major_category, minor_category) |>
  summarize_jobs_gender() |>
  ungroup()
```
```{r chunk_04}
by_minor_category_2016 |>
  mutate(minor_category = fct_reorder(minor_category, wage_percent_of_male)) |>
  ggplot(aes(minor_category, wage_percent_of_male, fill = major_category)) +
  geom_col() +
  coord_flip()
```
```{r chunk_05}
library(plotly)

p <- jobs_gender |>
  filter(year == 2016) |>
  filter(major_category == "Healthcare Practitioners and Technical") |>
  arrange(desc(wage_percent_of_male)) |>
  ggplot(aes(workers_female / total_workers,
             total_earnings,
             size = total_workers,
             label = occupation)) +
  geom_point() +
  scale_size_continuous(range = c(1, 10)) +
  labs(size = "Total # of workers",
       x = "% of workforce reported as female",
       y = "Median salary in the occupation") +
  scale_x_continuous(labels = percent_format()) +
  scale_y_continuous(labels = dollar_format()) +
  expand_limits(y = 0)

ggplotly(p)
```
```{r chunk_06}
p <- jobs_gender |>
  filter(year == 2016,
         total_workers >= 20000) |>
  filter(major_category == "Computer, Engineering, and Science") |>
  arrange(desc(wage_percent_of_male)) |>
  ggplot(aes(workers_female / total_workers,
             total_earnings_female / total_earnings_male,
             color = minor_category,
             size = total_workers,
             label = occupation)) +
  geom_point() +
  scale_size_continuous(range = c(1, 10)) +
  labs(size = "Total # of workers",
       x = "% of workforce reported as female",
       y = "% of median female salary / median male") +
  scale_x_continuous(labels = percent_format()) +
  scale_y_continuous(labels = percent_format())

ggplotly(p)
```

---

**→ [Hour 2: Modeling](h2_modeling_women_workplace.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
