---
title: "Hour 1 — EDA: Analyzing volcano eruptions in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 075 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`animated_transitions` · `fct_lump_infrequent` · `geom_histogram_density` · `glue_strings` · `grouped_summary_ggplot` · `log_scale` · `pivot_reshape` · `purrr_map_iterate` · `spline_regression` · `str_detect_filter` · `theme_set_global` · `tidytuesdayr_load`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(ggthemes)
library(leaflet)
library(glue)
library(htmlwidgets)
library(DT)
library(gganimate)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
eruptions <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-05-12/eruptions.csv", show_col_types = FALSE)
events <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-05-12/events.csv", show_col_types = FALSE)
sulfur <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-05-12/sulfur.csv", show_col_types = FALSE)
tree_rings <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-05-12/tree_rings.csv", show_col_types = FALSE)
volcano <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-05-12/volcano.csv", show_col_types = FALSE)
glimpse(eruptions)
glimpse(events)
```

## Explore
```{r chunk_02}
volcano <- volcano |>
  mutate(last_eruption_year = as.numeric(last_eruption_year))

eruptions <- eruptions
```
```{r chunk_03}
volcano |>
  count(evidence_category, sort = TRUE)

volcano |>
  mutate(years_ago = 2020 - last_eruption_year) |>
  ggplot(aes(years_ago + 1, fill = evidence_category)) +
  geom_histogram() +
  scale_x_log10()
```
```{r chunk_04}
volcano |>
  count(region, sort = TRUE)

volcano |>
  count(primary_volcano_type, sort = TRUE)
```
```{r chunk_05}
library(ggthemes)

volcano |>
  mutate(primary_volcano_type = str_remove(primary_volcano_type, "\\(.*\\)"),
         primary_volcano_type = fct_lump(primary_volcano_type, 6)) |>
  ggplot(aes(longitude, latitude)) +
  borders() +
  geom_point(aes(color = primary_volcano_type), size = .5) +
  theme_map() +
  labs(title = "Volcanoes of the World",
       color = "Type")
```
```{r chunk_06}
library(leaflet)
library(glue)
library(htmlwidgets)
library(DT)

template <- "<>{ volcano_name }</p><p>{ primary_volcano_type }</p>"

volcano |>
  mutate(transformed_pop = log2(population_within_5_km + 1),
         pop_color = colorNumeric(c("blue", "red"), transformed_pop)(transformed_pop)) |>
  gather(key, value,
         volcano_name, primary_volcano_type, last_eruption_year,
         country,
         tectonic_settings,
         population_within_5_km) |>
  mutate(key = str_to_title(str_replace_all(key, "_", " ")),
         key = paste0("<b>", key, "</b>")) |>
  replace_na(list(value = "Unknown")) |>
  nest(data = c(key, value)) |>
  mutate(html = map(data,
                    knitr::kable,
                    format = "html",
                    escape = FALSE,
                    col.names = c("", ""))) |>
  leaflet() |>
  addTiles() |>
  addCircleMarkers(lat = ~ latitude,
                   lng = ~ longitude,
                   color = ~ pop_color,
                   popup = ~ html,
                   radius = 1) |>
  addMeasure()
```
```{r chunk_07}
volcano |>
  ggplot(aes(population_within_5_km)) +
  geom_histogram() +
  scale_x_log10()
```
```{r chunk_08}
library(gganimate)

eruptions |>
  ggplot(aes(2020 - start_year)) +
  geom_histogram() +
  scale_x_log10()

eruptions |>
  filter(start_year >= 1900,
         eruption_category != "Discredited Eruption")
  ggplot(aes(2020 - start_year)) +
  geom_histogram() +
  scale_x_log10()
```
```{r chunk_09}
library(gganimate)

anim_graph <- eruptions |>
  filter(start_year >= 1900) |>
  mutate(volume = .00001 * 10 ^ vei) |>
  ggplot(aes(longitude, latitude)) +
  borders() +
  geom_point(aes(size = volume,
                 color = volume)) +
  theme_map() +
  scale_color_gradient2(low = "blue", high = "red",
                        midpoint = log10(.01),
                        trans = "log10",
                        guide = FALSE) +
  scale_size_continuous(range = c(1, 6)) +
  # transition_reveal(start_year) +
  transition_time(start_year) +
  labs(title = "Eruptions: Year { round(frame_time) }") +
  theme(legend.position = "none")

animate(anim_graph, 150, fps = 4)

anim_save("eruptions.gif")
```
```{r chunk_10}
eruptions
```

---

**→ [Hour 2: Modeling](h2_modeling_volcano_eruptions.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
