---
title: "Hour 1 — EDA: Analyzng scraping and analyzing ramen reviews in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 080 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`coord_flip_horizontal` · `error_bar_coef_plot` · `faceted_comparison` · `fct_lump_infrequent` · `fct_relevel_manual` · `fct_reorder_stat` · `grouped_summary_ggplot` · `janitor_clean` · `linear_model_broom` · `multi_table_join` · `network_graph_ggraph` · `pairwise_token_correlation`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- source HTML structure changes — selector-based scraping breaks silently
- text is multilingual without language detection
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- supplemental context is unavailable in the flat file and worth the scraping overhead
- text is long-form English and benefits from word-level decomposition
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(drlib)
library(broom)
library(tidytext)
library(rvest)
library(widyr)
library(igraph)
library(ggraph)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## R Markdown
```{r chunk_01}
library(tidyverse)
theme_set(theme_light())

ramen_ratings <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-06-04/ramen_ratings.csv")
```
```{r chunk_02}
library(drlib)

ramen_ratings_processed <- ramen_ratings |>
  mutate(style = fct_lump(style, 4),
         country = fct_lump(country, 12),
         brand = fct_lump(brand, 20)) |>
  replace_na(list(style = "Other")) |>
  mutate(brand = fct_relevel(brand, "Other"),
         country = fct_relevel(country, "Other"),
         style = fct_relevel(style, "Pack"))

ramen_ratings_processed |>
  gather(category, value, -review_number, -stars) |>
  count(category, value) |>
  group_by(category) |>
  top_n(20, n) |>
  ungroup() |>
  mutate(value = reorder_within(value, n, category)) |>
  ggplot(aes(value, n)) +
  geom_col() +
  facet_wrap(~ category, scales = "free_y") +
  scale_x_reordered() +
  coord_flip() +
  labs(title = "Categorical predictors (after processing)",
       x = "Predictor",
       y = "Count")
```
```{r chunk_03}
library(broom)

lm(stars ~ brand + country + style, ramen_ratings_processed) |>
  tidy(conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  arrange(desc(estimate)) |>
  extract(term, c("category", "term"), "^([a-z]+)([A-Z].*)") |>
  mutate(term = fct_reorder(term, estimate)) |>
  ggplot(aes(estimate, term, color = category)) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +
  geom_vline(lty = 2, xintercept = 0) +
  facet_wrap(~ category, ncol = 1, scales = "free_y") +
  theme(legend.position = "none") +
  labs(x = "Estimated effect on ramen rating",
       y = "",
       title = "Coefficients that predict ramen ratings",
       subtitle = "Less common brands and countries were lumped together as the reference level")
```
```{r chunk_04}
library(tidytext)

ramen_ratings_processed |>
  filter(!is.na(stars)) |>
  unnest_tokens(word, variety) |>
  group_by(word) |>
  summarize(avg_rating = mean(stars),
            n = n()) |>
  arrange(desc(n))
```

## Web scraping
```{r chunk_05}
library(rvest)

ramen_list <- read_html("https://www.theramenrater.com/resources-2/the-list/")

# How the original data was (probably) created
ramen_reviews <- ramen_list |>
  html_node("#myTable") |>
  html_table() |>
  tbl_df() |>
  janitor::clean_names() |>
  select(-t)
```
```{r chunk_06}
review_links <- read_html("https://www.theramenrater.com/resources-2/the-list/") |>
  html_nodes("#myTable a")

reviews <- tibble(review_number = parse_number(html_text(review_links)),
                  link = html_attr(review_links, "href"))
```
```{r chunk_07}
page <- read_html("https://www.theramenrater.com/2019/05/23/3180-yum-yum-moo-deng/")

get_review_text <- function(url) {
  message(url)
  
  read_html(url) |>
    html_nodes(".entry-content > p") |>
    html_text() |>
    str_subset(".")
}

review_text <- reviews |>
  head(250) |>
  mutate(text = map(link, possibly(get_review_text, character(0), quiet = FALSE)))
```
```{r chunk_08}
library(tidytext)

review_paragraphs <- review_text |> 
  filter(!map_lgl(text, is.null)) |>
  unnest() |>
  filter(str_detect(text, "Finished")) |>
  mutate(text = str_remove(text, "Finished.*?\\. "))

review_paragraphs_tokenized <- review_paragraphs |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = "word") |>
  filter(str_detect(word, "[a-z]")) |>
  inner_join(ramen_ratings, by = "review_number")

review_words <- review_paragraphs_tokenized |>
  filter(!is.na(stars)) |>
  group_by(word) |>
  summarize(number = n(),
            reviews = n_distinct(review_number),
            avg_rating = mean(stars)) |>
  arrange(desc(reviews))

review_words_filtered <- review_words |>
  filter(reviews < 200, reviews >= 10)

library(widyr)
word_cors <- review_paragraphs_tokenized |>
  semi_join(review_words_filtered, by = "word") |>
  distinct(review_number, word) |>
  pairwise_cor(word, review_number, sort = TRUE)
```
```{r chunk_09}
library(igraph)
library(ggraph)

set.seed(2019)

filtered_cors <- word_cors |>
  head(300)

nodes <- review_words_filtered |>
  filter(word %in% filtered_cors$item1 | word %in% filtered_cors$item2)

filtered_cors |>
  graph_from_data_frame(vertices = nodes) |>
  ggraph() +
  geom_edge_link() +
  geom_node_point(aes(size = reviews * 1.1)) +
  geom_node_point(aes(size = reviews, color = avg_rating)) +
  geom_node_text(aes(label = name), repel = TRUE) +
  scale_color_gradient2(low = "red", high = "blue", midpoint = 4) +
  theme_void() +
  labs(color = "Average rating",
       size = "# of reviews",
       title = "Network of words used together in ramen reviews",
       subtitle = "Based on 250 ramen reviews and their star ratings")
```

---

**→ [Hour 2: Modeling](h2_modeling_analyzng_scraping_ramen_reviews.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
