---
title: "Hour 2 — Modeling: Analyzing European energy in R"
subtitle: "Go deeper with statistics and models | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode: `european_energy` | [All episodes](https://thedatashrink.com.au/tidytuesday) | [YouTube](#) | [Instagram](#)
:::


::: {.callout-note collapse="true"}
## Atomic Concepts (DataX Index)
These reusable concepts are demonstrated in this episode. See `CONCEPT_INDEX.md`.

- `grouped_summary_ggplot`
- `pivot_reshape`
- `time_series_trend`
- `tokenization_pipeline`
:::


::: {.callout-warning collapse="true"}
## FAILS WHEN (DataX Stage 5)
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- text is multilingual without language detection
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN (DataX Stage 5)
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- text is long-form English and benefits from word-level decomposition
- time dimension is regular and spans multiple periods with meaningful change
:::


## Episode Context

- **Analysis type:** `EDA` — this shapes the modeling approach below
- **Pattern:** `load_inspect_reshape_tokenize_trend_compare_visualize`
- **Hour 2 goal:** Go deeper than David went in the original screencast

## Hour 2 Checklist
- [ ] TF-IDF or log-odds group comparison
- [ ] LDA / STM topic model (if corpus large enough)
- [ ] Bigram network graph with ggraph
- [ ] Sentiment arc if temporal dimension present


## TF-IDF Comparison

```{r}
#| label: tfidf
#| eval: false
library(tidytext)
# bind_tf_idf() on grouped token counts
```

## Log-Odds Ratio

```{r}
#| label: log-odds
#| eval: false
library(tidylo)
# bind_log_odds() for statistically grounded group comparison
```

## Topic Model (optional)

```{r}
#| label: topic-model
#| eval: false
library(stm)
# Fit STM / LDA if corpus is large enough
# k = 5-10 topics, display top terms per topic
```


## Model Story

> What does the model reveal that raw EDA couldn't?
> One paragraph for the YouTube "deep dive" section.

*[Write model narrative here]*

## Cross-Reference

- [← Hour 1 EDA](h1_eda_european_energy.qmd)
- [Hour 3 Communicate →](h3_communicate_european_energy.qmd)
