---
title: "Hour 1 — EDA: Analyzing college major & income data in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 020 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `boxplot_jitter` · `coord_flip_horizontal` · `error_bar_coef_plot` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_smooth_loess` · `grouped_summary_ggplot` · `interactive_plotly` · `label_repulsion_ggrepel` · `linear_model_broom` · `log_scale`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
:::

```{r setup}
library(here)
library(tidyverse)
library(scales)
library(ggrepel)
library(plotly)
library(broom)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_02}
recent_grads <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018-10-16/recent-grads.csv")

majors_processed <- recent_grads |>
  arrange(desc(Median)) |>
  mutate(Major = str_to_title(Major),
         Major = fct_reorder(Major, Median))
```
```{r by_major_category}
by_major_category <- majors_processed |>
  filter(!is.na(Total)) |>
  group_by(Major_category) |>
  summarize(Men = sum(Men),
            Women = sum(Women),
            Total = sum(Total),
            MedianSalary = sum(Median * Sample_size) / sum(Sample_size)) |>
  mutate(ShareWomen = Women / Total) |>
  arrange(desc(ShareWomen))
```

## What categories of majors make more money than others?
```{r chunk_03}
majors_processed |>
  mutate(Major_category = fct_reorder(Major_category, Median)) |>
  ggplot(aes(Major_category, Median, fill = Major_category)) +
  geom_boxplot() +
  scale_y_continuous(labels = dollar_format()) +
  expand_limits(y = 0) +
  coord_flip() +
  theme(legend.position = "none")
```

## What are the highest earning majors?
```{r chunk_04}
majors_processed |>
  filter(Sample_size >= 100) |>
  head(20) |>
  ggplot(aes(Major, Median, color = Major_category)) +
  geom_point() +
  geom_errorbar(aes(ymin = P25th, ymax = P75th)) +
  expand_limits(y = 0) +
  scale_y_continuous(labels = dollar_format()) +
  coord_flip() +
  labs(title = "What are the highest-earning majors?",
       subtitle = "Top 20 majors with at least 100 graduates surveyed. Bars represent the 25th to 75th percentile.",
       x = "",
       y = "Median salary of gradates")
```

## How does gender breakdown relate to typical earnings?
```{r chunk_05}
majors_processed |>
  arrange(desc(Total)) |>
  head(20) |>
  mutate(Major = fct_reorder(Major, Total)) |>
  gather(Gender, Number, Men, Women) |>
  ggplot(aes(Major, Number, fill = Gender)) +
  geom_col() +
  coord_flip()
```
```{r chunk_06}
library(ggrepel)

by_major_category |>
  mutate(Major_category = fct_lump(Major_category, 6)) |>
  ggplot(aes(ShareWomen, MedianSalary, color = Major_category)) +
  geom_point() +
  geom_smooth(method = "lm") +
  geom_text_repel(aes(label = Major_category), force = .2) +
  expand_limits(y = 0)
```
```{r chunk_07}
library(plotly)

g <- majors_processed |>
  mutate(Major_category = fct_lump(Major_category, 4)) |>
  ggplot(aes(ShareWomen, Median, color = Major_category, size = Sample_size, label = Major)) +
  geom_point() +
  geom_smooth(aes(group = 1), method = "lm") +
  scale_x_continuous(labels = percent_format()) +
  scale_y_continuous(labels = dollar_format()) +
  expand_limits(y = 0)

ggplotly(g)
```
```{r chunk_08}
majors_processed |>
  select(Major, Total, ShareWomen, Sample_size, Median) |>
  lm(Median ~ ShareWomen, data = ., weights = Sample_size) |>
  summary()
```
```{r chunk_09}
library(broom)

majors_processed |>
  select(Major, Major_category, Total, ShareWomen, Sample_size, Median) |>
  add_count(Major_category) |>
  filter(n >= 10) |>
  nest(-Major_category) |>
  mutate(model = map(data, ~ lm(Median ~ ShareWomen, data = ., weights = Sample_size)),
         tidied = map(model, tidy)) |>
  unnest(tidied) |>
  filter(term == "ShareWomen") |>
  arrange(estimate) |>
  mutate(fdr = p.adjust(p.value, method = "fdr"))
```
```{r chunk_10}
majors_processed |>
  filter(Sample_size >= 100) |>
  mutate(IQR = P75th - P25th) |>
  arrange(desc(IQR))
```

## Appendix
```{r chunk_11}
majors_processed |>
  ggplot(aes(Sample_size, Median)) +
  geom_point() +
  geom_text(aes(label = Major), check_overlap = TRUE, vjust = 1, hjust = 1) +
  scale_x_log10()
```
```{r chunk_12}
knitr::knit_exit()
```
```{r chunk_13}
majors_processed |>
  mutate(Major = fct_reorder(Major, Total)) |>
  arrange(desc(Total)) |>
  head(20) |>
  ggplot(aes(Major, Total, fill = Major_category)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = comma_format()) +
  labs(x = "",
       y = "Total # of graduates")
```
```{r chunk_14}
majors_processed |>
  group_by(Major_category) |>
  summarize(Median = median(Median)) |>
  mutate(Major_category = fct_reorder(Major_category, Median)) |>
  ggplot(aes(Major_category, Median)) +
  geom_col() +
  scale_y_continuous(labels = dollar_format()) +
  coord_flip()
```
```{r chunk_15}
majors_processed |>
  filter(Sample_size >= 100) |>
  tail(20) |>
  ggplot(aes(Major, Median, color = Major_category)) +
  geom_point() +
  geom_errorbar(aes(ymin = P25th, ymax = P75th)) +
  expand_limits(y = 0) +
  coord_flip()
```

---

**→ [Hour 2: Modeling](h2_modeling_college_major_income.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
