---
title: "Hour 1 — EDA: Analyzing extinct plants in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 028 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`faceted_comparison` · `fct_lump_infrequent` · `fct_relevel_manual` · `fct_reorder_stat` · `grouped_summary_ggplot` · `multi_table_join` · `purrr_map_iterate` · `reorder_within_facets` · `spline_regression` · `theme_set_global` · `tidytuesdayr_load` · `tokenization_pipeline`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- source HTML structure changes — selector-based scraping breaks silently
- text is multilingual without language detection
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- supplemental context is unavailable in the flat file and worth the scraping overhead
- text is long-form English and benefits from word-level decomposition
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(rvest)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
actions <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-08-18/actions.csv", show_col_types = FALSE)
plants <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-08-18/plants.csv", show_col_types = FALSE)
threats <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-08-18/threats.csv", show_col_types = FALSE)
glimpse(actions)
glimpse(plants)
```

## Readme
```{r Readme}
tt

plants <- tt$plants |>
  mutate(year_last_seen = fct_relevel(year_last_seen, "Before 1900")) |>
  separate(binomial_name, c("genus", "species"), sep = " ", remove = FALSE)

threats <- tt$threats |>
  filter(threatened == 1) |>
  select(-threatened) |>
  mutate(year_last_seen = fct_relevel(year_last_seen, "Before 1900")) |>
  separate(binomial_name, c("genus", "species"), sep = " ", remove = FALSE)

actions <- tt$actions |>
  filter(action_taken == 1) |>
  select(-action_taken) |>
  mutate(year_last_seen = fct_relevel(year_last_seen, "Before 1900")) |>
  separate(binomial_name, c("genus", "species"), sep = " ", remove = FALSE) |>
  filter(action_type != "Unknown")
```
```{r chunk_01}
plants |>

plants |>
  count(country = fct_lump(country, 10), sort = TRUE) |>
  mutate(country = fct_reorder(country, n)) |>
  ggplot(aes(n, country)) +
  geom_col()

plants |>
  count(continent, sort = TRUE) |>
  mutate(continent = fct_reorder(continent, n)) |>
  ggplot(aes(n, continent)) +
  geom_col()

plants |>
  count(red_list_category)

plants |>
  count(group, sort = TRUE)

plants |>
  filter(!is.na(year_last_seen)) |>
  count(year_last_seen, continent) |>
  ggplot(aes(year_last_seen, n, fill = continent)) +
  geom_col()

plants

by_continent_threat <- threats |>
  count(threat_type, continent, sort = TRUE) |>
  mutate(threat_type = fct_reorder(threat_type, n, sum))

by_continent_threat |>
  ggplot(aes(n, threat_type, fill = continent)) +
  geom_col() +
  labs(x = "# of plants facing this threat",
       y = "",
       fill = "Continent",
       title = "What are the most common threats to plants?")

by_continent_threat |>
  mutate(threat_type = reorder_within(threat_type, n, continent)) |>
  ggplot(aes(n, threat_type)) +
  geom_col() +
  scale_y_reordered() +
  facet_wrap(~ continent, scales = "free") +
  labs(x = "# of plants facing this threat",
       y = "",
       fill = "Continent",
       title = "What are the most common threats to plants by continent?")

threats |>
  filter(!is.na(year_last_seen)) |>
  count(year_last_seen, threat_type, continent) |>
  # filter(fct_lump(threat_type, 9, w = n) != "Other") |>
  mutate(threat_type = fct_reorder(threat_type, -n, sum)) |>
  ggplot(aes(year_last_seen, n, fill = continent)) +
  geom_col() +
  facet_wrap(~ threat_type) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Last seen",
       y = "# of plants extinct with this threat")
```
```{r chunk_02}
plants |>
  select(binomial_name) |>
  separate(binomial_name, c("genus", "species"), sep = " ") |>
  count(genus, sort = TRUE)

plants |>
  count(genus = fct_lump(genus, 9), year_last_seen) |>
  filter(genus != "Other") |>
  ggplot(aes(year_last_seen, n, fill = genus)) +
  geom_col() +
  facet_wrap(~ genus)
```
```{r chunk_03}
plants |>
  filter(red_list_category == "Extinct in the Wild") |>
  count(genus, sort = TRUE)

plants |>
  group_by(genus) |>
  summarize(extinct_in_the_wild = mean(red_list_category == "Extinct in the Wild"),
            n = n()) |>
  arrange(desc(n))
```
```{r chunk_04}
actions |>
  filter(action_type != "Unknown") |>
  count(action_type, sort = TRUE)

actions |>
  count(binomial_name, sort = TRUE) |>

plants |>
  count(continent, group, sort = TRUE) |>
  ggplot(aes(continent, n)) +
  geom_col() +
  facet_wrap(~ group, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

actions |>
  distinct(binomial_name, genus) |>
  count(genus, sort = TRUE)
```
```{r chunk_05}
library(rvest)

links <- read_html("http://www.orchidspecies.com/indexbulb.htm") |>
  html_nodes("li a")

link_text <- html_text(links)

bulbophyllum_links <- tibble(text = html_text(links),
       link = paste0("http://www.orchidspecies.com/", html_attr(links, "href"))) |>
  mutate(link_text = str_trim(link_text)) |>
  separate(link_text, c("genus", "species", "citation"), sep = " ", extra = "merge") |>
  filter(!is.na(citation))

plants |>
  inner_join(bulbophyllum_links, by = c("genus", "species")) |>
  select(binomial_name, country, continent, link) |>
  mutate(html = map(link, read_html)) |>
  mutate(image = map_chr(html, ~ html_attr(html_node(., "a img"), "src"))) |>
  transmute(binomial_name, image = paste0("http://www.orchidspecies.com/", image))
```

---

**→ [Hour 2: Modeling](h2_modeling_extinct_plants.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
