---
title: "Hour 1 — EDA: Analyzing train delays in France in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 069 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`boxplot_jitter` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_histogram_density` · `grouped_summary_ggplot` · `heatmap_tile` · `scales_axis_labels` · `spline_regression` · `theme_set_global` · `time_series_trend`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(tidyverse)
library(scales)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
library(scales)
theme_set(theme_light())

full_trains <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-26/full_trains.csv") |>
  mutate(pct_late_at_departure = num_late_at_departure / total_num_trips,
         arrival_station = str_to_title(arrival_station),
         departure_station = str_to_title(departure_station),
         date = as.Date(sprintf("%d-%02d-01", year, month))) |>
  arrange(departure_station, arrival_station, month) |>
  fill(service)
```
```{r chunk_02}
november_2018 <- full_trains |>
  filter(year == 2018, month == 11)

november_2018 |>
  ggplot(aes(pct_late_at_departure)) +
  geom_histogram(binwidth = .05) +
  scale_x_continuous(labels = percent_format())

november_2018 |>
  mutate(departure_station = fct_lump(departure_station, 3)) |>
  ggplot(aes(departure_station, pct_late_at_departure)) +
  geom_boxplot() +
  scale_y_continuous(labels = percent_format())

november_2018 |>
#  mutate(arrival_station = fct_infreq(fct_lump(arrival_station, prop = .01))) |>
#  mutate(departure_station = fct_infreq(fct_lump(departure_station, prop = .01))) |>
  mutate(arrival_station = fct_reorder(fct_lump(arrival_station, prop = .01), pct_late_at_departure)) |>
  mutate(departure_station = fct_reorder(fct_lump(departure_station, prop = .01), pct_late_at_departure)) |>
  group_by(arrival_station, departure_station) |>
  summarize(pct_late_at_departure = sum(num_late_at_departure) / sum(total_num_trips)) |>
  ggplot(aes(arrival_station, departure_station, fill = pct_late_at_departure)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", midpoint = .25, labels = percent_format()) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Arrival station",
       y = "Departure station",
       fill = "% late at departure",
       title = "Which routes have the most delayed trains in November 2018?",
       subtitle = "Stations with only one arriving/departing route were lumped into 'Other'")
```

## Over time
```{r chunk_03}
full_trains |>
  filter(departure_station == "Lyon Part Dieu") |>
  ggplot(aes(date, pct_late_at_departure, color = arrival_station)) +
  geom_line() +
  scale_y_continuous(labels = percent_format()) +
  expand_limits(y = 0)

by_departure_station_month <- full_trains |>
  group_by(departure_station = fct_lump(departure_station, prop = .01),
           date) |>
  summarize_at(vars(contains("num")), sum) |>
  ungroup() |>
  mutate(pct_late_at_departure = num_late_at_departure / total_num_trips)

by_departure_station_month |>
  mutate(departure_station = fct_reorder(departure_station, -pct_late_at_departure, last)) |>
  ggplot(aes(date, pct_late_at_departure, color = departure_station)) +
  geom_line() +
  scale_y_continuous(labels = percent_format()) +
  labs(x = "Month",
       y = "% late at departure",
       color = "Departure station")
```
```{r chunk_04}
by_departure_station_month <- full_trains |>
  group_by(departure_station = ifelse(service == "International",
                                      paste0(departure_station, " (International)"),
                                      departure_station),
           service,
           year,
           month = fct_reorder(month.name[month], month)) |>
  summarize_at(vars(contains("num")), sum) |>
  ungroup() |>
  mutate(pct_late_at_departure = num_late_at_departure / total_num_trips)

by_departure_station_month |>
  mutate(departure_station = fct_reorder(departure_station, (service != "International") + pct_late_at_departure, mean)) |>
  ggplot(aes(month, departure_station, fill = pct_late_at_departure)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", midpoint = .25, labels = percent_format()) +
  facet_wrap(~ year, nrow = 1, scales = "free_x") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  labs(fill = "% late at departure") +
  labs(x = "Month",
       y = "Departure station",
       title = "Which stations had delays in which months?",
       subtitle = "Ordered by the average delay, with international routes on the bottom")
```

---

**→ [Hour 2: Modeling](h2_modeling_train_delays_france.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
