---
title: "Hour 1 — EDA: Analyzing caribou locations in R"
subtitle: "Reproduce + extend David Robinson's original analysis | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode: `caribou_locations` | [All episodes](https://thedatashrink.com.au/tidytuesday) | [YouTube](#) | [Instagram](#)
:::


::: {.callout-note collapse="true"}
## Atomic Concepts (DataX Index)
These reusable concepts are demonstrated in this episode. See `CONCEPT_INDEX.md`.

- `geospatial_choropleth`
- `grouped_summary_ggplot`
- `time_series_trend`
:::


::: {.callout-warning collapse="true"}
## FAILS WHEN (DataX Stage 5)
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- region name spelling differs between data and shapefile — silent dropped rows
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN (DataX Stage 5)
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- geographic column can be matched to a standard ISO/FIPS shapefile without fuzzy joins
- time dimension is regular and spans multiple periods with meaningful change
:::


## Episode Context

- **Original:** `Analyzing caribou locations in R`
- **Analysis type:** `EDA`
- **Pattern:** `load_inspect_map_trend_compare_visualize`
- **Data:** TidyTuesday 2020-06-23 dataset
- **Hour 1 goal:** Reproduce David's core story, then extend with one new angle

::: {.callout-note collapse="true"}
## Original Transcript (preview)
```

```
:::

## Hour 1 Checklist
- [ ] Load data with load_tidyweek_data()
- [ ] Inspect schema and types
- [ ] Identify main grouping + metric columns
- [ ] Join to shapefile, plot choropleth
- [ ] Build 3-4 core ggplot charts telling the main story
- [ ] Write 150-word narrative summary

## Setup

```{r}
#| label: setup
#| eval: false
library(geosphere)
library(here)
library(lubridate)
library(sf)
library(tidyverse)
```

## Load Data

```{r}
#| label: load-data
#| eval: false
# Local TidyTuesday data
individuals <- read_csv(here::here("tidytuesday/data/2020/2020-06-23/individuals.csv"))
locations <- read_csv(here::here("tidytuesday/data/2020/2020-06-23/locations.csv"))
```

## Schema Inspection

```{r}
#| label: inspect
#| eval: false
# What columns do we have? What are the types? What is the grain?
glimpse(YOUR_DATA_HERE)
skimr::skim(YOUR_DATA_HERE)
```

## Core EDA

> Reproduce David's 3-4 key charts here. Follow the pattern_signature:
> `load_inspect_map_trend_compare_visualize`

```{r}
#| label: eda-core
#| eval: false
# YOUR EDA CODE HERE
# Follow the episode_source.R in this folder as your scaffold
```

## Extension: One New Angle

- Add a rolling 12-month average to smooth the trend line
- Add a second map showing year-over-year change (delta choropleth)

```{r}
#| label: eda-extension
#| eval: false
# YOUR EXTENSION CODE HERE
```

## Hour 1 Narrative

> Write 150 words: what does the data show? One sentence per chart.
> This becomes the YouTube intro and Instagram carousel slide 2.

*[Write narrative here]*

## Next Steps

- Hour 2: `h2_modeling_caribou_locations.qmd` — go deeper with a model
- Hour 3: `h3_communicate_caribou_locations.qmd` — publish to TheDataShrink™
