---
title: "Hour 1 — EDA: Analyzing cocktail recipes in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 017 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `grouped_summary_ggplot` · `label_repulsion_ggrepel` · `log_scale` · `network_graph_ggraph` · `pairwise_token_correlation` · `reorder_within_facets` · `scales_axis_labels` · `spline_regression`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- text is multilingual without language detection
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- text is long-form English and benefits from word-level decomposition
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(ggrepel)
library(widyr)
library(tidytext)
library(ggraph)
library(igraph)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
all_drinks <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-05-26/all_drinks.csv", show_col_types = FALSE)
boston_cocktails <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-05-26/boston_cocktails.csv", show_col_types = FALSE)
cocktails <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-05-26/cocktails.csv", show_col_types = FALSE)
mr-boston-flattened <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-05-26/mr-boston-flattened.csv", show_col_types = FALSE)
glimpse(all_drinks)
glimpse(boston_cocktails)
```

## Explore
```{r chunk_03}
mr_boston <- boston_cocktails

mr_boston |>
  count(name, sort = TRUE)

mr_boston |>
  count(ingredient, sort = TRUE) |>
  head(20) |>
  mutate(ingredient = fct_reorder(ingredient, n)) |>
  ggplot(aes(n, ingredient)) +
  geom_col() +
  labs(title = "Most common ingredients in Mr. Boston recipes")

n_recipes <- n_distinct(mr_boston$name)

mr_boston |>
  count(category, ingredient, sort = TRUE) |>
  mutate(category = fct_lump(category, 4),
         ingredient = fct_lump(ingredient, 20)) |>
  filter(ingredient != "Other") |>
  mutate(ingredient = fct_reorder(ingredient, n, sum)) |>
  ggplot(aes(n / n_recipes, ingredient, fill = category)) +
  geom_col() +
  scale_x_continuous(labels = percent_format()) +
  labs(title = "Most common ingredients in Mr. Boston recipes",
       x = "% of all recipes",
       y = "Ingredient",
       fill = "Category")
```

## Primary vs Secondary Ingredients
```{r chunk_04}
mr_boston |>
  count(measure, sort = TRUE)

# Data cleaning to turn this into a number
mr_boston_parsed <- mr_boston |>
  extract(measure, "amount", regex = "(.*) oz", remove = FALSE) |>
  extract(amount, "ones", regex = "(^\\d+$|^\\d+ )", convert = TRUE, remove = FALSE) |>
  extract(amount, c("numerator", "denominator"),
          regex = "(\\d+)\\/(\\d+)", convert = TRUE, remove = FALSE) |>
  replace_na(list(ones = 0, numerator = 0, denominator = 1)) |>
  mutate(oz = ones + numerator / denominator,
         oz = na_if(oz, 0))

ingredients_summarized <- mr_boston_parsed |>
  group_by(name) |>
  mutate(percentile = row_number() / n()) |>
  group_by(ingredient) |>
  summarize(n = n(),
            n_with_oz = sum(!is.na(oz)),
            avg_position = mean(percentile),
            avg_serving = mean(oz, na.rm = TRUE)) |>
  arrange(desc(n))
```
```{r chunk_05}
library(ggrepel)

ingredients_summarized |>
  head(30) |>
  ggplot(aes(avg_position, avg_serving)) +
  geom_point(aes(size = n / n_recipes)) +
  geom_text_repel(aes(label = ingredient)) +
  scale_x_continuous(labels = percent_format()) +
  scale_size_continuous(labels = percent_format()) +
  labs(y = "Average serving size",
       x = "Average position within drink",
       size = "% of recipes")
```
```{r chunk_06}
ingredients_summarized |>
  filter(n >= 10) |>
  ggplot(aes(n, avg_serving)) +
  geom_point() +
  scale_x_log10() +
  geom_text(aes(label = ingredient), check_overlap = TRUE,
            vjust = 1, hjust = 1) +
  expand_limits(y = 0)

ingredients_summarized |>
  filter(n_with_oz >= 10) |>
  arrange(desc(avg_serving)) |>
  head(12) |>
  mutate(ingredient = fct_reorder(ingredient, avg_serving)) |>
  ggplot(aes(avg_serving, ingredient)) +
  geom_col() +
  labs(title = "What ingredients have the largest serving size?",
       subtitle = "Only ones that appear in at least 10 recipes",
       x = "Average serving (oz)",
       y = "Ingredient")
```

## Clustering
```{r chunk_07}
library(widyr)
library(tidytext)

ingredient_pairs <- mr_boston |>
  add_count(ingredient) |>
  filter(n >= 10) |>
  pairwise_cor(ingredient, name, sort = TRUE)

ingredient_pairs |>
  filter(item1 %in% c("Gin", "Tequila", "Absinthe",
                      "Mezcal", "Bourbon whiskey",
                      "Vodka")) |>
  group_by(item1) |>
  top_n(10, correlation) |>
  mutate(item2 = reorder_within(item2, correlation, item1)) |>
  ggplot(aes(correlation, item2)) +
  geom_col() +
  facet_wrap(~ item1, scales = "free_y") +
  scale_y_reordered() +
  labs(title = "What ingredients are most correlated with particular ingredients?")
```
```{r chunk_08}
library(ggraph)
library(igraph)

top_cors <- ingredient_pairs |>
  head(150)

ingredient_info <- ingredients_summarized |>
  filter(ingredient %in% top_cors$item1)

top_cors |>
  graph_from_data_frame(vertices = ingredient_info) |>
  ggraph(layout = "fr") +
  geom_edge_link() +
  geom_node_text(aes(label = name), repel = TRUE) +
  geom_node_point(aes(size = 1.1 * n)) +
  geom_node_point(aes(size = n, color = avg_position)) +
  scale_color_gradient2(low = "red", high = "blue", midpoint = .5,
                        labels = scales::percent_format()) +
  labs(size = "# of recipes",
       color = "Avg position in drink",
       title = "The constellations of cocktail ingredients",
       subtitle = "Connected ingredients tend to appear in the same recipes. Red ingredients are early in the recipe, blue tend to be later")
```

## PCA
```{r chunk_09}
ingredient_svd <- mr_boston |>
  distinct(ingredient, name) |>
  mutate(value = 1) |>
  widely_svd(ingredient, name, value)

ingredient_svd |>
  filter(dimension > 1, dimension <= 5) |>
  mutate(dimension = paste0("PC", dimension)) |>
  group_by(dimension) |>
  top_n(16, abs(value)) |>
  mutate(ingredient = reorder_within(ingredient, value, dimension)) |>
  ggplot(aes(value, ingredient, fill = value > 0)) +
  geom_col(show.legend = FALSE) +
  scale_y_reordered() +
  facet_wrap(~ dimension, scales = "free_y") +
  labs(x = "Principal component value",
       y = "Ingredient",
       title = "What are the sources of variation in ingredients?")
```
```{r chunk_10}
recipe_svd <- mr_boston |>
  distinct(name, ingredient) |>
  mutate(value = 1) |>
  widely_svd(name, ingredient, value)

recipe_svd |>
  filter(dimension > 1, dimension <= 5) |>
  mutate(dimension = paste0("PC", dimension)) |>
  group_by(dimension) |>
  top_n(16, abs(value)) |>
  mutate(recipe = reorder_within(name, value, dimension)) |>
  ggplot(aes(value, recipe, fill = value > 0)) +
  geom_col(show.legend = FALSE) +
  scale_y_reordered() +
  facet_wrap(~ dimension, scales = "free_y") +
  labs(x = "Principal component value",
       y = "Ingredient",
       title = "What are the sources of variation in recipes?")
```

---

**→ [Hour 2: Modeling](h2_modeling_cocktail_recipes.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
