---
title: "Hour 1 — EDA: Analyzing Animal Crossing in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 004 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`broom_model_output` · `correlation_matrix` · `faceted_comparison` · `geom_histogram_density` · `log_scale` · `lubridate_dates` · `multi_table_join` · `pairwise_token_correlation` · `pivot_reshape` · `reorder_within_facets` · `sparse_matrix_text` · `spline_regression`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- text is multilingual without language detection
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- text is long-form English and benefits from word-level decomposition
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(tidytext)
library(lubridate)
library(widyr)
library(stm)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
theme_set(theme_light())

critic <- readr::read_tsv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-05/critic.tsv')
user_reviews <- readr::read_tsv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-05/user_reviews.tsv')
items <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-05/items.csv')
villagers <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-05/villagers.csv')
```
```{r chunk_02}
villagers
items |>
```

## Text Analysis
```{r chunk_03}
critic |>
  ggplot(aes(grade)) +
  geom_histogram()

user_reviews |>
  ggplot(aes(grade)) +
  geom_histogram()

head(user_reviews$text)
```
```{r chunk_04}
library(tidytext)
library(lubridate)

user_review_words <- user_reviews |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = "word") |>
  count(user_name, date, grade, word)
```
```{r chunk_05}
by_week <- user_reviews |>
  group_by(week = floor_date(date, "week", week_start = 1)) |>
  summarize(nb_reviews = n(),
            avg_grade = mean(grade),
            pct_zero = mean(grade == 0),
            pct_ten = mean(grade == 10))

by_week |>
  ggplot(aes(week, avg_grade)) +
  geom_line() +
  geom_point(aes(size = nb_reviews)) +
  expand_limits(y = 0) +
  labs(x = "Time",
       y = "Average grade",
       size = "# of reviews")

user_reviews |>
  ggplot(aes(grade)) +
  geom_histogram() +
  labs(title = "Most reviews were very low or very high")

by_week |>
  gather(type, value, contains("pct")) |>
  mutate(type = ifelse(type == "pct_zero", "% rated 0", "% rated 10")) |>
  ggplot(aes(week, value, color = type)) +
  geom_line() +
  geom_point(aes(size = nb_reviews)) +
  scale_y_continuous(labels = scales::percent) +
  expand_limits(y = 0) +
  labs(x = "Time",
       y = "% of reviews",
       size = "Total reviews in week",
       title = "Reviews got more polarizing in middle of game")
```
```{r chunk_06}
# user_reviews |>
#   mutate(start_text = str_sub(text, 1, pmin(30, str_length(text))),
#          regex = paste0(".", start_text, ".*")) |>
#   mutate(new_text = map2_chr(text, regex, str_remove)) |>
#   filter(str_length(new_text) != str_length(text))

by_word <- user_review_words |>
  group_by(word) |>
  summarize(avg_grade = mean(grade),
            nb_reviews = n()) |>
  arrange(desc(nb_reviews)) |>
  filter(nb_reviews >= 25) |>
  arrange(desc(avg_grade))
  
by_word |>
  filter(nb_reviews >= 75) |>
  ggplot(aes(nb_reviews, avg_grade)) +
  geom_point() +
  geom_text(aes(label = word), vjust = 1, hjust = 1, check_overlap = TRUE) +
  scale_x_log10()
```
```{r chunk_07}
by_word |>
  top_n(20, -avg_grade) |>
  ggplot(aes(nb_reviews, avg_grade)) +
  geom_point() +
  geom_text(aes(label = word), vjust = 1, hjust = 1, check_overlap = TRUE) +
  scale_x_log10() +
  labs(title = "What words were associated with low-grade reviews?",
       subtitle = "20 most negative words; only words in at least 25 reviews")
```
```{r chunk_08}
library(widyr)
library(stm)

review_matrix <- user_review_words |>
  group_by(word) |>
  filter(n() >= 25) |>
  cast_sparse(user_name, word, n)

topic_model_6 <- stm(review_matrix,
                     K = 6,
                     verbose = TRUE,
                     init.type = "Spectral",
                     emtol = 5e-5)
```
```{r chunk_09}
tidy(topic_model_6) |>
  group_by(topic) |>
  top_n(12, beta) |>
  mutate(term = reorder_within(term, beta, topic)) |>
  ggplot(aes(beta, term, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  scale_y_reordered() +
  facet_wrap(~ topic, scales = "free_y")

tidy(topic_model_6) |>
  filter(term == "progress")

topic_model_gamma <- tidy(topic_model_6, matrix = "gamma") |>
  mutate(user_name = rownames(review_matrix)[document]) |>
  inner_join(user_reviews, by = "user_name")

topic_model_gamma |>
  group_by(topic) |>
  top_n(1, gamma)

topic_model_gamma |>
  group_by(topic) |>
  summarize(correlation = cor(gamma, grade),
            spearman_correlation = cor(gamma, grade, method = "spearman"))

topic_model_gamma |>
  group_by(week = floor_date(date, "week", week_start = 1),
           topic) |>
  summarize(avg_gamma = mean(gamma)) |>
  ggplot(aes(week, avg_gamma, color = factor(topic))) +
  geom_line() +
  expand_limits(y = 0) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Time",
       y = "Average gamma (document-topic association)")
```

---

**→ [Hour 2: Modeling](h2_modeling_animal_crossing.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
