---
title: "Hour 1 — EDA: Analyzing historical phones in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 036 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`animated_transitions` · `confidence_ribbon` · `faceted_comparison` · `fct_relevel_manual` · `fuzzy_join` · `log_scale` · `multi_table_join` · `pivot_reshape` · `purrr_map_iterate` · `spline_regression` · `theme_set_global` · `tidytuesdayr_load`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- region name spelling differs between data and shapefile — silent dropped rows
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- geographic column can be matched to a standard ISO/FIPS shapefile without fuzzy joins
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(WDI)
library(fuzzyjoin)
library(gganimate)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
fixed-landline-telephone-subscriptions-vs-gdp-per-capita <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-11-10/fixed-landline-telephone-subscriptions-vs-gdp-per-capita.csv", show_col_types = FALSE)
landline <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-11-10/landline.csv", show_col_types = FALSE)
mobile-phone-subscriptions-vs-gdp-per-capita <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-11-10/mobile-phone-subscriptions-vs-gdp-per-capita.csv", show_col_types = FALSE)
mobile <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-11-10/mobile.csv", show_col_types = FALSE)
glimpse(fixed-landline-telephone-subscriptions-vs-gdp-per-capita)
glimpse(landline)
```

## Load the weekly Data
```{r chunk_01}
phones |>
  filter(country == "Sudan") |>
  ggplot(aes(year, subscriptions, color = type)) +
  geom_line()

summarize_subscriptions <- . |>
  filter(!is.na(subscriptions)) |>
  summarize(avg_subscriptions = mean(subscriptions),
            median_subscriptions = median(subscriptions),
            q25 = quantile(subscriptions, .25),
            q75 = quantile(subscriptions, .75))

phones |>
  filter(!is.na(subscriptions)) |>
  group_by(year, continent, type) |>
  summarize_subscriptions() |>
  ggplot(aes(year,
             avg_subscriptions,
             color = type)) +
  geom_line() +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = .25) +
  facet_wrap(~ continent) +
  labs(x = "Year",
       y = "Median subscriptions per person",
       color = "",
       title = "How do mobile and landline adoption differ between continents?",
       subtitle = "Ribbon shows the 25th-75th percentile range")
```
```{r chunk_02}
library(WDI)

country_incomes <- WDI(indicator = c(gdp_per_capita = "NY.GDP.PCAP.PP.KD",
                                     pop = "SP.POP.TOTL"),
                       start = 2005, end = 2005, extra = TRUE) |>
  as_tibble() |>
  select(code = iso3c, income, gdp_per_capita, pop) |>
  filter(!is.na(income)) |>
  mutate(income = fct_relevel(income, "Low income", "Lower middle income", "Upper middle income"))

by_year_income <- phones |>
  inner_join(country_incomes, by = "code") |>
  group_by(year, income, type) |>
  summarize_subscriptions()

by_year_income |>
  ggplot(aes(year,
             median_subscriptions,
             color = type)) +
  geom_line() +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = .25) +
  facet_wrap(~ income) +
  labs(x = "Year",
       y = "Median subscriptions per person",
       color = "",
       title = "How do mobile and landline adoption differ between income categories?",
       subtitle = "Ribbon shows the 25th-75th percentile range")

by_year_income |>
  mutate(income = fct_rev(income)) |>
  ggplot(aes(year,
             median_subscriptions,
             color = income)) +
  geom_line() +
  facet_wrap(~ type, ncol = 1) +
  labs(y = "Median subscriptions / 100 people", color = "Income")
```
```{r chunk_03}
mobile |>
  inner_join(landline |> select(code, year, total_pop, gdp_per_cap),
             by = c("code", "year"),
             suffix = c("_mobile", "_landline")) |>
  ggplot(aes(total_pop_mobile, total_pop_landline)) +
  geom_point() +
  geom_abline(color = "red")
```
```{r chunk_04}
country_sizes <- phones |>
  group_by(country) |>
  summarize(avg_population = mean(total_pop, na.rm = TRUE)) |>
  arrange(desc(avg_population))

phones |>
  semi_join(country_sizes |> top_n(50, avg_population), by = "country") |>
  ggplot(aes(year,
             subscriptions,
             color = type,
             group = interaction(type, country))) +
  geom_line() +
  facet_wrap(~ continent) +
  geom_hline(yintercept = 50, lty = 2) +
  labs(color = "")

countries_summarized <- phones |>
  filter(!is.na(subscriptions)) |>
  select(-total_pop, gdp_per_cap) |>
  pivot_wider(names_from = type, values_from = subscriptions) |>
  group_by(continent, country, code) |>
  summarize(year_past_50_mobile = na_if(min(year[Mobile >= 50]), Inf),
            peak_landline = max(Landline, na.rm = TRUE),
            peak_mobile = max(Mobile),
            n_mobile = sum(!is.na(Mobile))) |>
  ungroup() |>
  inner_join(country_incomes, by = "code") |>
  filter(n_mobile >= 25) |>
  arrange(desc(year_past_50_mobile))

countries_summarized |>
  ggplot(aes(gdp_per_capita,
             year_past_50_mobile,
             color = continent,
             size = pop)) +
  geom_point() +
  scale_x_log10(labels = dollar) +
  labs(x = "GDP per capita (in 2005)",
       y = "Year mobile subscriptions passed 50/100 people")

countries_summarized |>
  filter(!is.na(gdp_per_capita)) |>
  ggplot(aes(gdp_per_capita,
             peak_landline,
             color = continent,
             size = pop)) +
  geom_point() +
  scale_x_log10(labels = dollar) +
  labs(x = "GDP per capita (in 2005)",
       y = "Peak landline") +
  facet_wrap(~ continent) +
  theme(legend.position = "none")

countries_summarized |>
  filter(continent == "Africa") |>
  arrange(desc(gdp_per_capita)) |>
```
```{r chunk_05}
library(fuzzyjoin)

world_map_mobile <- map_data("world") |>
  as_tibble() |>
  regex_left_join(maps::iso3166, c(region = "mapname")) |>
  left_join(mobile, by = c(a3 = "code"))

library(gganimate)

world_map_mobile |>
  ggplot(aes(long, lat, group = group, fill = subscriptions)) +
  geom_polygon() +
  transition_manual(year) +
  coord_fixed(1.5) +
  scale_fill_gradient2(low = "blue", high = "red",
                       midpoint = 30) +
  ggthemes::theme_map() +
  labs(fill = "# of subscriptions")
```
```{r chunk_06}
phones |>
  arrange(desc(subscriptions)) |>
```

## Readme
```{r Readme}
tt
```

## Glimpse Data
```{r Glimpse}
tt |> 
  map(glimpse)
```

## Save Image
```{r chunk_07}
# This will save your most recent plot
ggsave(
  filename = "My TidyTuesday Plot.png",
  device = "png")
```

---

**→ [Hour 2: Modeling](h2_modeling_historical_phones.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
