---
title: "Hour 1 — EDA: Analyzing government spending on kids in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 033 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`faceted_comparison` · `fct_reorder_stat` · `grouped_summary_ggplot` · `pairwise_token_correlation` · `pivot_reshape` · `purrr_map_iterate` · `spline_regression` · `theme_set_global` · `tidytuesdayr_load` · `time_series_trend` · `us_census_tidycensus`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(ggthemes)
library(widyr)
library(tidycensus)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
kids <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-09-15/kids.csv", show_col_types = FALSE)
glimpse(kids)
```

## Load the weekly Data
```{r chunk_01}
library(ggthemes)

kids |>
  filter(variable == "PK12ed") |>
  group_by(year) |>
  summarize(total = 1000 * sum(inf_adj)) |>
  ggplot(aes(year, total)) +
  geom_line() +
  expand_limits(y = 0) +
  scale_y_continuous(labels = dollar)

kids |>
  filter(variable == "PK12ed") |>
        # state %in% sample(unique(state), 20)) |>
  mutate(state = fct_reorder(state, inf_adj_perchild, max, .desc = TRUE)) |>
  ggplot(aes(year, 1000 * inf_adj_perchild)) +
  geom_line() +
  geom_vline(xintercept = 2009, color = "red", lty = 2) +
  scale_y_continuous(labels = dollar) +
  expand_limits(y = 0) +
  facet_wrap(~ state) +
  theme_tufte() +
  labs(x = "Year",
       y = "Inflation-adjusted spending per child")

plot_faceted <- function(tbl, y_axis) {
  tbl |>
    mutate(state = fct_reorder(state, {{ y_axis }}, max, .desc = TRUE)) |>
    ggplot(aes(year, {{ y_axis }})) +
    geom_hline(yintercept = 0, color = "gray") +
    geom_line() +
    facet_wrap(~ state)
}

plot_change_faceted <- function(tbl) {
  tbl |>
    group_by(state, variable) |>
    mutate(change = inf_adj_perchild / first(inf_adj_perchild) - 1) |>
    ungroup() |>
    plot_faceted(change) +
    scale_y_continuous(labels = percent)
}
```
```{r chunk_02}
theme_set(theme_tufte())

kids |>
  filter(variable == "PK12ed") |>
  plot_faceted(inf_adj_perchild * 1000) +
  geom_vline(xintercept = 2009, color = "red", lty = 2) +
  scale_y_continuous(labels = dollar_format()) +
  labs(x = "",
       y = "Inflation-adjusted spending per child relative to 1997",
       title = "How has per-student K-12 spending changed per state?",
       subtitle = "Red line shows 2009 (global financial crisis)")

kids |>
  filter(variable == "PK12ed") |>
  plot_change_faceted() +
  geom_vline(xintercept = 2009, color = "red", lty = 2) +
  labs(x = "",
       y = "Increase in inflation-adjusted spending per child relative to 1997",
       title = "How has per-student K-12 spending changed per state?",
       subtitle = "Red line shows 2009 (global financial crisis)")

kids |>
  filter(variable == "highered") |>
  plot_faceted(inf_adj_perchild * 1000) +
  labs(y = "Spending per child")

kids |>
  filter(variable == "highered") |>
  plot_change_faceted() +
  labs(x = "",
       y = "Increase in inflation-adjusted spending per child relative to 1997",
       title = "How has higher education spending changed per state?",
       subtitle = "Red line shows 2009 (global financial crisis)")
```

## 2016 values
```{r chunk_03}
kids |>
  filter(year == 2016,
         variable %in% c("PK12ed", "highered")) |>
  arrange(desc(inf_adj_perchild)) |>
  mutate(state = fct_reorder(state, inf_adj_perchild, max)) |>
  ggplot(aes(inf_adj_perchild * 1000, state, fill = variable)) +
  geom_col(position = "dodge") +
  scale_x_continuous(labels = dollar) +
  labs(x = "Spending on K-12 per child in 2016") +
  scale_fill_discrete(guide = guide_legend(reverse = TRUE))
```
```{r chunk_04}
library(widyr)

kids |>
  filter(year == 2016,
         variable %in% c("PK12ed", "highered")) |>
  pivot_wider(names_from = variable, values_from = raw:inf_adj_perchild) |>
  ggplot(aes(inf_adj_perchild_PK12ed, inf_adj_perchild_highered)) +
  geom_point() +
  geom_text(aes(label = state), vjust = 1, hjust = 1) +
  scale_x_continuous(labels = dollar) +
  scale_y_continuous(labels = dollar) +
  labs(x = "K-12 spending per child (2016)",
       y = "Higher ed spending per child (2016)") +
  expand_limits(y = 0, x = 0)

kids |>
  filter(year == 2016) |>
  pairwise_cor(variable, state, inf_adj_perchild, sort = TRUE) |>
  filter(item1 == "PK12ed") |>
  mutate(item2 = fct_reorder(item2, correlation)) |>
  ggplot(aes(correlation, item2)) +
  geom_col() +
  labs(y = "Spending metric",
       x = "Correlation with K-12 education spending")
```
```{r chunk_05}
read_custom <- function(filename) {
  read_csv(filename) |>
    mutate(across(starts_with("20"), as.numeric))
}

gdp_by_state_raw <- dir("~/Downloads/SQGDP/", pattern = "SQGDP2.*.csv", full.names = TRUE) |>
  set_names(.) |>
  map_df(read_custom, .id = "filename")

by_year_state_gdp <- read_csv("~/Downloads/SQGDP/SQGDP1__ALL_AREAS_2005_2020.csv") |>
  pivot_longer(cols = starts_with("20"),
               names_to = "year_quarter",
               values_to = "gdp_millions") |>
  separate(year_quarter, c("year", "quarter"), sep = ":", convert = TRUE) |>
  group_by(state = GeoName, year, Description) |>
  summarize(gdp_millions = sum(gdp_millions))
```
```{r chunk_07}
# Extra credit! (Don't pay attention to this)
gdp_by_state_raw |>
  filter(Description == "All industry total") |>
  pivot_longer(cols = starts_with("20"),
               names_to = "year_quarter",
               values_to = "gdp_millions") |>
  separate(year_quarter, c("year", "quarter"), sep = ":", convert = TRUE) |>
  extract(filename, "state", "SQGDP2_(.*?)_20") |>
  filter(str_length(state) == 2) |>
  group_by(state, year) |>
  summarize(gdp_millions = sum(gdp_millions))
```

---

**→ [Hour 2: Modeling](h2_modeling_government_spending_kids.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
