---
title: "Hour 1 — EDA: Analyzing NCAA Women's Basketball"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 043 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`boxplot_jitter` · `correlation_matrix` · `faceted_comparison` · `fct_lump_infrequent` · `fct_relevel_manual` · `fct_reorder_stat` · `geom_smooth_loess` · `glue_strings` · `grouped_summary_ggplot` · `heatmap_tile` · `spline_regression` · `theme_set_global`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(glue)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
tournament <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-10-06/tournament.csv", show_col_types = FALSE)
glimpse(tournament)
```

## Load the weekly Data
```{r chunk_01}
tournament |>
  count(year) |> 
  ggplot(aes(year, n)) +
  geom_col()
```
```{r chunk_02}
tournament |>
  ggplot(aes(full_w, tourney_w + reg_w)) +
  geom_point(alpha = .1)

tournament |>
  filter(conf_l > reg_l)
```
```{r chunk_03}
by_seed <- tournament |>
  group_by(seed) |>
  summarize(n = n(),
            pct_win = mean(tourney_finish == "Champ"),
            pct_final_four = mean(tourney_finish %in% c("Champ", "N2nd", "NSF")),
            avg_round = mean(as.integer(tourney_finish)))

by_seed |>
  ggplot(aes(seed, avg_round)) +
  geom_line() +
  labs(x = "Starting seed",
       y = "On average, eliminated in round...")

by_seed |>
  ggplot(aes(seed, pct_final_four)) +
  geom_line() +
  labs(x = "Starting seed",
       y = "% of times making the Final Four")
```
```{r chunk_04}
tournament |>
  count(seed, tourney_finish) |>
  group_by(seed) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(tourney_finish, seed, fill = pct)) +
  geom_tile() +
  geom_text(aes(label = percent(pct))) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_reverse(breaks = seq(1, 16), expand = c(0, 0)) +
  scale_fill_gradient2(high = "blue", labels = percent) +
  theme(panel.grid = element_blank(),
        axis.ticks = element_blank()) +
  labs(x = "Tournament finish (worst to best)",
       y = "Starting seed",
       fill = "% of seed",
       title = "When a team starts in a seed, how do they end up?")
```
```{r chunk_05}
tournament |>
  ggplot(aes(seed, n_rounds)) +
  geom_point(alpha = .1)

tournament |>
  group_by(year) |>
  summarize(correlation = cor(seed, n_rounds)) |>
  ggplot(aes(year, correlation)) +
  geom_line()
```
```{r chunk_06}
tournament |>
  ggplot(aes(seed, reg_percent)) +
  geom_boxplot(aes(group = seed)) +
  geom_smooth(method = "loess")

tournament |>
  mutate(conference = fct_lump(conference, 8)) |>
  ggplot(aes(seed, reg_percent)) +
  geom_jitter(width = .05) +
  # geom_boxplot(aes(group = seed)) +
  geom_smooth(method = "loess") +
  facet_wrap(~ conference)
```
```{r chunk_07}
tournament |>
  ggplot(aes(reg_percent, n_rounds)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ seed)

tournament |>
  group_by(seed) |>
  summarize(correlation = cor(reg_percent, n_rounds))
```

## Conferences
```{r chunk_08}
summarize_entries <- . |>
  summarize(n_entries = n(),
            avg_seed = mean(seed),
            pct_win = mean(tourney_finish == "Champ"),
            pct_final_four = mean(tourney_finish %in% c("Champ", "N2nd", "NSF")),
            avg_round = mean(n_rounds)) |>
  arrange(desc(n_entries))

by_conference <- tournament |>
  group_by(conference) |>
  summarize_entries()

by_conference |>
  filter(n_entries >= 25) |>
  ggplot(aes(n_entries, avg_round)) +
  geom_point() +
  geom_text(aes(label = conference), vjust = 1, hjust = 1,
            check_overlap = TRUE) +
  expand_limits(x = 0)
```
```{r chunk_09}
library(glue)

by_school <- tournament |>
  group_by(school) |>
  summarize_entries()

by_school |>
  filter(n_entries >= 15) |>
  mutate(school = glue("{ school } ({ n_entries})"),
         school = fct_reorder(school, avg_round)) |>
  ggplot(aes(avg_round, school)) +
  geom_col() +
  labs(title = "Of schools that frequently play, which perform best?",
       subtitle = "Parentheses show the # of years out of the last 25 they played",
       x = "Average # of rounds won",
       y = "")
```

---

**→ [Hour 2: Modeling](h2_modeling_ncaa_women_s_basketball.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
