---
title: "Hour 1 — EDA: Analyzing space launches in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 056 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `boxplot_jitter` · `coord_flip_horizontal` · `countrycode_standardize` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `grouped_summary_ggplot` · `multi_table_join` · `spline_regression` · `theme_set_global` · `time_series_trend`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(countrycode)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
theme_set(theme_light())

agencies <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-01-15/agencies.csv")
launches <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-01-15/launches.csv")
```
```{r chunk_02}
launches |>
  count(launch_year, agency_type) |>
  ggplot(aes(launch_year, n, color = agency_type)) +
  geom_line() +
  labs(x = "Time",
       y = "# of launches this year",
       color = "Agency type")
```
```{r chunk_03}
library(countrycode)

agencies |>

launches |>
  count(agency_type, agency, sort = TRUE)

agencies |>
  count(state_code, wt = count, sort = TRUE) |>

launches_processed <- launches |>
  filter(launch_date <= Sys.Date()) |>
  mutate(state_code_cleaned = fct_collapse(
    state_code,
    "RU" = c("SU", "RU"),
    "FR" = "F",
    "JP" = "J",
    "IT" = "I"
  )) |>
  mutate(state_name = countrycode(state_code_cleaned, "iso2c", "country.name"),
         state_name = fct_lump(state_name, 6)) |>
  replace_na(list(state_name = "Other"))

launches_processed |>
  count(launch_year, state_name) |>
  mutate(state_name = fct_reorder(state_name, -n, sum)) |>
  ggplot(aes(launch_year, n, color = state_name)) +
  geom_line() +
  labs(x = "Time",
       y = "Launches per year",
       color = "Responsible state",
       title = "Launches per year per country",
       subtitle = "Combines Soviet Union (pre-1990) with Russia")
```

## Focus on private + startup launches
```{r chunk_04}
agencies |>
  filter(agency_type %in% c("private", "startup")) |>

private_startup_launches <- launches_processed |>
  filter(agency_type %in% c("private", "startup")) |>
  inner_join(agencies |>
               select(agency, agency_name = name, short_name, parent), by = "agency") |>
  mutate(agency_name_lumped = fct_lump(agency_name, 6),
         agency_name_lumped = if_else(agency_name_lumped == "Other" & state_name == "United States",
                                     "Other US", as.character(agency_name_lumped)))

private_startup_launches |>
  count(agency_name_lumped, state_name, sort = TRUE) |>
  mutate(agency_name_lumped = fct_reorder(agency_name_lumped, n, sum)) |>
  ggplot(aes(agency_name_lumped, n, fill = state_name)) +
  geom_col() +
  coord_flip() +
  labs(x = "",
       y = "# of launches overall",
       title = "What private/startup agencies have had the most launches?",
       fill = "Country")

private_startup_launches |>
  count(agency_name_lumped,
        decade = 5 * (launch_year %/% 5)) |>
  complete(agency_name_lumped, decade, fill = list(n = 0)) |>
  mutate(agency_name_lumped = fct_reorder(agency_name_lumped, -n, sum)) |>
  ggplot(aes(decade, n, color = agency_name_lumped)) +
  geom_line() +
  facet_wrap(~ agency_name_lumped) +
  theme(legend.position = "none") +
  labs(x = "Time",
       y = "# of launches in 5 year period")
```
```{r chunk_05}
vehicles <- launches_processed |>
  group_by(type, state_name) |>
  summarize(first_launch = min(launch_year),
            last_launch = max(launch_year),
            launches = n()) |>
  ungroup()

russian_vehicles <- vehicles |>
  filter(state_name == "Russia") |>
  arrange(desc(launches)) |>
  filter(launches >= 30)

launches_processed |>
  semi_join(russian_vehicles, by = "type") |>
  mutate(type = fct_reorder(type, launch_date, min)) |>
  ggplot(aes(launch_date, type)) +
  geom_jitter(color = "blue", alpha = .25, width = 0, height = .2) +
  theme(legend.position = "none") +
  labs(title = "Timeline of Soviet/Russian space vehicles",
       x = "Launch date",
       y = "Vehicle type",
       subtitle = "Only vehicles with at least 30 launches")

launches_processed |>
  filter(state_code == "US") |>
  add_count(type) |>
  filter(n >= 20) |>
  mutate(type = fct_reorder(type, launch_date, min),
         agency_type = str_to_title(agency_type)) |>
  ggplot(aes(launch_date, type, color = agency_type)) +
  geom_jitter(alpha = .25, width = 0, height = .2) +
  labs(title = "Timeline of US space vehicles",
       x = "Launch date",
       y = "Vehicle type",
       color = "Agency type",
       subtitle = "Only vehicles with at least 20 launches")

by_type |>
  arrange(desc(launches)) |>
  count(type, sort = TRUE)
```

---

**→ [Hour 2: Modeling](h2_modeling_space_launches.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
