---
title: "Hour 1 — EDA: Analyzing pet names in Seattle in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 049 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`coord_flip_horizontal` · `crossing_simulation_grid` · `faceted_comparison` · `fct_reorder_stat` · `geom_histogram_density` · `grouped_summary_ggplot` · `lubridate_dates` · `multi_table_join` · `scales_axis_labels` · `spline_regression` · `theme_set_global` · `time_series_trend`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(tidyverse)
library(lubridate)
library(scales)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
library(lubridate)
theme_set(theme_light())

seattle_pets <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-03-26/seattle_pets.csv") |>
  mutate(license_issue_date = mdy(license_issue_date)) |>
  rename(animal_name = animals_name)
```
```{r chunk_02}
seattle_pets |>
  filter(license_issue_date >= "2017-01-01") |>
  count(species, primary_breed, sort = TRUE) |>
  filter(species %in% c("Cat", "Dog")) |>
  mutate(percent = n / sum(n)) |>
  group_by(species) |>
  top_n(10, percent) |>
  ungroup() |>
  mutate(primary_breed = fct_reorder(primary_breed, percent)) |>
  ggplot(aes(primary_breed, percent, fill = species)) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent_format()) +
  facet_wrap(~ species, scales = "free_y", ncol = 1) +
  coord_flip() +
  labs(x = "Primary breed",
       y = "% of this species",
       title = "Most common cat and dog breeds",
       subtitle = "Of licensed pets in Seattle 2017-2018")
```
```{r chunk_03}
dogs <- seattle_pets |>
  filter(species == "Dog")

name_counts <- dogs |>
  group_by(animal_name) |>
  summarize(name_total = n()) |>
  filter(name_total >= 100)

breed_counts <- dogs |>
  group_by(primary_breed) |>
  summarize(breed_total = n()) |>
  filter(breed_total >= 200)

total_dogs <- nrow(dogs)

name_breed_counts <- dogs |>
  count(primary_breed, animal_name) |>
  complete(primary_breed, animal_name, fill = list(n = 0)) |>
  inner_join(name_counts, by = "animal_name") |>
  inner_join(breed_counts, by = "primary_breed")

# One-sided hypergeometric p-value
hypergeom_test <- name_breed_counts |>
  mutate(percent_of_breed = n / breed_total,
         percent_overall = name_total / total_dogs) |>
  mutate(overrepresented_ratio = percent_of_breed / percent_overall) |>
  arrange(desc(overrepresented_ratio)) |>
  mutate(hypergeom_p_value = 1 - phyper(n, name_total, total_dogs - name_total, breed_total),
         holm_p_value = p.adjust(hypergeom_p_value),
         fdr = p.adjust(hypergeom_p_value, method = "fdr"))

hypergeom_test |>
  filter(fdr < .05)

ggplot(hypergeom_test, aes(hypergeom_p_value)) +
  geom_histogram(binwidth = .1) +
  labs(x = "One-sided hypergeometric p-values for overrepresented name")
```
```{r chunk_04}
crossing(name_total = c(100, 200, 300),
         breed_total = seq(200, 1000, 25)) |>
  mutate(max_p_value = 1 - phyper(0, name_total, total_dogs - name_total, breed_total)) |>
  ggplot(aes(breed_total, max_p_value, color = factor(name_total))) +
  geom_line() +
  labs(x = "Total # of dogs in breed",
       y = "Maximum one-sided p-value",
       color = "# of dogs with name")
```
```{r chunk_05}
library(scales)

hypergeom_test |>
  filter(fdr <= .05) |>
  arrange(fdr) |>
  transmute(`Breed` = primary_breed,
            `Name` = animal_name,
            `# of dogs with name` = n,
            `% of breed` = percent(percent_of_breed),
            `% overall` = percent(percent_overall),
            `FDR-adjusted one-sided p-value` = fdr) |>
  knitr::kable()
```

---

**→ [Hour 2: Modeling](h2_modeling_pet_names_seattle.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
