---
title: "Super QMD: Analyzing Broadway shows in R"
subtitle: "Cognition-first blueprint built from the episode guide and KER"
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show scaffold code"
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

## What This Document Is

This super QMD is the bridge between:

- the reproduction artifact in `output/episodes/013_broadway_shows/episode_guide.qmd`
- the KER entry for `broadway_shows`
- a final guided analysis document like `bechdel_ker_guided_analysis.qmd`

The point is to preserve not only the code path, but the author's cognition:

- what question gets asked first
- how the dataset is inspected
- when transformation happens
- why one comparison is chosen before another
- which pieces should become reusable helpers

## Episode Snapshot

- `queue_position`: `13`
- `function_name`: `broadway_shows`
- `analysis_type`: `EDA`
- `pattern_signature`: `load_inspect_trend_compare_visualize`
- `data_source`: TidyTuesday 2020-04-28 dataset
- `reproduction_status`: `ready_local`

::: {.callout-note}
## Analysis Intent
Explore distributions, group differences, and notable patterns in Broadway shows through cleaning, summarization, and visualization
:::



## Transcript Preview

Analyzing Broadway shows in R David Robinson 15.9k subscribers 103 Share Save Clip 2,906 views Streamed live on 29 Apr 2020 Tidy Tuesday R Screencasts I'll analyze a dataset on Broadway shows in R, without looking at the data in advance. I'll be analyzing the dataset as a live stream instead of prerecording it. The dataset comes from the Tidy Tuesday project: https://github.com/rfordatascience/ti... Search in video Introduction 0:06 hi everyone I'm Dave Robinson and welcome to another live screencast where 0:12 I'l...

## Cognitive Signature

- Start by naming the question and the unit of analysis before choosing plots or metrics.
- Use schema inspection to decide which table carries the main story and which tables are support tables.
- Start broad with time or trend structure when it helps explain which later comparisons are fair.
- End with one communication-ready chart or table that carries the episode's main claim.

## Data and Execution Context

### Local Assets

- `tidytuesday/data/2020/2020-04-28/cpi.csv`
- `tidytuesday/data/2020/2020-04-28/grosses.csv`
- `tidytuesday/data/2020/2020-04-28/pre-1985-starts.csv`
- `tidytuesday/data/2020/2020-04-28/synopses.csv`

### Execution Notes

- local TidyTuesday files were matched, so you can use `load_tidyweek_data()` and avoid a GitHub PAT

### External Dependencies

- No external dependencies were captured.

## Setup Scaffold

```{r}
#| label: setup
#| eval: false

library(tidyverse)
source("R/screencast_helpers.R")

# Optional episode packages:
# library(scales)
# library(shinymetrics)
# library(tidymetrics)

tuesdata <- load_tidyweek_data("2020-04-28")
names(tuesdata)
grosses <- tuesdata$grosses
```

## Thought -> Code -> Data Bridge

### Thought Process Map

- `T1 | Context | Frame the episode as exploratory analysis of Broadway shows with limited advance knowledge of the data. | Transcript title and opening context`
- `T2 | Inspect | Load the data, inspect table shape, and identify the main entities, measures, and comparison axes. | Initial data load and early counts or prints`
- `T6 | Compare | Summarize group differences, rates, rankings, or trajectories to surface the main story in the data. | Grouped summaries and comparative plots`
- `T7 | Visualize | Turn the cleaned summaries into interpretable visual comparisons and iterate toward the clearest view. | ggplot workflow`
- `T8 | Reflect | Identify reusable workflow pieces that can transfer to future datasets with a similar shape or question. | End-state of the analysis pattern`

### Code Mapping Table

- `T1 | Data load and inspection | tt_load | Establish source shape, units, and comparison candidates`
- `T6 | Group comparison | summarize, mutate | Compare categories, time periods, or entities`
- `T7 | Visualization | ggplot, geom_col, geom_line | Communicate the core result visually`

### Data Transformation Flow

- `Raw | Imported source tables or files | Load source data into tibbles or lists | Raw episode objects | Establish the starting analysis surface`
- `Raw -> Clean | Raw source tables | Filter, rename, and derive analysis columns | Cleaned table | Make the data comparable and plot-ready`
- `Clean -> Analytical | Cleaned table | Group, summarize, rank, or compare entities | Summary tables or rate calculations | Surface the core analytical story`
- `Analytical -> Visual | Summary or model outputs | Reorder factors and prepare plotting fields | Plot-ready table | Turn the analytical result into an interpretable chart`

## Guided Build Sequence

## Stage 1: Context

Frame the episode as exploratory analysis of Broadway shows with limited advance knowledge of the data.

::: {.callout-note}
## Cognitive Move
Define the analytical question, likely unit of analysis, and what would count as an interesting result.

Trigger:

- Transcript title and opening context


:::

```{r}
#| label: stage_01_context
#| eval: false

# Write the one-sentence analytical question for this episode.
# Topic: Broadway shows
# Unit of analysis:
# Main comparison axis:
# What would make the result interesting?
```
## Stage 2: Inspect

Load the data, inspect table shape, and identify the main entities, measures, and comparison axes.

::: {.callout-note}
## Cognitive Move
Inspect shape, columns, missingness, and table roles before committing to transformations.

Trigger:

- Initial data load and early counts or prints


:::

```{r}
#| label: stage_02_inspect
#| eval: false

# Start here once data is loaded.
# names(tuesdata)
# purrr::imap_dfr(tuesdata, ~ tibble(table = .y, rows = nrow(.x), cols = ncol(.x)))
# glimpse(analysis_tbl)
```
## Stage 3: Compare

Summarize group differences, rates, rankings, or trajectories to surface the main story in the data.

::: {.callout-note}
## Cognitive Move
Summarize groups or entities in a way that reveals differences, rankings, or trends without overfitting the story.

Trigger:

- Grouped summaries and comparative plots


:::

```{r}
#| label: stage_03_compare
#| eval: false

# Summarize the comparison that best answers the question.
# comparison_tbl <- analysis_tbl %>%
#   group_by(...) %>%
#   summarise(...) %>%
#   arrange(desc(...))
```
## Stage 4: Visualize

Turn the cleaned summaries into interpretable visual comparisons and iterate toward the clearest view.

::: {.callout-note}
## Cognitive Move
Choose the plot after the summary table exists, then optimize for clarity rather than novelty.

Trigger:

- ggplot workflow

Additional guidance:

- When time is present, lead with the broad trend before showing narrower subgroup comparisons.
:::

```{r}
#| label: stage_04_visualize
#| eval: false

# Turn the comparison table into one clear communication plot.
# ggplot(comparison_tbl, aes(...)) +
#   geom_col() +
#   labs(title = ..., subtitle = ...)
```
## Stage 5: Reflect

Identify reusable workflow pieces that can transfer to future datasets with a similar shape or question.

::: {.callout-note}
## Cognitive Move
Translate the episode-specific move into a reusable function, helper, or workflow note.

Trigger:

- End-state of the analysis pattern


:::

```{r}
#| label: stage_05_reflect
#| eval: false

# Capture the reusable move after the analysis is clear.
# Candidate helper:
# - input:
# - output:
# - why it is reusable:
```

## Reusable Layer

### Reusable Patterns

- `inspect_source_schema | Inspect source tables, field meanings, and likely comparison axes | Raw dataset or dataset list | Initial analysis plan | Y`
- `plot_comparative_story | Reorder entities and visualize the main comparison cleanly | Summary table or model output | Communication-ready plot | Y`

### Function Candidates

- `plot_comparative_story(data, x_col, y_col, group_col = NULL)`

## How To Finish This Super QMD

1. Replace each scaffold chunk with the real episode-specific analysis code.
2. After each major result, add a short callout explaining what changed in your thinking.
3. Keep one broad chart early, one comparison chart in the middle, and one final communication chart at the end.
4. Move any repeated transformation or comparison logic into `R/screencast_helpers.R`.
5. Once the document reads like a guided walkthrough instead of a reproduction note, promote it into a finished KER-guided analysis.

## Adaptation Layer

- `1 | inspect_source_schema | Start by mapping entities, measures, and likely comparison axes before transforming anything`
- `8 | plot_comparative_story | Choose the clearest visual comparison once the analytical table is ready`
