---
title: "Super QMD: Analyzing squirrels in NYC in R"
subtitle: "Cognition-first blueprint built from the episode guide and KER"
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show scaffold code"
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

## What This Document Is

This super QMD is the bridge between:

- the reproduction artifact in `output/episodes/057_squirrels_nyc/episode_guide.qmd`
- the KER entry for `squirrels_nyc`
- a final guided analysis document like `bechdel_ker_guided_analysis.qmd`

The point is to preserve not only the code path, but the author's cognition:

- what question gets asked first
- how the dataset is inspected
- when transformation happens
- why one comparison is chosen before another
- which pieces should become reusable helpers

## Episode Snapshot

- `queue_position`: `57`
- `function_name`: `squirrels_nyc`
- `analysis_type`: `Modeling`
- `pattern_signature`: `load_inspect_map_model_evaluate_visualize`
- `data_source`: TidyTuesday 2019-10-29 source files
- `reproduction_status`: `ready_local`

::: {.callout-note}
## Analysis Intent
Explore geographic variation in squirrels in NYC through joins, summarization, and visualization
:::



## Transcript Preview

analyzing squirrels in NYC David Robinson 15.9k subscribers 141 Share Save Clip 4,837 views 2 Nov 2019 Tidy Tuesday R Screencasts I analyze a dataset of squirrels in NYC as an example of exploratory data analysis and machine learning in R, performed without looking at the data in advance. This leads to some visualization of spatial patterns in squirrel behavior, including importing shapefiles and plotting them with ggplot2. This data comes from the #tidytuesday project: https://github.com/rfordatascience/ti... You...

## Cognitive Signature

- Start by naming the question and the unit of analysis before choosing plots or metrics.
- Use schema inspection to decide which table carries the main story and which tables are support tables.
- Do descriptive work before modeling; the model should organize what the exploratory work already suggested.
- Aggregate to the geographic unit before drawing maps so the visual carries a clean analytical message.
- End with one communication-ready chart or table that carries the episode's main claim.

## Data and Execution Context

### Local Assets

- `tidytuesday/data/2019/2019-10-29/nyc_squirrels.csv`

### Execution Notes

- local TidyTuesday files were matched, so you can use `load_tidyweek_data()` and avoid a GitHub PAT

### External Dependencies

- No external dependencies were captured.

## Setup Scaffold

```{r}
#| label: setup
#| eval: false

library(tidyverse)
source("R/screencast_helpers.R")

# Optional episode packages:
# library(sf)

tuesdata <- load_tidyweek_data("2019-10-29")
names(tuesdata)
```

## Thought -> Code -> Data Bridge

### Thought Process Map

- `T1 | Context | Frame the episode as exploratory analysis of squirrels in NYC with limited advance knowledge of the data. | Transcript title and opening context`
- `T2 | Inspect | Load the data, inspect table shape, and identify the main entities, measures, and comparison axes. | Initial data load and early counts or prints`
- `T6 | Compare | Move from descriptive patterns into statistical or predictive modeling once the feature space is ready. | Modeling functions and evaluation code`
- `T7 | Visualize | Translate the analytical result into geographic views to reveal regional variation and outliers. | Spatial joins or mapping code`
- `T8 | Reflect | Identify reusable workflow pieces that can transfer to future datasets with a similar shape or question. | End-state of the analysis pattern`

### Code Mapping Table

- `T1 | Data load and inspection | read_csv, count | Establish source shape, units, and comparison candidates`
- `T6 | Modeling and evaluation | lm, glm | Test explanatory or predictive relationships after EDA`
- `T7 | Visualization | ggplot, geom_point, geom_sf | Communicate the core result visually`

### Data Transformation Flow

- `Raw | Imported source tables or files | Load source data into tibbles or lists | Raw episode objects | Establish the starting analysis surface`
- `Raw -> Clean | Raw source tables | Filter, rename, and derive analysis columns | Cleaned table | Make the data comparable and plot-ready`
- `Clean -> Analytical | Feature table | Engineer predictors and fit models | Model outputs, predictions, or coefficients | Quantify relationships beyond descriptive summaries`
- `Analytical -> Visual | Analytical summaries with geographic identifiers | Join map geometry or region lookup data | Map-ready plotting table | Display geographic variation`

## Guided Build Sequence

## Stage 1: Context

Frame the episode as exploratory analysis of squirrels in NYC with limited advance knowledge of the data.

::: {.callout-note}
## Cognitive Move
Define the analytical question, likely unit of analysis, and what would count as an interesting result.

Trigger:

- Transcript title and opening context


:::

```{r}
#| label: stage_01_context
#| eval: false

# Write the one-sentence analytical question for this episode.
# Topic: squirrels in NYC
# Unit of analysis:
# Main comparison axis:
# What would make the result interesting?
```
## Stage 2: Inspect

Load the data, inspect table shape, and identify the main entities, measures, and comparison axes.

::: {.callout-note}
## Cognitive Move
Inspect shape, columns, missingness, and table roles before committing to transformations.

Trigger:

- Initial data load and early counts or prints


:::

```{r}
#| label: stage_02_inspect
#| eval: false

# Start here once data is loaded.
# names(tuesdata)
# purrr::imap_dfr(tuesdata, ~ tibble(table = .y, rows = nrow(.x), cols = ncol(.x)))
# glimpse(analysis_tbl)
```
## Stage 3: Compare

Move from descriptive patterns into statistical or predictive modeling once the feature space is ready.

::: {.callout-note}
## Cognitive Move
Summarize groups or entities in a way that reveals differences, rankings, or trends without overfitting the story.

Trigger:

- Modeling functions and evaluation code

Additional guidance:

- If you plan to model, use this step to identify which predictors deserve to enter the first baseline model.
:::

```{r}
#| label: stage_03_compare
#| eval: false

# Summarize the comparison that best answers the question.
# comparison_tbl <- analysis_tbl %>%
#   group_by(...) %>%
#   summarise(...) %>%
#   arrange(desc(...))
```
## Stage 4: Visualize

Translate the analytical result into geographic views to reveal regional variation and outliers.

::: {.callout-note}
## Cognitive Move
Choose the plot after the summary table exists, then optimize for clarity rather than novelty.

Trigger:

- Spatial joins or mapping code


:::

```{r}
#| label: stage_04_visualize
#| eval: false

# Turn the comparison table into one clear communication plot.
# ggplot(comparison_tbl, aes(...)) +
#   geom_col() +
#   labs(title = ..., subtitle = ...)
```
## Stage 5: Reflect

Identify reusable workflow pieces that can transfer to future datasets with a similar shape or question.

::: {.callout-note}
## Cognitive Move
Translate the episode-specific move into a reusable function, helper, or workflow note.

Trigger:

- End-state of the analysis pattern


:::

```{r}
#| label: stage_05_reflect
#| eval: false

# Capture the reusable move after the analysis is clear.
# Candidate helper:
# - input:
# - output:
# - why it is reusable:
```

## Reusable Layer

### Reusable Patterns

- `inspect_source_schema | Inspect source tables, field meanings, and likely comparison axes | Raw dataset or dataset list | Initial analysis plan | Y`
- `build_predictive_or_explanatory_model | Move from cleaned features to fitted models and evaluation outputs | Feature table and outcome field | Model object and diagnostics | Y`
- `prepare_geographic_summary | Aggregate metrics by region and join map geometry or lookup tables | Regional metrics and identifiers | Map-ready data | Y`
- `plot_comparative_story | Reorder entities and visualize the main comparison cleanly | Summary table or model output | Communication-ready plot | Y`

### Function Candidates

- `build_predictive_pipeline(data, formula, ...)`
- `evaluate_episode_model(model, test_data, outcome_col)`
- `prepare_map_summary(data, region_col, value_col)`
- `plot_geographic_metric(map_data, value_col)`
- `plot_comparative_story(data, x_col, y_col, group_col = NULL)`

## How To Finish This Super QMD

1. Replace each scaffold chunk with the real episode-specific analysis code.
2. After each major result, add a short callout explaining what changed in your thinking.
3. Keep one broad chart early, one comparison chart in the middle, and one final communication chart at the end.
4. Move any repeated transformation or comparison logic into `R/screencast_helpers.R`.
5. Once the document reads like a guided walkthrough instead of a reproduction note, promote it into a finished KER-guided analysis.

## Adaptation Layer

- `1 | inspect_source_schema | Start by mapping entities, measures, and likely comparison axes before transforming anything`
- `6 | build_predictive_or_explanatory_model | Fit and evaluate a baseline model after the core feature engineering is stable`
- `7 | prepare_geographic_summary | Aggregate to the geographic unit available in the new dataset and rebuild the map join`
- `8 | plot_comparative_story | Choose the clearest visual comparison once the analytical table is ready`
