---
title: "Super QMD: Analyzing UN votes in R"
subtitle: "Cognition-first blueprint built from the episode guide and KER"
author: "Andrei Stoian"
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show scaffold code"
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

## What This Document Is

This super QMD is the bridge between:

- the reproduction artifact in `output/episodes/071_un_votes/episode_guide.qmd`
- the KER entry for `un_votes`
- a final guided analysis document like `bechdel_ker_guided_analysis.qmd`

The point is to preserve not only the code path, but the author's cognition:

- what question gets asked first
- how the dataset is inspected
- when transformation happens
- why one comparison is chosen before another
- which pieces should become reusable helpers

## Episode Snapshot

::: {.callout-note}
## Analysis Intent
Explore and compare language patterns in UN votes using tokenization, grouped counts, and discriminative text analysis
:::

| Field | Value |
|---|---|
| Queue | #71 |
| Function | `un_votes` |
| Analysis type | EDA |
| Data source | TidyTuesday 2021-03-23 dataset |
| Status | `ready_local` |

**Packages:** `WDI`, `countrycode`, `fuzzyjoin`, `ggthemes`, `lubridate`, `scales`, `tidytext`, `tidytuesdayR`, `tidyverse`, `widyr`

**Features:** ggplot, joins, reshape, text_analysis, maps, time, multi_table

**Execution notes:**
- interactive `View()` calls are present and should be skipped or replaced in automated runs
- local TidyTuesday files were matched, so you can use `load_tidyweek_data()` and avoid a GitHub PAT

**Analytical pattern:** `load_inspect_reshape_join_tokenize_map_trend_compare_visualize`





## Cognitive Signature

- Start by naming the question and the unit of analysis before choosing plots or metrics.
- Use schema inspection to decide which table carries the main story and which tables are support tables.
- Build one analysis-ready table before comparing groups so the narrative rests on a stable grain.
- Lock the grouping variable first, then tokenize text and compare language only after the corpus structure is clear.
- Aggregate to the geographic unit before drawing maps so the visual carries a clean analytical message.
- Start broad with time or trend structure when it helps explain which later comparisons are fair.
- End with one communication-ready chart or table that carries the episode's main claim.

## Data and Execution Context

### Local Assets

- `tidytuesday/data/2021/2021-03-23/issues.csv`
- `tidytuesday/data/2021/2021-03-23/roll_calls.csv`
- `tidytuesday/data/2021/2021-03-23/unvotes.csv`

### Execution Notes

- interactive `View()` calls are present and should be skipped or replaced in automated runs
- local TidyTuesday files were matched, so you can use `load_tidyweek_data()` and avoid a GitHub PAT

### External Dependencies

- No external dependencies were captured.

## Setup Scaffold

```{r}
#| label: setup
#| eval: false

library(tidyverse)
source("R/screencast_helpers.R")

# Optional episode packages:
# library(WDI)
# library(countrycode)
# library(fuzzyjoin)
# library(ggthemes)
# library(lubridate)
# library(scales)
# library(tidytext)
# library(tidytuesdayR)
# library(widyr)

tuesdata <- load_tidyweek_data("2021-03-23")
names(tuesdata)
```

## Thought -> Code -> Data Bridge

### Thought Process Map

- `T1 | Context | Frame the episode as exploratory analysis of UN votes with limited advance knowledge of the data. | Transcript title and opening context`
- `T2 | Inspect | Load the data, inspect table shape, and identify the main entities, measures, and comparison axes. | Initial data load and early counts or prints`
- `T4 | Transform | Reshape fields, standardize column meanings, and join supporting tables to create analysis-ready records. | Join and reshaping operations in the Rmd`
- `T5 | Compare | Tokenize text, remove noise terms, and compare vocabulary across groups or documents. | tidytext workflow in the episode`
- `T6 | Compare | Summarize group differences, rates, rankings, or trajectories to surface the main story in the data. | Grouped summaries and comparative plots`
- `T7 | Visualize | Translate the analytical result into geographic views to reveal regional variation and outliers. | Spatial joins or mapping code`
- `T8 | Reflect | Identify reusable workflow pieces that can transfer to future datasets with a similar shape or question. | End-state of the analysis pattern`

```{r}
#| label: thought_execution_log
#| eval: false

# Fill in `finding` after completing each stage.
thought_log <- tibble::tribble(
  ~step,  ~move,        ~finding,
  "T1", "Context", "",
  "T2", "Inspect", "",
  "T4", "Transform", "",
  "T5", "Compare", "",
  "T6", "Compare", "",
  "T7", "Visualize", "",
  "T8", "Reflect", "",
)
thought_log
```


### Code Mapping Table

- `T1 | Data load and inspection | tt_load, count | Establish source shape, units, and comparison candidates`
- `T4 | Structuring and enrichment | spread, unnest, left_join, inner_join | Convert messy inputs into analysis-ready tables`
- `T5 | Text comparison | unnest_tokens, anti_join | Identify terms that distinguish groups or documents`
- `T6 | Group comparison | group_by, summarize, count, mutate | Compare categories, time periods, or entities`
- `T7 | Visualization | ggplot, geom_col, geom_point, geom_line, facet_wrap | Communicate the core result visually`

### Data Transformation Flow

- `Raw | Imported source tables or files | Load source data into tibbles or lists | Raw episode objects | Establish the starting analysis surface`
- `Raw -> Clean | Raw source tables | Rename, reshape, unnest, and join supporting data | Analysis-ready rows with consistent keys | Standardize the schema before comparison`
- `Clean -> Analytical | Clean text corpus | Tokenize text, remove stop words, and count terms by group | Term-frequency comparison tables | Surface language differences and signatures`
- `Analytical -> Visual | Analytical summaries with geographic identifiers | Join map geometry or region lookup data | Map-ready plotting table | Display geographic variation`

## Guided Build Sequence

## Stage 1: Context

Frame the episode as exploratory analysis of UN votes with limited advance knowledge of the data.

::: {.callout-note}
## Cognitive Move
Define the analytical question, likely unit of analysis, and what would count as an interesting result.

Trigger:

- Transcript title and opening context


:::

```{r}
#| label: stage_01_context
#| eval: false

# Write the one-sentence analytical question for this episode.
# Topic: UN votes
# Unit of analysis:
# Main comparison axis:
# What would make the result interesting?
```
## Stage 2: Inspect

Load the data, inspect table shape, and identify the main entities, measures, and comparison axes.

::: {.callout-note}
## Cognitive Move
Inspect shape, columns, missingness, and table roles before committing to transformations.

Trigger:

- Initial data load and early counts or prints


:::

```{r}
#| label: stage_02_inspect
#| eval: false

# Start here once data is loaded.
# names(tuesdata)
# purrr::imap_dfr(tuesdata, ~ tibble(table = .y, rows = nrow(.x), cols = ncol(.x)))
# glimpse(analysis_tbl)
```
## Stage 3: Transform

Reshape fields, standardize column meanings, and join supporting tables to create analysis-ready records.

::: {.callout-note}
## Cognitive Move
Reshape, join, or normalize only enough to make the next comparison possible.

Trigger:

- Join and reshaping operations in the Rmd


:::

```{r}
#| label: stage_03_transform
#| eval: false

# Build the canonical analysis table for this episode.
# Source assets: tidytuesday/data/2021/2021-03-23/issues.csv, tidytuesday/data/2021/2021-03-23/roll_calls.csv, tidytuesday/data/2021/2021-03-23/unvotes.csv
# Suggested verbs: mutate(), separate_rows(), pivot_longer(), left_join(), unnest()
# analysis_tbl <- raw_tbl %>%
# ...
```
## Stage 4: Compare

Tokenize text, remove noise terms, and compare vocabulary across groups or documents.

::: {.callout-note}
## Cognitive Move
Summarize groups or entities in a way that reveals differences, rankings, or trends without overfitting the story.

Trigger:

- tidytext workflow in the episode

Additional guidance:

- For text work, compare grouped term patterns only after you know what counts as a document and a group.
:::

```{r}
#| label: stage_04_compare
#| eval: false

# Summarize the comparison that best answers the question.
# comparison_tbl <- analysis_tbl %>%
# group_by(...) %>%
# summarise(...) %>%
# arrange(desc(...))
```
## Stage 5: Compare

Summarize group differences, rates, rankings, or trajectories to surface the main story in the data.

::: {.callout-note}
## Cognitive Move
Summarize groups or entities in a way that reveals differences, rankings, or trends without overfitting the story.

Trigger:

- Grouped summaries and comparative plots

Additional guidance:

- For text work, compare grouped term patterns only after you know what counts as a document and a group.
:::

```{r}
#| label: stage_05_compare
#| eval: false

# Summarize the comparison that best answers the question.
# comparison_tbl <- analysis_tbl %>%
# group_by(...) %>%
# summarise(...) %>%
# arrange(desc(...))
```
## Stage 6: Visualize

Translate the analytical result into geographic views to reveal regional variation and outliers.

::: {.callout-note}
## Cognitive Move
Choose the plot after the summary table exists, then optimize for clarity rather than novelty.

Trigger:

- Spatial joins or mapping code

Additional guidance:

- When time is present, lead with the broad trend before showing narrower subgroup comparisons.
:::

```{r}
#| label: stage_06_visualize
#| eval: false

# Turn the comparison table into one clear communication plot.
# ggplot(comparison_tbl, aes(...)) +
# geom_col() +
# labs(title = ..., subtitle = ...)
```
## Stage 7: Reflect

Identify reusable workflow pieces that can transfer to future datasets with a similar shape or question.

::: {.callout-note}
## Cognitive Move
Translate the episode-specific move into a reusable function, helper, or workflow note.

Trigger:

- End-state of the analysis pattern


:::

```{r}
#| label: stage_07_reflect
#| eval: false

# Capture the reusable move after the analysis is clear.
# Candidate helper:
# - input:
# - output:
# - why it is reusable:
```

## Reusable Layer

### Reusable Patterns

- `inspect_source_schema | Inspect source tables, field meanings, and likely comparison axes | Raw dataset or dataset list | Initial analysis plan | Y`
- `reshape_analysis_inputs | Convert wide or nested source fields into normalized long-form records | Raw tabular fields | Tidy analysis table | Y`
- `join_context_tables | Enrich the main table with lookup or metadata tables | Primary data plus keys | Context-enriched analysis table | Y`
- `compare_group_vocabulary | Tokenize text and compute grouped term distinctiveness | Text, grouping variable, document or entity ids | Ranked term comparison tables | Y`
- `prepare_geographic_summary | Aggregate metrics by region and join map geometry or lookup tables | Regional metrics and identifiers | Map-ready data | Y`
- `plot_comparative_story | Reorder entities and visualize the main comparison cleanly | Summary table or model output | Communication-ready plot | Y`

### Function Candidates

- `reshape_episode_inputs(data, ...)`
- `join_episode_context(primary_data, lookup_data, by)`
- `prepare_text_tokens(data, text_column)`
- `compare_group_vocabulary(data, group_col, term_col, count_col = 'n')`
- `prepare_map_summary(data, region_col, value_col)`
- `plot_geographic_metric(map_data, value_col)`
- `plot_comparative_story(data, x_col, y_col, group_col = NULL)`

## How To Finish This Super QMD

1. Replace each scaffold chunk with the real episode-specific analysis code.
2. After each major result, add a short callout explaining what changed in your thinking.
3. Keep one broad chart early, one comparison chart in the middle, and one final communication chart at the end.
4. Move any repeated transformation or comparison logic into `R/screencast_helpers.R`.
5. Once the document reads like a guided walkthrough instead of a reproduction note, promote it into a finished KER-guided analysis.

## Adaptation Layer

- `1 | inspect_source_schema | Start by mapping entities, measures, and likely comparison axes before transforming anything`
- `3 | reshape_analysis_inputs | Normalize wide, nested, or repeated fields into one analysis-ready row structure`
- `4 | join_context_tables | Enrich the main analysis table with keys, metadata, or auxiliary measurements`
- `5 | compare_group_vocabulary | Tokenize the new corpus and compare terms across groups or documents`
- `7 | prepare_geographic_summary | Aggregate to the geographic unit available in the new dataset and rebuild the map join`
- `8 | plot_comparative_story | Choose the clearest visual comparison once the analytical table is ready`
