---
title: "Super QMD: Analyzing wealth and income in R"
subtitle: "Cognition-first blueprint built from the episode guide and KER"
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show scaffold code"
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

## What This Document Is

This super QMD is the bridge between:

- the reproduction artifact in `output/episodes/077_wealth_income/episode_guide.qmd`
- the KER entry for `wealth_income`
- a final guided analysis document like `bechdel_ker_guided_analysis.qmd`

The point is to preserve not only the code path, but the author's cognition:

- what question gets asked first
- how the dataset is inspected
- when transformation happens
- why one comparison is chosen before another
- which pieces should become reusable helpers

## Episode Snapshot

- `queue_position`: `77`
- `function_name`: `wealth_income`
- `analysis_type`: `Predictive`
- `pattern_signature`: `load_inspect_trend_simulate_model_evaluate_visualize`
- `data_source`: TidyTuesday 2021-05-18 dataset
- `reproduction_status`: `needs_external_sources`

::: {.callout-note}
## Analysis Intent
Predict or estimate outcomes in wealth and income using cleaned features, exploratory analysis, and model evaluation
:::

::: {.callout-important}
## Status Note
This episode currently depends on external data retrieval. The cognitive structure is ready, but the execution path still needs remote source handling before it becomes locally reproducible.
:::


## Transcript Preview

Analyzing wealth and income in R David Robinson 15.9k subscribers 148 Share Save Clip 4,552 views Streamed live on 10 Feb 2021 Tidy Tuesday R Screencasts I'll analyze a dataset about wealth and income inequality over time, without looking at the dataset in advance. The dataset comes from the Tidy Tuesday project: https://github.com/rfordatascience/ti... Code: https://github.com/dgrtwo/data-screen... Timestamped annotations of specific tricks, tips and tools used: https://github.com/dgrtwo/data-screen... Introductio...

## Cognitive Signature

- Start by naming the question and the unit of analysis before choosing plots or metrics.
- Use schema inspection to decide which table carries the main story and which tables are support tables.
- Do descriptive work before modeling; the model should organize what the exploratory work already suggested.
- Start broad with time or trend structure when it helps explain which later comparisons are fair.
- End with one communication-ready chart or table that carries the episode's main claim.

## Data and Execution Context

### Local Assets

- No local assets were captured yet.

### Execution Notes

- this episode depends on `tidytuesdayR::tt_load()`, so set `GITHUB_PAT` if GitHub rate limits appear

### External Dependencies

- tidytuesdayR::tt_load is used and no local TidyTuesday files were matched

## Setup Scaffold

```{r}
#| label: setup
#| eval: false

library(tidyverse)
source("R/screencast_helpers.R")

# Optional episode packages:
# library(broom)
# library(lubridate)
# library(scales)
# library(tidymodels)
# library(tidytuesdayR)

tuesdata <- load_tidyweek_data("2021-05-18")
names(tuesdata)
```

## Thought -> Code -> Data Bridge

### Thought Process Map

- `T1 | Context | Frame the episode as exploratory analysis of wealth and income with limited advance knowledge of the data. | Transcript title and opening context`
- `T2 | Inspect | Load the data, inspect table shape, and identify the main entities, measures, and comparison axes. | Initial data load and early counts or prints`
- `T6 | Compare | Move from descriptive patterns into statistical or predictive modeling once the feature space is ready. | Modeling functions and evaluation code`
- `T7 | Visualize | Turn the cleaned summaries into interpretable visual comparisons and iterate toward the clearest view. | ggplot workflow`
- `T8 | Reflect | Identify reusable workflow pieces that can transfer to future datasets with a similar shape or question. | End-state of the analysis pattern`

### Code Mapping Table

- `T1 | Data load and inspection | tt_load, count | Establish source shape, units, and comparison candidates`
- `T6 | Modeling and evaluation | lm, recipe, fit, vfold_cv, tune_grid | Test explanatory or predictive relationships after EDA`
- `T7 | Visualization | ggplot, geom_col, geom_line | Communicate the core result visually`

### Data Transformation Flow

- `Raw | Imported source tables or files | Load source data into tibbles or lists | Raw episode objects | Establish the starting analysis surface`
- `Raw -> Clean | Raw source tables | Filter, rename, and derive analysis columns | Cleaned table | Make the data comparable and plot-ready`
- `Clean -> Analytical | Feature table | Engineer predictors and fit models | Model outputs, predictions, or coefficients | Quantify relationships beyond descriptive summaries`
- `Analytical -> Visual | Summary or model outputs | Reorder factors and prepare plotting fields | Plot-ready table | Turn the analytical result into an interpretable chart`

## Guided Build Sequence

## Stage 1: Context

Frame the episode as exploratory analysis of wealth and income with limited advance knowledge of the data.

::: {.callout-note}
## Cognitive Move
Define the analytical question, likely unit of analysis, and what would count as an interesting result.

Trigger:

- Transcript title and opening context


:::

```{r}
#| label: stage_01_context
#| eval: false

# Write the one-sentence analytical question for this episode.
# Topic: wealth and income
# Unit of analysis:
# Main comparison axis:
# What would make the result interesting?
```
## Stage 2: Inspect

Load the data, inspect table shape, and identify the main entities, measures, and comparison axes.

::: {.callout-note}
## Cognitive Move
Inspect shape, columns, missingness, and table roles before committing to transformations.

Trigger:

- Initial data load and early counts or prints


:::

```{r}
#| label: stage_02_inspect
#| eval: false

# Start here once data is loaded.
# names(tuesdata)
# purrr::imap_dfr(tuesdata, ~ tibble(table = .y, rows = nrow(.x), cols = ncol(.x)))
# glimpse(analysis_tbl)
```
## Stage 3: Compare

Move from descriptive patterns into statistical or predictive modeling once the feature space is ready.

::: {.callout-note}
## Cognitive Move
Summarize groups or entities in a way that reveals differences, rankings, or trends without overfitting the story.

Trigger:

- Modeling functions and evaluation code

Additional guidance:

- If you plan to model, use this step to identify which predictors deserve to enter the first baseline model.
:::

```{r}
#| label: stage_03_compare
#| eval: false

# Summarize the comparison that best answers the question.
# comparison_tbl <- analysis_tbl %>%
#   group_by(...) %>%
#   summarise(...) %>%
#   arrange(desc(...))
```
## Stage 4: Visualize

Turn the cleaned summaries into interpretable visual comparisons and iterate toward the clearest view.

::: {.callout-note}
## Cognitive Move
Choose the plot after the summary table exists, then optimize for clarity rather than novelty.

Trigger:

- ggplot workflow

Additional guidance:

- When time is present, lead with the broad trend before showing narrower subgroup comparisons.
:::

```{r}
#| label: stage_04_visualize
#| eval: false

# Turn the comparison table into one clear communication plot.
# ggplot(comparison_tbl, aes(...)) +
#   geom_col() +
#   labs(title = ..., subtitle = ...)
```
## Stage 5: Reflect

Identify reusable workflow pieces that can transfer to future datasets with a similar shape or question.

::: {.callout-note}
## Cognitive Move
Translate the episode-specific move into a reusable function, helper, or workflow note.

Trigger:

- End-state of the analysis pattern


:::

```{r}
#| label: stage_05_reflect
#| eval: false

# Capture the reusable move after the analysis is clear.
# Candidate helper:
# - input:
# - output:
# - why it is reusable:
```

## Reusable Layer

### Reusable Patterns

- `inspect_source_schema | Inspect source tables, field meanings, and likely comparison axes | Raw dataset or dataset list | Initial analysis plan | Y`
- `build_predictive_or_explanatory_model | Move from cleaned features to fitted models and evaluation outputs | Feature table and outcome field | Model object and diagnostics | Y`
- `plot_comparative_story | Reorder entities and visualize the main comparison cleanly | Summary table or model output | Communication-ready plot | Y`

### Function Candidates

- `build_predictive_pipeline(data, formula, ...)`
- `evaluate_episode_model(model, test_data, outcome_col)`
- `plot_comparative_story(data, x_col, y_col, group_col = NULL)`

## How To Finish This Super QMD

1. Replace each scaffold chunk with the real episode-specific analysis code.
2. After each major result, add a short callout explaining what changed in your thinking.
3. Keep one broad chart early, one comparison chart in the middle, and one final communication chart at the end.
4. Move any repeated transformation or comparison logic into `R/screencast_helpers.R`.
5. Once the document reads like a guided walkthrough instead of a reproduction note, promote it into a finished KER-guided analysis.

## Adaptation Layer

- `1 | inspect_source_schema | Start by mapping entities, measures, and likely comparison axes before transforming anything`
- `6 | build_predictive_or_explanatory_model | Fit and evaluate a baseline model after the core feature engineering is stable`
- `8 | plot_comparative_story | Choose the clearest visual comparison once the analytical table is ready`
