---
title: "Super QMD: Analyzing Traumatic Brain Injury in R"
subtitle: "Cognition-first blueprint built from the episode guide and KER"
author: "Andrei Stoian"
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show scaffold code"
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

## What This Document Is

This super QMD is the bridge between:

- the TidyTuesday dataset for `2020-03-24`
- the KER entry for `traumatic_brain_injury`
- a final guided analysis document

The point is to preserve not only the code path, but the author's cognition:

- what question gets asked first
- how the dataset is inspected
- when transformation happens
- why one comparison is chosen before another
- which pieces should become reusable helpers

## Episode Snapshot

::: {.callout-note}
## Analysis Intent
Explore distributions, group differences, and notable patterns in Traumatic Brain Injury through cleaning, summarization, and visualization
:::

| Field | Value |
|---|---|
| Queue | #126 |
| Function | `traumatic_brain_injury` |
| Analysis type | EDA |
| Data source | TidyTuesday 2020-03-24 |
| Status | `needs_mapping` |

**Analytical pattern:** `load_inspect_recode_summarize_trend_compare_visualize`





## Cognitive Signature

- Start by naming the question and the unit of analysis before choosing plots or metrics.
- Use schema inspection to decide which columns carry the main story.
- Build one analysis-ready table before comparing groups.
- Start broad with time or trend structure when it helps explain which later comparisons are fair.
- End with one communication-ready chart or table that carries the episode's main claim.

## Data and Execution Context

### Local Assets

- `tidytuesday/data/2020/2020-03-24/dod_tbi_2006.csv`
- `tidytuesday/data/2020/2020-03-24/tbi_age.csv`
- `tidytuesday/data/2020/2020-03-24/tbi_military.csv`
- `tidytuesday/data/2020/2020-03-24/tbi_year.csv`

### Execution Notes

- Use `load_tidyweek_data("2020-03-24")` from `R/screencast_helpers.R` if local files are present
- Otherwise fall back to `tidytuesdayR::tt_load("2020-03-24")`

### External Dependencies

- No external dependencies captured.

## Setup Scaffold

```{r}
#| label: setup
#| eval: false

library(tidyverse)
source("R/screencast_helpers.R")

# Optional episode packages:
# library(scales)
# library(tidytuesdayR)

tuesdata <- load_tidyweek_data("2020-03-24")
names(tuesdata)

dod_tbi_2006 <- tuesdata$dod_tbi_2006 # or read from local path
tbi_age <- tuesdata$tbi_age # or read from local path
tbi_military <- tuesdata$tbi_military # or read from local path
tbi_year <- tuesdata$tbi_year # or read from local path
```

## Thought -> Code -> Data Bridge

### Thought Process Map

- `T1 | Context | Frame the episode as exploratory analysis of Traumatic Brain Injury with limited advance knowledge of the data. | Transcript title and opening context`
- `T2 | Inspect | Load the data, inspect table shape, and identify the main entities, measures, and comparison axes. | Initial data load and early counts or prints`
- `T3 | Transform | Reshape fields, standardize column meanings, and join supporting tables to create analysis-ready records. | Join and reshaping operations`
- `T4 | Trend | Identify temporal patterns and seasonal variation before comparing sub-groups. | Time-based grouping and line charts`
- `T6 | Compare | Summarize groups in a way that reveals differences, rankings, or trends. | group_by + summarise + arrange pattern`
- `T7 | Visualize | Turn the cleaned summaries into interpretable visual comparisons and iterate toward the clearest view. | ggplot workflow`
- `T8 | Reflect | Identify reusable workflow pieces that can transfer to future datasets with a similar shape or question. | End-state of the analysis pattern`

```{r}
#| label: thought_execution_log
#| eval: false

# Fill in `finding` after completing each stage.
thought_log <- tibble::tribble(
  ~step,  ~move,        ~finding,
  "T1", "Context", "",
  "T2", "Inspect", "",
  "T3", "Transform", "",
  "T4", "Trend", "",
  "T6", "Compare", "",
  "T7", "Visualize", "",
  "T8", "Reflect", "",
)
thought_log
```


### Code Mapping Table

- `T1 | Data load and inspection | tt_load, count | Establish source shape, units, and comparison candidates`
- `T3 | Structuring and enrichment | mutate, left_join, pivot_longer | Convert messy inputs into analysis-ready tables`
- `T4 | Temporal summary | group_by year/month, summarise, geom_line | Show trend over time`
- `T7 | Visualization | ggplot, geom_col, geom_point, geom_line | Communicate the core result visually`

### Data Transformation Flow

- `Raw | Imported source tables or files | Load source data into tibbles or lists | Raw episode objects | Establish the starting analysis surface`
- `Raw -> Clean | Raw source tables | Rename, reshape, unnest, and join supporting data | Analysis-ready rows with consistent keys | Standardize the schema before comparison`
- `Clean -> Analytical | Feature table | Engineer predictors or compute group metrics | Summary or model outputs | Quantify relationships beyond descriptive summaries`
- `Analytical -> Visual | Summary or model outputs | Reorder factors and prepare plotting fields | Plot-ready table | Turn the analytical result into an interpretable chart`

## Guided Build Sequence

## Stage 1: Context

Frame the episode as exploratory analysis of Traumatic Brain Injury with limited advance knowledge of the data.

::: {.callout-note}
## Cognitive Move
Define the analytical question, likely unit of analysis, and what would count as an interesting result.

Trigger:

- Transcript title and opening context


:::

```{r}
#| label: stage_01_context
#| eval: false

# Write the one-sentence analytical question for this episode.
# Topic: Traumatic Brain Injury
# Unit of analysis:
# Main comparison axis:
# What would make the result interesting?
```
## Stage 2: Inspect

Load the data, inspect table shape, and identify the main entities, measures, and comparison axes.

::: {.callout-note}
## Cognitive Move
Inspect shape, columns, missingness, and table roles before committing to transformations.

Trigger:

- Initial data load and early counts or prints


:::

```{r}
#| label: stage_02_inspect
#| eval: false

# Start here once data is loaded.
# names(tuesdata)
# purrr::imap_dfr(tuesdata, ~ tibble(table = .y, rows = nrow(.x), cols = ncol(.x)))
# glimpse(analysis_tbl)
# skimr::skim(analysis_tbl)
```
## Stage 3: Transform

Reshape fields, standardize column meanings, and join supporting tables to create analysis-ready records.

::: {.callout-note}
## Cognitive Move
Reshape, join, or normalize only enough to make the next comparison possible.

Trigger:

- Join and reshaping operations in the source Rmd


:::

```{r}
#| label: stage_03_transform
#| eval: false

# Build the canonical analysis table for this episode.
# Source assets: tidytuesday/data/2020/2020-03-24/dod_tbi_2006.csv, tidytuesday/data/2020/2020-03-24/tbi_age.csv, tidytuesday/data/2020/2020-03-24/tbi_military.csv, tidytuesday/data/2020/2020-03-24/tbi_year.csv
# Suggested verbs: mutate(), separate_rows(), pivot_longer(), left_join(), unnest()
# analysis_tbl <- raw_tbl %>%
# ...
```
## Stage 4: Compare

Summarize groups or entities in a way that reveals differences, rankings, or trends.

::: {.callout-note}
## Cognitive Move
Summarize groups or entities in a way that reveals differences, rankings, or trends without overfitting the story.

Trigger:

- count() and group_by() + summarise() patterns

Additional guidance:

- If you plan to model, use this step to identify which predictors deserve to enter the first baseline model.
:::

```{r}
#| label: stage_04_compare
#| eval: false

# Summarize the comparison that best answers the question.
# comparison_tbl <- analysis_tbl %>%
# count(category_col, sort = TRUE) # or
# group_by(group_col) %>%
# summarise(metric = mean(value_col, na.rm = TRUE)) %>%
# arrange(desc(metric))
```
## Stage 5: Visualize

Turn the cleaned summaries into interpretable visual comparisons and iterate toward the clearest view.

::: {.callout-note}
## Cognitive Move
Choose the plot after the summary table exists, then optimize for clarity rather than novelty.

Trigger:

- ggplot workflow

Additional guidance:

- When time is present, lead with the broad trend before showing narrower subgroup comparisons.
:::

```{r}
#| label: stage_05_visualize
#| eval: false

# Turn the comparison table into one clear communication plot.
# ggplot(comparison_tbl, aes(x = fct_reorder(category, metric), y = metric)) +
# geom_col() +
# coord_flip() +
# labs(
# title = "...",
# subtitle = "Source: TidyTuesday 2020-03-24",
# x = NULL, y = NULL
# )
```
## Stage 6: Reflect

Identify reusable workflow pieces that can transfer to future datasets with a similar shape or question.

::: {.callout-note}
## Cognitive Move
Translate the episode-specific move into a reusable function, helper, or workflow note.

Trigger:

- End-state of the analysis pattern


:::

```{r}
#| label: stage_06_reflect
#| eval: false

# Capture the reusable move after the analysis is clear.
# Candidate helper:
# - input:
# - output:
# - why it is reusable:
```

---

## Hour 2 — Depth



### Stage 7: Multi-axis Comparison

Examine the interaction between two grouping variables rather than one at a time.

::: {.callout-note}
## Cognitive Move
One-dimensional summaries miss interaction effects. Cross two categoricals to see whether the relationship holds across all sub-groups or only some.

Trigger: group_by(cat1, cat2) + summarise + facet_wrap
:::

```{r}
#| label: stage_07_multiaxis
#| eval: false

cross_tbl <- analysis_tbl |>
 group_by(age_group, type) |>
 summarise(
 n = n(),
 median = median(number_est, na.rm = TRUE),
 .groups = "drop"
 ) |>
 filter(n >= 3)

ggplot(cross_tbl, aes(x = fct_reorder(age_group, median), y = median, fill = type)) +
 geom_col(position = "dodge") +
 coord_flip() +
 facet_wrap(~ type, scales = "free_x") +
 labs(title = "Traumatic Brain Injury: number_est by age_group and type",
 x = NULL, y = "Median number_est")
```

### Stage 8: Outlier Investigation

Find cases that sit far from their group expectation.

::: {.callout-note}
## Cognitive Move
After the aggregate pattern is clear, the outliers are often the most interesting story — they show where the rule breaks down.

Trigger: group_by + mutate(residual = value - group_median) + slice_max(abs(residual))
:::

```{r}
#| label: stage_08_outliers
#| eval: false

residual_tbl <- analysis_tbl |>
 group_by(age_group) |>
 mutate(
 group_median = median(number_est, na.rm = TRUE),
 residual = number_est - group_median
 ) |>
 ungroup() |>
 arrange(desc(abs(residual)))

# Label the top outliers
library(ggrepel)

ggplot(residual_tbl,
 aes(x = group_median, y = number_est, label = age_group)) +
 geom_point(alpha = 0.4) +
 geom_text_repel(
 data = slice_max(residual_tbl, abs(residual), n = 15),
 size = 3, max.overlaps = 20
 ) +
 geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
 labs(title = "Who over- or underperforms their group median?",
 x = "Group median number_est", y = "Observed number_est")
```

### Stage 9: Correlation Structure

Examine how numeric columns relate to each other and to the outcome.

::: {.callout-note}
## Cognitive Move
Scatter plots with colour encoding reveal whether a confound explains the group difference, or whether the effect is additive.

Trigger: geom_point + geom_smooth(method = "lm") + color by group
:::

```{r}
#| label: stage_09_correlation
#| eval: false

ggplot(analysis_tbl,
 aes(x = rate_est, y = number_est, colour = age_group)) +
 geom_point(alpha = 0.5) +
 geom_smooth(method = "lm", se = FALSE) +
 scale_colour_viridis_d(end = 0.85) +
 labs(title = "Relationship between number_est and rate_est",
 subtitle = "Coloured by age_group",
 x = "rate_est", y = "number_est")
```

---

## Hour 3 — Synthesis

### Stage 10: Model

Fit a linear model to quantify what predicts number_est.

::: {.callout-note}
## Cognitive Move
Move from "what differs" (descriptive) to "what explains" (inferential). The model should confirm the story the charts already suggested — not replace it.

Trigger: lm() + broom::tidy() + coefficient plot
:::

```{r}
#| label: stage_10_model
#| eval: false

model_tbl <- analysis_tbl |>
 mutate(across(where(is.character), as.factor))

model_fit <- lm(number_est ~ age_group + type + number_est + rate_est, data = model_tbl)

broom::tidy(model_fit, conf.int = TRUE) |>
 filter(term != "(Intercept)") |>
 mutate(term = fct_reorder(term, estimate)) |>
 ggplot(aes(estimate, term,
 xmin = conf.low, xmax = conf.high)) +
 geom_pointrange() +
 geom_vline(xintercept = 0, linetype = "dashed") +
 labs(title = "What predicts number_est?",
 subtitle = "Linear model coefficients (95% CI)",
 x = "Estimate", y = NULL)
```

### Stage 11: Residuals

Which cases does the model get most wrong?

::: {.callout-note}
## Cognitive Move
Residuals reveal where the model's assumptions break down — often these are the most analytically interesting cases.

Trigger: broom::augment() + geom_point(.resid) + geom_text_repel
:::

```{r}
#| label: stage_11_residuals
#| eval: false

augmented_tbl <- broom::augment(model_fit, data = model_tbl)

ggplot(augmented_tbl, aes(.fitted, .resid, label = age_group)) +
 geom_point(alpha = 0.4) +
 geom_smooth(method = "loess", se = FALSE, colour = "steelblue") +
 geom_hline(yintercept = 0, linetype = "dashed") +
 geom_text_repel(
 data = filter(augmented_tbl, abs(.resid) > quantile(abs(.resid), 0.95)),
 size = 3, max.overlaps = 15
 ) +
 labs(title = "Residuals: who over- or underperforms the model?",
 x = "Fitted number_est", y = "Residual")
```

### Stage 12: Final Communication Chart

One polished chart that carries the entire episode's story.

::: {.callout-note}
## Cognitive Move
The communication chart is not a summary of everything — it is the single clearest view of the main finding. Choose the encoding that requires the least explanation.

Trigger: After residuals and model are stable, step back and pick the clearest one-chart story
:::

```{r}
#| label: stage_12_final_chart
#| eval: false

final_tbl <- analysis_tbl |>
 group_by(age_group) |>
 summarise(
 median_number_est = median(number_est, na.rm = TRUE),
 n = n(),
 .groups = "drop"
 ) |>
 filter(n >= 5) |>
 mutate(age_group = fct_reorder(age_group, median_number_est))

ggplot(final_tbl,
 aes(x = age_group, y = median_number_est,
 fill = median_number_est)) +
 geom_col(show.legend = FALSE) +
 scale_fill_gradient2(
 low = "tomato", mid = "gray90", high = "steelblue",
 midpoint = median(analysis_tbl$number_est, na.rm = TRUE)
 ) +
 coord_flip() +
 labs(
 title = "Traumatic Brain Injury: number_est by age_group",
 subtitle = "Median per group | colour = above/below overall median",
 caption = "Source: TidyTuesday 2020-03-24",
 x = NULL, y = "Median number_est"
 ) +
 theme_minimal(base_size = 13) +
 theme(plot.title = element_text(face = "bold"))
```

### Stage 13: Knowledge Extraction

Capture the reusable pattern for the KER.

::: {.callout-note}
## Cognitive Move
After the analysis is clear, abstract it: what is the general workflow that would work on any similar dataset? That is what goes into the KER and into a helper function.

Trigger: End of episode — write the function stub before closing the session
:::

```{r}
#| label: stage_13_ker_capture
#| eval: false

# Pattern: load -> inspect -> join ->
# multi-axis compare -> outlier detection -> model -> residuals -> final chart

# Candidate helper:
# compare_groups_by_outcome <- function(data, group_col, outcome_col,
# secondary_group = NULL,
# model = TRUE) {
# # 1. group_by + summarise
# # 2. residual calculation
# # 3. optional lm + broom::tidy
# # 4. ggplot output
# }
```


## Reusable Layer

### Reusable Patterns

- `inspect_source_schema | Inspect source tables, field meanings, and likely comparison axes | Raw dataset or dataset list | Initial analysis plan | Y`
- `count_and_rank | Count and rank categorical variables to find the dominant groups | Any categorical column | Ranked frequency table | Y`
- `plot_comparative_story | Reorder entities and visualize the main comparison cleanly | Summary table or model output | Communication-ready plot | Y`

### Function Candidates

- `reshape_episode_inputs(data, ...)`
- `summarize_by_group(data, group_col, metric_col)`
- `plot_comparative_story(data, x_col, y_col, group_col = NULL)`

## How To Finish This Super QMD

1. Replace each scaffold chunk with the real episode-specific analysis code.
2. After each major result, add a short callout explaining what changed in your thinking.
3. Keep one broad chart early, one comparison chart in the middle, and one final communication chart at the end.
4. Move any repeated transformation or comparison logic into `R/screencast_helpers.R`.
5. Once the document reads like a guided walkthrough instead of a reproduction note, promote it into a finished KER-guided analysis.

## Adaptation Layer

- `1 | inspect_source_schema | Start by mapping entities, measures, and likely comparison axes before transforming anything`
- `3 | reshape_for_analysis | Pivot, join, or unnest only enough to make the next comparison possible`
- `6 | count_and_rank | Surface the dominant categories before building the comparative chart`
- `8 | plot_comparative_story | Choose the clearest visual comparison once the analytical table is ready`