---
title: "Super QMD: Analyzing Stranger Things Dialogue in R"
subtitle: "Cognition-first blueprint built from the episode guide and KER"
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show scaffold code"
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

## What This Document Is

This super QMD is the bridge between:

- the TidyTuesday dataset for `2022-10-18`
- the KER entry for `stranger_things_dialogue`
- a final guided analysis document

The point is to preserve not only the code path, but the author's cognition:

- what question gets asked first
- how the dataset is inspected
- when transformation happens
- why one comparison is chosen before another
- which pieces should become reusable helpers

## Episode Snapshot

- `queue_position`: `213`
- `function_name`: `stranger_things_dialogue`
- `analysis_type`: `Text`
- `pattern_signature`: `load_inspect_tokenize_trend_compare_visualize`
- `data_source`: TidyTuesday 2022-10-18
- `reproduction_status`: `needs_mapping`

::: {.callout-note}
## Analysis Intent
Explore and compare language patterns in Stranger Things Dialogue using tokenization, grouped counts, and discriminative text analysis
:::



## Transcript Preview

Analyzing Stranger Things Dialogue in R David Robinson 15.9k subscribers 72 Share Save Clip 4,521 views Streamed live on 2022-10-18 Tidy Tuesday R Screencasts I'll analyze a dataset about stranger things dialogue without looking at the dataset in advance. The dataset comes from the Tidy Tuesday proj...

## Cognitive Signature

- Start by naming the question and the unit of analysis before choosing plots or metrics.
- Use schema inspection to decide which columns carry the main story.
- Build one analysis-ready table before comparing groups.
- Start broad with time or trend structure when it helps explain which later comparisons are fair.
- Tokenize first at the word level, then zoom in on discriminative terms using tf-idf or log-odds.
- End with one communication-ready chart or table that carries the episode's main claim.

## Data and Execution Context

### Local Assets

- `tidytuesday/data/2022/2022-10-18/episodes.csv`
- `tidytuesday/data/2022/2022-10-18/stranger_things_all_dialogue.csv`

### Execution Notes

- Use `load_tidyweek_data("2022-10-18")` from `R/screencast_helpers.R` if local files are present
- Otherwise fall back to `tidytuesdayR::tt_load("2022-10-18")`

### External Dependencies

- No external dependencies captured.

## Setup Scaffold

```{r}
#| label: setup
#| eval: false

library(tidyverse)
source("R/screencast_helpers.R")

# Optional episode packages:
# library(scales)
# library(tidytuesdayR)

tuesdata <- load_tidyweek_data("2022-10-18")
names(tuesdata)

episodes <- tuesdata$episodes  # or read from local path
stranger_things_all_dialogue <- tuesdata$stranger_things_all_dialogue  # or read from local path
```

## Thought -> Code -> Data Bridge

### Thought Process Map

- `T1 | Context | Frame the episode as exploratory analysis of Stranger Things Dialogue with limited advance knowledge of the data. | Transcript title and opening context`
- `T2 | Inspect | Load the data, inspect table shape, and identify the main entities, measures, and comparison axes. | Initial data load and early counts or prints`
- `T3 | Transform | Reshape fields, standardize column meanings, and join supporting tables to create analysis-ready records. | Join and reshaping operations`
- `T4 | Trend | Identify temporal patterns and seasonal variation before comparing sub-groups. | Time-based grouping and line charts`
- `T4 | Tokenize | Split text into tokens, remove stop words, and compute term frequencies or tf-idf. | tidytext or tokenizers workflow`
- `T6 | Compare | Summarize groups in a way that reveals differences, rankings, or trends. | group_by + summarise + arrange pattern`
- `T7 | Visualize | Turn the cleaned summaries into interpretable visual comparisons and iterate toward the clearest view. | ggplot workflow`
- `T8 | Reflect | Identify reusable workflow pieces that can transfer to future datasets with a similar shape or question. | End-state of the analysis pattern`

### Code Mapping Table

- `T1 | Data load and inspection | tt_load, count | Establish source shape, units, and comparison candidates`
- `T3 | Structuring and enrichment | mutate, left_join, pivot_longer | Convert messy inputs into analysis-ready tables`
- `T4 | Temporal summary | group_by year/month, summarise, geom_line | Show trend over time`
- `T4 | Tokenization | unnest_tokens, count, bind_tf_idf | Surface discriminative vocabulary`
- `T7 | Visualization | ggplot, geom_col, geom_point, geom_line | Communicate the core result visually`

### Data Transformation Flow

- `Raw | Imported source tables or files | Load source data into tibbles or lists | Raw episode objects | Establish the starting analysis surface`
- `Raw -> Clean | Raw source tables | Rename, reshape, unnest, and join supporting data | Analysis-ready rows with consistent keys | Standardize the schema before comparison`
- `Clean -> Analytical | Feature table | Engineer predictors or compute group metrics | Summary or model outputs | Quantify relationships beyond descriptive summaries`
- `Analytical -> Visual | Summary or model outputs | Reorder factors and prepare plotting fields | Plot-ready table | Turn the analytical result into an interpretable chart`

## Guided Build Sequence

## Stage 1: Context

Frame the episode as exploratory analysis of Stranger Things Dialogue with limited advance knowledge of the data.

::: {.callout-note}
## Cognitive Move
Define the analytical question, likely unit of analysis, and what would count as an interesting result.

Trigger:

- Transcript title and opening context


:::

```{r}
#| label: stage_01_context
#| eval: false

# Write the one-sentence analytical question for this episode.
# Topic: Stranger Things Dialogue
# Unit of analysis:
# Main comparison axis:
# What would make the result interesting?
```
## Stage 2: Inspect

Load the data, inspect table shape, and identify the main entities, measures, and comparison axes.

::: {.callout-note}
## Cognitive Move
Inspect shape, columns, missingness, and table roles before committing to transformations.

Trigger:

- Initial data load and early counts or prints


:::

```{r}
#| label: stage_02_inspect
#| eval: false

# Start here once data is loaded.
# names(tuesdata)
# purrr::imap_dfr(tuesdata, ~ tibble(table = .y, rows = nrow(.x), cols = ncol(.x)))
# glimpse(analysis_tbl)
# skimr::skim(analysis_tbl)
```
## Stage 3: Transform

Reshape fields, standardize column meanings, and join supporting tables to create analysis-ready records.

::: {.callout-note}
## Cognitive Move
Reshape, join, or normalize only enough to make the next comparison possible.

Trigger:

- Join and reshaping operations in the source Rmd


:::

```{r}
#| label: stage_03_transform
#| eval: false

# Build the canonical analysis table for this episode.
# Source assets: tidytuesday/data/2022/2022-10-18/episodes.csv, tidytuesday/data/2022/2022-10-18/stranger_things_all_dialogue.csv
# Suggested verbs: mutate(), separate_rows(), pivot_longer(), left_join(), unnest()
# analysis_tbl <- raw_tbl %>%
#   ...
```
## Stage 4: Compare

Summarize groups or entities in a way that reveals differences, rankings, or trends.

::: {.callout-note}
## Cognitive Move
Summarize groups or entities in a way that reveals differences, rankings, or trends without overfitting the story.

Trigger:

- count() and group_by() + summarise() patterns

Additional guidance:

- If you plan to model, use this step to identify which predictors deserve to enter the first baseline model.
:::

```{r}
#| label: stage_04_compare
#| eval: false

# Summarize the comparison that best answers the question.
# comparison_tbl <- analysis_tbl %>%
#   count(category_col, sort = TRUE)   # or
#   group_by(group_col) %>%
#   summarise(metric = mean(value_col, na.rm = TRUE)) %>%
#   arrange(desc(metric))
```
## Stage 5: Visualize

Turn the cleaned summaries into interpretable visual comparisons and iterate toward the clearest view.

::: {.callout-note}
## Cognitive Move
Choose the plot after the summary table exists, then optimize for clarity rather than novelty.

Trigger:

- ggplot workflow

Additional guidance:

- When time is present, lead with the broad trend before showing narrower subgroup comparisons.
:::

```{r}
#| label: stage_05_visualize
#| eval: false

# Turn the comparison table into one clear communication plot.
# ggplot(comparison_tbl, aes(x = fct_reorder(category, metric), y = metric)) +
#   geom_col() +
#   coord_flip() +
#   labs(
#     title = "...",
#     subtitle = "Source: TidyTuesday 2022-10-18",
#     x = NULL, y = NULL
#   )
```
## Stage 6: Reflect

Identify reusable workflow pieces that can transfer to future datasets with a similar shape or question.

::: {.callout-note}
## Cognitive Move
Translate the episode-specific move into a reusable function, helper, or workflow note.

Trigger:

- End-state of the analysis pattern


:::

```{r}
#| label: stage_06_reflect
#| eval: false

# Capture the reusable move after the analysis is clear.
# Candidate helper:
# - input:
# - output:
# - why it is reusable:
```

---

## Hour 2 — Depth

### Stage 7: Tokenize

Split text into words, remove stop words, and compute term frequencies or tf-idf.

::: {.callout-note}
## Cognitive Move
Tokenize before any group comparison — vocabulary-level signals often explain group differences more cleanly than aggregate counts.

Trigger: unnest_tokens / tidytext workflow
:::

```{r}
#| label: stage_07_tokenize
#| eval: false

library(tidytext)

tokens_tbl <- analysis_tbl |>
  unnest_tokens(word, text_col) |>
  anti_join(stop_words, by = "word") |>
  count(raw_text, word, sort = TRUE)

# tf-idf
tokens_tbl |>
  bind_tf_idf(word, raw_text, n) |>
  arrange(desc(tf_idf))
```

### Stage 8: Discriminative Vocabulary

Which words are most characteristic of each group?

::: {.callout-note}
## Cognitive Move
Use log-odds or tf-idf to find vocabulary that discriminates groups, not just frequency.

Trigger: bind_log_odds or bind_tf_idf
:::

```{r}
#| label: stage_08_discriminative_vocab
#| eval: false

library(tidylo)

tokens_tbl |>
  bind_log_odds(raw_text, word, n) |>
  group_by(raw_text) |>
  slice_max(log_odds_weighted, n = 10) |>
  ggplot(aes(log_odds_weighted, fct_reorder(word, log_odds_weighted), fill = raw_text)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ raw_text, scales = "free_y") +
  labs(title = "Most characteristic words per group",
       subtitle = "Log-odds weighted",
       x = "Log-odds", y = NULL)
```

---

## Hour 3 — Synthesis

### Stage 9: Model

Fit a linear model to quantify what predicts season.

::: {.callout-note}
## Cognitive Move
Move from "what differs" (descriptive) to "what explains" (inferential). The model should confirm the story the charts already suggested — not replace it.

Trigger: lm() + broom::tidy() + coefficient plot
:::

```{r}
#| label: stage_09_model
#| eval: false

model_tbl <- analysis_tbl |>
  mutate(across(where(is.character), as.factor))

model_fit <- lm(season ~ raw_text + stage_direction + season + episode, data = model_tbl)

broom::tidy(model_fit, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(term = fct_reorder(term, estimate)) |>
  ggplot(aes(estimate, term,
             xmin = conf.low, xmax = conf.high)) +
  geom_pointrange() +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(title = "What predicts season?",
       subtitle = "Linear model coefficients (95% CI)",
       x = "Estimate", y = NULL)
```

### Stage 10: Residuals

Which cases does the model get most wrong?

::: {.callout-note}
## Cognitive Move
Residuals reveal where the model's assumptions break down — often these are the most analytically interesting cases.

Trigger: broom::augment() + geom_point(.resid) + geom_text_repel
:::

```{r}
#| label: stage_10_residuals
#| eval: false

augmented_tbl <- broom::augment(model_fit, data = model_tbl)

ggplot(augmented_tbl, aes(.fitted, .resid, label = raw_text)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", se = FALSE, colour = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_text_repel(
    data = filter(augmented_tbl, abs(.resid) > quantile(abs(.resid), 0.95)),
    size = 3, max.overlaps = 15
  ) +
  labs(title = "Residuals: who over- or underperforms the model?",
       x = "Fitted season", y = "Residual")
```

### Stage 11: Final Communication Chart

One polished chart that carries the entire episode's story.

::: {.callout-note}
## Cognitive Move
The communication chart is not a summary of everything — it is the single clearest view of the main finding. Choose the encoding that requires the least explanation.

Trigger: After residuals and model are stable, step back and pick the clearest one-chart story
:::

```{r}
#| label: stage_11_final_chart
#| eval: false

final_tbl <- analysis_tbl |>
  group_by(raw_text) |>
  summarise(
    median_season = median(season, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  filter(n >= 5) |>
  mutate(raw_text = fct_reorder(raw_text, median_season))

ggplot(final_tbl,
       aes(x = raw_text, y = median_season,
           fill = median_season)) +
  geom_col(show.legend = FALSE) +
  scale_fill_gradient2(
    low = "tomato", mid = "gray90", high = "steelblue",
    midpoint = median(analysis_tbl$season, na.rm = TRUE)
  ) +
  coord_flip() +
  labs(
    title    = "Stranger Things Dialogue: season by raw_text",
    subtitle = "Median per group | colour = above/below overall median",
    caption  = "Source: TidyTuesday 2022-10-18",
    x = NULL, y = "Median season"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

### Stage 12: Knowledge Extraction

Capture the reusable pattern for the KER.

::: {.callout-note}
## Cognitive Move
After the analysis is clear, abstract it: what is the general workflow that would work on any similar dataset? That is what goes into the KER and into a helper function.

Trigger: End of episode — write the function stub before closing the session
:::

```{r}
#| label: stage_12_ker_capture
#| eval: false

# Pattern: load -> inspect -> join ->
#          multi-axis compare -> outlier detection -> model -> residuals -> final chart

# Candidate helper:
# compare_groups_by_outcome <- function(data, group_col, outcome_col,
#                                        secondary_group = NULL,
#                                        model = TRUE) {
#   # 1. group_by + summarise
#   # 2. residual calculation
#   # 3. optional lm + broom::tidy
#   # 4. ggplot output
# }
```


## Reusable Layer

### Reusable Patterns

- `inspect_source_schema | Inspect source tables, field meanings, and likely comparison axes | Raw dataset or dataset list | Initial analysis plan | Y`
- `count_and_rank | Count and rank categorical variables to find the dominant groups | Any categorical column | Ranked frequency table | Y`
- `plot_comparative_story | Reorder entities and visualize the main comparison cleanly | Summary table or model output | Communication-ready plot | Y`
- `tokenize_and_tfidf | Tokenize text, remove stop words, compute tf-idf | Text column | Term-importance table | Y`

### Function Candidates

- `reshape_episode_inputs(data, ...)`
- `summarize_by_group(data, group_col, metric_col)`
- `plot_comparative_story(data, x_col, y_col, group_col = NULL)`
- `tokenize_and_clean(data, text_col, stop_words)`
- `compute_tfidf_by_group(tokens_tbl, group_col)`

## How To Finish This Super QMD

1. Replace each scaffold chunk with the real episode-specific analysis code.
2. After each major result, add a short callout explaining what changed in your thinking.
3. Keep one broad chart early, one comparison chart in the middle, and one final communication chart at the end.
4. Move any repeated transformation or comparison logic into `R/screencast_helpers.R`.
5. Once the document reads like a guided walkthrough instead of a reproduction note, promote it into a finished KER-guided analysis.

## Adaptation Layer

- `1 | inspect_source_schema | Start by mapping entities, measures, and likely comparison axes before transforming anything`
- `3 | reshape_for_analysis | Pivot, join, or unnest only enough to make the next comparison possible`
- `4 | tokenize_and_tfidf | Tokenize text before any group comparison to enable vocabulary-level analysis`
- `6 | count_and_rank | Surface the dominant categories before building the comparative chart`
- `8 | plot_comparative_story | Choose the clearest visual comparison once the analytical table is ready`