---
title: "Hour 1 — EDA: Analyzing code in CRAN packages in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 018 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`boxplot_jitter` · `coord_flip_horizontal` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_histogram_density` · `geom_smooth_loess` · `grouped_summary_ggplot` · `janitor_clean` · `log_scale` · `multi_table_join` · `pivot_reshape`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- text is multilingual without language detection
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- text is long-form English and benefits from word-level decomposition
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(tidytext)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
loc_cran_packages <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2019/2019-11-12/loc_cran_packages.csv", show_col_types = FALSE)
glimpse(loc_cran_packages)
```

## Explore
```{r chunk_03}
cran_code |>
  count(language, sort = TRUE)
```
```{r chunk_04}
by_language <- cran_code |>
  group_by(language) |>
  summarize(packages = n(),
            code = sum(code),
            comments = sum(comment),
            files = sum(file),
            lines_per_package = code / packages,
            files_per_package = files / packages,
            comment_code_ratio = comments / code) |>
  arrange(desc(packages))
```
```{r chunk_05}
by_language |>
  head(20) |>
  mutate(language = fct_reorder(language, packages)) |>
  ggplot(aes(language, packages)) +
  geom_col() +
  coord_flip() +
  labs(x = "",
       y = "# of packages that have code from this language")

library(tidytext)

by_language |>
  gather(metric, value, packages, code, files) |>
  group_by(metric) |>
  top_n(8, value) |>
  ungroup() |>
  mutate(language = reorder_within(language, value, metric),
         metric = str_to_title(metric)) |>
  ggplot(aes(language, value)) +
  geom_col() +
  coord_flip() +
  scale_x_reordered() +
  scale_y_continuous(labels = scales::comma) +
  facet_wrap(~ metric, scales = "free", ncol = 1) +
  labs(x = "Value (# of lines of code, files, or packages)")
```
```{r chunk_06}
by_language |>
  filter(packages >= 20) |>
  ggplot(aes(packages, comment_code_ratio)) +
  geom_point() +
  geom_text(aes(label = language), check_overlap = TRUE, vjust = 1, hjust = 1) +
  scale_x_log10() +
  expand_limits(x = 10) +
  labs(x = "# of packages language is used in",
       y = "Comment/Code ratio")
```
```{r chunk_07}
by_language |>
  filter(packages >= 20) |>
  ggplot(aes(packages, lines_per_package)) +
  geom_point() +
  geom_text(aes(label = language), check_overlap = TRUE, vjust = 1, hjust = 1) +
  scale_x_log10() +
  expand_limits(x = 10) +
  labs(x = "# of packages language is used in",
       y = "Lines per package")
```
```{r chunk_08}
cran_code |>
  filter(language == "R") |>
  ggplot(aes(code)) +
  geom_histogram() +
  scale_x_log10(labels = scales::comma)
```
```{r chunk_09}
cran_code |>
  filter(language == "R") |>
  arrange(desc(code))
```
```{r chunk_10}
packages <- tidyverse_packages() |>
  str_extract("[a-z\\d]+")

cran_code |>
  filter(pkg_name %in% packages) |>
  mutate(pkg_name = fct_reorder(pkg_name, code, sum),
         language = fct_lump(language, 6),
         language = fct_reorder(language, code, sum)) |>
  ggplot(aes(pkg_name, code, fill = language)) +
  geom_col() +
  guides(fill = guide_legend(reverse = TRUE)) +
  coord_flip() +
  labs(title = "How much code does each tidyverse package have?",
       x = "",
       y = "# of lines of code",
       fill = "Language")
```
```{r chunk_11}
cran_code |>
  filter(pkg_name %in% packages) |>
  filter(language == "R") |>
  mutate(comment_code_ratio = comment / code) |>
  arrange(desc(comment_code_ratio)) |>
```
```{r chunk_12}
cran_code |>
  filter(language == "R", code >= 100) |>
  mutate(tidyverse = ifelse(pkg_name %in% packages, "Tidyverse", "Other")) |>
  ggplot(aes(code / comment)) +
  geom_histogram() +
  scale_x_log10(labels = scales::number_format(accuracy = .1)) +
  facet_wrap(~ tidyverse, ncol = 1, scales = "free_y") +
  labs(x = "Code to comment ratio")
```
```{r chunk_13}
cran_code |>
  filter(code >= 100, language == "R", comment > 0) |>
  mutate(code_comment_ratio = code / comment) |>
  arrange(desc(code_comment_ratio))
```
```{r chunk_14}
pkgs <- available.packages()

head(pkgs)

# Take package downloads from yesterday
package_downloads <- read_csv("http://cran-logs.rstudio.com/2019/2019-12-19.csv.gz")

downloads_by_package <- package_downloads |>
  distinct(package, ip_id) |>
  count(pkg_name = package, sort = TRUE, name = "downloads")
```
```{r chunk_15}
cran_code |>
  filter(language == "R") |>
  inner_join(downloads_by_package, by = "pkg_name") |>
  arrange(desc(downloads)) |>
  filter(downloads >= 10) |>
  ggplot(aes(downloads, code)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  scale_y_log10()

cran_code |>
  filter(language == "R") |>
  inner_join(downloads_by_package, by = "pkg_name") |>
  arrange(desc(downloads)) |>
  filter(downloads >= 10, code >= 10) |>
  ggplot(aes(downloads, code / comment)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  scale_y_log10()
```
```{r chunk_16}
package_metadata <- available.packages() |>
  as_tibble() |>
  janitor::clean_names() |>
  select(-version, -file)
  
cran_code |>
  filter(language == "R") |>
  inner_join(package_metadata, by = c(pkg_name = "package")) |>
  mutate(license = fct_lump(license, 10),
         license = fct_reorder(license, code)) |>
  ggplot(aes(license, code)) +
  geom_boxplot() +
  coord_flip() +
  scale_y_log10()

n_import <- package_metadata |>
  select(package, imports) |>
  separate_rows(imports, sep = ",") |>
  extract(imports, "import", "([A-Za-z\\d\\.]+)") |>
  filter(!is.na(import)) |>
  count(import, sort = TRUE, name = "n_reverse_import")
```
```{r chunk_17}
cran_code |>
  filter(language == "R") |>
  left_join(n_import, by = c(pkg_name = "import")) |>
  replace_na(list(n_reverse_import = 0)) |>
  filter(n_reverse_import >= 10) |>
  ggplot(aes(n_reverse_import, code)) +
  geom_point() +
  geom_text(aes(label = pkg_name), vjust = 1, hjust = 1, check_overlap = TRUE) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Number of packages that IMPORT this",
       y = "Lines of R code in package")
```

---

**→ [Hour 2: Modeling](h2_modeling_code_cran_packages.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
