---
title: "Hour 2 — Modeling: Analyzing code in CRAN packages in R"
subtitle: "Going deeper with statistical models | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 018 · Hour 2 · [← Hour 1 EDA](h1_eda_code_cran_packages.qmd)
:::


::: {.callout-note collapse="true"}
## Atomic Concepts — Hour 2
`boxplot_jitter` · `coord_flip_horizontal` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_histogram_density` · `geom_smooth_loess` · `grouped_summary_ggplot` · `janitor_clean` · `log_scale`
:::

## Setup

```{r setup}
library(tidyverse)
library(scales)
library(broom)
library(here)
theme_set(theme_light(base_size = 13))
```

## Reload data

```{r load-data}
# Re-load from Hour 1 — same local paths
source_qmd <- here::here("output/episodes/code_cran_packages/h1_eda_code_cran_packages.qmd")
# Data files are re-read directly below if needed
```

## Model 1 — Sentiment analysis

```{r text-setup}
library(tidytext)
df <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2019/2019-11-12/loc_cran_packages.csv", show_col_types = FALSE)
```

```{r sentiment}
afinn <- get_sentiments("afinn")

# Adjust column names to match your data
# df |> unnest_tokens(word, text_column) |> inner_join(afinn, by = "word") |>
#   group_by(group_col) |> summarise(mean_sentiment = mean(value), .groups = "drop") |>
#   mutate(group_col = fct_reorder(group_col, mean_sentiment)) |>
#   ggplot(aes(mean_sentiment, group_col)) + geom_col()
cat("Sentiment analysis: customise the column names above to your dataset\n")
```

## Model 2 — TF-IDF: most distinctive terms by group

```{r tfidf}
# df |> unnest_tokens(word, text_column) |>
#   anti_join(stop_words, by = "word") |>
#   count(group_col, word) |>
#   bind_tf_idf(word, group_col, n) |>
#   group_by(group_col) |>
#   slice_max(tf_idf, n = 8, with_ties = FALSE) |>
#   mutate(word = reorder_within(word, tf_idf, group_col)) |>
#   ggplot(aes(tf_idf, word, fill = group_col)) +
#   geom_col(show.legend = FALSE) +
#   facet_wrap(~group_col, scales = "free_y") +
#   scale_y_reordered()
cat("TF-IDF: customise column names above\n")
```


---

**→ [Hour 3: Communicate](h3_communicate_code_cran_packages.qmd)**
