---
title: "Hour 1 — EDA: Analyzing HBCU enrollment in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 034 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`cumsum_running_total` · `faceted_comparison` · `fct_relevel_manual` · `fct_reorder_stat` · `grouped_summary_ggplot` · `multi_table_join` · `pivot_reshape` · `spline_regression` · `str_detect_filter` · `theme_set_global` · `tidytuesdayr_load` · `time_series_trend`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
104.10 <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-02-02/104.10.xlsx", show_col_types = FALSE)
bach_students <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-02-02/bach_students.csv", show_col_types = FALSE)
female_bach_students <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-02-02/female_bach_students.csv", show_col_types = FALSE)
female_hs_students <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-02-02/female_hs_students.csv", show_col_types = FALSE)
hbcu_all <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-02-02/hbcu_all.csv", show_col_types = FALSE)
hbcu_black <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-02-02/hbcu_black.csv", show_col_types = FALSE)
hs_students <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-02-02/hs_students.csv", show_col_types = FALSE)
male_bach_students <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-02-02/male_bach_students.csv", show_col_types = FALSE)
male_hs_students <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-02-02/male_hs_students.csv", show_col_types = FALSE)
tabn313.20 <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-02-02/tabn313.20.xls", show_col_types = FALSE)
glimpse(104.10)
glimpse(bach_students)
```

## Load the weekly Data
```{r chunk_01}
hbcu_black_long <- tt$hbcu_black |>
  gather(metric, black_enrollment, -Year) |>
  rename(year = Year)

hbcu_compare_long <- hbcu_all_long |>
  full_join(hbcu_black_long, by = c("year", "metric")) |>
  mutate(pct_black = black_enrollment / enrollment)

hbcu_compare_long |>
  filter(metric == "Total enrollment") |>
  ggplot(aes(year, pct_black)) +
  geom_line() +
  scale_y_continuous(labels = percent) +
  expand_limits(y = 0) +
  labs(y = "% of HBCU enrollment that is Black")

hbcu_compare_long |>
  filter(metric %in% c("Males", "Females")) |>
  ggplot(aes(year, pct_black, color = metric)) +
  geom_line() +
  scale_y_continuous(labels = percent) +
  expand_limits(y = 0) +
  labs(y = "% of HBCU enrollment that is Black")

hbcu_compare_long |>
  filter(str_detect(metric, "Total -")) |>
  mutate(metric = str_remove(metric, "Total - ")) |>
  ggplot(aes(year, pct_black, color = metric)) +
  geom_line() +
  scale_y_continuous(labels = percent) +
  expand_limits(y = 0) +
  labs(y = "% of HBCU enrollment that is Black",
       color = "")
```
```{r chunk_02}
gather_race_ethnicity <- function(tbl) {
  tbl |>
    mutate_if(is.character, parse_number) |>
    rename(year = Total) |>
    filter(!is.na(year)) |>
    gather(race_ethnicity, value, -year) |>
    mutate(column = ifelse(str_detect(race_ethnicity, "Standard Errors - "), "standard_error", "percent"),
           race_ethnicity = str_remove(race_ethnicity, "Standard Errors - ")) |>
    spread(column, value) |>
    mutate(standard_error = abs(standard_error)) |>
    filter(!is.na(percent)) |>
    mutate(race_ethnicity = str_remove(race_ethnicity, "1$"),
           percent = percent / 100,
           standard_error = standard_error / 100)
}

hs_over_time <- tt$hs_students |>
  slice(-(1:3)) |>
  gather_race_ethnicity()

bach_over_time <- tt$bach_students |>
  gather_race_ethnicity()

education_over_time <- bind_rows(hs_over_time |> mutate(degree = "High School"),
                                 bach_over_time |> mutate(degree = "Bachelor's"))

hs_over_time |>
  mutate(race_ethnicity = fct_reorder(race_ethnicity, -percent)) |>
  ggplot(aes(year, percent, color = race_ethnicity)) +
  geom_line() +
  scale_y_continuous(labels = percent) +
  labs(color = "Race/ethnicity",
       y = "% of people aged >=25 who graduated HS") +
  expand_limits(y = 0)

bach_over_time |>
  mutate(race_ethnicity = fct_reorder(race_ethnicity, -percent)) |>
  ggplot(aes(year, percent, color = race_ethnicity)) +
  geom_line() +
  scale_y_continuous(labels = percent) +
  labs(color = "Race/ethnicity",
       y = "% of people aged >=25 who graduated a bachelor's program") +
  expand_limits(y = 0)

education_over_time |>
  filter(year >= 1940,
         !str_detect(race_ethnicity, "Islander -")) |>
  mutate(degree = fct_relevel(degree, "High School"),
         race_ethnicity = str_remove(race_ethnicity, "Total - ")) |>
  mutate(race_ethnicity = fct_reorder(race_ethnicity, percent, last, .desc = TRUE)) |>
  ggplot(aes(year, percent, color = race_ethnicity)) +
  geom_line() +
  facet_wrap(~ degree) +
  scale_y_continuous(labels = percent) +
  labs(x = "Year",
       color = "Race/ethnicity",
       y = "% of people aged >=25 who have this degree") +
  expand_limits(y = 0)
```
```{r chunk_03}
a25 <- readxl::read_excel("~/Downloads/A-25.xls")

a25_cleaned <- a25 |>
  select(-starts_with("...")) |>
  rename(field_gender = 1) |>
  mutate(group = cumsum(is.na(field_gender))) |>
  filter(!is.na(field_gender)) |>
  select(group, everything()) |>
  mutate(field_gender = str_remove(field_gender, " \\.\\.\\..*")) |>
  group_by(group) |>
  mutate(field = first(field_gender),
         gender = ifelse(field_gender %in% c("Men", "Women"), field_gender, "Total")) |>
  ungroup() |>
  select(field, gender, everything()) |>
  select(-field_gender, -group)

a25_cleaned |>
  select(field, gender, contains("HBCU")) |>
  rename(pct_hbcu_total = 3,
         pct_hbcu_black = 4) |>
  filter(gender != "Total") |>
  mutate(field = fct_reorder(field, pct_hbcu_black, na.rm = TRUE),
         pct_hbcu_black = pct_hbcu_black / 100) |>
  ggplot(aes(pct_hbcu_black, field, fill = gender)) +
  geom_col(position = "dodge") +
  scale_x_continuous(labels = percent) +
  labs(x = "% of first degrees from an HBCU, among Black students",
       y = "Field",
       fill = "")
```

---

**→ [Hour 2: Modeling](h2_modeling_hbcu_enrollment.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
