---
title: "Hour 1 — EDA: Analyzing Super Bowl ads in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 058 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`boxplot_jitter` · `correlation_matrix` · `faceted_comparison` · `fct_reorder_stat` · `geom_histogram_density` · `grouped_summary_ggplot` · `heatmap_tile` · `linear_model_broom` · `log_scale` · `logistic_regression` · `multi_table_join` · `pivot_reshape`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- text is multilingual without language detection
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- text is long-form English and benefits from word-level decomposition
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(tidytext)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
youtube <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-03-02/youtube.csv", show_col_types = FALSE)
glimpse(youtube)
```

## Load the weekly Data
```{r chunk_01}
youtube |>
  count(brand, sort = TRUE) |>
  head(20) |>
  mutate(brand = fct_reorder(brand, n)) |>
  ggplot(aes(n, brand)) +
  geom_col()
```
```{r chunk_02}
youtube |>
  ggplot(aes(year, fill = brand)) +
  geom_bar() +
  facet_wrap(~ brand) +
  theme(legend.position = "none")
```
```{r chunk_03}
youtube |>
  ggplot(aes(view_count)) +
  geom_histogram(binwidth = .5) +
  scale_x_log10(labels = comma) +
  labs(x = "# of views")
```
```{r chunk_04}
youtube |>
  gather(metric, value, contains("_count")) |>
  ggplot(aes(value)) +
  geom_histogram(binwidth = .5) +
  scale_x_log10(labels = comma) +
  labs(x = "# of views") +
  facet_wrap(~ metric)
```
```{r chunk_05}
youtube |>
  filter(!is.na(view_count)) |>
  mutate(brand = fct_reorder(brand, view_count)) |>
  ggplot(aes(view_count, brand, fill = funny)) +
  geom_boxplot() +
  scale_x_log10(labels = comma)

youtube |>
  ggplot(aes(year, view_count, group = year)) +
  geom_boxplot()

youtube |>
  filter(!is.na(view_count)) |>
  group_by(year) |>
  summarize(n = n(),
            median_views = median(view_count)) |>
  filter(n >= 7) |>
  ggplot(aes(year, median_views)) +
  geom_line() +
  geom_point(aes(size = n)) +
  scale_y_continuous(labels = comma) +
  theme(legend.position = "none") +
  labs(y = "Median # of views of Super Bowl ads")

gathered_categories <- youtube |>
  gather(category, value, funny:use_sex) |>
  mutate(category = str_to_title(str_replace_all(category, "_", " ")))

gathered_categories |>
  ggplot(aes(category, view_count, fill = value)) +
  geom_boxplot() +
  scale_y_log10()

gathered_categories |>
  filter(!is.na(view_count)) |>
  group_by(category, value) |>
  summarize(n = n(),
            median_view_count = median(view_count)) |>
  ggplot(aes(median_view_count, category, fill = value)) +
  geom_col(position = "dodge")

gathered_categories |>
  filter(!is.na(view_count)) |>
  group_by(category) |>
  summarize(correlation = cor(value, log(view_count + 1))) |>
  arrange(desc(correlation))

lm(log2(view_count) ~ danger + patriotic + funny + show_product_quickly + celebrity + animals + use_sex,
   data = youtube) |>
  summary()
```
```{r chunk_06}
glm(celebrity ~ year,
    data = youtube,
    family = "binomial") |>
  summary()

coefficients <- gathered_categories |>
  group_by(category) |>
  summarize(model = list(glm(value ~ year, family = "binomial"))) |>
  mutate(td = map(model, broom::tidy)) |>
  unnest(td) |>
  filter(term != "(Intercept)") |>
  arrange(desc(estimate))

gathered_categories |>
  group_by(category,
           year = 2 * (year %/% 2)) |>
  summarize(pct = mean(value),
            n = n()) |>
  inner_join(coefficients, by = "category") |>
  filter(p.value <= .01) |>
  ggplot(aes(year, pct, color = category)) +
  geom_line() +
  scale_y_continuous(labels = percent) +
  facet_wrap(~ category) +
  theme(legend.position = "none") +
  labs(x = "Time (rounded to 2-years)",
       y = "% of ads with this quality")
```
```{r chunk_07}
library(tidytext)

by_brand_category <- gathered_categories |>
  group_by(brand, category) |>
  summarize(pct = mean(value)) |>
  ungroup()

by_brand_category |>
  mutate(brand = reorder_within(brand, pct, category)) |>
  ggplot(aes(pct, brand)) +
  geom_col() +
  scale_x_continuous(labels = percent) +
  scale_y_reordered() +
  facet_wrap(~ category, scales = "free_y") +
  labs(y = "",
       x = "% of brand's ads have this quality")

by_brand_category |>
  mutate(category = reorder_within(category, pct, brand)) |>
  ggplot(aes(pct, category)) +
  geom_col() +
  scale_x_continuous(labels = percent) +
  scale_y_reordered() +
  facet_wrap(~ brand, scales = "free_y") +
  labs(y = "",
       x = "% of brand's ads have this quality") +
  labs(title = "What is each brand's 'fingerprint' in terms of the type of ads it produces?")
```
```{r chunk_08}
by_brand_category |>
  ggplot(aes(category, brand, fill = pct)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", midpoint = .5)

by_brand_category |>
  reshape2::acast(brand ~ category, value.var = "pct") |>
  heatmap()
```
```{r chunk_09}
likes_dislikes <- youtube |>
  mutate(dislike_pct = dislike_count / view_count,
         like_ratio = like_count / dislike_count) |>
  mutate(like_dislike_total = like_count + dislike_count) |>
  filter(like_dislike_total >= 500) |>
  filter(!is.na(dislike_pct)) |>
  select(brand, year, title, description, view_count, like_dislike_total, like_count,
         dislike_count, like_ratio) |>
  mutate(dislike_pct = dislike_count / (like_count + dislike_count)) |>
  arrange(desc(view_count))

likes_dislikes |>
  filter(like_dislike_total >= 1000) |>
  ggplot(aes(view_count, dislike_pct)) +
  geom_point(aes(color = brand)) +
  scale_x_log10(labels = comma) +
  scale_y_continuous(labels = percent) +
  geom_text(aes(label = str_trunc(title, 60)),
            check_overlap = TRUE, vjust = 1,
            hjust = 1, size = 3) +
  labs(x = "Total views",
       y = "Dislikes / (Likes + Dislikes)")

likes_dislikes |>
  mutate(brand = fct_reorder(brand, dislike_pct, mean)) |>
  ggplot(aes(dislike_pct, brand)) +
  geom_boxplot() +
  scale_x_continuous(labels = percent) +
  labs(title = "What brands tend to produce polarizing ads in terms of Youtube likes?",
       x = "Dislikes / (Likes + Dislikes)",
       y = "Brand")
```

---

**→ [Hour 2: Modeling](h2_modeling_super_bowl_ads.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
