---
title: "Hour 1 — EDA: Analyzing bike frequencies in Seattle in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 009 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`case_when_conditional_col` · `coord_flip_horizontal` · `faceted_comparison` · `fct_relevel_manual` · `fct_reorder_stat` · `geom_histogram_density` · `grouped_summary_ggplot` · `log_scale` · `lubridate_dates` · `pivot_reshape` · `scales_axis_labels` · `spline_regression`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(tidyverse)
library(lubridate)
library(scales)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
library(lubridate)
library(scales)
theme_set(theme_light())

bike_traffic_raw <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-04-02/bike_traffic.csv")

bike_traffic <- bike_traffic_raw |>
  mutate(date = mdy_hms(date)) |>
  filter(bike_count < 2000) |>
  select(-ped_count)
```
```{r chunk_02}
bike_traffic |>
  count(crossing, direction)
```
```{r chunk_03}
bike_traffic |>
  ggplot(aes(date, fill = is.na(bike_count))) +
  geom_histogram() +
  facet_grid(crossing ~ direction)
```

## When in the day do we see bikers?
```{r chunk_04}
bike_traffic |>
  group_by(crossing,
           hour = hour(date)) |>
  summarize(bike_count = sum(bike_count, na.rm = TRUE)) |>
  mutate(pct_bike = bike_count / sum(bike_count)) |>
  ggplot(aes(hour, pct_bike, color = crossing)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "When in the day do people bike through these Seattle crossings?",
       subtitle = "Based on crossings from 2014-February 2019",
       color = "Crossing",
       x = "Time of day (local time)",
       y = "% of bike crossings that happen in this hour")
```
```{r chunk_05}
bike_by_time_window <- bike_traffic |>
  mutate(hour = hour(date)) |>
  mutate(time_window = case_when(
    between(hour, 7, 10) ~ "Morning Commute",
    between(hour, 11, 15) ~ "Midday",
    between(hour, 16, 18) ~ "Evening Commute",
    TRUE ~ "Night"
  )) |>
  group_by(crossing,
           time_window) |>
  summarize(number_missing = sum(is.na(bike_count)),
            bike_count = sum(bike_count, na.rm = TRUE)) |>
  mutate(pct_bike = bike_count / sum(bike_count))

bike_by_time_window |>
  select(-number_missing, -bike_count) |>
  spread(time_window, pct_bike) |>
  mutate(TotalCommute = `Evening Commute` + `Morning Commute`) |>
  arrange(desc(TotalCommute))

bike_by_time_window |>
  ggplot(aes(time_window, pct_bike)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ crossing)

bike_by_time_window |>
  group_by(crossing) |>
  summarize(total_bikes = sum(bike_count),
            pct_commute = sum(bike_count[str_detect(time_window, "Commute")]) / total_bikes) |>
  ggplot(aes(total_bikes, pct_commute)) +
  geom_point() +
  scale_x_log10()
```
```{r chunk_06}
bike_traffic |>
  group_by(crossing,
           weekday = wday(date, label = TRUE),
           hour = hour(date)) |>
  summarize(total_bikes = sum(bike_count, na.rm = TRUE)) |>
  group_by(crossing) |>
  mutate(pct_bike = total_bikes / sum(total_bikes)) |>
  ggplot(aes(hour, pct_bike, color = crossing)) +
  geom_line(show.legend = FALSE) +
  facet_grid(crossing ~ weekday) +
  scale_y_continuous(labels = percent_format()) +
  labs(x = "Time of week",
       y = "% of bike crossings happening in this hour",
       title = "When in the week do people in Seattle bike?",
       subtitle = "Based on crossings from 2014-February 2019")
```
```{r chunk_07}
bike_traffic |>
  filter(date < "2018-01-01") |>
  group_by(crossing,
           month = fct_relevel(month.name[month(date)], month.name)) |>
  summarize(total_bikes = sum(bike_count, na.rm = TRUE)) |>
  mutate(pct_bike = total_bikes / sum(total_bikes)) |>
  ggplot(aes(month, pct_bike, color = crossing, group = crossing)) +
  geom_line() +
  expand_limits(y = 0) +
  scale_y_continuous(labels = percent_format()) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "What time of year do people bike?",
       subtitle = "Based on 2014-2017 bike crossings",
       y = "% of yearly trips in this month",
       x = "")
```

## What directions do people commute by bike?
```{r chunk_08}
bike_by_direction_hour_crossing <- bike_traffic |>
  filter(crossing != "MTS Trail",
         !wday(date, label = TRUE) %in% c("Sat", "Sun"),
         direction %in% c("North", "South")) |>
  mutate(hour = hour(date)) |>
  group_by(crossing,
           direction,
           hour) |>
  summarize(bike_count = sum(bike_count, na.rm = TRUE)) |>
  mutate(pct_bike = bike_count / sum(bike_count))

bike_by_direction_hour_crossing |>
  group_by(crossing) |>
  mutate(average_hour = sum((hour * pct_bike)[direction == "North"])) |>
  ungroup() |>
  mutate(crossing = fct_reorder(crossing, average_hour)) |>
  ggplot(aes(hour, pct_bike, color = direction)) +
  geom_line() +
  facet_grid(crossing ~ .) +
  scale_y_continuous(labels = percent_format()) +
  labs(x = "Time of day",
       y = "% of bike crossings happening in this hour",
       title = "In which directions do people commute by bike?",
       subtitle = "Based on weekday crossings at six Seattle locations from 2014-February 2019",
       color = "Direction")
```

---

**→ [Hour 2: Modeling](h2_modeling_bike_frequencies_seattle.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
