---
title: "Hour 1 — EDA: Analyzing Chopped epiodes in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 016 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `boxplot_jitter` · `fct_lump_infrequent` · `fct_relevel_manual` · `fct_reorder_stat` · `geom_histogram_density` · `glue_strings` · `grouped_summary_ggplot` · `janitor_clean` · `linear_model_broom` · `log_scale` · `network_graph_ggraph`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(glue)
library(widyr)
library(ggraph)
library(tidygraph)
library(tidymodels)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
chopped-raw <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-08-25/chopped-raw.csv", show_col_types = FALSE)
chopped <- read_tsv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-08-25/chopped.tsv", show_col_types = FALSE)
glimpse(chopped-raw)
glimpse(chopped)
```

## Load the weekly Data
```{r chunk_01}
chopped |>
  ggplot(aes(episode_rating)) +
  geom_histogram()

chopped |>
  arrange(episode_rating) |>

chopped |>
  filter(!is.na(episode_rating)) |>
  ggplot(aes(series_episode, episode_rating)) +
  geom_line(alpha = .5, color = "gray") +
  geom_point(aes(color = factor(season))) +
  geom_text(aes(label = episode_name), hjust = 1,
            check_overlap = TRUE) +
  theme(legend.position = "none")

chopped |>
  filter(!is.na(episode_rating)) |>
  group_by(season) |>
  summarize(n_episodes = n(),
            avg_rating = mean(episode_rating)) |>
  ggplot(aes(season, avg_rating)) +
  geom_line() +
  geom_point(aes(size = n_episodes)) +
  theme(legend.position = "none") +
  labs(x = "Season",
       y = "Average Rating")
```
```{r chunk_02}
library(glue)

chopped |>
  arrange(desc(episode_rating)) |>
  head(25) |>
  mutate(name = glue("{ season }.{season_episode} { episode_name }"),
         name = fct_reorder(name, episode_rating)) |>
  ggplot(aes(episode_rating, name)) +
  geom_point()
```

## Ingredients!
```{r chunk_03}
ingredients <- chopped |>
  select(season, season_episode, series_episode, episode_name,
         episode_rating, appetizer:dessert) |>
  pivot_longer(cols = c(appetizer:dessert),
               names_to = "course",
               values_to = "ingredient") |>
  separate_rows(ingredient, sep = ", ") |>
  mutate(course = fct_relevel(course, c("appetizer", "entree")))

ingredients |>
  count(course, ingredient, sort = TRUE) |>
  filter(fct_lump(ingredient, 25, w = n) != "Other") |>
  mutate(ingredient = fct_reorder(ingredient, n, sum),
         course = fct_rev(course)) |>
  ggplot(aes(n, ingredient, fill = course)) +
  geom_col() +
  scale_fill_discrete(guide = guide_legend(reverse = TRUE)) +
  labs(x = "# of episodes",
       y = "",
       title = "Most common ingredients in Chopped",
       fill = "Course")
```
```{r chunk_04}
library(widyr)
library(ggraph)
library(tidygraph)

ingredients_filtered <- ingredients |>
  add_count(ingredient) |>
  filter(n >= 8)

ingredient_correlations <- ingredients_filtered |>
  pairwise_cor(ingredient, series_episode, sort = TRUE)

ingredients_filtered |>
  pairwise_count(ingredient, series_episode, sort = TRUE)

# Not sure this is useful since they appear across courses
ingredient_correlations |>
  head(75) |>
  ggraph(layout = "fr") +
  geom_edge_link(aes(edge_alpha = correlation)) +
  geom_node_point() +
  geom_node_text(aes(label = name), repel = TRUE)

# Do any pairs of ingredients appear together in the same course
# across episodes?
ingredients_filtered |>
  unite(episode_course, series_episode, course) |>
  pairwise_count(ingredient, episode_course, sort = TRUE)
```

## Are there ingredients that got more popular in later seasons?
```{r chunk_05}
early_late_ingredients <- ingredients_filtered |>
  group_by(ingredient) |>
  summarize(first_season = min(season),
            avg_season = mean(season),
            last_season = max(season),
            n_appearances = n()) |>
  arrange(desc(avg_season)) |>
  slice(c(1:6, tail(row_number())))

ingredients_filtered |>
  semi_join(early_late_ingredients, by = "ingredient") |>
  mutate(ingredient = fct_reorder(ingredient, season)) |>
  ggplot(aes(season, ingredient)) +
  geom_boxplot()
```

## What ingredients are "popular"
```{r chunk_06}
ingredients_wide <- ingredients_filtered |>
  select(season, series_episode, episode_rating, ingredient) |>
  mutate(value = 1) |>
  pivot_wider(names_from = ingredient,
              values_from = value,
              values_fill = list(value = 0)) |>
  select(-series_episode) |>
  janitor::clean_names()

lm(episode_rating ~ season, data = ingredients_wide) |>
  summary()

library(tidymodels)

set.seed(2020)
split_data <- ingredients_wide |>
  filter(!is.na(episode_rating)) |>
  initial_split()

training_set <- training(split_data)
```
```{r chunk_07}
cv_samples <- training_set |>
  vfold_cv(v = 10)
```
```{r chunk_08}
model_spec <- linear_reg(penalty = tune()) |>
  set_engine("glmnet")

parameter_search <- model_spec |>
  tune_grid(episode_rating ~ ., resamples = cv_samples)

parameter_search |>
  collect_metrics() |>
  filter(.metric == "rmse") |>
  ggplot(aes(penalty, mean)) +
  geom_line() +
  scale_x_log10() +
  labs(y = "Mean Squared Error")
```
```{r chunk_09}
rf_spec <- rand_forest(mode = "regression",
                       trees = tune()) |>
  set_engine("ranger")

cv_samples <- training_set |>
  vfold_cv(v = 10)

parameter_search <- rf_spec |>
  tune_grid(episode_rating ~ ., resamples = cv_samples)

parameter_search |>
  collect_metrics() |>
  filter(.metric == "rmse") |>
  ggplot(aes(trees, mean)) +
  geom_line() +
#   scale_x_log10() +
  labs(x = "# of trees in random forest",
       y = "Mean Squared Error")

model <- rand_forest(mode = "regression", mtry = 3, trees = 500) |>
  set_engine("ranger") |>
  fit(episode_rating ~ ., training_set)

test_set <- testing(split_data)

predict(model, test_set) |>
  bind_cols(test_set) |>
  rmse(.pred, episode_rating)

test_set |>
  mutate(average = mean(episode_rating)) |>
  rmse(average, episode_rating)
```
```{r chunk_10}
rec <- recipe(episode_rating ~ season, training_set) |>
  step_ns(season, deg_free = tune())

parameter_search_df <- linear_reg() |>
  set_engine("lm") |>
  tune_grid(rec, resamples = cv_samples)

parameter_search_df |>
  collect_metrics() |>
  filter(.metric == "rmse") |>
  ggplot(aes(deg_free, mean)) +
  geom_line()

training_data_processed <- recipe(episode_rating ~ season, training_set) |>
  step_ns(season, deg_free = 2) |>
  prep() |>
  juice()

spline_model <- linear_reg() |>
  set_engine("lm") |>
  fit(episode_rating ~ season, data = juice(training_data_processed))
```

---

**→ [Hour 2: Modeling](h2_modeling_chopped_epiodes.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
