---
title: "Hour 1 — EDA: Analyzing caribou locations in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 015 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`choropleth_sf` · `faceted_comparison` · `geom_histogram_density` · `grouped_summary_ggplot` · `log_scale` · `lubridate_dates` · `spline_regression` · `theme_set_global` · `tidytuesdayr_load` · `time_series_trend`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- region name spelling differs between data and shapefile — silent dropped rows
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- geographic column can be matched to a standard ISO/FIPS shapefile without fuzzy joins
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(sf)
library(lubridate)
library(geosphere)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
individuals <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-06-23/individuals.csv", show_col_types = FALSE)
locations <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-06-23/locations.csv", show_col_types = FALSE)
glimpse(individuals)
glimpse(locations)
```

## Explore
```{r chunk_02}
individuals <- individuals
locations <- locations
```
```{r chunk_03}
individuals |>
  summarize(across(sex:study_site, list(~ mean(!is.na(.)))))

individuals |>
  filter(deploy_off_type == "dead") |>
  count(death_cause, sort = TRUE)
```
```{r chunk_04}
library(sf)
province_sf <- read_sf("~/Downloads/province")

bc <- province_sf |>
  filter(PROV == "BC")
```
```{r chunk_05}
individuals |>
  filter(deploy_on_latitude > 40) |>
  count(study_site, deploy_on_longitude, deploy_on_latitude, sort = TRUE) |>
  ggplot() +
  # geom_sf(data = bc) +
  # borders("world", regions = "canada") +
  geom_point(aes(deploy_on_longitude, deploy_on_latitude,
                 size = n,
             color = study_site)) +
  scale_size_continuous(guide = FALSE)
```
```{r chunk_06}
individuals |>
  count(animal_id, sort = TRUE)

individuals |>
  filter(animal_id == "MO_car150") |>
```

## Looking at locations
```{r chunk_07}
locations |>
  ggplot(aes(longitude, latitude, color = study_site)) +
  geom_point()
```
```{r chunk_08}
by_animal <- locations |>
  group_by(animal_id, study_site) |>
  summarize(start = min(timestamp),
            end = max(timestamp),
            num_points = n()) |>
  ungroup() |>
  arrange(desc(num_points))
```
```{r chunk_09}
library(lubridate)

example_animal <- locations |>
  arrange(timestamp) |>
  filter(animal_id == sample(unique(animal_id), 1))

example_animal |>
  mutate(quarter = as.Date(floor_date(timestamp, "quarter"))) |>
  ggplot(aes(longitude, latitude, color = timestamp)) +
  geom_point(alpha = .5) +
  geom_path(alpha = .5) +
  facet_wrap(~ quarter) +
  labs(title = "One caribou over time")
```
```{r chunk_10}
library(geosphere)

locations_with_gaps <- locations |>
  group_by(animal_id) |>
  mutate(last_lon = lag(longitude),
         last_lat = lag(latitude),
         hours = as.numeric(difftime(timestamp, lag(timestamp), unit = "hours")),
         km = distHaversine(cbind(longitude, latitude), cbind(last_lon, last_lat)) / 1000,
         kph = km / hours) |>
  ungroup()

locations_with_gaps |>
  filter(hours <= 8) |>
  ggplot(aes(kph)) +
  geom_histogram() +
  scale_x_log10(labels = scales::comma) +
  labs(title = "On average, how fast do caribou travel?")

by_animal <- locations_with_gaps |>
  filter(hours <= 8,
         hours >= .5) |>
  group_by(animal_id, study_site) |>
  summarize(start = min(timestamp),
            end = max(timestamp),
            num_points = n(),
            avg_speed = mean(kph, na.rm = TRUE)) |>
  ungroup() |>
  arrange(desc(num_points)) |>
  filter(num_points >= 10)
```
```{r chunk_11}
by_animal |>
  ggplot(aes(num_points, avg_speed)) +
  geom_point() +
  scale_x_log10() +
  expand_limits(y = 0)

by_animal |>
  arrange(desc(avg_speed))

locations_with_gaps |>
  filter(animal_id == "QU_car107") |>
  mutate(quarter = as.Date(floor_date(timestamp, "quarter"))) |>
  ggplot(aes(longitude, latitude, color = kph)) +
  geom_point(alpha = .5) +
  geom_path(alpha = .5) +
  facet_wrap(~ quarter) +
  labs(title = "One caribou over time")

locations_with_gaps |>
  filter(animal_id == "SC_car171") |>
  arrange(desc(kph)) |>
```
```{r chunk_12}
locations_with_gaps |>
  filter(study_site != "Hart Ranges") |>
  filter(hours <= 8,
         hours >= .5) |>
  group_by(month = month(timestamp, label = TRUE),
           study_site) |>
  summarize(avg_speed = median(kph),
            n = n()) |>
  ggplot(aes(month, avg_speed, group = study_site,
             color = study_site)) +
  geom_line() +
  geom_point(aes(size = n)) +
  expand_limits(y = 0) +
  facet_wrap(~ study_site) +
  theme(legend.position = "none") +
  labs(title = "Seasonal trend in Caribou speed",
       y = "Average speed (kph)")
```
```{r chunk_13}
by_animal |>
  filter(num_points >= 10) |>
  arrange(desc(avg_speed))
```
```{r chunk_14}
locations |>
  arrange(timestamp) |>
  group_by(animal_id) |>
  mutate(gap = round(difftime(timestamp, lag(timestamp), unit = "hours"))) |>
  ungroup() |>
  filter(!is.na(gap)) |>
  filter(gap <= 24) |>
  ggplot(aes(gap)) +
  geom_histogram(binwidth = 2)
```

---

**→ [Hour 2: Modeling](h2_modeling_caribou_locations.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
