---
title: "Hour 1 — EDA: Analyzing data on R downloads"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 021 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`coord_flip_horizontal` · `countrycode_standardize` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_histogram_density` · `grouped_summary_ggplot` · `log_scale` · `lubridate_dates` · `pivot_reshape` · `scales_axis_labels` · `spline_regression`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(tidyverse)
library(scales)
library(countrycode)
library(lubridate)
library(cranlogs)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
library(scales)
library(countrycode)
theme_set(theme_light())

r_downloads_year_raw <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018-10-30/r_downloads_year.csv")

r_downloads_year <- r_downloads_year_raw |>
  select(-X1) |>
  mutate(country_name = countrycode(country, "iso2c", "country.name"))
```
```{r chunk_02}
r_downloads_year |>
  count(date) |>
  ggplot(aes(date, n)) +
  geom_line() +
  expand_limits(y = 0) +
  labs(y = "# of R downloads per day")

library(lubridate)

r_downloads_year |>
  count(date) |>
  group_by(weekday = wday(date, label = TRUE)) |>
  summarize(average = mean(n)) |>
  ggplot(aes(weekday, average)) +
  geom_line(group = 1) +
  expand_limits(y = 0) +
  labs(y = "Average downloads per weekday")
```
```{r chunk_03}
r_downloads_year |>
  group_by(week = floor_date(date, "week")) |>
  summarize(n = n_distinct(ip_id)) |>
  filter(week > min(week)) |>
  ggplot(aes(week, n)) +
  geom_line() +
  expand_limits(y = 0) +
  labs(y = "# of R downloads per week (distinct IPs)")
```
```{r chunk_04}
r_downloads_year |>
  mutate(country = countrycode(country, "iso2c", "country.name")) |>
  filter(!is.na(country)) |>
  count(hour = hour(time),
        country = fct_lump(country, 8)) |>
  ggplot(aes(hour, n)) +
  geom_line() +
  expand_limits(y = 0) +
  facet_wrap(~ country, scales = "free_y")
```
```{r chunk_05}
library(countrycode)

r_downloads_year |>
  count(country = countrycode(country, "iso2c", "country.name"), sort = TRUE) |>
  mutate(percent = n / sum(n)) |>
  filter(!is.na(country)) |>
  head(16) |>
  mutate(country = fct_reorder(country, percent)) |>
  ggplot(aes(country, percent)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "What countries install the most R?")
```
```{r chunk_06}
r_downloads_year |>
  mutate(version = fct_lump(version, 8)) |>
  count(date, version) |>
  ggplot(aes(date, n, color = version)) +
  geom_line()
```
```{r chunk_07}
r_downloads_year |>
  count(country = fct_lump(country, 8),
        week = floor_date(date, "week")) |>
  filter(week > min(week)) |>
  ggplot(aes(week, n, color = country)) +
  geom_line()
```

## R package downloads
```{r chunk_08}
package_downloads <- read_csv("http://cran-logs.rstudio.com/2018/2018-10-27.csv.gz")
```
```{r chunk_09}
package_downloads |>
  filter(country %in% c("US", "IN")) |>
  group_by(country, package, sort = TRUE) |>
  summarize(n = n_distinct(ip_id)) |>
  spread(country, n, fill = 0) |>
  ungroup() |>
  mutate(total = US + IN,
         IN = (IN + 1) / sum(IN + 1),
         US = (US + 1) / sum(US + 1),
         ratio = US / IN) |>
  filter(total >= 1000) |>
  arrange((ratio)) |>
```
```{r chunk_10}
library(cranlogs)

cranlogs::cran_downloads(packages = c("tidyverse", "broom"), when = "last-week")
```

## Appendix: Why count only distinct IPs?
```{r chunk_11}
r_download_gaps <- r_downloads_year |>
  mutate(datetime = as.POSIXlt(date) + time) |>
  arrange(datetime) |>
  group_by(ip_id) |>
  mutate(gap = as.numeric(datetime - lag(datetime))) |>
  filter(!is.na(gap))
```
```{r chunk_12}
ip_counts <- r_downloads_year |>
  count(ip_id, sort = TRUE)
```
```{r chunk_13}
r_download_gaps |>
  ggplot(aes(gap)) +
  geom_histogram() +
  geom_vline(color = "red", lty = 2, xintercept = 86400) +
  scale_x_log10(breaks = 60 ^ (0:4),
                labels = c("Second", "Minute", "Hour", "2.5 Days", "120 Days"))
```
```{r chunk_14}
r_download_gaps
```

---

**→ [Hour 2: Modeling](h2_modeling_downloads.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
