---
title: "Hour 2 — Modeling: Analyzing Dog Breeds in R"
subtitle: "Go deeper with statistics and models | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode: `dog_breeds` | [All episodes](https://thedatashrink.com.au/tidytuesday) | [YouTube](#) | [Instagram](#)
:::


::: {.callout-note collapse="true"}
## Atomic Concepts (DataX Index)
These reusable concepts are demonstrated in this episode. See `CONCEPT_INDEX.md`.

- `grouped_summary_ggplot`
:::


::: {.callout-warning collapse="true"}
## FAILS WHEN (DataX Stage 5)
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
:::

::: {.callout-important collapse="true"}
## WORKS WHEN (DataX Stage 5)
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
:::


## Episode Context

- **Analysis type:** `EDA` — this shapes the modeling approach below
- **Pattern:** `load_inspect_compare`
- **Hour 2 goal:** Go deeper than David went in the original screencast

## Hour 2 Checklist
- [ ] Fit linear model on main numeric outcome
- [ ] Compute bootstrap confidence intervals for key estimates
- [ ] Test for interaction effects between top 2 grouping vars
- [ ] Summarise effect sizes with broom::tidy()


## Statistical Model

```{r}
#| label: model
#| eval: false
# Fit linear model on main numeric outcome
# Use broom::tidy() + broom::glance() for clean output
```

## Bootstrap Confidence Intervals

```{r}
#| label: bootstrap
#| eval: false
library(rsample)
# boot <- bootstraps(data, times = 1000)
# Calculate statistic of interest, extract CI
```

## Interaction Effects

```{r}
#| label: interaction
#| eval: false
# Test top 2 grouping variable interactions
# Plot with interaction plots or faceted model output
```


## Model Story

> What does the model reveal that raw EDA couldn't?
> One paragraph for the YouTube "deep dive" section.

*[Write model narrative here]*

## Cross-Reference

- [← Hour 1 EDA](h1_eda_dog_breeds.qmd)
- [Hour 3 Communicate →](h3_communicate_dog_breeds.qmd)
