---
title: "Hour 1 — EDA: Analyzing and predicting horror movie ratings in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 003 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `boxplot_jitter` · `broom_model_output` · `coord_flip_horizontal` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_histogram_density` · `geom_smooth_loess` · `grouped_summary_ggplot` · `lasso_glmnet` · `linear_model_broom` · `log_scale`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- text is multilingual without language detection
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- text is long-form English and benefits from word-level decomposition
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(tidytext)
library(glmnet)
library(Matrix)
library(broom)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
theme_set(theme_light())

horror_movies_raw <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-10-22/horror_movies.csv")
```
```{r chunk_02}
horror_movies <- horror_movies_raw |>
  arrange(desc(review_rating)) |>
  extract(title, "year", "\\((\\d\\d\\d\\d)\\)$", remove = FALSE, convert = TRUE) |>
  mutate(budget = parse_number(budget)) |>
  separate(plot, c("director", "cast_sentence", "plot"), extra = "merge", sep = "\\. ", fill = "right") |>
  distinct(title, .keep_all = TRUE)
```
```{r chunk_03}
horror_movies |>
  count(genres, sort = TRUE)

horror_movies |>
  count(language, sort = TRUE)

horror_movies |>
  ggplot(aes(budget)) +
  geom_histogram() +
  scale_x_log10(labels = scales::dollar)
```
```{r chunk_04}
horror_movies |>
  ggplot(aes(budget, review_rating)) +
  geom_point() +
  scale_x_log10(labels = scales::dollar) +
  geom_smooth(method = "lm")
```
```{r chunk_05}
horror_movies |>
  mutate(movie_rating = fct_lump(movie_rating, 5),
         movie_rating = fct_reorder(movie_rating, review_rating, na.rm = TRUE)) |>
  ggplot(aes(movie_rating, review_rating)) +
  geom_boxplot() +
  coord_flip()

horror_movies |>
  filter(!is.na(movie_rating)) |>
  mutate(movie_rating = fct_lump(movie_rating, 5)) |>
  lm(review_rating ~ movie_rating, data = .) |>
  anova()
```
```{r chunk_06}
horror_movies |>
  separate_rows(genres, sep = "\\| ") |>
  mutate(genre = fct_lump(genres, 5)) |>
  ggplot(aes(genre, review_rating)) +
  geom_boxplot()
```
```{r chunk_07}
library(tidytext)

horror_movies_unnested <- horror_movies |>
  unnest_tokens(word, plot) |>
  anti_join(stop_words, by = "word") |>
  filter(!is.na(word))

horror_movies_unnested |>
  filter(!is.na(review_rating)) |>
  group_by(word) |>
  summarize(movies = n(),
            avg_rating = mean(review_rating)) |>
  arrange(desc(movies)) |>
  filter(movies >= 100) |>
  mutate(word = fct_reorder(word, avg_rating)) |>
  ggplot(aes(avg_rating, word)) +
  geom_point()
```

## Lasso regression for predicing review rating based on words in plot
```{r chunk_08}
library(glmnet)
library(Matrix)

movie_word_matrix <- horror_movies_unnested |>
  filter(!is.na(review_rating)) |>
  add_count(word) |>
  filter(n >= 20) |>
  count(title, word) |>
  cast_sparse(title, word, n)

rating <- horror_movies$review_rating[match(rownames(movie_word_matrix), horror_movies$title)]

lasso_model <- cv.glmnet(movie_word_matrix, rating)
```
```{r chunk_09}
library(broom)

tidy(lasso_model$glmnet.fit) |>
  filter(term %in% c("quickly", "seek", "army", "teacher", "unexpected", "friends", "evil")) |>
  ggplot(aes(lambda, estimate, color = term)) +
  geom_line() +
  scale_x_log10() +
  geom_vline(xintercept = lasso_model$lambda.min) +
  geom_hline(yintercept = 0, lty = 2)
```
```{r chunk_10}
plot(lasso_model)

tidy(lasso_model$glmnet.fit) |>
  filter(lambda == lasso_model$lambda.min,
         term != "(Intercept)") |>
  mutate(term = fct_reorder(term, estimate)) |>
  ggplot(aes(term, estimate)) +
  geom_col() +
  coord_flip()
```
```{r chunk_11}
features <- horror_movies |>
  filter(!is.na(review_rating)) |>
  select(title, genres, director, cast, movie_rating, language, release_country) |>
  mutate(director = str_remove(director, "Directed by ")) |>
  gather(type, value, -title) |>
  filter(!is.na(value)) |>
  separate_rows(value, sep = "\\| ?") |>
  unite(feature, type, value, sep = ": ") |>
  mutate(n = 1)

movie_feature_matrix <- horror_movies_unnested |>
  filter(!is.na(review_rating)) |>
  count(title, feature = paste0("word: ", word)) |>
  bind_rows(features) |>
  add_count(feature) |>
  filter(n >= 10) |>
  cast_sparse(title, feature)

rating <- horror_movies$review_rating[match(rownames(movie_feature_matrix), horror_movies$title)]

feature_lasso_model <- cv.glmnet(movie_feature_matrix, rating)
```
```{r chunk_12}
plot(feature_lasso_model)

tidy(feature_lasso_model$glmnet.fit) |>
  filter(lambda == feature_lasso_model$lambda.1se,
         term != "(Intercept)") |>
  mutate(term = fct_reorder(term, estimate)) |>
  ggplot(aes(term, estimate)) +
  geom_col() +
  coord_flip() +
  labs(x = "",
       y = "Coefficient for predicting movie rating",
       title = "What affects a horror movie rating?",
       subtitle = "Based on a lasso regression to predict IMDb ratings of ~3000 movies")
```
```{r chunk_13}
horror_movies |>
  filter(str_detect(genres, "Comedy"),
         !is.na(movie_rating),
         !is.na(budget),
         movie_rating != "PG") |>
  arrange(desc(review_rating)) |>
  select(title, review_rating, movie_rating, plot, director, budget, language) |>
```

---

**→ [Hour 2: Modeling](h2_modeling_predicting_horror_movie_ratings.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
