---
title: "Hour 1 — EDA: Analyzing Video Games in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 074 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_histogram_density` · `grouped_summary_ggplot` · `interactive_plotly` · `log_scale` · `lubridate_dates` · `multi_table_join` · `parse_number_dirty` · `pivot_reshape` · `purrr_map_iterate`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(plotly)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
games <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-03-16/games.csv", show_col_types = FALSE)
sliced-tidytuesday <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-03-16/sliced-tidytuesday.csv", show_col_types = FALSE)
glimpse(games)
glimpse(sliced-tidytuesday)
```

## Load the weekly Data
```{r chunk_01}
games |>
  count(date) |>
  ggplot(aes(date, n)) +
  geom_col()

summarize_games <- function(tbl) {
  tbl |>
    summarize(median_avg = median(avg),
              median_peak = median(peak))
}

games |>
  group_by(date) |>
  summarize_games() |>
  ggplot(aes(date, median_avg)) +
  geom_line() +
  expand_limits(y = 0) +
  labs(x = "Month",
       y = "Median popularity of a game on Steam")

games |>
  filter(date < "2020-01-01") |>
  group_by(month) |>
  summarize_games() |>
  ggplot(aes(month, median_avg)) +
  geom_line(group = 1) +
  expand_limits(y = 0) +
  labs(x = "Month of year",
       y = "Median popularity of a game on Steam",
       title = "Seasonal trend in Steam games",
       subtitle = "From 2012-2019")

games |>
  group_by(year, month) |>
  summarize_games() |>
  ggplot(aes(month, median_avg)) +
  geom_line(aes(color = factor(year), group = year)) +
  expand_limits(y = 0) +
  labs(x = "Month of year",
       y = "Median popularity of a game on Steam",
       title = "Seasonal trend in Steam games",
       color = "year")

games |>
  group_by(date) |>
  summarize_games() |>
  ggplot(aes(date, median_peak)) +
  geom_line() +
  expand_limits(y = 0)
```
```{r chunk_02}
games |>
  filter(date == max(date)) |>
  ggplot(aes(avg)) +
  geom_histogram() +
  scale_x_log10(labels = comma,
                breaks = 10 ^ seq(0, 5)) +
  labs(x = "Average # of players across Feb 2021")

games |>
  filter(avg >= 100) |>
  filter(date == max(date)) |>
  ggplot(aes(avg_peak_perc)) +
  geom_histogram()

games |>
  filter(avg >= 100) |>
  filter(date == max(date)) |>
  arrange((avg_peak_perc))

games |>
  filter(avg >= 1000) |>
  filter(date == "2021-02-01") |>
  arrange((avg_peak_perc)) |>
  ggplot(aes(avg, 1 / avg_peak_perc, label = gamename)) +
  geom_point() +
  geom_text(vjust = 1, hjust = 1, check_overlap = TRUE) +
  scale_y_log10() +
  scale_x_log10(labels = comma) +
  labs(x = "Average # of players in Feb 2021",
       y = "Ratio of Peak / Average")

games |>
  filter(avg >= 1000,
         date == max(date)) |>
  arrange((avg_peak_perc)) |>
  ggplot(aes(avg, 1 / avg_peak_perc, label = gamename)) +
  geom_point() +
  geom_text(vjust = 1, hjust = 1, check_overlap = TRUE) +
  scale_y_log10() +
  scale_x_log10(labels = comma) +
  labs(x = "Average # of players in Feb 2021",
       y = "Ratio of Peak / Average")

# library(plotly)
# ggplotly(g)

games |>
  filter(date == max(date)) |>
  arrange(desc(avg))
```
```{r chunk_03}
games |>
  filter(fct_lump(gamename, 16, w = avg) != "Other") |>
  mutate(gamename = fct_reorder(gamename, -avg)) |>
  ggplot(aes(date, avg)) +
  geom_line() +
  expand_limits(y = 0) +
  scale_y_continuous(labels = comma) +
  facet_wrap(~ gamename, scales = "free_y") +
  labs(x = "Month",
       y = "Average players of this game in this month")
```

## What became popular in the pandemic specifically?
```{r chunk_04}
apr_feb_ratios <- games |>
  filter(year == 2020) |>
  select(gamename, month, avg, peak) |>
  pivot_wider(names_from = month, values_from = c(avg, peak)) |>
  select(gamename, contains("January"), contains("April"), contains("February")) |>
  mutate(apr_feb_ratio = avg_April / avg_February) |>
  filter(avg_January >= 100,
         avg_February >= 100) |>
  arrange(desc(apr_feb_ratio))

games |>
  filter(date >= "2018-01-01") |>
  inner_join(apr_feb_ratios |>
               top_n(12, apr_feb_ratio), by = "gamename") |>
  mutate(gamename = as.character(gamename)) |>
  complete(gamename, date, fill = list(avg = 0)) |>
  mutate(gamename = fct_reorder(gamename, -avg)) |>
  ggplot(aes(date, avg)) +
  geom_line() +
  expand_limits(y = 0) +
  geom_vline(xintercept = as.Date("2020-03-01"),
             color = "red", lty = 2) +
  scale_y_continuous(labels = comma) +
  facet_wrap(~ gamename, scales = "free_y") +
  labs(x = "Month",
       y = "Average players of this game in this month")
```

## Readme
```{r Readme}
tt
```

## Glimpse Data
```{r Glimpse}
tt |> 
  map(glimpse)
```

## Save Image
```{r chunk_05}
# This will save your most recent plot
ggsave(
  filename = "My TidyTuesday Plot.png",
  device = "png")
```

---

**→ [Hour 2: Modeling](h2_modeling_video_games.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
