---
title: "Hour 1 — EDA: Analyzing Friends transcripts in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 030 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `boxplot_jitter` · `correlation_matrix` · `faceted_comparison` · `fct_reorder_stat` · `geom_smooth_loess` · `glue_strings` · `grouped_summary_ggplot` · `linear_model_broom` · `log_odds_ratio` · `log_scale` · `multi_table_join`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- text is multilingual without language detection
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- text is long-form English and benefits from word-level decomposition
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(glue)
library(tidytext)
library(tidylo)
library(widyr)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
friends <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-09-08/friends.csv", show_col_types = FALSE)
friends_emotions <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-09-08/friends_emotions.csv", show_col_types = FALSE)
friends_info <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-09-08/friends_info.csv", show_col_types = FALSE)
glimpse(friends)
glimpse(friends_emotions)
```

## Load the weekly Data
```{r chunk_01}
library(glue)

episodes <- tt$friends_info |>
  mutate(full_title = glue("{ season }.{ episode } { title }"),
         full_title = fct_reorder(full_title, season + .001 * episode))

friends <- tt$friends |>
  inner_join(episodes, by = c("season", "episode"))
  
friends |>
  count(season, full_title)

tt$friends |>
  count(speaker, sort = TRUE)
```
```{r chunk_02}
episodes |>
  ggplot(aes(as.integer(full_title), us_views_millions)) +
  geom_line() +
  geom_point(aes(color = factor(season))) +
  geom_text(aes(label = title), vjust = 1, hjust = 1,
            check_overlap = TRUE,
            size = 2) +
  expand_limits(y = 0) +
  labs(x = "Episode number",
       color = "Season")

episodes |>
  ggplot(aes(as.integer(full_title), imdb_rating)) +
  geom_line() +
  geom_point(aes(color = factor(season))) +
  geom_text(aes(label = title), vjust = 1, hjust = 1,
            check_overlap = TRUE,
            size = 2) +
  labs(x = "Episode number",
       y = "IMDB Rating",
       color = "Season")
```

## Predict a rating based on the characters who speak?
```{r chunk_03}
speaker_lines_per_episode <- friends |>
  count(speaker, title, imdb_rating, season) |>
  complete(speaker, title, fill = list(n = 0)) |>
  group_by(title) |>
  fill(imdb_rating, season, .direction = "downup") |>
  ungroup() |>
  add_count(title, wt = n, name = "episode_total") |>
  mutate(pct = n / episode_total)

speaker_lines_per_episode |>
  semi_join(main_cast, by = "speaker") |>
  mutate(speaker = fct_reorder(speaker, pct)) |>
  ggplot(aes(pct, speaker)) +
  geom_boxplot() +
  scale_x_log10()

speaker_lines_per_episode |>
  semi_join(main_cast, by = "speaker") |>
  group_by(speaker) |>
  summarize(correlation = cor(log2(pct), imdb_rating))

friends |>
  count(season, episode, title) |>
  inner_join(distinct(episodes, season, episode, imdb_rating)) |>
  filter(n <= 500) |>
  ggplot(aes(n, imdb_rating)) +
  geom_point() +
  geom_smooth(method = "lm")
```
```{r chunk_04}
speakers_per_episode_wide <- speaker_lines_per_episode |>
  semi_join(main_cast, by = "speaker") |>
  select(-n) |>
  spread(speaker, pct) |>
  select(-title)

speakers_per_episode_wide |>
  select(-season) |>
  mutate(across(.cols = -c(imdb_rating), log2)) |>
  lm(imdb_rating ~ ., data = .) |>
  summary()
```
```{r chunk_05}
library(tidytext)

words_unnested <- friends |>
  select(text, speaker, season, episode, full_title) |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = "word") |>
  filter(!word %in% c("hey", "yeah", "gonna", "uh"))

library(tidylo)

by_speaker_word <- words_unnested |>
  semi_join(main_cast, by = "speaker") |>
  count(speaker, word) |>
  add_count(word, wt = n, name = "word_total") |>
  filter(word_total >= 50)

by_speaker_word |>
  bind_log_odds(speaker, word, n) |>
  group_by(speaker) |>
  top_n(16, log_odds_weighted) |>
  ungroup() |>
  mutate(word = reorder_within(word, log_odds_weighted, speaker)) |>
  ggplot(aes(log_odds_weighted, word)) +
  geom_col(aes(fill = speaker), show.legend = FALSE) +
  scale_y_reordered() +
  facet_wrap(~ speaker, scales = "free") +
  labs(x = "Log odds (this character) / (other characters)",
       title = "What words are most characteristic of each character")

words_unnested |>
  count(word, sort = TRUE)
```
```{r chunk_06}
library(widyr)

friends |>
  unite(scene_id, season, episode, scene) |>
  count(speaker, scene_id) |>
  semi_join(main_cast, by = "speaker") |>
  pairwise_cor(speaker, scene_id, n, sort = TRUE) |>
  mutate(item2 = reorder_within(item2, correlation, item1)) |>
  ggplot(aes(correlation, item2)) +
  geom_col(aes(fill = item1), show.legend = FALSE) +
  scale_y_reordered() +
  facet_wrap(~ item1, scales = "free_y") +
  labs(x = "Correlation between characters appearing in a scene",
       y = "Other character")
```
```{r chunk_07}
friends |>
  unite(scene_id, season, episode, scene) |>
  add_count(speaker, name = "speaker_total") |>
  filter(speaker_total >= 50) |>
  count(speaker, scene_id) |>
  pairwise_cor(speaker, scene_id, n, sort = TRUE) |>
  filter(item1 == "Ross Geller")
```
```{r chunk_08}
friends |>
  count(season, episode, sort = TRUE)

main_cast <- friends |>
  count(speaker, sort = TRUE) |>
  head(6)

main_cast |>
  mutate(speaker = fct_reorder(speaker, n)) |>
  ggplot(aes(n, speaker)) +
  geom_col()

friends |>
  semi_join(main_cast, by = "speaker") |>
  group_by(speaker) |>
  summarize(n_episodes = n_distinct())

friends |>
  count(season)

friends |>
  count(speaker, season) |>
  filter()
```

## Readme
```{r Readme}
tt
```

## Glimpse Data
```{r Glimpse}
tt |> 
  map(glimpse)
```

## Save Image
```{r chunk_09}
# This will save your most recent plot
ggsave(
  filename = "My TidyTuesday Plot.png",
  device = "png")
```

---

**→ [Hour 2: Modeling](h2_modeling_friends_transcripts.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
