---
title: "Hour 1 — EDA: Analyzing post offices in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 052 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `animated_transitions` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_area_stacked` · `grouped_summary_ggplot` · `multi_table_join` · `purrr_map_iterate` · `spline_regression` · `theme_set_global` · `tidytuesdayr_load`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(ggthemes)
library(historydata)
library(gganimate)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Load the weekly Data
```{r Load}
tt <- tt_load("2021-04-13")

tt$post_offices |>
  add_count(name, orig_name, state, county1, sort = TRUE) |>
  filter(n >= 2) |>
  arrange(name, orig_name, state, county1) |>

post_offices <- tt$post_offices |>
  select(name, state, county1, established, discontinued, continuous,
         stamp_index, id,
         coordinates,
         latitude,
         longitude,
         gnis_dist,
         gnis_county,
         gnis_state) |>
  filter(established >= 1639,
         is.na(discontinued) | discontinued >= established)
```
```{r chunk_01}
post_office_years <- post_offices |>
  select(name, state, established, discontinued, latitude, longitude) |>
  replace_na(list(discontinued = 2003)) |>
  filter(established >= 1750,
         discontinued <= 2021) |>
  mutate(year = map2(established, discontinued, seq)) |>
  unnest(year)
```
```{r chunk_02}
post_office_years |>
  count(year, name = "n_post_offices") |>
  arrange(desc(year)) |>
  ggplot(aes(year, n_post_offices)) +
  geom_area() +
  labs(x = "Year",
       y = "# of post offices currently active in US")
```
```{r chunk_03}
post_offices_cumulative <- post_office_years |>
  count(year,
        state = fct_lump(state, 16),
        name = "n_post_offices")

post_offices_cumulative |>
  mutate(state = fct_reorder(state, -n_post_offices, sum)) |>
  filter(state != "Other") |>
  ggplot(aes(year, n_post_offices, fill = state)) +
  geom_area() +
  labs(x = "Year",
       y = "# of post offices currently active in US") +
  facet_wrap(~ state) +
  theme(legend.position = "none")
```
```{r chunk_04}
post_office_closures <- post_offices |>
  filter(!is.na(discontinued)) |>
  filter(discontinued >= 1750,
         established >= 1750,
         discontinued <= 2021) |>
  count(state = fct_lump(state, 16),
        decade = 10 * (discontinued %/% 10),
        name = "n_closures")

post_office_closures |>
  mutate(state = fct_reorder(state, -n_closures, sum)) |>
  filter(state != "Other") |>
  ggplot(aes(decade, n_closures, fill = state)) +
  geom_area() +
  labs(x = "Decade",
       y = "# of post offices closed in this decade") +
  facet_wrap(~ state) +
  theme(legend.position = "none") +
  labs(title = "When and where were the most post offices closed?")
```
```{r chunk_05}
post_office_closures |>
  filter(decade < 2000) |>
  inner_join(post_offices_cumulative, by = c("state", "decade" = "year")) |>
  mutate(pct_closed = n_closures / n_post_offices) |>
  filter(n_post_offices >= 50) |>
  filter(state %in% c("KY", "PA")) |>
  ggplot(aes(decade, pct_closed, color = state)) +
  geom_line() +
  scale_y_continuous(labels = percent) +
  labs(y = "% of post offices open at start of decade that closed")
```

## Maps
```{r chunk_06}
library(ggthemes)

states_map <- map_data("state") |>
  as_tibble() |>
  mutate(state = state.abb[match(region, str_to_lower(state.name))]) |>
  replace_na(list(state = "DC"))

library(historydata)

state_pops <- us_state_populations |>
  mutate(state = state.abb[match(state, state.name)]) |>
  replace_na(list(state = "DC"))

post_office_years |>
  filter(year == 2003) |>
  count(state, sort = TRUE) |>
  inner_join(states_map, by = "state") |>
  ggplot(aes(long, lat, group = group, fill = n)) +
  geom_polygon() +
  scale_fill_gradient2(low = "blue", high = "red",
                       midpoint = 750) +
  theme_map() +
  labs(fill = "# of post offices")

post_office_years |>
  filter(year == 2000) |>
  count(state, sort = TRUE) |>
  inner_join(state_pops |> filter(year == 2000), by = "state") |>
  mutate(post_office_density = n / (population / 1e6)) |>
  inner_join(states_map, by = "state") |>
  ggplot(aes(long, lat, group = group, fill = post_office_density)) +
  geom_polygon() +
  scale_fill_gradient2(low = "blue", high = "red",
                       midpoint = 250) +
  theme_map() +
  labs(fill = "Post offices / M people")
```
```{r chunk_07}
post_office_cumulative_all <- post_office_years |>
  count(year,
        state,
        name = "n_post_offices")

by_state_year <- post_office_cumulative_all |>
  left_join(state_pops, by = c("year", "state")) |>
  mutate(post_office_density = n_post_offices / (population / 1e6))

by_state_year |>
  filter(year %in% c(1800, 1850, 1900, 1950)) |>
  inner_join(states_map, by = "state") |>
  ggplot(aes(long, lat, group = group, fill = n_post_offices)) +
  geom_polygon() +
  scale_fill_gradient2(low = "blue", high = "red",
                       mid = "pink",
                       midpoint = 2000) +
  facet_wrap(~ year) +
  theme_map() +
  labs(fill = "Post offices")
```
```{r chunk_08}
library(gganimate)

by_state_year |>
  inner_join(states_map, by = "state") |>
  ggplot(aes(long, lat, group = group, fill = n_post_offices)) +
  geom_polygon() +
  scale_fill_gradient2(low = "blue", high = "red",
                       mid = "pink",
                       midpoint = 2000) +
  transition_manual(year) +
  theme_map() +
  labs(fill = "Post offices",
       title = "{ current_frame }")

anim_save("~/Desktop/post_offices_over_time.gif")
```
```{r chunk_09}
post_office_years |>
  filter(year == 1900,
         !state %in% c("HI", "AK")) |>
  ggplot(aes(longitude, latitude)) +
  borders("state") +
  geom_point(size = .01, alpha = .25) +
  theme_map() +
  coord_map()

anim <- post_office_years |>
  filter(year %% 4 == 0) |>
  filter(!state %in% c("HI", "AK")) |>
  ggplot(aes(longitude, latitude)) +
  borders("state", colour = "lightgray") +
  geom_point(size = .01, alpha = .25, color = "#004B87") +
  transition_manual(year) +
  labs(title = "Post office locations in the continental US: { current_frame }") +
  theme_map() +
  coord_map()

animate(
  anim,
  nframes = 30,
  fps = 6
)

anim_save("~/Desktop/post-office-dots-blue.gif")
```

---

**→ [Hour 2: Modeling](h2_modeling_post_offices.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
