---
title: "Hour 1 — EDA: Analyzing transit costs in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 070 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `boxplot_jitter` · `countrycode_standardize` · `error_bar_coef_plot` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_histogram_density` · `glue_strings` · `grouped_summary_ggplot` · `spline_regression` · `theme_set_global` · `tidytuesdayr_load`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
Merged Costs (1.0) - Sheet1 <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-01-05/Merged Costs (1.0) - Sheet1.csv", show_col_types = FALSE)
transit_cost <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-01-05/transit_cost.csv", show_col_types = FALSE)
glimpse(Merged Costs (1.0) - Sheet1)
glimpse(transit_cost)
```

## Load the weekly Data
```{r chunk_01}
transit_cost |>
  count(city, country, sort = TRUE)

transit_cost |>
  filter(country == "United States") |>
  mutate(line = fct_reorder(line, year)) |>
  ggplot(aes(xmin = start_year, xmax = end_year, y = line,
             color = city,
             size = real_cost)) +
  geom_errorbarh(height = .1) +
  labs(x = "Year",
       y = "",
       color = "City")

transit_cost |>
  ggplot(aes(cost_km_millions)) +
  geom_histogram() +
  scale_x_continuous(labels = dollar) +
  labs(x = "Cost / KM (Millions of USD)")

transit_cost |>
  filter(!is.na(cost_km_millions),
         tunnel_per == 1) |>
  mutate(country = fct_lump(country, 10)) |>
  add_count(country) |>
  mutate(country = glue("{ country } ({ n })"),
         country = fct_reorder(country, cost_km_millions)) |>
  ggplot(aes(cost_km_millions, country)) +
  geom_boxplot() +
  scale_x_continuous(labels = dollar) +
  labs(x = "Cost / KM (Millions of USD)",
       y = "")
```
```{r chunk_02}
transit_cost |>
  filter(!is.na(cost_km_millions),
         tunnel_per == 1,
         country == "China") |>
  mutate(city = fct_lump(city, 10)) |>
  add_count(city) |>
  mutate(city = glue("{ city } ({ n })"),
         city = fct_reorder(city, cost_km_millions)) |>
  ggplot(aes(cost_km_millions, city)) +
  geom_boxplot() +
  scale_x_continuous(labels = dollar) +
  labs(x = "Cost / KM (Millions of USD)",
       y = "") +
  expand_limits(x = 0)

transit_cost |>
  filter(!is.na(cost_km_millions),
         tunnel_per == 1,
         country == "China")
```
```{r chunk_03}
transit_cost |>
  filter(country == "China",
         city == "Shanghai",
         !is.na(start_year),
         !is.na(end_year)) |>
  mutate(city = fct_lump(city, 5)) |>
  mutate(line = fct_reorder(line, year)) |>
  ggplot(aes(xmin = start_year, xmax = end_year, y = line,
             color = city,
             size = real_cost)) +
  geom_errorbarh(height = .1) +
  labs(x = "Year",
       y = "",
       color = "City")

transit_cost |>
  filter(tunnel_per == 1,
         end_year <= 2020,
         country == "China") |>
  group_by(year = (year %/% 5) * 5) |>
  summarize(median_cost_km = median(cost_km_millions),
            n = n()) |>
  ggplot(aes(year, median_cost_km)) +
  geom_line() +
  geom_point(aes(size = n))

transit_cost |>
  filter(tunnel_per == 1,
         end_year <= 2020,
         country == "China") |>
  mutate(year = (year %/% 5) * 5,
         city = fct_lump(city, 5)) |>
  ggplot(aes(year, cost_km_millions, group = year)) +
  geom_boxplot(outlier.size = -1) +
  geom_jitter(aes(color = city), height = 0, width = 1) +
  expand_limits(y = 0) +
  labs(y = "Cost / km (Real USD, Millions)",
       x = "Year",
       title = "Cost distribution / km in China")

transit_cost |>
  filter(tunnel_per == 1,
         end_year <= 2020,
         country == "China") |>
  mutate(city = fct_lump(city, 4)) |>
  ggplot(aes(stations / length, cost_km_millions, size = length,
             color = city)) +
  geom_point() +
  expand_limits(x = 0, y = 0) +
  labs(x = "Stations / km", "Cost / kilometer",
       y = "Cost / km")
```
```{r chunk_04}
transit_cost |>
  ggplot(aes(cost_km_millions, real_cost / length)) +
  geom_point()
```

---

**→ [Hour 2: Modeling](h2_modeling_transit_costs.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
