---
title: "Hour 1 — EDA: Analyzing European energy in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 026 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `grouped_summary_ggplot` · `log_scale` · `pivot_reshape` · `reorder_within_facets` · `spline_regression` · `theme_set_global` · `tidytuesdayr_load` · `time_series_trend` · `tokenization_pipeline`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- text is multilingual without language detection
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- text is long-form English and benefits from word-level decomposition
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(tidytext)
library(ggflags)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
country_totals <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-08-04/country_totals.csv", show_col_types = FALSE)
Electricity_generation_statistics_2019 <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-08-04/Electricity_generation_statistics_2019.xlsx", show_col_types = FALSE)
energy_types <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-08-04/energy_types.csv", show_col_types = FALSE)
glimpse(country_totals)
glimpse(Electricity_generation_statistics_2019)
```

## Load the weekly Data
```{r chunk_01}
tt$country_totals |>
  count(type, sort = TRUE)
```

## Total energy consumption in Europe
```{r chunk_02}
europe_totals <- energy_types |>
  filter(level == "Level 1") |>
  group_by(year, type) |>
  summarize(total_power = sum(gigawatt_hours)) |>
  ungroup() |>
  mutate(type = fct_reorder(type, total_power, sum))

europe_totals |>
  ggplot(aes(year, total_power, fill = type)) +
  geom_col() +
  scale_y_continuous(labels = comma) +
  labs(x = "Year",
       y = "Total power production (gigawatt-hours)",
       fill = "Type")

europe_totals |>
  ggplot(aes(year, total_power)) +
  geom_col() +
  scale_y_continuous(labels = comma) +
  labs(x = "Year",
       y = "Total power production (gigawatt-hours)") +
  facet_wrap(~ type, scales = "free_y")
```
```{r chunk_03}
europe_totals |>
  filter(year == 2018) |>
  mutate(type = fct_reorder(type, total_power, sum)) |>
  ggplot(aes(total_power, type)) +
  geom_col() +
  scale_x_continuous(labels = comma) +
  labs(x = "Total power production in 2018 in Europe (gigawatt-hours)")
```
```{r chunk_04}
library(tidytext)
# devtools::install_github("rensa/ggflags")

library(ggflags)

data_prepared <- energy_types |>
  filter(level == "Level 1",
         year == 2018,
         gigawatt_hours > 0,
         type != "Other") |>
  group_by(type) |>
  mutate(country_name = fct_lump(country_name, 10, w = gigawatt_hours),
         country = as.character(fct_lump(country, 10, w = gigawatt_hours))) |>
  mutate(country_name = reorder_within(country_name, gigawatt_hours, type, fun = sum),
         type = fct_reorder(type, -gigawatt_hours, sum))

data_prepared |>
  filter(country != "Other") |>
  mutate(country = str_to_lower(country)) |>
  ggplot(aes(gigawatt_hours, country_name)) +
  geom_col(width = .1) +
  geom_flag(aes(country = country)) +
  scale_y_reordered() +
  facet_wrap(~ type, scales = "free") +
  scale_x_continuous(labels = comma) +
  scale_country(guide = FALSE) +
  theme_minimal() +
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank()) +
  labs(x = "Total power production in 2018 (gigawatt-hours)",
       y = "")
```
```{r chunk_05}
plot_slopegraph <- function(tbl) {
  tbl |>
    filter(gigawatt_hours > 1) |>
    ggplot(aes(year, gigawatt_hours)) +
    geom_line(aes(group = country)) +
    geom_flag(aes(country = country)) +
    geom_text(aes(label = ifelse(year == 2017, NA, country_name),
                  hjust = ifelse(year == 2016, 1.2, -.2)),
              check_overlap = TRUE) +
    scale_x_continuous(breaks = c(2016, 2018),
                       limits = c(2015, 2019)) +
    theme(panel.grid = element_blank()) +
    labs(x = "",
         y = "Gigawatt hours produced in this year")
}

energy_types |>
  filter(type == "Wind") |>
  plot_slopegraph() +
  scale_y_continuous(labels = comma) +
  # scale_y_log10(labels = comma) +
  labs(title = "Wind power production over time by country")

energy_types |>
  filter(type == "Solar") |>
  plot_slopegraph() +
  scale_y_continuous(labels = comma) +
  # scale_y_log10(labels = comma) +
  labs(title = "Solar power production over time by country")

energy_types |>
  filter(type %in% c("Solar", "Wind", "Conventional thermal", "Hydro")) |>
  group_by(type) |>
  filter(fct_lump(country, 10, w = gigawatt_hours) != "Other") |>
  ungroup() |>
  plot_slopegraph() +
  scale_y_continuous(labels = comma) +
  facet_wrap(~ type, scales = "free_y") +
  # scale_y_log10(labels = comma) +
  labs(title = "Solar power production over time by country")
```
```{r chunk_06}
energy_types |>
  filter(level == "Level 1",
         year == 2018,
         gigawatt_hours > 0) |>
  ggplot(aes(type, gigawatt_hours)) +
  geom_point() +
  geom_flag(aes(country = country))
```

---

**→ [Hour 2: Modeling](h2_modeling_european_energy.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
