---
title: "Hour 1 — EDA: Analyzing board games and predicting ratings in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 011 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `boxplot_jitter` · `coord_flip_horizontal` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_histogram_density` · `grouped_summary_ggplot` · `lasso_glmnet` · `linear_model_broom` · `log_scale` · `multi_table_join`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- text is multilingual without language detection
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- text is long-form English and benefits from word-level decomposition
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(drlib)
library(broom)
library(glmnet)
library(tidytext)
library(Matrix)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
theme_set(theme_light())

board_games_raw <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-03-12/board_games.csv")

holdout_set <- board_games_raw |>
  filter(game_id %% 5 == 0)

board_games <- board_games_raw |>
  filter(game_id %% 5 != 0)
```

## EDA
```{r chunk_02}
board_games |>
  count(publisher, sort = TRUE)

board_games |>
  count(year_published) |>
  arrange(desc(year_published))
  ggplot(aes(year_published, n)) +
  geom_line()
  
board_games |>
  ggplot(aes(average_rating)) +
  geom_histogram()

board_games |>
  filter(max_playtime > 5, max_playtime < 1000) |>
  ggplot(aes(max_playtime / 60)) +
  geom_histogram(binwidth = .25) +
  scale_x_log10(breaks = 2 ^ seq(-2, 4))

board_games |>
  count(year_published) |>
  ggplot(aes(year_published, n)) +
  geom_line()
```
```{r chunk_03}
categorical_variables <- board_games |>
  select(game_id, name, family, category, artist, designer, mechanic) |>
  gather(type, value, -game_id, -name) |>
  filter(!is.na(value)) |>
  separate_rows(value, sep = ",") |>
  arrange(game_id)

categorical_counts <- categorical_variables |>
  count(type, value, sort = TRUE)

library(drlib)

categorical_counts |>
  group_by(type) |>
  top_n(10, n) |>
  ungroup() |>
  mutate(value = reorder_within(value, n, type),
         type = fct_reorder(type, n, .desc = TRUE)) |>
  ggplot(aes(value, n, fill = type)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ type, scales = "free_y") +
  coord_flip() +
  scale_x_reordered() +
  labs(title = "Most common categories")
```

## Predict average rating
```{r chunk_04}
board_games |>
  group_by(decade = 10 * (year_published %/% 10)) |>
  summarize(average_rating = mean(average_rating)) |>
  ggplot(aes(decade, average_rating)) +
  geom_line()
```
```{r chunk_05}
library(broom)

lm(average_rating ~
     log2(max_players + 1) +
     log2(max_playtime + 1) +
     year_published, board_games) |>
  tidy()
```
```{r chunk_06}
by_categorical <- board_games |>
  inner_join(categorical_variables, by = c("game_id", "name")) |>
  select(type, value, average_rating) |>
  group_by(type, value) |>
  summarize(games = n(),
            average_rating = mean(average_rating)) |>
  arrange(desc(games))

board_games |>
  inner_join(categorical_variables, by = c("game_id", "name")) |>
  filter(type == "category") |>
  mutate(value = fct_lump(value, 15),
         value = fct_reorder(value, average_rating)) |>
  ggplot(aes(value, average_rating)) +
  geom_boxplot() +
  coord_flip()

board_games |>
  inner_join(categorical_variables, by = c("game_id", "name")) |>
  filter(type == "family") |>
  mutate(value = fct_lump(value, 15),
         value = fct_reorder(value, average_rating)) |>
  ggplot(aes(value, average_rating)) +
  geom_boxplot() +
  coord_flip()
```
```{r chunk_07}
non_categorical_features <- board_games |>
  transmute(game_id,
            name,
            year = year_published - 1950,
            log2_max_players = log2(max_players + 1),
            log2_max_playtime = log2(max_playtime + 1)) |>
  gather(feature, value, -game_id, -name)

features <- categorical_variables |>
  unite(feature, type, value, sep = ": ") |>
  add_count(feature) |>
  filter(n >= 20) |>
  mutate(value = 1) |>
  bind_rows(non_categorical_features)

library(glmnet)
library(tidytext)
library(Matrix)

# Predictor
feature_matrix <- features |>
  cast_sparse(game_id, feature, value)

# What I'm predicting
ratings <- board_games$average_rating[match(rownames(feature_matrix), board_games$game_id)]

cv_lasso <- cv.glmnet(feature_matrix, ratings)
plot(cv_lasso)

cv_lasso$glmnet.fit |>
  tidy() |>
  filter(lambda == cv_lasso$lambda.1se) |>
  arrange(desc(estimate)) |>
  filter(term != "(Intercept)") |>
  top_n(25, abs(estimate)) |>
  mutate(term = fct_reorder(term, estimate)) |>
  ggplot(aes(term, estimate)) +
  geom_col() +
  coord_flip() +
  labs(title = "Largest coefficients in our predictive model",
       subtitle = "Based on a LASSO regression",
       x = "",
       y = "Coefficient")
```

---

**→ [Hour 2: Modeling](h2_modeling_board_games_predicting_ratings.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
