---
title: "Hour 1 — EDA: Analyzing UN votes in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 071 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `correlation_matrix` · `countrycode_standardize` · `error_bar_coef_plot` · `faceted_comparison` · `fct_relevel_manual` · `fct_reorder_stat` · `fuzzy_join` · `grouped_summary_ggplot` · `lubridate_dates` · `multi_table_join` · `pairwise_token_correlation`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- region name spelling differs between data and shapefile — silent dropped rows
- text is multilingual without language detection
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- geographic column can be matched to a standard ISO/FIPS shapefile without fuzzy joins
- text is long-form English and benefits from word-level decomposition
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(lubridate)
library(ggthemes)
library(fuzzyjoin)
library(countrycode)
library(WDI)
library(widyr)
library(tidytext)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
issues <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-03-23/issues.csv", show_col_types = FALSE)
roll_calls <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-03-23/roll_calls.csv", show_col_types = FALSE)
unvotes <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-03-23/unvotes.csv", show_col_types = FALSE)
glimpse(issues)
glimpse(roll_calls)
```

## Explore
```{r chunk_01}
unvotes <- tt$unvotes |>
  mutate(vote_number = match(vote, c("no", "abstain", "yes")) - 2) |>
  left_join(tt$roll_calls |>
              select(rcid, date, amend), by = "rcid")

unvotes |>
  count(country, sort = TRUE)
```
```{r chunk_02}
summarize_votes <- function(tbl, min_votes = 10) {
  tbl |>
    summarize(n_votes = n(),
              n_yes = sum(vote == "yes"),
              pct_yes = n_yes / n_votes,
              .groups = "drop") |>
  filter(n_votes >= min_votes) |>
  arrange(desc(pct_yes))
}

by_country <- unvotes |>
  group_by(country, country_code) |>
  summarize_votes()

by_country |>
  slice(c(1:10, (n() - 10):n())) |>
  mutate(country = fct_reorder(country, pct_yes)) |>
  ggplot(aes(pct_yes, country)) +
  geom_point(aes(size = n_votes)) +
  scale_x_continuous(labels = percent) +
  labs(x = "% of yes votes in UN",
       title = "What countries voted yes the least")
```
```{r chunk_03}
library(lubridate)
unvote

by_year <- unvotes |>
  group_by(year = year(date)) |>
  summarize_votes()

by_year |>
  ggplot(aes(year, pct_yes)) +
  geom_line() +
  expand_limits(y = 0)

by_country_year <- unvotes |>
  bind_rows(unvotes |> mutate(country = "Overall")) |>
  group_by(year = year(date), country, country_code) |>
  summarize_votes()

by_country_year |>
  filter(country %in% c("United States", "Canada", "Mali", "Israel")) |>
  mutate(country = fct_reorder(country, pct_yes)) |>
  ggplot(aes(year, pct_yes)) +
  geom_line(data = by_year, size = 2) +
  geom_line(aes(color = country)) +
  scale_y_continuous(labels = percent) +
  scale_color_discrete(guide = guide_legend(reverse = TRUE)) +
  expand_limits(y = 0) +
  labs(y = "% yes votes")

by_country_year |>
  filter(country %in% c("United States", "Canada", "Mali", "Israel",
                        "Germany", "France")) |>
  mutate(country = fct_reorder(country, pct_yes)) |>
  ggplot(aes(year, pct_yes, color = country)) +
  geom_line() +
  scale_y_continuous(labels = percent) +
  expand_limits(y = 0) +
  facet_wrap(~ country) +
  theme(legend.position = "none") +
  labs(y = "% yes votes")
```
```{r chunk_04}
library(ggthemes)
library(fuzzyjoin)

world_data <- map_data("world") |>
  as_tibble() |>
  regex_left_join(maps::iso3166 |>
                    select(mapname, country_code = a2),
                  c(region = "mapname")) |>
  filter(region != "Antarctica")

world_data |>
  left_join(by_country, by = "country_code") |>
  ggplot(aes(long, lat, group = group, fill = pct_yes)) +
  geom_polygon() +
  theme_map() +
  scale_fill_gradient2(low = "red",
                       high = "blue",
                       midpoint = .6,
                       labels = percent) +
  labs(fill = "% yes votes")
```
```{r chunk_05}
library(countrycode)

plot_by <- function(tbl, category) {
  tbl |>
    filter(!is.na({{ category }})) |>
    mutate(category = fct_reorder({{ category }}, pct_yes)) |>
    ggplot(aes(year, pct_yes)) +
    geom_line(aes(color = category)) +
    scale_y_continuous(labels = percent) +
    scale_color_discrete(guide = guide_legend(reverse = TRUE)) +
    expand_limits(y = 0) +
    labs(y = "% yes votes",
         x = "Year")
}

unvotes |>
  group_by(continent = countrycode(country_code, "iso2c", "continent"),
           year = year(date)) |>
  summarize_votes() |>
  plot_by(continent)
```
```{r chunk_06}
library(WDI)

country_incomes <- WDI(indicator = c(gdp_per_capita = "NY.GDP.PCAP.PP.KD",
                                     pop = "SP.POP.TOTL"),
                       start = 2005, end = 2005, extra = TRUE) |>
  as_tibble() |>
  select(country_code = iso2c, income, gdp_per_capita, pop) |>
  filter(!is.na(income)) |>
  mutate(income = fct_relevel(income, "Low income", "Lower middle income", "Upper middle income"))

unvotes |>
  inner_join(country_incomes, by = "country_code") |>
  group_by(income,
           year = year(date)) |>
  summarize_votes() |>
  plot_by(income)
```

## Correlation between countries
```{r chunk_07}
unvotes |>
  filter(country %in% c("India", "Canada")) |>
  select(rcid, country, vote_number) |>
  spread(country, vote_number, fill = 0) |>
  summarize(correlation = cor(Canada, India))

library(widyr)

country_correlations <- unvotes |>
  filter(country != "Zanzibar") |>
  pairwise_cor(country, rcid, vote_number, sort = TRUE) |>
  mutate(continent1 = countrycode(item1, "country.name", "continent"),
         continent2 = countrycode(item2, "country.name", "continent"))

country_correlations |>
  filter(item1 == "South Africa") |>
  slice(c(1:16, n() - 0:15)) |>
  mutate(country = fct_reorder(item2, correlation)) |>
  ggplot(aes(correlation, country)) +
  geom_errorbarh(height = 0, aes(xmin = correlation, xmax = 0)) + 
  geom_point() +
  labs(x = "How much this country's voting is correlated with the US",
       y = "")

country_correlations |>
  group_by(country = item1) |>
  summarize(median_correlation = median(correlation),
            avg_correlation = mean(correlation)) |>
  arrange(desc(median_correlation)) |>

country_correlations |>
  filter(!is.na(continent1), !is.na(continent2)) |>
  group_by(continent1, continent2) |>
  summarize(avg_correlation = mean(correlation)) |>
  arrange(desc(avg_correlation))

country_correlations |>
  filter(!is.na(continent1), !is.na(continent2)) |>
  filter(continent1 == continent2) |>
  group_by(item1) |>
  summarize(avg_intracontinent_correlation = mean(correlation)) |>
  arrange((avg_intracontinent_correlation))

country_correlations |>
  filter("German Democratic Republic" == item1)
```
```{r chunk_08}
tt$issues |>
  count(issue)

library(tidytext)

rc_words <- tt$roll_calls |>
  filter(!is.na(short)) |>
  unnest_tokens(word, short) |>
  anti_join(stop_words, by = "word") |>
  distinct(rcid, word) |>
  add_count(word, name = "word_count") |>
  filter(word_count >= 100)

unvotes |>
  inner_join(rc_words, by = "rcid") |>
  filter(country %in% c("Israel", "United States")) |>
  group_by(word, country) |>
  summarize_votes(min_votes = 100) |>
  mutate(word = fct_reorder(word, pct_yes)) |>
  ggplot(aes(pct_yes, word)) +
  geom_point(aes(size = n_votes, color = country)) +
  expand_limits(x = 0) +
  scale_x_continuous(labels = percent) +
  labs(x = "% yes",
       y = "")
```
```{r chunk_09}
by_country_word <- unvotes |>
  inner_join(rc_words, by = "rcid") |>
  group_by(word, country) |>
  summarize_votes(min_votes = 0)

library(tidytext)

by_country_word |>
  widely_svd(word, country, pct_yes) |>
  filter(dimension == 2) |>
  mutate(word = reorder_within(word, value, dimension)) |>
  top_n(30, abs(value)) |>
  ggplot(aes(value, word)) +
  geom_col() +
  scale_y_reordered()
```

---

**→ [Hour 2: Modeling](h2_modeling_un_votes.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
