---
title: "Hour 1 — EDA: Analyzing tidying and analyzing US PhDs in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 066 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`faceted_comparison` · `fct_reorder_stat` · `grouped_summary_ggplot` · `pivot_reshape` · `scales_axis_labels` · `spline_regression` · `str_detect_filter` · `theme_set_global` · `time_series_trend`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
theme_set(theme_light())

# Major field of study
major_fields_raw <- readxl::read_xlsx("~/Downloads/data_tables/sed17-sr-tab012.xlsx",
                  skip = 3)

major_fields <- major_fields_raw |>
  rename(field = `Field of study`) |>
  gather(key, value, -field) |>
  mutate(year = as.numeric(ifelse(str_detect(key, "X__"), NA, key)),
         type = ifelse(!str_detect(value, "Number|Percent"), NA, value),
         value = as.numeric(value)) |>
  fill(year, type) |>
  select(-key) |>
  filter(!is.na(value)) |>
  spread(type, value)
```
```{r chunk_02}
fine_fields <- readxl::read_xlsx("~/Downloads/data_tables/sed17-sr-tab013.xlsx",
                  skip = 3) |>
  rename(field = 1) |>
  gather(year, number, -field) |>
  mutate(year = as.numeric(year),
         number = as.numeric(number)) |>
  filter(!is.na(number))

fine_fields |>
  filter(field %in% sample(unique(field), 6)) |>
  ggplot(aes(year, number, color = field)) +
  geom_line()
```
```{r chunk_03}
# get the broad field names and the major field names

sex <- c("All", "Male", "Female", "All doctorate recipientsa", "All fieldsa")

broad_fields <- readxl::read_xlsx("~/Downloads/data_tables/sed17-sr-tab014.xlsx", skip = 4) |>
  rename(field = 1) |>
  filter(!field %in% sex) |>
  mutate(field = fct_recode(field,
                            "Life sciences" = "Life sciencesb",
                            "Other" = "Otherc")) |>
  pull(field) |>
  as.character()
```
```{r chunk_04}
recipients_year_field_sex <- readxl::read_xlsx("~/Downloads/data_tables/sed17-sr-tab015.xlsx", skip = 3) |>
  rename(field = 1) |>
  select(-contains("change")) |>
  mutate(field = as.character(fct_recode(field, "All" = "All doctorate recipientsa",
                                         "Other" = "Otherb")),
         sex = if_else(field %in% sex, field, NA_character_),
         broad_field = ifelse(field %in% broad_fields, field, NA)) |>
  fill(sex, broad_field) |>
  gather(year, number, -sex, -broad_field, -field) |>
  mutate(year = as.numeric(year),
         number = as.numeric(number)) |>
  filter(!field %in% sex) |>
  filter(!is.na(number))

recipients_year_field_sex |>
  filter(sex != "All",
         broad_field == "Mathematics and computer sciences") |>
  ggplot(aes(year, number, color = sex)) +
  geom_line() +
  expand_limits(y = 0) +
  facet_wrap(~ field)

recipients_year_field_sex |>
  spread(sex, number) |>
  mutate(pct_male = Male / All) |>
  filter(broad_field == "Engineering") |>
  mutate(field = fct_reorder(field, -pct_male)) |>
  ggplot(aes(year, pct_male, color = field)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Year",
       y = "% of PhD recipients reporting as male",
       color = "Major field",
       title = "Breakdown by sex over time within Engineering major fields")

recipients_year_field_sex |>
  spread(sex, number) |>
  mutate(pct_male = Male / All) |>
  filter(broad_field == "Humanities and arts") |>
  mutate(field = fct_reorder(field, -pct_male)) |>
  ggplot(aes(year, pct_male, color = field)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Year",
       y = "% of PhD recipients reporting as male",
       color = "Major field",
       title = "Breakdown by sex over time within Humanities & Arts major fields")

recipients_year_field_sex |>
  spread(sex, number) |>
  mutate(pct_male = Male / All) |>
  filter(broad_field == "Education") |>
  mutate(field = fct_reorder(field, -pct_male)) |>
  ggplot(aes(year, pct_male, color = field)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Year",
       y = "% of PhD recipients reporting as male",
       color = "Major field",
       title = "Breakdown by sex over time within Education major fields")
```

---

**→ [Hour 2: Modeling](h2_modeling_tidying_phds.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
