---
title: "Hour 1 — EDA: Analyzing coffee ratings in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 019 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`boxplot_jitter` · `correlation_matrix` · `error_bar_coef_plot` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_histogram_density` · `geom_smooth_loess` · `grouped_summary_ggplot` · `linear_model_broom` · `network_graph_ggraph` · `pairwise_token_correlation`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- text is multilingual without language detection
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- text is long-form English and benefits from word-level decomposition
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(ggridges)
library(widyr)
library(ggraph)
library(igraph)
library(tidytext)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
coffee_ratings <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-07-07/coffee_ratings.csv", show_col_types = FALSE)
glimpse(coffee_ratings)
```

## Explore
```{r chunk_02}
coffee_ratings |>
  count(species, sort = TRUE)

coffee_lumped <- coffee_ratings |>
  filter(!is.na(variety)) |>
  mutate(variety = fct_lump(variety, 12), sort = TRUE)

coffee_lumped |>
  mutate(variety = fct_reorder(variety, total_cup_points)) |>
  ggplot(aes(total_cup_points, variety)) +
  geom_boxplot()

coffee_lumped |>
  ggplot(aes(total_cup_points, fill = variety)) +
  geom_histogram(binwidth = 2) +
  facet_wrap(~ variety, scale = "free_y") +
  theme(legend.position = "none")
```
```{r chunk_03}
coffee_ratings |>
  summarize(across(everything(), ~ mean(!is.na(.)))) |>
  gather() |>

coffee_ratings |>
  count(producer, sort = TRUE)

coffee_ratings |>
  count(company, sort = TRUE)

coffee_ratings |>
  count(color, sort = TRUE)

coffee_ratings |>
  count(country = fct_lump(country_of_origin, 12), sort = TRUE) |>
  filter(!is.na(country)) |>
  mutate(country = fct_reorder(country, n)) |>
  ggplot(aes(n, country)) +
  geom_col()

coffee_ratings |>
  filter(!is.na(country_of_origin)) |>
  mutate(country = fct_lump(country_of_origin, 12),
         country = fct_reorder(country, total_cup_points)) |>
  ggplot(aes(total_cup_points, country)) +
  geom_boxplot()
```
```{r chunk_04}
library(ggridges)

coffee_metrics <- coffee_ratings |>
  select(coffee_id, total_cup_points, variety, company,
         country_of_origin,
         altitude_mean_meters,
         aroma:moisture) |>
  pivot_longer(aroma:cupper_points, names_to = "metric", values_to = "value")

coffee_metrics |>
  mutate(metric = fct_reorder(metric, value)) |>
  ggplot(aes(value, metric)) +
  geom_density_ridges()

coffee_metrics |>
  group_by(metric) |>
  summarize(average = mean(value),
            sd = sd(value)) |>
  arrange(desc(average))
```
```{r chunk_05}
library(widyr)
library(ggraph)
library(igraph)
library(tidytext)

correlations <- coffee_metrics |>
  pairwise_cor(metric, coffee_id, value, sort = TRUE)

correlations |>
  head(50) |>
  graph_from_data_frame() |>
  ggraph() +
  geom_edge_link(aes(edge_alpha = correlation)) +
  geom_node_point() +
  geom_node_text(aes(label = name), repel = TRUE)

coffee_metrics |>
  filter(!metric %in% c("sweetness", "clean_cup", "uniformity")) |>
  group_by(metric) |>
  mutate(centered = value - mean(value)) |>
  ungroup() |>
  widely_svd(metric, coffee_id, value) |>
  filter(between(dimension, 2, 5)) |>
  mutate(metric = reorder_within(metric, value, dimension)) |>
  ggplot(aes(value, metric)) +
  geom_col() +
  scale_y_reordered() +
  facet_wrap(~ dimension, scales = "free_y")
```
```{r chunk_06}
coffee_ratings |>
  filter(altitude_mean_meters < 10000,
         altitude != 1) |>
  mutate(altitude_mean_meters = pmin(altitude_mean_meters, 3000)) |>
  ggplot(aes(altitude_mean_meters, total_cup_points)) +
  geom_point() +
  geom_smooth(method = "lm")

coffee_metrics |>
  filter(altitude_mean_meters < 10000) |>
  mutate(altitude_mean_meters = pmin(altitude_mean_meters, 3000)) |>
  mutate(km = altitude_mean_meters / 1000) |>
  group_by(metric) |>
  summarize(correlation = cor(altitude_mean_meters, value),
            model = list(lm(value ~ km))) |>
  mutate(tidied = map(model, broom::tidy, conf.int = TRUE)) |>
  unnest(tidied) |>
  filter(term == "km") |>
  ungroup() |>
  mutate(metric = fct_reorder(metric, estimate)) |>
  ggplot(aes(estimate, metric, color = p.value < .05)) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = .1) +
  labs(y = "Evaluation of coffee",
       x = "Each kilometer of altitude contributes this much to score (95% confidence interval)")
```

---

**→ [Hour 2: Modeling](h2_modeling_coffee_ratings.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
