---
title: "Hour 1 — EDA: Analyzing Bob Ross paintings in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 012 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `broom_model_output` · `coord_flip_horizontal` · `faceted_comparison` · `fct_reorder_stat` · `grouped_summary_ggplot` · `janitor_clean` · `linear_model_broom` · `network_graph_ggraph` · `pairwise_token_correlation` · `pivot_reshape` · `reorder_within_facets`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- text is multilingual without language detection
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- text is long-form English and benefits from word-level decomposition
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(widyr)
library(ggraph)
library(igraph)
library(reshape2)
library(broom)
library(tidytext)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
theme_set(theme_light())

bob_ross <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-08-06/bob-ross.csv")
```
```{r chunk_02}
bob_ross_gathered <- bob_ross |>
  janitor::clean_names() |>
  gather(element, present, -episode, -title) |>
  filter(present == 1) |>
  mutate(title = str_to_title(str_remove_all(title, '"')),
         element = str_to_title(str_replace(element, "_", " "))) |>
  select(-present) |>
  extract(episode, c("season", "episode_number"), "S(.*)E(.*)", convert = TRUE, remove = FALSE) |>
  arrange(season, episode_number)
```

## Exploring the paintings
```{r chunk_03}
bob_ross_gathered |>
  count(element, sort = TRUE) |>
  head(25) |>
  mutate(element = fct_reorder(element, n)) |>
  ggplot(aes(element, n)) +
  geom_col() +
  coord_flip()
```
```{r chunk_04}
bob_ross_gathered |>
  add_count(episode) |>
  arrange(desc(n))
```
```{r chunk_05}
by_season_element <- bob_ross_gathered |>
  filter(!element %in% c("Tree", "Trees")) |>
  group_by(season) |>
  mutate(number_episodes = n_distinct(episode)) |>
  count(season, element, number_episodes, sort = TRUE) |>
  mutate(percent_included = n / number_episodes) |>
  group_by(element) |>
  mutate(element_total = sum(n)) |>
  ungroup()

by_season_element |>
  filter(element_total >= 50) |>
  ggplot(aes(season, percent_included, color = element)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent_format()) +
  expand_limits(y = 0) +
  facet_wrap(~ element)
```

## Clustering
```{r chunk_06}
library(widyr)

correlations <- bob_ross_gathered |>
  add_count(element) |>
  filter(n >= 5) |>
  pairwise_cor(element, episode, sort = TRUE)

correlations |>
  filter(item1 == "River") |>
  mutate(item2 = fct_reorder(item2, correlation)) |>
  ggplot(aes(item2, correlation)) +
  geom_col() +
  coord_flip() +
  labs(title = "What tends to appear with a river?",
       subtitle = "Among elements that appeared in at least 10 paintings")

correlations |>
  filter(item1 == "Snow") |>
  mutate(item2 = fct_reorder(item2, correlation)) |>
  ggplot(aes(item2, correlation)) +
  geom_col() +
  coord_flip() +
  labs(title = "What tends to appear with snow?",
       subtitle = "Among elements that appeared in at least 10 paintings")
```
```{r chunk_07}
library(ggraph)
library(igraph)

set.seed(2019)

correlations |>
  head(100) |>
  graph_from_data_frame() |>
  ggraph() +
  geom_edge_link(aes(alpha = correlation)) +
  geom_node_point() +
  geom_node_text(aes(label = name), vjust = 1, hjust = 1) +
  theme_void()
```

## Principal Component Analysis
```{r chunk_08}
library(reshape2)
library(broom)
library(tidytext)

binary_matrix <- bob_ross_gathered |>
  acast(title ~ element)

# Center the columns
centered_matrix <- t(t(binary_matrix) - colMeans(binary_matrix))

svd_result <- svd(centered_matrix)

element_weights <- tidy(svd_result, matrix = "v") |>
  mutate(element = colnames(binary_matrix)[column])

element_weights |>
  filter(PC <= 4) |>
  group_by(PC) |>
  top_n(16, abs(value)) |>
  ungroup() |>
  mutate(element = reorder_within(element, value, PC)) |>
  ggplot(aes(element, value, fill = factor(PC))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ PC, scales = "free") +
  scale_x_reordered() +
  coord_flip() +
  labs(title = "First four principal components of elements in Bob Ross paintings")
```
```{r chunk_09}
painting_weights <- broom::tidy(svd_result, matrix = "u") |>
  mutate(painting = rownames(binary_matrix)[row])
```
```{r chunk_10}
painting_weights |>
  filter(PC == 1) |>
  arrange((value))

bob_ross_gathered |>
  filter(title == "Frozen Solitude")

painting_weights |>
  filter(PC <= 4) |>
  group_by(PC) |>
  top_n(20, abs(value)) |>
  ungroup() |>
  mutate(painting = reorder_within(painting, value, PC)) |>
  ggplot(aes(painting, value, fill = factor(PC))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ PC, scales = "free") +
  scale_x_reordered() +
  coord_flip() +
  labs(title = "First four principal components of Bob Ross paintings")
```
```{r chunk_11}
broom::tidy(svd_result, matrix = "d") |>
  ggplot(aes(PC, percent)) +
  geom_point()
```

---

**→ [Hour 2: Modeling](h2_modeling_bob_ross_paintings.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
