---
title: "Hour 1 — EDA: Analyzing GDPR violations in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 031 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `boxplot_jitter` · `fct_lump_infrequent` · `fct_reorder_stat` · `grouped_summary_ggplot` · `log_scale` · `lubridate_dates` · `multi_table_join` · `rolling_window_metrics` · `shiny_app` · `spline_regression` · `str_detect_filter`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(tidymetrics)
library(shinymetrics)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
gdpr_text <- read_tsv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-04-21/gdpr_text.tsv", show_col_types = FALSE)
gdpr_violations <- read_tsv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-04-21/gdpr_violations.tsv", show_col_types = FALSE)
glimpse(gdpr_text)
glimpse(gdpr_violations)
```

## Explore
```{r chunk_02}
gdpr_violations |>
  summarize(total = sum(price))

gdpr_violations |>
  count(country = fct_lump(country, 8, w = price),
        sort = TRUE, wt = price, name = "total_price") |>
  mutate(country = fct_reorder(country, total_price)) |>
  ggplot(aes(total_price, country)) +
  geom_col() +
  scale_x_continuous(labels = dollar_format())
```
```{r chunk_03}
gdpr_violations |>
  count(month = floor_date(date, "month"),
        country = fct_lump(country, 6, w = price),
        sort = TRUE, wt = price, name = "total_price") |>
  mutate(country = fct_reorder(country, -total_price, sum)) |>
  ggplot(aes(month, total_price, fill = country)) +
  geom_col() +
  scale_y_continuous(labels = dollar_format()) +
  labs(x = "Time",
       y = "Total fines",
       fill = "Country")
```
```{r chunk_04}
gdpr_violations |>
  count(source, country, sort = TRUE)

gdpr_violations |>
  select(controller, date, article_violated, type, summary, price) |>
  mutate(summary = str_trunc(summary, 140)) |>
  arrange(desc(price)) |>
  mutate(price = dollar(price)) |>
  head(10)
```
```{r chunk_05}
gdpr_violations |>
  mutate(type = fct_lump(type, 8, w = price),
         type = fct_reorder(type, price),
         country = fct_lump(country, 5)) |>
  ggplot(aes(price, type)) +
  geom_boxplot() +
  geom_jitter(aes(color = country), width = 0, height = .25) +
  scale_x_log10(labels = dollar_format())
```

## Which article was violated?
```{r chunk_06}
article_titles <- gdpr_text |>
  distinct(article, article_title)

separated_articles <- gdpr_violations |>
  separate_rows(article_violated, sep = "\\|") |>
  extract(article_violated, "article", "Art\\. ?(\\d+)", convert = TRUE, remove = FALSE) |>
  left_join(article_titles, by = "article") |>
  mutate(article_title = paste0(article, ". ", str_trunc(article_title, 50)),
         article_title = ifelse(is.na(article), "Unknown", article_title)) |>
  add_count(id) |>
  mutate(price_per_article = price / n)

separated_articles |>
  group_by(article = fct_lump(article_title, 8, w = price)) |>
  summarize(total_fine = sum(price_per_article),
            violations = n()) |>
  arrange(desc(total_fine)) |>
  mutate(article = fct_reorder(article, total_fine)) |>
  ggplot(aes(total_fine, article)) +
  geom_col() +
  scale_x_continuous(labels = dollar_format()) +
  labs(title = "What articles got the most fines?",
       y = "",
       x = "Total fine")
```
```{r chunk_07}
gdpr_text <- gdpr_text
```
```{r chunk_08}
gdpr_violations |>
  filter(str_detect(controller, "Vodafone")) |>
  group_by(date, country) |>
  summarize(violations = n(),
            total_fine = sum(price)) |>
  ggplot(aes(date, total_fine, size = violations, color = country)) +
  geom_point() +
  scale_y_continuous(labels = dollar_format()) +
  scale_size_continuous(guide = FALSE) +
  labs(title = "Vodafone's GDPR violations",
       color = "",
       x = "Time",
       y = "Total fine on this day")
```
```{r chunk_09}
library(tidymetrics)

summarized <- separated_articles |>
  filter(!is.na(date)) |>
  mutate(country = fct_lump(country, 6, w = price_per_article),
         article_title = fct_lump(article_title, 6, w = price_per_article),
         type = fct_lump(type, 6, w = price_per_article)) |>
  cross_by_dimensions(country, article_title, type) |>
  cross_by_periods(c("month", "quarter")) |>
  summarize(nb_violations = n_distinct(id),
            nb_total_fine = sum(price_per_article)) |>
  ungroup()

gdpr_metrics <- create_metrics(summarized)
```
```{r chunk_10}
library(shinymetrics)

preview_metric(gdpr_metrics$gdpr_violations_nb_violations)
```

---

**→ [Hour 2: Modeling](h2_modeling_gdpr_violations.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
