---
title: "Hour 1 — EDA: Analyzing African-American achievements in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 001 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`fct_reorder_stat` · `geom_histogram_density` · `glue_strings` · `grouped_summary_ggplot` · `interactive_plotly` · `janitor_clean` · `pivot_reshape` · `purrr_map_iterate` · `spline_regression` · `str_detect_filter` · `theme_set_global` · `tidytuesdayr_load`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- source HTML structure changes — selector-based scraping breaks silently
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- supplemental context is unavailable in the flat file and worth the scraping overhead
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(plotly)
library(glue)
library(rvest)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
firsts <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-06-09/firsts.csv", show_col_types = FALSE)
science <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-06-09/science.csv", show_col_types = FALSE)
glimpse(firsts)
glimpse(science)
```

## Explore
```{r chunk_02}
firsts |>

science |>
```
```{r chunk_03}
firsts <- firsts

firsts |>
  ggplot(aes(year)) +
  geom_histogram()

firsts |>
  count(category, sort = TRUE) |>
  mutate(category = fct_reorder(category, n)) |>
  ggplot(aes(n, category)) +
  geom_col()
```
```{r chunk_04}
firsts <- firsts |>
  mutate(person = str_remove(person, "[\\[\\(].*"),
         person = str_trim(person))
```
```{r chunk_05}
library(plotly)
library(glue)

g <- firsts |>
  ggplot(aes(year,
             category,
             color = category,
             text = glue("{ year }: { accomplishment }\n{ person }"))) +
  geom_point() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position = "none") +
  labs(title = "Timeline of some notable African-American achievements",
       subtitle = "Source: https://en.wikipedia.org/wiki/List_of_African-American_firsts",
       y = "Category",
       x = "Year")

ggplotly(g, tooltip = "text")
```

## Science
```{r chunk_06}
science |>
  ggplot(aes(birth)) +
  geom_histogram()

science |>
  separate_rows(occupation_s, sep = "; ") |>
  mutate(occupation = str_to_title(occupation_s)) |>
  count(occupation, sort = TRUE)

science |>
  filter(str_detect(occupation_s, regex("istician", ignore_case = TRUE))) |>
  pull(name)

science |>
  filter(str_detect(occupation_s, "statistician")) |>
```
```{r chunk_07}
library(rvest)

science_html <- science |>
  mutate(html = map(links, possibly(read_html, NULL, quiet = FALSE)))
```
```{r chunk_08}
extract_infobox <- . |>
  html_node(".vcard") |>
  html_table(header = FALSE) |>
  as_tibble()

infoboxes <- science_html |>
  filter(!map_lgl(html, is.null)) |>
  mutate(infobox = map(html, possibly(extract_infobox, NULL))) |>
  select(link = links, infobox) |>
  unnest(infobox) |>
  filter(X1 != "" | X2 != "", X1 != "Scientific career") |>
  rename(key = X1, value = X2)

science_infoboxes <- infoboxes |>
  group_by(link) |>
  mutate(name = first(key)) |>
  group_by(key) |>
  filter(n() >= 10) |>
  ungroup() |>
  distinct(name, key, .keep_all = TRUE) |>
  spread(key, value) |>
  janitor::clean_names()
```
```{r chunk_09}
science_infoboxes |>
  count(nationality, sort = TRUE)
```

---

**→ [Hour 2: Modeling](h2_modeling_african_american_achievements.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
