---
title: "Hour 1 — EDA: Analyzing Tour de France data in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 068 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`animated_transitions` · `boxplot_jitter` · `broom_model_output` · `cumsum_running_total` · `fct_reorder_stat` · `grouped_summary_ggplot` · `janitor_clean` · `linear_model_broom` · `lubridate_dates` · `multi_table_join` · `reorder_within_facets` · `spline_regression`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- text is multilingual without language detection
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- text is long-form English and benefits from word-level decomposition
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(survival)
library(broom)
library(gganimate)
library(tidytext)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
stage_data <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-04-07/stage_data.csv", show_col_types = FALSE)
tdf_stages <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-04-07/tdf_stages.csv", show_col_types = FALSE)
tdf_winners <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-04-07/tdf_winners.csv", show_col_types = FALSE)
glimpse(stage_data)
glimpse(tdf_stages)
```

## Explore
```{r chunk_04}
tdf_winners |>
  count(birth_country, sort = TRUE) |>
  mutate(birth_country = fct_reorder(birth_country, n)) |>
  ggplot(aes(n, birth_country)) +
  geom_col() +
  labs(y = "",
       title = "What countries were the most Tour de France winners born in?")

tdf_winners |>
  count(winner_name, birth_country, sort = TRUE)
```
```{r chunk_05}
by_decade <- tdf_winners |>
  group_by(decade = 10 * (year %/% 10)) |>
  summarize(winner_age = mean(age),
            winner_height = mean(height, na.rm = TRUE),
            winner_weight = mean(weight, na.rm = TRUE),
            winner_margin = mean(time_margin, na.rm = TRUE),
            winner_speed = mean(speed, na.rm = TRUE))

by_decade |>
  filter(decade >= 1910) |>
  ggplot(aes(decade, winner_margin * 60)) +
  geom_line() +
  expand_limits(y = 0) +
  labs(x = "Decade",
       y = "Average margin of winner (minutes)",
       title = "Tour de France races have been getting closer")

by_decade |>
  ggplot(aes(decade, winner_speed)) +
  geom_line() +
  expand_limits(y = 0) +
  labs(x = "Decade",
       y = "Average speed of winner (km/h)",
       title = "Tour de France winners have been getting faster")
```
```{r chunk_06}
library(survival)
library(broom)

surv_model <- tdf_winners |>
  distinct(winner_name, .keep_all = TRUE) |>
  transmute(winner_name,
            birth_year = year(born),
            death_year = year(died),
            dead = as.integer(!is.na(death_year))) |>
  mutate(age_at_death = coalesce(death_year, 2020) - birth_year) |>
  survfit(Surv(age_at_death, dead) ~ 1, data = .)

glance(surv_model)
```
```{r chunk_07}
stages_joined <- stage_data |>
  extract(stage_results_id, "stage", "stage-(.*)") |>
  inner_join(tdf_stages, by = c("year", "stage")) |>
  mutate(rank = as.integer(rank)) |>
  group_by(year, stage) |>
  mutate(finishers = sum(!is.na(rank))) |>
  ungroup() |>
  mutate(percentile = 1 - rank / finishers)

total_points <- stages_joined |>
  group_by(year, rider) |>
  summarize(total_points = sum(points, na.rm = TRUE)) |>
  mutate(points_rank = percent_rank(total_points)) |>
  ungroup()

stages_joined |>
  filter(stage == "1") |>
  group_by(winner_country) |>
  summarize(stages = n(),
            median_percentile = median(percentile, na.rm = TRUE)) |>
  arrange(desc(stages))
```
```{r chunk_08}
stages_joined |>
  filter(stage == "1") |>
  inner_join(total_points, by = c("year", "rider")) |>
  select(year, rider,
         percentile_first_stage = percentile,
         points_rank) |>
  mutate(first_stage_bin = cut(percentile_first_stage, seq(0, 1, .1),
                               include.lowest = TRUE)) |>
  filter(!is.na(first_stage_bin)) |>
  ggplot(aes(first_stage_bin, points_rank)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Decile performance in the first stage",
       y = "Overall points percentile")
```
```{r chunk_09}
library(gganimate)
library(tidytext)

top_10_2017 <- total_points |>
  filter(year == max(year)) |>
  top_n(10, total_points)

stages_joined |>
  filter(year == max(year)) |>
  semi_join(top_10_2017, by = "rider") |>
  mutate(stage = as.integer(stage),
         points = coalesce(points, 0)) |>
  arrange(stage) |>
  group_by(rider) |>
  mutate(cumulative_points = cumsum(points)) |>
  ungroup() |>
  # mutate(rider = reorder_within(rider, cumulative_points, stage)) |>
  ggplot(aes(cumulative_points, rider, fill = cumulative_points)) +
  geom_col() +
  transition_time(stage) +
  theme(legend.position = "none") +
  labs(title = "The 2017 Tour de France. Stage: { frame_time }",
       x = "Cumulative points at this stage",
       y = "")
```

---

**→ [Hour 2: Modeling](h2_modeling_tour_de_france.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
