---
title: "Hour 1 — EDA: Analyzing the Great American Beer Festival"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 064 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `error_bar_coef_plot` · `faceted_comparison` · `fct_lump_infrequent` · `fct_relevel_manual` · `fct_reorder_stat` · `glue_strings` · `grouped_summary_ggplot` · `linear_model_broom` · `log_odds_ratio` · `logistic_regression` · `multi_table_join`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- text is multilingual without language detection
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- text is long-form English and benefits from word-level decomposition
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(tidylo)
library(tidytext)
library(broom)
library(ggthemes)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
beer_awards <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-10-20/beer_awards.csv", show_col_types = FALSE)
glimpse(beer_awards)
```

## Load the weekly Data
```{r chunk_01}
beer_counts <- beer_awards |>
  count(beer_name, brewery, medal, city, state)

beer_counts |>
  mutate(beer_name = glue("{ beer_name } ({ brewery })")) |>
  filter(fct_lump(beer_name, 16, w = n) != "Other") |>
  mutate(beer_name = fct_reorder(beer_name, n, sum)) |>
  ggplot(aes(n, beer_name, fill = medal)) +
  geom_col() +
  labs(title = "Which beers have won the most awards?",
       x = "# of awards",
       y = "",
       fill = "")

beer_counts |>
  pivot_wider(names_from = medal, values_from = n,
              values_fill = list(n = 0)) |>
  mutate(total = Bronze + Silver + Gold) |>
  arrange(desc(total))
```
```{r chunk_02}
by_brewery <- beer_awards |>
  group_by(brewery) |>
  summarize(n_medals = n(),
            n_beers = n_distinct(beer_name),
            n_gold = sum(medal == "Gold"),
            weighted_medals = sum(as.integer(medal))) |>
  arrange(desc(n_medals))

beer_awards |>
  count(brewery, medal) |>
  filter(fct_lump(brewery, 16, w = n) != "Other") |>
  mutate(brewery = fct_reorder(brewery, n, sum)) |>
  ggplot(aes(n, brewery, fill = medal)) +
  geom_col() +
  labs(title = "Which breweries have won the most awards?",
       x = "# of awards",
       y = "",
       fill = "")

beer_awards |>
  count(state, medal) |>
  filter(fct_lump(state, 16, w = n) != "Other") |>
  mutate(state = fct_reorder(state, n, sum)) |>
  ggplot(aes(n, state, fill = medal)) +
  geom_col() +
  labs(title = "Which states have won the most awards?",
       x = "# of awards",
       y = "",
       fill = "")
```
```{r chunk_03}
library(tidylo)
library(tidytext)

beer_awards |>
  filter(fct_lump(state, 12) != "Other",
         fct_lump(category, 10) != "Other") |>
  count(state, category) |>
  complete(state, category, fill = list(n = 0)) |>
  bind_log_odds(state, category, n) |>
  mutate(state = fct_reorder(state, -n, sum)) |>
  mutate(category = reorder_within(category, log_odds_weighted, state)) |>
  ggplot(aes(log_odds_weighted, category, fill = log_odds_weighted > 0)) +
  geom_col() +
  scale_y_reordered() +
  facet_wrap(~ state, scales = "free_y") +
  theme(legend.position = "none") +
  labs(x = "Weighted log-odds",
       y = "Category",
       title = "What types of beer are over- or under-represented in awards each state?",
       subtitle = "For the 12 states and the 10 categories with the most awards")
```
```{r chunk_04}
by_year_state <- beer_awards |>
  add_count(year, name = "year_total") |>
  mutate(state = fct_lump(state, 9)) |>
  count(year, state, year_total, sort = TRUE) |>
  mutate(pct_year = n / year_total)

by_year_state |>
  filter(state != "Other") |>
  ggplot(aes(year, pct_year, color = state)) +
  geom_line() +
  expand_limits(y = 0) +
  scale_y_continuous(labels = percent) +
  facet_wrap(~ state)

by_year_state |>
  filter(state == "WI") |>
  glm(cbind(n, year_total - n) ~ year,
      data = .,
      family = "binomial") |>
  summary()

library(broom)

by_year_state |>
  filter(state != "Other") |>
  mutate(state = state.name[match(state, state.abb)]) |>
  group_by(state) |>
  summarize(model = list(glm(cbind(n, year_total - n) ~ year, family = "binomial"))) |>
  mutate(tidied = map(model, tidy, conf.int = TRUE)) |>
  unnest(tidied) |>
  filter(term == "year") |>
  mutate(p.value = format.pval(p.value),
         state = fct_reorder(state, estimate)) |>
  ggplot(aes(estimate, state)) +
  geom_point() +
  geom_vline(xintercept = 0, lty = 2) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = .1) +
  labs(x = "Estimated slope",
       title = "Which states become more or less frequent medal winners?",
       y = "")
```
```{r chunk_05}
library(ggthemes)
map_data("state")

beer_awards |>
  mutate(state = str_to_lower(state.name[match(state, state.abb)])) |>
  count(state) |>
  right_join(map_data("state"), by = c(state = "region")) |>
  ggplot(aes(long, lat, fill = n, group = group)) +
  geom_polygon() +
  theme_map() +
  coord_map() +
  labs(fill = "# of award-winning beers")
```
```{r chunk_06}
beer_awards

breweries <- read_csv("~/Downloads/archive/breweries.csv") |>
  rename(brewery_id = X1,
         brewery = name)

beers <- read_csv("~/Downloads/archive/beers.csv") |>
  rename(beer_name = name) |>
  inner_join(breweries, by = "brewery_id")

beers

beer_awards |>
  distinct(beer_name, brewery) |>
  semi_join(beers, by = c("beer_name"))

# DEAD END!
```

---

**→ [Hour 2: Modeling](h2_modeling_great_american_beer_festival.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
