---
title: "Hour 1 — EDA: Analyzing art collections in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 005 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `boxplot_jitter` · `case_when_conditional_col` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_area_stacked` · `geom_histogram_density` · `glue_strings` · `grouped_summary_ggplot` · `janitor_clean` · `log_scale`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(glue)
library(ggridges)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
artists <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-01-12/artists.csv", show_col_types = FALSE)
artwork <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-01-12/artwork.csv", show_col_types = FALSE)
glimpse(artists)
glimpse(artwork)
```

## Load the weekly Data
```{r chunk_01}
artwork |>
  separate(medium, c("medium", "surface"), sep = " on ",
           fill = "right", extra = "merge") |>
  count(medium, sort = TRUE) |>
  head(20) |>
  replace_na(list(medium = "Missing")) |>
  mutate(medium = fct_reorder(medium, n)) |>
  ggplot(aes(n, medium)) +
  geom_col() +
  labs(y = 'Medium (ignoring "on...")',
       x = "# of pieces in Tate Modern")

artwork |>
  filter(fct_lump(artist, 16) != "Other") |>
  mutate(medium = fct_lump(medium, 8)) |>
  count(artist, medium, sort = TRUE) |>
  mutate(artist = fct_reorder(artist, n, sum)) |>
  ggplot(aes(n, artist, fill = medium)) +
  geom_col()
```
```{r chunk_02}
by_decade_medium <- artwork |>
  # separate(medium, c("medium", "surface"), sep = " on ",
  #          fill = "right", extra = "merge") |>
  filter(year >= 1750) |>
  group_by(decade = round(year, -1)) |>
  mutate(n_artists_decade = n_distinct(artist)) |>
  ungroup() |>
  mutate(medium = fct_lump(medium, 11)) |>
  group_by(decade, medium, n_artists_decade) |>
  summarize(n = n(),
            n_artists = n_distinct(artist)) |>
  ungroup() |>
  complete(decade, medium, fill = list(n = 0, n_artists = 0)) |>
  group_by(decade) |>
  mutate(pct = n / sum(n),
         pct_artists = n_artists / n_artists_decade) |>
  ungroup()

by_decade_medium |>
  ggplot(aes(decade, n, fill = medium)) +
  geom_area()

by_decade_medium |>
  filter(!is.na(medium)) |>
  mutate(medium = fct_reorder(medium, n, sum)) |>
  ggplot(aes(decade, pct, fill = medium)) +
  geom_area() +
  scale_y_continuous(labels = percent) +
  labs(x = "Decade",
       y = "% of Tate Modern pieces",
       fill = "Medium (without \"on...\")")

by_decade_medium |>
  filter(!is.na(medium)) |>
  mutate(medium = fct_reorder(medium, pct, sum, .desc = TRUE)) |>
  ggplot(aes(decade, pct, fill = medium)) +
  geom_col() +
  facet_wrap(~ medium, scales = "free_y") +
  scale_y_continuous(labels = percent) +
  theme(legend.position = "none") +
  labs(x = "Decade",
       y = "% of Tate Modern pieces",
       fill = "Medium (without \"on...\")")

by_decade_medium |>
  filter(!is.na(medium)) |>
  mutate(medium = fct_reorder(medium, pct_artists, sum, .desc = TRUE)) |>
  ggplot(aes(decade, pct_artists, fill = medium)) +
  geom_col() +
  facet_wrap(~ medium, scales = "free_y") +
  scale_y_continuous(labels = percent) +
  theme(legend.position = "none") +
  labs(x = "Decade",
       y = "% of artists in this decade using this medium",
       fill = "Medium (without \"on...\")")
```
```{r chunk_03}
artwork_size <- artwork |>
  filter(units == "mm",
         !is.na(height),
         !is.na(width)) |>
  mutate(area = (height / 1000) * (width / 1000),
         ratio = width / height)

artwork_size |>
  ggplot(aes(area)) +
  geom_histogram() +
  scale_x_log10(labels = comma) +
  labs(x = "Area (square meters)")

artwork_size |>
  filter(ratio > .2, ratio < 5) |>
  ggplot(aes(ratio)) +
  geom_histogram(bins = 50) +
  scale_x_log10() +
  geom_vline(xintercept = c(3 / 4, 4 / 3),
             linetype = "dashed", color = "red") +
  labs(x = "Width / height ratio (log scale)",
       title = "What are typical shapes of Tate artwork?",
       subtitle = "Red lines show 3:4 and 4:3 ratios")

artwork_size |>
  mutate(shape = case_when(
    ratio > 1.05 ~ "Landscape",
    ratio < .95 ~ "Portrait",
    TRUE ~ "Square"
  )) |>
  count(shape,
        decade = round(year, -1)) |>
  filter(decade >= 1700) |>
  complete(decade, shape, fill = list(n = 0)) |>
  group_by(decade) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(decade, pct, fill = shape)) +
  geom_area()

size_by_decade <- artwork_size |>
  group_by(decade = round(year, -1)) |>
  summarize(median_ratio = median(ratio),
            median_area = median(area),
            n = n()) |>
  filter(decade >= 1700)

size_by_decade |>
  ggplot(aes(decade, median_ratio)) +
  geom_line() +
  geom_point(aes(size = n)) +
  labs(x = "Decade",
       y = "Median ratio (width / height)")

size_by_decade |>
  ggplot(aes(decade, median_area)) +
  geom_line() +
  geom_point(aes(size = n)) +
  labs(x = "Decade",
       y = "Median area in meters^2")

artwork_size |>
  mutate(decade = round(year, -1)) |>
  filter(decade >= 1700) |>
  ggplot(aes(decade, area, group = decade)) +
  geom_boxplot() +
  scale_y_log10(labels = comma,
                breaks = 10 ^ seq(-3, 2))
```

## Artists and size
```{r chunk_04}
artwork_size |>
  group_by(artist) |>
  summarize(n_pieces = n(),
            avg_year = mean(year, na.rm = TRUE),
            first_year = min(year, na.rm = TRUE),
            last_year = max(year, na.rm = TRUE),
            median_area = median(area),
            median_ratio = median(ratio)) |>
  arrange(desc(n_pieces))

library(glue)

artwork_size |>
  extract(medium, "medium_on", " on (.*)", remove = FALSE) |>
  filter(fct_lump(artist, 25) != "Other") |>
  add_count(artist) |>
  mutate(artist = glue("{ artist } ({ n })")) |>
  mutate(artist = fct_reorder(artist, ratio)) |>
  ggplot(aes(ratio, artist)) +
  geom_boxplot() +
  geom_vline(xintercept = c(3 / 4, 4 / 3),
             linetype = "dashed", color = "red") +
  scale_x_log10(limits = c(.25, 4)) +
  labs(x = "Ratio (width / height)",
       y = "",
       title = "What aspect ratio do artists work in?",
       subtitle = "For the 25 most common artists in the Tate")

library(ggridges)
artwork_size |>
  extract(medium, "medium_on", " on (.*)", remove = FALSE) |>
  filter(fct_lump(artist, 25) != "Other") |>
  add_count(artist) |>
  mutate(artist = glue("{ artist } ({ n })")) |>
  mutate(artist = fct_reorder(artist, ratio)) |>
  ggplot(aes(ratio, artist)) +
  geom_density_ridges() +
  geom_vline(xintercept = c(3 / 4, 4 / 3),
             linetype = "dashed", color = "red") +
  scale_x_log10(limits = c(.33, 3)) +
  labs(x = "Ratio (width / height)",
       y = "",
       title = "What aspect ratio do artists work in?",
       subtitle = "For the 25 most common artists in the Tate")

artwork_size |>
  filter(fct_lump(medium, 25) != "Other") |>
  add_count(medium) |>
  mutate(medium = glue("{ medium } ({ n })")) |>
  mutate(medium = fct_reorder(medium, ratio)) |>
  ggplot(aes(ratio, medium)) +
  geom_boxplot() +
  geom_vline(xintercept = c(3 / 4, 4 / 3),
             linetype = "dashed", color = "red") +
  scale_x_log10(limits = c(.25, 4)) +
  labs(x = "Ratio (width / height)",
       y = "",
       title = "What aspect ratio is a medium typically in?",
       subtitle = "For the 25 most common medium in the Tate")
```

## Readme
```{r Readme}
tt
```

## Glimpse Data
```{r Glimpse}
tt |> 
  map(glimpse)
```

## Save Image
```{r chunk_05}
# This will save your most recent plot
ggsave(
  filename = "My TidyTuesday Plot.png",
  device = "png")
```

---

**→ [Hour 2: Modeling](h2_modeling_art_collections.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
