---
title: "Hour 1 — EDA: Analyzing franchise revenue in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 029 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`coord_flip_horizontal` · `faceted_comparison` · `fct_reorder_stat` · `glue_strings` · `grouped_summary_ggplot` · `heatmap_tile` · `spline_regression` · `theme_set_global`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(glue)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
theme_set(theme_light())

media_franchises <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-07-02/media_franchises.csv")
```
```{r chunk_02}
media_franchises |>
  count(franchise, sort = TRUE)

franchises <- media_franchises |>
  group_by(franchise, original_media, year_created, creators, owners) |>
  summarize(categories = n(),
            total_revenue = sum(revenue),
            most_profitable = revenue_category[which.max(revenue)]) |>
  ungroup()

franchises

media_franchises
```

## Franchise-level
```{r chunk_03}
library(glue)

top_franchises <- franchises |>
  mutate(franchise = glue("{ franchise } ({ year_created })")) |>
  top_n(20, total_revenue)

media_franchises |>
  mutate(franchise = glue("{ franchise } ({ year_created })")) |>
  semi_join(top_franchises, by = "franchise") |>
  mutate(franchise = fct_reorder(franchise, revenue, sum),
         revenue_category = fct_reorder(revenue_category, revenue, sum)) |>
  ggplot(aes(franchise, revenue)) +
  geom_col(aes(fill = revenue_category)) +
  geom_text(aes(y = total_revenue,
                label = paste0(scales::dollar(total_revenue, accuracy = 1), "B")),
            data = top_franchises,
            hjust = 0) +
  scale_y_continuous(labels = scales::dollar) +
  expand_limits(y = 100) +
  coord_flip() +
  theme(panel.grid.major.y = element_blank()) +
  guides(fill = guide_legend(reverse = TRUE)) +
  labs(title = "What are the most profitable franchises of all time?",
       fill = "Category",
       x = "",
       y = "Revenue (Billions)")
```
```{r chunk_04}
media_franchises |>
  group_by(owners) |>
  filter(n_distinct(franchise) > 2) |>
  ungroup() |>
  mutate(franchise = fct_reorder(franchise, revenue, sum),
         owners = fct_reorder(owners, -revenue, sum),
         revenue_category = fct_reorder(revenue_category, revenue, sum)) |>
  ggplot(aes(franchise, revenue, fill = revenue_category)) +
  geom_col() +
  facet_wrap(~ owners, scales = "free_y") +
  guides(fill = guide_legend(reverse = TRUE)) +
  coord_flip() +
  labs(title = "What companies own at least 3 franchises?",
       fill = "Category",
       x = "",
       y = "Revenue (Billions)")
```
```{r chunk_05}
franchises |>
  ggplot(aes(year_created, total_revenue)) +
  geom_point(aes(size = total_revenue, color = original_media)) +
  geom_text(aes(label = franchise), check_overlap = TRUE, vjust = 1, hjust = 1) +
  expand_limits(x = 1910) +
  labs(title = "When were the 'great' franchises created?")
```
```{r chunk_06}
media_franchises |>
  group_by(original_media) |>
  summarize(revenue = sum(revenue)) 

original_media_revenue_categories <- media_franchises |>
  group_by(original_media) |>
  filter(sum(revenue) >= 45) |>
  group_by(original_media, revenue_category) |>
  summarize(revenue = sum(revenue)) |>
  ungroup() |>
  mutate(revenue_category = fct_reorder(revenue_category, revenue, sum),
         original_media = fct_reorder(original_media, -revenue, sum))

original_media_revenue_categories |>
  ggplot(aes(revenue_category, revenue)) +
  geom_col() +
  scale_y_continuous(labels = scales::dollar) +
  coord_flip() +
  facet_wrap(~ original_media) +
  labs(x = "Revenue category",
       y = "Revenue (Billions)",
       title = "What kinds of media lead to what types of revenue?")
```
```{r chunk_07}
original_media_revenue_categories |>
  mutate(revenue_category = fct_rev(revenue_category),
         original_media = fct_rev(original_media)) |>
  ggplot(aes(revenue_category, original_media, fill = revenue)) +
  geom_tile() +
  scale_fill_gradient2(low = "white", high = "red", labels = scales::dollar) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(fill = "Revenue (billions)")
```

---

**→ [Hour 2: Modeling](h2_modeling_franchise_revenue.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
