---
title: "Hour 2 — Modeling: Analyzing franchise revenue in R"
subtitle: "Going deeper with statistical models | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 029 · Hour 2 · [← Hour 1 EDA](h1_eda_franchise_revenue.qmd)
:::


::: {.callout-note collapse="true"}
## Atomic Concepts — Hour 2
`coord_flip_horizontal` · `faceted_comparison` · `fct_reorder_stat` · `glue_strings` · `grouped_summary_ggplot` · `heatmap_tile` · `spline_regression` · `theme_set_global`
:::

## Setup

```{r setup}
library(tidyverse)
library(scales)
library(broom)
library(here)
theme_set(theme_light(base_size = 13))
```

## Reload data

```{r load-data}
# Re-load from Hour 1 — same local paths
source_qmd <- here::here("output/episodes/franchise_revenue/h1_eda_franchise_revenue.qmd")
# Data files are re-read directly below if needed
```

## Model 1 — Summary statistics

```{r summary-setup}
df <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2019/2019-07-02/media_franchises.csv", show_col_types = FALSE)
glimpse(df)
```

```{r summary-stats}
# Numeric columns summary
df |>
  select(where(is.numeric)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  group_by(variable) |>
  summarise(
    n      = n(),
    mean   = mean(value, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    sd     = sd(value, na.rm = TRUE),
    .groups = "drop"
  )
```

## Model 2 — Correlation matrix

```{r correlations}
numeric_df <- df |> select(where(is.numeric)) |> drop_na()
if (ncol(numeric_df) >= 2) {
  cor_mat <- cor(numeric_df)
  cor_mat |>
    as.data.frame() |>
    rownames_to_column("var1") |>
    pivot_longer(-var1, names_to = "var2", values_to = "correlation") |>
    filter(var1 < var2) |>
    slice_max(abs(correlation), n = 20) |>
    mutate(pair = paste(var1, "↔", var2),
           pair = fct_reorder(pair, correlation)) |>
    ggplot(aes(correlation, pair, fill = correlation > 0)) +
    geom_col(show.legend = FALSE) +
    geom_vline(xintercept = 0, lty = 2) +
    labs(title = "Top correlations", x = "Pearson r", y = NULL)
} else {
  cat("Not enough numeric columns for correlation matrix\n")
}
```

## Model 3 — Top categories by frequency

```{r top-categories}
char_cols <- df |> select(where(is.character)) |> names()
if (length(char_cols) > 0) {
  col <- char_cols[1]
  df |>
    count(.data[[col]], sort = TRUE) |>
    slice_max(n, n = 20) |>
    mutate(!!col := fct_reorder(.data[[col]], n)) |>
    ggplot(aes(n, .data[[col]])) +
    geom_col(fill = "#1b6ca8") +
    labs(title = paste("Top", col, "by frequency"), x = "Count", y = NULL)
} else {
  cat("No character columns found\n")
}
```


---

**→ [Hour 3: Communicate](h3_communicate_franchise_revenue.qmd)**
