---
title: "Hour 1 — EDA: Analyzing Broadway shows in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 013 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`rolling_window_metrics` · `shiny_app` · `spline_regression` · `str_detect_filter` · `theme_set_global` · `tidytuesdayr_load` · `time_series_trend`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(tidymetrics)
library(shinymetrics)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
cpi <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-04-28/cpi.csv", show_col_types = FALSE)
grosses <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-04-28/grosses.csv", show_col_types = FALSE)
pre-1985-starts <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-04-28/pre-1985-starts.csv", show_col_types = FALSE)
synopses <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-04-28/synopses.csv", show_col_types = FALSE)
glimpse(cpi)
glimpse(grosses)
```

## Explore
```{r chunk_02}
grosses <- grosses

grosses |>
  filter(show %in% c("Hamilton", "The Lion King")) |>
  ggplot(aes(week_ending, weekly_gross, color = show)) +
  geom_line() +
  scale_y_continuous(labels = scales::dollar) +
  expand_limits(y = 0)
```
```{r chunk_03}
# devtools::install_github("ramnathv/tidymetrics")
library(tidymetrics)

shows_summarized <- grosses |>
  filter(show %in% c("Hamilton", "The Lion King",
                     "Les Miserables", "Rent",
                     "The Phantom of the Opera", "Wicked",
                     "Harry Potter and the Cursed Child, Parts One and Two",
                     "The Book of Mormon")) |>
  mutate(show = str_remove(show, "\\, Parts.*")) |>
  rename(date = week_ending) |>
  cross_by_dimensions(show) |>
  cross_by_periods(c("month", "quarter", "year"),
                   windows = 28) |>
  summarize(usd_gross = sum(weekly_gross),
            avg_ticket_price = mean(avg_ticket_price),
            pct_capacity = mean(pct_capacity)) |>
  ungroup()

show_metrics <- create_metrics(shows_summarized)
```
```{r chunk_04}
# devtools::install_github("ramnathv/shinymetrics")
library(shinymetrics)
saveRDS(show_metrics, "broadway-shinybones/show_metrics.rds")

preview_metric(show_metrics$broadway_revenue_usd_gross)
preview_metric(show_metrics$broadway_revenue_avg_ticket_price)
preview_metric(show_metrics$broadway_revenue_pct_capacity)
```
```{r chunk_05}
shows_summarized |>
  filter(period == "quarter",
         show != "All") |>
  ggplot(aes(date, usd_gross, fill = show)) +
  geom_col()

shows_summarized |>
  filter(period == "quarter",
         show != "All") |>
  ggplot(aes(date, avg_ticket_price, col = show)) +
  geom_line() +
  expand_limits(y = 0)
```

---

**→ [Hour 2: Modeling](h2_modeling_broadway_shows.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
