---
title: "Hour 1 — EDA: Analyzing Women's World Cup data in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 078 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`animated_transitions` · `coord_flip_horizontal` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_histogram_density` · `geom_smooth_loess` · `glue_strings` · `grouped_summary_ggplot` · `multi_table_join` · `purrr_map_iterate` · `spline_regression`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- source HTML structure changes — selector-based scraping breaks silently
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- supplemental context is unavailable in the flat file and worth the scraping overhead
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(ggthemes)
library(rvest)
library(StatsBombR)
library(gganimate)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
library(ggthemes)
theme_set(theme_light())

wwc_outcomes <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-07-09/wwc_outcomes.csv")
squads <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-07-09/squads.csv")
codes <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-07-09/codes.csv")

outcomes <- wwc_outcomes |>
  left_join(codes, by = "team") |>
  group_by(year, yearly_game_id) |>
  mutate(opposing_score = rev(score)) |>
  ungroup() |>
  mutate(won_by = score - opposing_score)
```
```{r chunk_02}
library(rvest)

# Leave this code as a web scraping example
fifa_country_codes <- read_html("https://simple.wikipedia.org/wiki/List_of_FIFA_country_codes") |>
  html_nodes("table") |>
  map(html_table, fill = TRUE) |>
  .[2:5] |>
  bind_rows() |>
  tbl_df() |>
  select(country = Country, team = Code)
```
```{r chunk_03}
wwc_outcomes |>
  ggplot(aes(score)) +
  geom_histogram() +
  facet_wrap(~ win_status)

outcomes |>
  filter(year == 2019) |>
  count(round, sort = TRUE)
```
```{r chunk_04}
# Of the 3 games each country plays in the "group" round, how much did they win by on average?
avg_group_scores <- outcomes |>
  filter(round == "Group") |>
  group_by(year, team) |>
  summarize(avg_group_score = mean(score),
            avg_group_won_by = mean(won_by)) |>
  ungroup()

outcomes |>
  inner_join(avg_group_scores, by = c("year", "team")) |>
  filter(round == "Final") |>
  ggplot(aes(country, avg_group_won_by, fill = win_status)) +
  geom_col() +
  facet_wrap(~ year, scales = "free_x") +
  labs(title = "Does performance in the group round predict the winner of the finals?",
       subtitle = "Yes in all years except 2011. (2015 had been tied)",
       y = "Average # of goals the team had won by in the Group round",
       x = "Country",
       fill = "Result")
```
```{r chunk_05}
outcomes |>
  inner_join(avg_group_scores, by = c("year", "team")) |>
  filter(round != "Group") |>
  group_by(year, yearly_game_id) |>
  mutate(difference = diff(avg_group_won_by)) |>
  ungroup() |>
  filter(team_num == 2) |>
  mutate(round = fct_reorder(round, round, length)) |>
  ggplot(aes(difference, won_by)) +
  geom_point() +
  geom_smooth(method = "lm") +
  geom_abline(color = "red") +
  facet_wrap(~ round) +
  labs(x = "Difference in the team's average Group score from their opponent",
       y = "How much they won by",
       title = "Can we use a team's performance in the Group round to predict finals performance?",
       subtitle = "From 1991-2019")
```

## Play by play data
```{r chunk_06}
library(StatsBombR)

wwc_matches <- FreeMatches(72)

extract_match <- function(index) {
  wwc_match1 <- tbl_df(get.matchFree(wwc_matches[index, ])) |>
    mutate(timestamp = as.difftime(timestamp))
}

matches <- map_df(1:12, extract_match, .id = "match_index")

matches |>
  mutate(match_index = as.integer(match_index)) |>
  inner_join(wwc_matches |> mutate(match_index = row_number())) |>
  transmute(index,
            location,
            possession,
            minute,
            second,
            type = fct_lump(type.name, 8)) |>
  mutate(x = map_dbl(location, 1, .default = NA),
         y = map_dbl(location, 2, .default = NA),
         timestamp = minute * 60 + second)

ggplot(wwc_match1, aes(as.numeric(timestamp) / 60)) +
  geom_histogram()

match <- wwc_match1 |>
  transmute(index,
            location,
            possession,
            minute,
            second,
            type = fct_lump(type.name, 8)) |>
  mutate(x = map_dbl(location, 1, .default = NA),
         y = map_dbl(location, 2, .default = NA),
         timestamp = minute * 60 + second)

match |>
  ggplot(aes(x, y, color = type)) +
  geom_point() +
  facet_wrap(~ type)
```
```{r chunk_07}
library(gganimate)

p <- match |>
  filter(minute < 5) |>
  ggplot(aes(x, y, color = type)) +
  geom_point() +
  coord_fixed() +
  transition_reveal(index, timestamp)

animate(p, nframes = 20)
```
```{r chunk_08}
wwc_matches <- FreeMatches(72)

extract_match <- function(index) {
  wwc_match1 <- tbl_df(get.matchFree(wwc_matches[index, ])) |>
    mutate(timestamp = as.difftime(timestamp))
}

matches <- map_df(52, extract_match)

plays <- matches |>
  inner_join(wwc_matches, by = "match_id") |>
  transmute(match_id,
            index,
            location,
            possession,
            minute,
            second,
            duration,
            player = player.name,
            position = position.name,
            type = fct_lump(type.name, 8)) |>
  mutate(x = map_dbl(location, 1, .default = NA),
         y = map_dbl(location, 2, .default = NA),
         timestamp = minute * 60 + second)

plays |>
  filter(!is.na(player)) |>
  mutate(player = glue::glue("{ player } ({ position })"),
         player = fct_lump(player, 9)) |>
  count(player, type, sort = TRUE) |>
  mutate(type = fct_reorder(type, n, sum),
         player = fct_reorder(player, -n, sum)) |>
  filter(player != "Other") |>
  ggplot(aes(type, n)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ player) +
  labs(x = "",
       y = "# of plays they were involved in",
       title = "Which players were involved in the most plays in the 2019 Final?")

plays |>
  filter(!is.na(position)) |>
  mutate(position = fct_lump(position, 15)) |>
  ggplot(aes(x, y, color = position)) +
  geom_point() +
  facet_wrap(~ position)
```

---

**→ [Hour 2: Modeling](h2_modeling_women_s_world_cup.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
