---
title: "Hour 1 — EDA: Analyzing dolphin data in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 024 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`boxplot_jitter` · `broom_model_output` · `confidence_ribbon` · `coord_flip_horizontal` · `fct_lump_infrequent` · `fct_reorder_stat` · `fuzzy_join` · `geom_area_stacked` · `geom_histogram_density` · `grouped_summary_ggplot` · `linear_model_broom` · `lubridate_dates`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(tidyverse)
library(lubridate)
library(scales)
library(fuzzyjoin)
library(survival)
library(broom)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
theme_set(theme_light())

cetaceans_raw <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018-12-18/allCetaceanData.csv")

cetaceans <- cetaceans_raw |>
  select(-X1) |>
  mutate(birthYear = as.integer(birthYear),
         age = if_else(status == "Died", year(statusDate) - birthYear, NA_real_))
```
```{r chunk_02}
library(lubridate)

cetaceans |>
  count(acquisition, sort = TRUE)

cetaceans |>
  count(status, sort = TRUE)

cetaceans |>
  filter(status == "Died") |>
  select(birthYear, statusDate) |>
  filter(!is.na(birthYear), !is.na(statusDate)) |>
  mutate(age = year(statusDate) - birthYear) |>
  ggplot(aes(age)) +
  geom_histogram()

cetaceans |>
  count(species = fct_lump(species, 5), sort = TRUE) |>
  mutate(species = fct_reorder(species, n)) |>
  ggplot(aes(species, n)) +
  geom_col() +
  coord_flip()

cetaceans |>
  mutate(species = fct_lump(species, 5), sort = TRUE) |>
  filter(!is.na(birthYear), !is.na(statusDate)) |>
  mutate(age = year(statusDate) - birthYear) |>
  ggplot(aes(species, age)) +
  geom_boxplot() +
  coord_flip()
```
```{r chunk_03}
cetaceans |>
  count(acquisition, originLocation, sort = TRUE) |>
  filter(originLocation != "Unknown") |>

cetaceans |>
  ggplot(aes(originDate, fill = acquisition)) +
  geom_histogram()

library(scales)

cetaceans |>
  filter(originDate >= "1960-01-01") |>
  count(acquisition,
        decade = 5 * (year(originDate) %/% 5)) |>
  complete(acquisition, decade, fill = list(n = 0)) |>
  mutate(acquisition = fct_reorder(acquisition, n, sum)) |>
  group_by(decade) |>
  mutate(percent = n / sum(n)) |>
  ggplot(aes(decade, percent, fill = acquisition)) +
  geom_area() +
  scale_y_continuous(labels = percent_format()) +
  theme_minimal() +
  labs(x = "year",
       y = "% of dolphins recorded")
```
```{r chunk_04}
library(fuzzyjoin)

regexes <- tribble(
  ~ regex, ~ category,
  "Unknown", "Unknown",
  "Gulf of Mexico", "Gulf of Mexico",
  "Florida|FL", "Florida",
  "Texas|TX", "Texas",
  "SeaWorld", "SeaWorld",
  "Pacific", "Pacific Ocean",
  "Atlantic", "Atlantic Ocean"
)

cetaceans_annotated <- cetaceans |>
  mutate(unique_id = row_number()) |>
  regex_left_join(regexes, c(originLocation = "regex")) |>
  distinct(unique_id, .keep_all = TRUE) |>
  mutate(category = coalesce(category, originLocation))

cetaceans_annotated |>
  filter(acquisition == "Capture") |>
  mutate(category = fct_lump(category, 4),
         category = fct_reorder(category, category, length)) |>
  ggplot(aes(originDate, fill = category)) +
  geom_histogram()

  count(category,
        decade = 5 * (year(originDate) %/% 5)) |>
  complete(category, decade, fill = list(n = 0)) |>
  mutate(category = fct_lump(category, 7),
         category = fct_reorder(category, n, sum)) |>
  group_by(decade) |>
  mutate(percent = n / sum(n)) |>
  ggplot(aes(decade, percent, fill = category)) +
  geom_area() +
  scale_y_continuous(labels = percent_format()) +
  theme_minimal() +
  labs(x = "year",
       y = "% of dolphins recorded",
       title = "Dolphins born in captivity")
```

## Survival analysis
```{r chunk_05}
library(survival)

dolphin_survival <- cetaceans |>
  filter(status %in% c("Alive", "Died")) |>
  mutate(deathYear = ifelse(status == "Alive", 2017, year(statusDate)),
         status = ifelse(status == "Alive", 0, 1),
         age = deathYear - birthYear) |>
  filter(!is.na(deathYear)) |>
  select(birthYear, deathYear, status, sex, age, acquisition, species) |>
  filter(deathYear >= birthYear) |>
  filter(sex != "U")

model <- survfit(Surv(age, status) ~ sex, dolphin_survival)

library(broom)

tidy(model) |>
  ggplot(aes(time, estimate, color = strata)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = .2) +
  scale_y_continuous(labels = percent_format()) +
  labs(y = "Estimated % survival")

coxph(Surv(age, status) ~ sex, dolphin_survival) |>
  tidy()
```
```{r chunk_06}
model <- survfit(Surv(age, status) ~ acquisition, dolphin_survival)

library(broom)

tidy(model) |>
  filter(strata != "acquisition=Unknown") |>
  ggplot(aes(time, estimate, color = strata)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = .2) +
  scale_y_continuous(labels = percent_format()) +
  labs(y = "Estimated % survival")
```

---

**→ [Hour 2: Modeling](h2_modeling_dolphin.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
