---
title: "Hour 1 — EDA: Analyzing US dairy consumption in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 072 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`confidence_ribbon` · `coord_flip_horizontal` · `crossing_simulation_grid` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `grouped_summary_ggplot` · `lubridate_dates` · `pivot_reshape` · `purrr_map_iterate` · `rolling_window_metrics` · `spline_regression`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(sweep)
library(timetk)
library(lubridate)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
theme_set(theme_light())

milk_products_facts <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-01-29/milk_products_facts.csv")
```
```{r chunk_02}
milk_products_tidied <- milk_products_facts |>
  gather(product, lbs_per_person, -year) |>
  separate(product, c("category", "product"), sep = "_",
           extra = "merge", fill = "right") |>
  mutate(product = coalesce(product, category),
         product = str_to_title(str_replace_all(product, "_", " ")),
         category = str_to_title(category),
         product = ifelse(product == "Other", paste(product, category), product))

milk_products_tidied |>
  group_by(category, year) |>
  summarize(lbs_per_person = sum(lbs_per_person)) |>
  ggplot(aes(year, lbs_per_person, color = category)) +
  geom_line() +
  labs(title = "Dairy consumption by category",
       subtitle = "Based on US consumption (source: USDA)",
       x = "Year",
       y = "Lbs per person")

milk_products_tidied |>
  group_by(product = fct_lump(product, 6, w = lbs_per_person),
           year) |>
  summarize(lbs_per_person = sum(lbs_per_person)) |>
  ggplot(aes(year, lbs_per_person, color = product)) +
  geom_line()

milk_products_tidied |>
  ggplot(aes(year, lbs_per_person)) +
  geom_line() +
  facet_wrap(~ product, scales = "free") +
  expand_limits(y = 0)

milk_products_tidied |>
  filter(year == max(year)) |>
  mutate(product = fct_reorder(product, lbs_per_person, sum)) |>
  ggplot(aes(product, lbs_per_person, fill = category)) +
  geom_col() +
  coord_flip() +
  labs(x = "",
       y = "Pounds consumed per US person in 2017")
```
```{r chunk_03}
library(sweep)
library(timetk)
library(lubridate)

milk_product_ts <- milk_products_tidied |>
  mutate(year = as.Date("0001-01-01") + years(year - 1)) |>
  nest(-category, -product) |>
  mutate(ts = map(data, tk_ts, start = 1975, freq = 1))

milk_product_ets <- milk_product_ts |>
  mutate(model = map(ts, ets))

milk_product_ets |>
  unnest(map(model, sw_glance))

milk_product_ts |>
  crossing(model_name = c("auto.arima", "ets")) |>
  mutate(model = map2(model_name, ts, ~ invoke(.x, list(.y))),
         forecast = map(model, forecast, h = 10)) |>
  unnest(map(forecast, sw_sweep)) |>
  ggplot(aes(index, lbs_per_person, color = model_name, lty = key)) +
  geom_line() +
  geom_ribbon(aes(ymin = lo.80, ymax = hi.80), alpha = .5) +
  facet_wrap(~ product, scales = "free_y") +
  expand_limits(y = 0) +
  scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  scale_linetype_discrete(guide = FALSE) +
  labs(x = "Year",
       y = "Average US consumption (lbs per person)",
       title = "Forecasted consumption of dairy products",
       subtitle = "Based on USDA data 1975-2017. Showing 80% prediction intervals.",
       color = "Model")
```
```{r chunk_04}
cheese <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-01-29/clean_cheese.csv")

cheese_tidied <- cheese |>
  gather(type, lbs_per_person, -Year) |>
  rename(year = Year) |>
  mutate(type = str_to_title(type),
         type = fct_recode(type, "Total American Cheese" = "Total American Chese"))

cheese_tidied |>
  ggplot(aes(year, lbs_per_person)) +
  geom_line() +
  facet_wrap(~ type, scales = "free_y") +
  expand_limits(y = 0)

cheese_ts <- cheese_tidied |>
  mutate(year = as.Date("0001-01-01") + years(year - 1)) |>
  nest(-type) |>
  mutate(ts = map(data, tk_ts, start = 1970, freq = 1))

cheese_ts |>
  crossing(model_name = c("auto.arima", "ets")) |>
  mutate(model = map2(model_name, ts, ~ invoke(.x, list(.y))),
         forecast = map(model, forecast, h = 10)) |>
  unnest(map(forecast, sw_sweep)) |>
  ggplot(aes(index, lbs_per_person, color = model_name, lty = key)) +
  geom_line() +
  geom_ribbon(aes(ymin = lo.80, ymax = hi.80), alpha = .5) +
  facet_wrap(~ type, scales = "free_y") +
  expand_limits(y = 0) +
  scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  scale_linetype_discrete(guide = FALSE) +
  labs(x = "Year",
       y = "Average US consumption (lbs per person)",
       title = "Forecasted consumption of dairy products",
       subtitle = "Based on USDA data 1975-2017. Showing 80% prediction intervals.",
       color = "Model")
```

---

**→ [Hour 2: Modeling](h2_modeling_dairy_consumption.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
