---
title: "Hour 1 — EDA: Analyzing global crop yields in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 032 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `boxplot_jitter` · `countrycode_standardize` · `faceted_comparison` · `fct_reorder_stat` · `grouped_summary_ggplot` · `janitor_clean` · `label_repulsion_ggrepel` · `log_scale` · `pivot_reshape` · `spline_regression` · `theme_set_global`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(ggrepel)
library(countrycode)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
arable_land_pin <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-09-01/arable_land_pin.csv", show_col_types = FALSE)
cereal_crop_yield_vs_fertilizer_application <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-09-01/cereal_crop_yield_vs_fertilizer_application.csv", show_col_types = FALSE)
cereal_yields_vs_tractor_inputs_in_agriculture <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-09-01/cereal_yields_vs_tractor_inputs_in_agriculture.csv", show_col_types = FALSE)
key_crop_yields <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-09-01/key_crop_yields.csv", show_col_types = FALSE)
land_use_vs_yield_change_in_cereal_production <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-09-01/land_use_vs_yield_change_in_cereal_production.csv", show_col_types = FALSE)
glimpse(arable_land_pin)
glimpse(cereal_crop_yield_vs_fertilizer_application)
```

## Load the weekly Data
```{r chunk_01}
tt$key_crop_yields |>

yields <- tt$key_crop_yields |>
  clean_names() |>
  rename_all(str_remove, "_tonnes.*")

arable_land <- tt$arable_land_pin |>
  clean_names() |>
  rename(arable_land_needed = 4)

arable_land |>
  filter(entity == "Afghanistan") |>
  ggplot(aes(year, arable_land_needed)) +
  geom_line()

fertilizer <- tt$cereal_crop_yield_vs_fertilizer_application |>
  janitor::clean_names() |>
  rename(yield = 4,
         fertilizer_use = 5)
```
```{r chunk_02}
yields_tidy <- yields |>
  pivot_longer(wheat:bananas, names_to = "crop", values_to = "yield") |>
  filter(!is.na(yield)) |>
  mutate(crop = str_replace_all(crop, "_", " "),
         crop = str_to_title(crop))

yields_tidy |>
  write_rds("crop-yields-shiny/yields_tidy.rds")
```
```{r chunk_03}
yields_tidy |>
  filter(code == "USA") |>
  mutate(crop = fct_reorder(crop, -yield)) |>
  ggplot(aes(year, yield)) +
  geom_line() +
  facet_wrap(~ crop)

yields_tidy |>
  filter(code == "USA") |>
  mutate(crop = fct_reorder(crop, -yield)) |>
  ggplot(aes(year, yield, color = crop)) +
  geom_line() +
  labs(x = "Year",
       y = "Yield (tonnes per hectare)",
       title = "Crop yields in the US over time",
       color = "Crop")

yields_tidy |>
  filter(entity == "India") |>
  mutate(crop = fct_reorder(crop, -yield)) |>
  ggplot(aes(year, yield, color = crop)) +
  geom_line() +
  labs(x = "Year",
       y = "Yield (tonnes per hectare)",
       title = "Crop yields in the US over time",
       color = "Crop")
```
```{r chunk_04}
yields_tidy |>
  filter(crop == "Wheat") |>
  add_count(entity) |>
  filter(n == max(n)) |>
  filter(entity %in% sample(unique(entity), 25)) |>
  ggplot(aes(year, yield)) +
  geom_line() +
  facet_wrap(~ entity)

crop_yields_50_years <- yields_tidy |>
  arrange(entity, year) |>
  filter(year >= 1968) |>
  group_by(entity, code, crop) |>
  summarize(year_start = min(year),
            year_end = max(year),
            yield_start = first(yield),
            yield_end = last(yield)) |>
  ungroup() |>
  filter(year_start == 1968) |>
  mutate(yield_ratio = yield_end / yield_start)

crop_yields_50_years |>
  filter(!is.na(code)) |>
  ggplot(aes(yield_start, yield_end)) +
  geom_abline(color = "red") +
  geom_point() +
  facet_wrap(~ crop, scales = "free")

crop_yields_50_years |>
  mutate(crop = fct_reorder(crop, yield_ratio)) |>
  ggplot(aes(yield_ratio, crop)) +
  geom_boxplot() +
  scale_x_log10()

crop_yields_50_years |>
  group_by(crop) |>
  summarize(median_yield_ratio = median(yield_ratio)) |>
  mutate(crop = fct_reorder(crop, median_yield_ratio)) |>
  ggplot(aes(median_yield_ratio, crop)) +
  geom_col() +
  labs(title = "How much has the average country improved at producing this crop?",
       x = "(2018 yield) / (1968 yield)",
       y = "")

crop_yields_50_years |>
  filter(is.na(code)) |>
  filter(entity %in% c("Africa", "Asia", "Northern America", "South America", "Oceania")) |>
  ggplot(aes(yield_start, yield_end, color = entity)) +
  geom_abline(color = "red") +
  geom_point() +
  expand_limits(y = 0, x = 0) +
  facet_wrap(~ crop, scales = "free") +
  labs(x = "Tonnes per hectare in 1968",
       y = "Tonnes per hectare in 2018",
       color = "Continent")
```
```{r chunk_05}
library(ggrepel)
library(countrycode)

crop_yields_50_years |>
  filter(crop == "Wheat",
         !is.na(code)) |>
  mutate(continent = countrycode(code, "iso3c", "continent")) |>
  filter(!is.na(continent)) |>
  ggplot(aes(yield_start, yield_ratio)) +
  geom_point(aes(color = continent)) +
  scale_x_log10() +
  scale_y_log10(breaks = c(.25, .5, 1, 2, 4),
                labels = c("1/4X", "1/2X", "Same", "2X", "4X")) +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_text_repel(aes(label = entity), force = .1,
                  size = 2.5) +
  labs(x = "1968 yield (tonnes per hectare), log scale",
       y = "(2018 yield) / (1968 yield), log scale",
       color = "Continent",
       title = "How has wheat efficiency changed across countries?")

countrycode("USA", "iso3c", "continent")
```

---

**→ [Hour 2: Modeling](h2_modeling_global_crop_yields.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
