---
title: "Hour 1 — EDA: Analyzing Thanksgiving dinners in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 060 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`confidence_ribbon` · `coord_flip_horizontal` · `empirical_bayes` · `faceted_comparison` · `fct_reorder_stat` · `grouped_summary_ggplot` · `multi_table_join` · `network_graph_ggraph` · `pairwise_token_correlation` · `parse_number_dirty` · `pivot_reshape` · `scales_axis_labels`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(ebbr)
library(widyr)
library(ggraph)
library(igraph)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
theme_set(theme_light())

thanksgiving_survey <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018-11-20/thanksgiving_meals.csv") |>
  mutate(family_income = fct_reorder(family_income, parse_number(family_income)))
```
```{r chunk_02}
thanksgiving_survey |>
  count(celebrate)
```
```{r chunk_03}
thanksgiving_survey |>
  count(main_dish, sort = TRUE)
```
```{r chunk_04}
thanksgiving_survey |>
  count(main_dish, main_prep, sort = TRUE)
```
```{r chunk_05}
thanksgiving_survey |>
  count(cranberry, sort = TRUE)

thanksgiving_survey |>
  count(gravy, sort = TRUE)
```

## Relationship with income
```{r chunk_06}
thanksgiving_survey |>
  group_by(family_income) |>
  summarize(celebrate = sum(celebrate == "Yes"),
            total = n(),
            low = qbeta(0.025, celebrate + .5, total - celebrate + .5),
            high = qbeta(0.975, celebrate + .5, total - celebrate + .5)) |>
  ggplot(aes(family_income, celebrate / total, group = 1)) +
  geom_line() +
  geom_ribbon(aes(ymin = low, ymax = high), alpha = .2) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Family income",
       y = "% celebrating Thanksgiving")
```
```{r chunk_07}
thanksgiving_survey |>
  filter(cranberry %in% c("Canned", "Homemade")) |>
  group_by(family_income) |>
  summarize(homemade = sum(cranberry == "Homemade"),
            total = n(),
            low = qbeta(0.025, homemade + .5, total - homemade + .5),
            high = qbeta(0.975, homemade + .5, total - homemade + .5)) |>
  ggplot(aes(family_income, homemade / total, group = 1)) +
  geom_line() +
  geom_ribbon(aes(ymin = low, ymax = high), alpha = .2) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Family income",
       y = "% serving homemade")
```
```{r chunk_08}
food_gathered <- thanksgiving_survey |>
  select(id, starts_with("side"),
         starts_with("pie"),
         starts_with("dessert")) |>
  select(-side15, -pie13, -dessert12) |>
  gather(type, value, -id) |>
  filter(!is.na(value),
         !value %in% c("None", "Other (please specify)")) |>
  mutate(type = str_remove(type, "\\d+"))

n_respondents <- n_distinct(food_gathered$id)
```
```{r chunk_09}
food_gathered |>
  count(type, value, sort = TRUE) |>
  mutate(value = fct_reorder(value, n)) |>
  ggplot(aes(value, n / n_respondents, fill = type)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format()) +
  facet_wrap(~ type, scales = "free_y", ncol = 1) +
  labs(x = "",
       y = "% of respondents")
```
```{r chunk_10}
food_gathered |>
  inner_join(thanksgiving_survey, by = "id") |>
  mutate(age_number = parse_number(age)) |>
  group_by(value) |>
  summarize(average_age = mean(age_number, na.rm = TRUE),
            total = n()) |>
  arrange(desc(average_age))

food_by_region <- food_gathered |>
  inner_join(thanksgiving_survey, by = "id") |>
  group_by(us_region) |>
  mutate(respondents = n_distinct(id)) |>
  count(us_region, respondents, type, value) |>
  ungroup() |>
  mutate(percent = n / respondents)

food_by_region |>
  filter(value == "Cornbread") |>
  arrange(desc(percent))
```
```{r chunk_11}
library(ebbr)

food_gathered |>
  inner_join(thanksgiving_survey, by = "id") |>
  filter(!is.na(prayer)) |>
  group_by(type, value) |>
  summarize(prayer = sum(prayer == "Yes"),
            total = n()) |>
  add_ebb_estimate(prayer, total) |>
  arrange(desc(.fitted)) |>
```

## Network of foods
```{r chunk_12}
library(widyr)

food_cors <- food_gathered |>
  pairwise_cor(value, id, sort = TRUE)
```
```{r chunk_13}
library(ggraph)
library(igraph)

set.seed(2018)

food_types <- food_gathered |>
  count(value, type, sort = TRUE)

food_cors |>
  head(75) |>
  graph_from_data_frame(vertices = food_types) |>
  ggraph() +
  geom_edge_link() +
  geom_node_point(aes(color = type, size = n / n_respondents)) +
  geom_node_text(aes(label = name), vjust = 1, hjust = 1, repel = TRUE) +
  scale_size_continuous(labels = scales::percent_format()) +
  theme_void() +
  labs(title = "What foods get served together at Thanksgiving?",
       color = "",
       size = "% of respondents")
```

---

**→ [Hour 2: Modeling](h2_modeling_thanksgiving_dinners.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
