---
title: "Hour 1 — EDA: Analyzing squirrels in NYC in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 057 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `choropleth_sf` · `geom_smooth_loess` · `grouped_summary_ggplot` · `logistic_regression` · `skim_inspect` · `spline_regression` · `theme_set_global` · `time_series_trend`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- region name spelling differs between data and shapefile — silent dropped rows
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- geographic column can be matched to a standard ISO/FIPS shapefile without fuzzy joins
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(sf)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
theme_set(theme_light())

nyc_squirrels <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-10-29/nyc_squirrels.csv")
```
```{r chunk_02}
nyc_squirrels |>
  count(zip_codes, sort = TRUE)
```
```{r chunk_03}
nyc_squirrels |>
  ggplot(aes(long, lat)) +
  geom_point()

by_hectare <- nyc_squirrels |>
  filter(!is.na(primary_fur_color)) |>
  group_by(hectare) |>
  summarize(long = mean(long),
            lat = mean(lat),
            pct_gray = mean(primary_fur_color == "Gray", na.rm = TRUE),
            n = n())

by_hectare |>
  filter(n >= 10) |>
  ggplot(aes(long, lat, size = n, color = pct_gray)) +
  geom_point() +
  theme_void()

by_hectare |>
  filter(n >= 10) |>
  ggplot(aes(lat, pct_gray)) +
  geom_point() +
  geom_smooth()

by_hectare |>
  mutate(n_gray = round(pct_gray * n)) |>
  glm(cbind(n_gray, n - n_gray) ~ lat, data = ., family = "binomial") |>
  summary()
```
```{r chunk_04}
nyc_squirrels |>
  count(highlight_fur_color, sort = TRUE)

nyc_squirrels |>
  count(approaches, indifferent, runs_from, sort = TRUE)
```
```{r chunk_05}
glm(runs_from ~ lat, data = nyc_squirrels, family = "binomial") |>
  summary()
```
```{r chunk_06}
library(sf)

central_park_sf <- read_sf("~/Downloads/CentralAndProspectParks/")

by_hectare <- nyc_squirrels |>
  add_count(hectare) |>
  mutate(above_ground = !is.na(location) & location == "Above Ground") |>
  group_by(hectare, n) |>
  summarize_at(vars(long, lat, approaches:runs_from,  ends_with("ing"), above_ground), mean) |>
  ungroup()

by_hectare |>
  filter(n >= 10) |>
  ggplot() +
  geom_sf(data = central_park_sf) +  
  geom_point(aes(long, lat, size = n, color = runs_from)) +
  theme_void() +
  scale_color_gradient2(low = "blue", high = "red", mid = "pink",
                        midpoint = .3, labels = scales::percent) +
  labs(color = "% of squirrels run",
       size = "# of squirrels",
       title = "Squirrels in the northwest corner of Central Park are more likely to run away") +
  coord_sf(datum = NA)
```
```{r chunk_07}
central_park_sf |>
  count(lanes, sort = TRUE)

ggplot(central_park_sf) +
  geom_sf() +
  geom_point(aes(long, lat, color = runs_from), data = by_hectare) +
  coord_sf(datum = NA)
```
```{r chunk_08}
central_park_sf |>
  ggplot() +
  geom_sf(aes(color = bicycle)) +
  coord_sf(datum = NA)
```

---

**→ [Hour 2: Modeling](h2_modeling_squirrels_nyc.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
