---
title: "Hour 1 — EDA: Analyzing beach volleyball in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 007 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`fct_lump_infrequent` · `geom_smooth_loess` · `grouped_summary_ggplot` · `log_scale` · `lubridate_dates` · `multi_table_join` · `pivot_reshape` · `skim_inspect` · `spline_regression` · `str_detect_filter` · `theme_set_global` · `tidytuesdayr_load`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
vb_matches <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-05-19/vb_matches.csv", show_col_types = FALSE)
glimpse(vb_matches)
```

## Explore
```{r chunk_02}
vb_matches |>
  count(circuit, tournament, date, sort = TRUE)

vb_matches |>
  count(gender, sort = TRUE)

vb_matches |>
  count(year) |>
```
```{r chunk_03}
vb_long <- vb_matches |>
  rename(w_p1_name = w_player1, w_p2_name = w_player2,
         l_p1_name = l_player1, l_p2_name = l_player2,
         w_team_rank = w_rank, l_team_rank = l_rank) |>
  mutate_at(vars(starts_with("w_"), starts_with("l_")), as.character) |>
  pivot_longer(cols = c(starts_with("w_"), starts_with("l_"))) |>
  separate(name, c("winner_loser", "player", "name"),
           sep = "_",
           extra = "merge",
           fill = "right") |>
  mutate(winner_loser = str_to_upper(winner_loser))

vb_player_matches <- vb_long |>
  filter(name != "rank") |>
  spread(name, value) |>
  type_convert()
```
```{r chunk_04}
vb_sets <- vb_matches |>
  select(match_id, circuit:match_num, score) |>
  separate_rows(score, sep = ", ") |>
  mutate(score = str_remove(score, " retired")) |>
  mutate(score = na_if(score, "Forfeit or other")) |>
  separate(score, c("w_score", "l_score"), convert = TRUE)
```

## Look at players
```{r chunk_05}
by_player <- vb_player_matches |>
  group_by(name, gender) |>
  summarize(n_matches = n(),
            pct_winner = mean(winner_loser == "W"),
            first_game = min(date),
            last_game = max(date)) |>
  arrange(desc(n_matches)) |>
  ungroup()

by_player |>
  filter(n_matches >= 200) |>
  ggplot(aes(n_matches, pct_winner, color = gender)) +
  geom_point() +
  scale_x_log10() +
  scale_y_continuous(labels = percent) +
  labs(x = "# of matches since 2000 (log scale)",
       y = "% of matches won")

by_player |>
  filter(n_matches >= 200) |>
  arrange(desc(pct_winner))
```
```{r chunk_06}
vb_player_matches |>
  summarize_all(~ mean(!is.na(.))) |>
  gather() |>
```
```{r chunk_07}
vb_player_matches |>
  group_by(tournament) |>
  summarize(pct_has_attacks = mean(!is.na(tot_attacks)),
            n = n()) |>
  arrange(desc(n))
```

## How would we predict if a player will win in 2019?
```{r chunk_08}
summarize_players <- . |>
  summarize(n_matches = n(),
            pct_winner = mean(winner_loser == "W"),
            avg_attacks = mean(tot_attacks, na.rm = TRUE),
            avg_errors = mean(tot_errors, na.rm = TRUE),
            avg_serve_errors = mean(tot_serve_errors, na.rm = TRUE),
            avg_kills = mean(tot_kills, na.rm = TRUE),
            avg_aces = mean(tot_aces, na.rm = TRUE),
            n_with_data = sum(!is.na(tot_attacks))) |>
  ungroup() |>
  arrange(desc(n_matches))

players_before_2019 <- vb_player_matches |>
  filter(year < 2019) |>
  group_by(name, gender, hgt, birthdate, country) |>
  summarize_players() |>
  filter(!is.na(avg_attacks))

players_2019 <- vb_player_matches |>
  filter(year == 2019) |>
  group_by(name, gender, hgt, birthdate, country, year,
           age = year - year(birthdate)) |>
  summarize_players()
```
```{r chunk_09}
performance_joined <- players_before_2019 |>
  inner_join(players_2019 |>
               select(name, n_matches, pct_winner),
             by = "name",
             suffix = c("", "_2019"))

performance_joined |>
  filter(n_matches >= 10,
         n_matches_2019 >= 5) |>
  ggplot(aes(pct_winner, pct_winner_2019)) +
  geom_point() +
  geom_abline(color = "red") +
  geom_smooth(method = "lm")

performance_joined |>
  mutate(n_wins_2019 = n_matches_2019 * pct_winner_2019,
         country = fct_lump(country, 3)) |>
  glm(cbind(n_wins_2019, n_matches_2019 - n_wins_2019) ~
        pct_winner + avg_errors + avg_serve_errors,
      data = .,
      family = "binomial") |>
  summary()
```
```{r chunk_10}
players_before_2019 |>
  filter(n_with_data >= 20) |>
  ggplot(aes(avg_serve_errors, avg_aces, size = n_with_data)) +
  geom_point() +
  labs(size = "Matches",
       title = "DON'T TRUST THIS")
```
```{r chunk_13}
player_first_year_summarized |>
  filter(!is.nan(avg_attacks))
```

---

**→ [Hour 2: Modeling](h2_modeling_beach_volleyball.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
