---
title: "Hour 1 — EDA: Analyzing deforestation in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 023 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `fuzzy_join` · `geom_area_stacked` · `grouped_summary_ggplot` · `multi_table_join` · `pivot_reshape` · `reorder_within_facets` · `spline_regression` · `theme_set_global` · `tidytuesdayr_load`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- region name spelling differs between data and shapefile — silent dropped rows
- text is multilingual without language detection
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- geographic column can be matched to a standard ISO/FIPS shapefile without fuzzy joins
- text is long-form English and benefits from word-level decomposition
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(tidytext)
library(fuzzyjoin)
library(ggthemes)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Load the weekly Data
```{r Load}
tt <- tt_load("2021-04-06")

tt$forest |>
  count(entity, sort = TRUE) |>

forest_area |>
  filter(entity == "United States") |>
  ggplot(aes(year, forest_area)) +
  geom_line() +
  expand_limits(y = 0)

forest_area_country <- tt$forest_area |>
  filter(str_length(code) == 3,
         year >= 1992) |>
  rename(country = entity) |>
  mutate(forest_area = forest_area / 100)

forest <- tt$forest |>
  filter(str_length(code) == 3) |>
  rename(country = entity)

forest_area_country |>
  filter(country %in% c("United States", "Canada", "China", "India",
                        "Senegal", "Germany", "Brazil", "Poland")) |>
  mutate(country = fct_reorder(country, -forest_area)) |>
  ggplot(aes(year, forest_area, color = country)) +
  geom_line() +
  scale_y_continuous(labels = percent) +
  expand_limits(y = 0) +
  labs(x = "Year",
       y = "% of global forest area")
```
```{r chunk_01}
forest_area_country |>
  mutate(country = fct_lump(country, 9, w = forest_area)) |>
  group_by(country, year) |>
  summarize(forest_area = sum(forest_area),
            .groups = "drop") |>
  mutate(country = fct_reorder(country, -forest_area)) |>
  ggplot(aes(year, forest_area, fill = country)) +
  geom_area() +
  scale_y_continuous(labels = percent) +
  expand_limits(y = 0) +
  labs(x = "Year",
       y = "% of global forest area")

forest_area_country |>
  filter(year %in% c(1992, 2020)) |>
  mutate(year = paste0("forest_area_", year)) |>
  spread(year, forest_area) |>
  arrange(desc(forest_area_1992))
```
```{r chunk_02}
forest |>
  group_by(year) |>
  summarize(net_forest_conversion = sum(net_forest_conversion))

forest |>
  filter(year == 2015) |>
  arrange((net_forest_conversion)) |>
  slice_max(abs(net_forest_conversion), n = 20) |>
  mutate(country = fct_reorder(country, net_forest_conversion)) |>
  ggplot(aes(net_forest_conversion, country,
             fill = net_forest_conversion > 0)) +
  geom_col() +
  scale_x_continuous(labels = comma) +
  theme(legend.position = "none") +
  labs(x = "Net change in forest in 2015 (hectares)",
       y = "")

library(tidytext)

forest |>
  group_by(year) |>
  slice_max(abs(net_forest_conversion), n = 10) |>
  ungroup() |>
  mutate(country = reorder_within(country, net_forest_conversion, year)) |>
  ggplot(aes(net_forest_conversion, country,
             fill = net_forest_conversion > 0)) +
  geom_col() +
  facet_wrap(~ year, scales = "free_y") +
  scale_x_continuous(label = comma) +
  scale_y_reordered() +
  theme(legend.position = "none") +
  labs(x = "Net change in forest in 2015 (hectares)",
       y = "")

forest |>
  mutate(country = fct_lump(country, 8, w = abs(net_forest_conversion))) |>
  group_by(country, year) |>
  summarize(net_forest_conversion = sum(net_forest_conversion),
            .groups = "drop") |>
  mutate(country = fct_reorder(country, -net_forest_conversion)) |>
  ggplot(aes(year, net_forest_conversion, color = country)) +
  geom_line() +
  scale_y_continuous(labels = comma) +
  labs(y = "Net change in forest (hectares)")
```
```{r chunk_03}
brazil_loss <- tt$brazil_loss |>
  pivot_longer(commercial_crops:small_scale_clearing,
               names_to = "cause",
               values_to = "loss") |>
  mutate(cause = str_to_sentence(str_replace_all(cause, "_", " ")))

brazil_loss |>
  filter(year == max(year)) |>
  arrange(desc(loss)) |>
  mutate(cause = fct_reorder(cause, loss)) |>
  ggplot(aes(loss, cause)) +
  geom_col() +
  scale_x_continuous(labels = comma) +
  labs(x = "Loss of forest in 2013 (hectares)",
       y = "")

brazil_loss |>
  mutate(cause = fct_reorder(cause, -loss)) |>
  ggplot(aes(year, loss, color = cause)) +
  geom_line() +
  scale_y_continuous(labels = comma) +
  labs(y = "Loss of forest (hectares)",
       x = "")

brazil_loss |>
  mutate(cause = fct_lump(cause, 6, w = loss)) |>
  group_by(cause, year) |>
  summarize(loss = sum(loss), .groups = "drop") |>
  mutate(cause = fct_reorder(cause, loss)) |>
  ggplot(aes(year, loss, fill = cause)) +
  geom_area() +
  scale_y_continuous(labels = comma) +
  labs(y = "Loss of forest (hectares)",
       x = "")
```
```{r chunk_04}
soybean_use <- tt$soybean_use |>
  filter(str_length(code) == 3) |>
  rename(country = entity) |>
  mutate(total = human_food + animal_feed + processed) |>
  pivot_longer(human_food:processed,
               names_to = "use",
               values_to = "amount") |>
  replace_na(list(amount = 0)) |>
  arrange(desc(total)) |>
  mutate(use = str_to_sentence(str_replace_all(use, "_", " ")))

vegetable_oil <- tt$vegetable_oil |>
  rename(country = entity)

soybean_use |>
  filter(country %in% c("Brazil", "United States", "China", "Indonesia")) |>
  ggplot(aes(year, amount, fill = use)) +
  geom_area() +
  scale_y_continuous(labels = comma) +
  labs(y = "Soybeans (tonnes)",
       fill = "") +
  facet_wrap(~ country)

vegetable_oil |>
  filter(!is.na(production)) |>
  mutate(crop_oil = fct_lump(crop_oil, 5, w = production)) |>
  group_by(country, crop_oil, year) |>
  summarize(production = sum(production)) |>
  filter(country %in% c("United States", "India", "China", "Indonesia")) |>
  ggplot(aes(year, production, fill = crop_oil)) +
  geom_area() +
  scale_y_continuous(labels = comma) +
  facet_wrap(~ country)
```
```{r chunk_05}
library(fuzzyjoin)
library(ggthemes)

country_data <- forest |>
  filter(year == 2010) |>
  inner_join(maps::iso3166, by = c(code = "a3"))

map_data("world") |>
  as_tibble() |>
  filter(region != "Antarctica") |>
  regex_left_join(country_data, by = c(region = "mapname")) |>
  ggplot(aes(long, lat, group = group, fill = net_forest_conversion)) +
  geom_polygon(color = "black", size = .05) +
  scale_fill_gradient2(low = "red", high = "green",
                       labels = comma) +
  theme_map() +
  labs(fill = "Net forest change (2010)")
```

---

**→ [Hour 2: Modeling](h2_modeling_deforestation.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
