---
title: "Hour 1 — EDA: Analyzing malaria incidence in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 040 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`animated_transitions` · `countrycode_standardize` · `faceted_comparison` · `fuzzy_join` · `multi_table_join` · `pivot_reshape` · `scales_axis_labels` · `spline_regression` · `str_detect_filter` · `theme_set_global` · `time_series_trend`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- region name spelling differs between data and shapefile — silent dropped rows
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- geographic column can be matched to a standard ISO/FIPS shapefile without fuzzy joins
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(malariaAtlas)
library(fuzzyjoin)
library(stringr)
library(gganimate)
library(countrycode)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_02}
kenya_pr <- tbl_df(malariaAtlas::getPR(ISO = "KEN", species = "BOTH")) |>
  filter(!is.na(pr))
```
```{r chunk_03}
kenya_pr |>
  group_by(year_start) |>
  summarize(examined = sum(examined),
            positive = sum(positive),
            studies = n()) |>
  mutate(pr = positive / examined) |>
  ggplot(aes(year_start, pr)) +
  geom_line()

kenya_pr |>
  mutate(decade = 10 * (year_start %/% 10)) |>
  arrange(pr) |>
  ggplot(aes(longitude, latitude, color = pr)) +
  borders("world", regions = "Kenya") +
  geom_point() +
  scale_color_gradient2(low = "blue", high = "red", midpoint = .5, labels = scales::percent_format()) +
  facet_wrap(~ decade) +
  theme_void() +
  coord_map() +
  labs(color = "Prevalence")
```

## Aggregated across countries
```{r chunk_04}
malaria_inc <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018-11-13/malaria_inc.csv")

malaria_inc_processed <- malaria_inc |>
  setNames(c("country", "code", "year", "incidence")) |>
  mutate(incidence = incidence / 1000)
```
```{r chunk_05}
malaria_inc_processed |>
  filter(country %in% sample(unique(country), 6)) |>
  ggplot(aes(year, incidence, color = country)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent_format())
```
```{r chunk_06}
malaria_spread <- malaria_inc_processed |>
  mutate(year = paste0("Y", year)) |>
  spread(year, incidence)

malaria_spread |>
  filter(country != "Turkey",
         !is.na(code)) |>
  mutate(current = Y2015,
         change =  Y2015 - Y2000) |>
  ggplot(aes(current, change)) +
  geom_point() +
  geom_text(aes(label = code), vjust = 1, hjust = 1)
```
```{r chunk_07}
world <- map_data("world") |>
  filter(region != "Antarctica")

malaria_inc_processed |>
  filter(incidence < 1) |>
  inner_join(maps::iso3166 |>
               select(a3, mapname), by = c(code = "a3")) |>
  inner_join(world, by = c(mapname = "region")) |>
  ggplot(aes(long, lat, group = group, fill = incidence)) +
  geom_polygon() +
  scale_fill_gradient2(low = "blue", high = "red", midpoint = .20, labels = scales::percent_format()) +
  coord_map() +
  facet_wrap(~ year) +
  theme_void() +
  labs(title = "Malaria incidence over time around the world")
```

## Malaria deaths over time
```{r chunk_08}
library(tidyverse)

malaria_deaths <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018-11-13/malaria_deaths.csv")

malaria_deaths_processed <- malaria_deaths |>
  setNames(c("country", "code", "year", "deaths"))
```
```{r chunk_09}
malaria_deaths_processed |>
  filter(country %in% sample(unique(country), 6)) |>
  ggplot(aes(year, deaths, color = country)) +
  geom_line() +
  labs(y = "Deaths per 100,000")
```
```{r chunk_10}
library(fuzzyjoin)
library(stringr)

malaria_country_data <- malaria_deaths_processed |>
  inner_join(maps::iso3166 |>
               select(a3, mapname), by = c(code = "a3")) |>
  mutate(mapname = str_remove(mapname, "\\(.*"))

malaria_map_data <- map_data("world") |>
  filter(region != "Antarctica") |>
  tbl_df() |>
  inner_join(malaria_country_data, by = c(region = "mapname"))

malaria_map_data |>
  ggplot(aes(long, lat, group = group, fill = deaths)) +
  geom_polygon() +
  scale_fill_gradient2(low = "blue", high = "red", midpoint = 100) +
  theme_void() +
  labs(title = "Malaria deaths over time around the world",
       fill = "Deaths per 100,000")
```
```{r chunk_11}
library(gganimate)
library(countrycode)

malaria_map_data |>
  mutate(continent = countrycode(code, "iso3c", "continent")) |>
  filter(continent == "Africa") |>
  ggplot(aes(long, lat, group = group, fill = deaths)) +
  geom_polygon() +
  scale_fill_gradient2(low = "blue", high = "red", midpoint = 100) +
  theme_void() +
  labs(title = "Malaria deaths over time in Africa ({ current_frame })",
       fill = "Deaths per 100,000") +
  transition_manual(year)

anim_save("malaria_map.gif")
```

---

**→ [Hour 2: Modeling](h2_modeling_malaria_incidence.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
