---
title: "Hour 1 — EDA: Analyzing pizza ratings in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 050 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `boxplot_jitter` · `error_bar_coef_plot` · `faceted_comparison` · `fct_lump_infrequent` · `fct_relevel_manual` · `fct_reorder_stat` · `geom_smooth_loess` · `glue_strings` · `grouped_summary_ggplot` · `linear_model_broom` · `multi_table_join`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(broom)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## R Markdown
```{r chunk_02}
pizza_jared <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-10-01/pizza_jared.csv")
pizza_barstool <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-10-01/pizza_barstool.csv")
pizza_datafiniti <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-10-01/pizza_datafiniti.csv")
```
```{r chunk_03}
answer_orders <- c("Never Again", "Poor", "Average", "Good", "Excellent")

by_place_answer <- pizza_jared |>
  mutate(time = as.POSIXct(time, origin = "1970-01-01"),
         date = as.Date(time),
         answer = fct_relevel(answer, answer_orders)) |>
  group_by(place, answer) |>
  summarize(votes = sum(votes)) |>
  mutate(total = sum(votes),
         percent = votes / total,
         answer_integer = as.integer(answer),
         average = sum(answer_integer * percent)) |>
  ungroup()

by_place <- by_place_answer |>
  distinct(place, total, average)
```
```{r chunk_04}
by_place_answer |>
  filter(as.integer(fct_reorder(place, total, .desc = TRUE)) <= 16,
         answer != "Fair") |>
  mutate(place = glue::glue("{ place } ({ total })"),
         place = fct_reorder(place, average)) |>
  ggplot(aes(answer, percent)) +
  geom_col() +
  facet_wrap(~ place) +
  scale_y_continuous(labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "",
       y = "% of respondents",
       title = "What is the most popular pizza place in Open Stats meetup?",
       subtitle = "Only the 16 pizza places with the most respondents. # respondents shown in parentheses.")
```
```{r chunk_05}
library(broom)

# THIS TRICK DOESN'T WORK
# tidy(lm(c(1, 2, 3, 4, 5) ~ 1, weights = c(100, 300, 100, 200, 150)), conf.int = TRUE)

t_test_repeated <- function(x, frequency) {
  tidy(t.test(rep(x, frequency)))
}

by_place_answer |>
  filter(total >= 3) |>
  group_by(place, total) |>
  summarize(t_test_result = list(t_test_repeated(answer_integer, votes))) |>
  ungroup() |>
  unnest(t_test_result) |>
  select(place, total, average = estimate, low = conf.low, high = conf.high) |>
  top_n(16, total) |>
  mutate(place = fct_reorder(place, average)) |>
  ggplot(aes(average, place)) +
  geom_point(aes(size = total)) +
  geom_errorbarh(aes(xmin = low, xmax = high)) +
  labs(x = "Average score (1-5 Likert Scale)",
       y = "",
       title = "What is the most popular pizza place in Open Stats meetup?",
       subtitle = "Only the 16 pizza places with the most respondents.",
       size = "# of respondents")
```
```{r chunk_06}
# Don't bother comparing them, this is a bad graph
pizza_barstool |>
  select(place = name,
         barstool_total = review_stats_all_count,
         barstool_average = review_stats_all_average_score) |>
  inner_join(by_place, by = "place") |>
  group_by(place) |>
  filter(n() == 1) |>
  ungroup() |>
  filter(barstool_total >= 5,
         total >= 5) |>
  ggplot(aes(average, barstool_average)) +
  geom_point() +
  labs(x = "Meetup",
       y = "Barstool")
```
```{r chunk_07}
pizza_barstool |>
  top_n(50, review_stats_all_count) |>
  ggplot(aes(price_level, review_stats_all_average_score, group = price_level)) +
  geom_boxplot()

pizza_barstool |>
  filter(review_stats_all_count >= 50) |>
  mutate(name = fct_reorder(name, review_stats_all_average_score)) |>
  ggplot(aes(review_stats_all_average_score, name, size = review_stats_all_count)) +
  geom_point() +
  labs(x = "Average rating",
       y = "",
       size = "# of reviews",
       title = "Barstool Sports ratings of pizza places",
       subtitle = "Only places with at least 50 reviews")
```
```{r chunk_08}
pizza_barstool |>
  filter(review_stats_all_count >= 20) |>
  mutate(city = fct_lump(city, 3)) |>
  add_count(city) |>
  mutate(city = glue::glue("{ city } ({ n })")) |>
  ggplot(aes(city, review_stats_all_average_score)) +
  geom_boxplot() +
  labs(title = "Do pizza ratings differ across cities?",
       subtitle = "Only pizza places with at least 20 reviews")
```
```{r chunk_09}
pizza_cleaned <- pizza_barstool |>
  select(place = name,
         price_level,
         contains("review")) |>
  rename_all(~ str_remove(., "review_stats_")) |>
  select(-contains("provider"))

pizza_cleaned |>
  filter(critic_count > 0) |>
  ggplot(aes(critic_average_score, dave_average_score)) +
  geom_point() +
  geom_abline(color = "red") +
  geom_smooth(method = "lm") +
  labs(title = "Does Barstool Sports' Dave agree with the critics?",
       x = "Critic average score",
       y = "Dave score")
```
```{r chunk_10}
pizza_cleaned |>
  filter(community_count >= 20) |>
  ggplot(aes(community_average_score, dave_average_score)) +
  geom_point(aes(size = community_count)) +
  geom_abline(color = "red") +
  geom_smooth(method = "lm") +
  labs(size = "# of community reviews",
       x = "Community score",
       y = "Dave score")
```

---

**→ [Hour 2: Modeling](h2_modeling_pizza_ratings.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
