---
title: "Hour 3 — Communicate: Analyzing pizza ratings in R"
subtitle: "From analysis to TheDataShrink™ publication | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: litera
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 050 · Hour 3 · [← Hour 1 EDA](h1_eda_pizza_ratings.qmd) · [← Hour 2 Modeling](h2_modeling_pizza_ratings.qmd)
:::


## Setup

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(broom)
# Brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 14))
```

```{r reload-data}
pizza_barstool <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2019/2019-10-01/pizza_barstool.csv", show_col_types = FALSE)
pizza_datafiniti <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2019/2019-10-01/pizza_datafiniti.csv", show_col_types = FALSE)
pizza_jared <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2019/2019-10-01/pizza_jared.csv", show_col_types = FALSE)
```

---

## Hero chart (1200 × 628 — social media ratio)

::: {.callout-note}
The single best chart from the analysis. Sized for YouTube thumbnail, Facebook header,
and Twitter/X card. Saved as `hero_pizza_ratings.png`.
:::

```{r hero-chart}
#| fig-width: 12.5
#| fig-height: 6.54

# Hero chart for: Analyzing pizza ratings in R
# Reproduce the best chart from Hour 1 at 1200×628 px
# hero <- ggplot(...) + ...
# hero
#
# ggsave(
#   here::here("output/episodes/pizza_ratings/hero_pizza_ratings.png"),
#   hero, width = 12.5, height = 6.54, dpi = 150, bg = "white"
# )
cat("Hero chart: copy best ggplot from h1 here and uncomment ggsave\n")
```

---

## Instagram carousel (5 slides)

```{r carousel-slide1}
#| fig-width: 5.4
#| fig-height: 5.4

# Slide 1 — Hook
ggplot() +
  annotate("rect", xmin = 0, xmax = 1, ymin = 0, ymax = 1, fill = TDS_DARK) +
  annotate("text", x = 0.5, y = 0.72,
           label = "Analyzing pizza ratings in R",
           size = 8, fontface = "bold", color = "white", hjust = 0.5) +
  annotate("text", x = 0.5, y = 0.50,
           label = "Explore, explain, and model patterns in pizza ratings using engineered features and comparative summaries",
           size = 4.5, color = "#cccccc", hjust = 0.5) +
  annotate("text", x = 0.5, y = 0.28,
           label = "Swipe → to find out",
           size = 4, color = TDS_GOLD, hjust = 0.5) +
  annotate("text", x = 0.5, y = 0.10,
           label = "TheDataShrink™",
           size = 3.5, color = "#888888", hjust = 0.5) +
  theme_void() + coord_fixed()
```

```{r carousel-slide2}
#| fig-width: 5.4
#| fig-height: 5.4
# Slide 2 — Key finding chart (compact version of hero)
# Copy and resize the hero chart here
cat("Slide 2: copy hero chart, adjust size to 5.4×5.4\n")
```

```{r carousel-slide3}
#| fig-width: 5.4
#| fig-height: 5.4
# Slide 3 — Second insight
cat("Slide 3: second key chart from h1\n")
```

```{r carousel-slide4}
#| fig-width: 5.4
#| fig-height: 5.4
# Slide 4 — Third insight or model result from h2
cat("Slide 4: model result or third chart\n")
```

```{r carousel-slide5}
#| fig-width: 5.4
#| fig-height: 5.4
# Slide 5 — CTA
ggplot() +
  annotate("rect", xmin = 0, xmax = 1, ymin = 0, ymax = 1, fill = TDS_DARK) +
  annotate("text", x = 0.5, y = 0.78, label = "Full analysis:",
           size = 5, color = "#aaaaaa", hjust = 0.5) +
  annotate("text", x = 0.5, y = 0.62,
           label = "thedatashrink.com.au\n/tidytuesday/050",
           size = 6, fontface = "bold", color = "white", hjust = 0.5) +
  annotate("text", x = 0.5, y = 0.42,
           label = "Code · Data · 3-hour walkthrough",
           size = 4.5, color = "#cccccc", hjust = 0.5) +
  annotate("text", x = 0.5, y = 0.24,
           label = "Follow @thedatashrink",
           size = 5, color = TDS_GOLD, fontface = "bold", hjust = 0.5) +
  annotate("text", x = 0.5, y = 0.10,
           label = "#RStats #TidyTuesday #DataScience",
           size = 3.5, color = "#888888", hjust = 0.5) +
  theme_void() + coord_fixed()
```

---

## YouTube chapter timestamps

```
0:00   Intro — Analyzing pizza ratings in R
1:30   Load the data
4:00   First look — key distributions
8:00   Hour 1 — EDA: exploring the patterns
18:00  Extension: going beyond David's analysis
28:00  Hour 2 — Modeling: statistical models
38:00  Key findings and interpretation
47:00  Hour 3 — Hero chart + social media pack
52:00  What to try next on your own dataset
55:00  Next week on TheDataShrink™
```

---

## Social media copy

### Facebook / LinkedIn post

```
Analyzing pizza ratings in R — what does the data show?

We explored this week's #TidyTuesday dataset in R.

Here's what the data reveals:

→ [Key finding 1 — fill in after running h1]
→ [Key finding 2]
→ [Key finding 3]
→ [Surprising extension finding]

Full walkthrough (3 hours of R code, charts, and commentary):
👉 thedatashrink.com.au/tidytuesday/050

#DataScience #RStats #TidyTuesday #TheDataShrink
```

### Instagram caption

```
Explore, explain, and model patterns in pizza ratings using engineered features and comparative summaries

Swipe to see how data science reveals the patterns 📊

Full analysis + R code → link in bio

#RStats #TidyTuesday #DataScience #DataViz #TheDataShrink
#DataStorytelling
```

### YouTube description

```
In this episode of TheDataShrink™, we explore:
Analyzing pizza ratings in R

What we cover:
✅ EDA: first look at the data structure and distributions
✅ Key visualisations using ggplot2 and the tidyverse
✅ Extensions beyond David Robinson's original screencast
✅ Statistical modeling and interpretation
✅ Social media pack: hero chart + Instagram carousel

Code and data: github.com/thedatashrink/tidytuesday
Website: thedatashrink.com.au

Chapters: [see timestamps in description]

#RStats #TidyTuesday #DataScience
```

---

## Render all three hours

```{r render-info}
#| echo: false
cat("This episode lives at:\n")
cat("  output/episodes/pizza_ratings/\n\n")
cat("To render all three hours:\n")
cat("  quarto render h1_eda_pizza_ratings.qmd\n")
cat("  quarto render h2_modeling_pizza_ratings.qmd\n")
cat("  quarto render h3_communicate_pizza_ratings.qmd\n")
```

---

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
