---
title: "Hour 1 — EDA: Analyzing IKEA furniture in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 038 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `boxplot_jitter` · `error_bar_coef_plot` · `fct_lump_infrequent` · `fct_relevel_manual` · `fct_reorder_stat` · `geom_smooth_loess` · `glue_strings` · `grouped_summary_ggplot` · `linear_model_broom` · `log_scale` · `ridge_joy_plot`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(glue)
library(ggridges)
library(broom)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
ikea <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-11-03/ikea.csv", show_col_types = FALSE)
glimpse(ikea)
```

## Load the weekly Data
```{r chunk_01}
ikea <- tt$ikea |>
  select(-X1) |>
  mutate(price_usd = 0.27 * price,
         short_description = str_squish(short_description)) |>
  add_count(category, name = "category_total")

ikea |>
  count(category, sort = TRUE) |>
  mutate(category = fct_reorder(category, n)) |>
  ggplot(aes(n, category)) +
  geom_col() +
  labs(x = "# of items",
       y = "",
       title = "Most common categories of IKEA items")
```
```{r chunk_02}
library(glue)

ikea |>
  mutate(category = glue("{ category } ({ category_total })"),
         category = fct_reorder(category, price_usd)) |>
  ggplot(aes(price_usd, category)) +
  geom_boxplot() +
  # geom_jitter(width = 0, height = .1, alpha = .25) +
  scale_x_log10(labels = dollar) +
  labs(x = "Price (USD)",
       y = "",
       title = "How much do items in each category cost?")
```
```{r chunk_03}
library(ggridges)

ikea |>
  mutate(category = glue("{ category } ({ category_total })"),
         category = fct_reorder(category, price_usd)) |>
  ggplot(aes(price_usd, category)) +
  geom_density_ridges() +
  # geom_jitter(width = 0, height = .1, alpha = .25) +
  scale_x_log10(labels = dollar) +
  labs(x = "Price (USD)",
       y = "",
       title = "How much do items in each category cost?")
```
```{r chunk_04}
ikea |>
  mutate(category = glue("{ category } ({ category_total })"),
         category = fct_reorder(category, price_usd)) |>
  ggplot(aes(price_usd, category, fill = other_colors)) +
  geom_density_ridges(alpha = .5) +
  # geom_jitter(width = 0, height = .1, alpha = .25) +
  scale_x_log10(labels = dollar) +
  labs(x = "Price (USD)",
       y = "",
       title = "How much do items in each category cost?")
```
```{r chunk_05}
ikea |>
  mutate(name = fct_lump(name, 20)) |>
  filter(name != "Other") |>
  count(name, category, sort = TRUE) |>
  mutate(name = fct_reorder(name, n, sum),
         category = fct_reorder(category, n, sum)) |>
  ggplot(aes(n, name, fill = category)) +
  geom_col() +
  scale_fill_discrete(guide = guide_legend(reverse = TRUE)) +
  labs(x = "# of items",
       y = "Name of item")
```
```{r chunk_06}
ikea |>
  separate(short_description,
           c("main_description", "rest"),
           sep = ", ",
           extra = "merge",
           fill = "right",
           remove = FALSE) |>
  extract(rest, "description_cm", "([\\d\\-xX]+) cm", remove = FALSE) |>
  unite(category_and_description, category, main_description, sep = " - ") |>
  count(category_and_description, sort = TRUE)
```

## Volume
```{r chunk_07}
ikea_volume <- ikea |>
  mutate(volume_m3 = depth * height * width / 1e6) |>
  filter(!is.na(volume_m3),
         volume_m3 >= .001) |>
  arrange(desc(volume_m3)) |>
  add_count(category, name = "category_total")

ikea_volume |>
  arrange(desc(volume_m3))

ikea_volume |>
  mutate(category = glue("{ category } ({ category_total })"),
         category = fct_reorder(category, volume_m3)) |>
  ggplot(aes(volume_m3, category)) +
  geom_boxplot() +
  # geom_jitter(width = 0, height = .1, alpha = .25) +
  scale_x_log10() +
  labs(x = "Volume of furniture (cubic meters)",
       y = "",
       title = "How much space do items in each category take up?")
```
```{r chunk_08}
ikea_volume |>
  mutate(category = fct_lump(category, 6)) |>
  ggplot(aes(volume_m3, price_usd, color = category)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  scale_y_log10()

ikea_volume |>
  mutate(dollar_per_m3 = price_usd / volume_m3) |>
  arrange(desc(dollar_per_m3)) |>
```
```{r chunk_09}
ikea |>
  group_by(designer) |>
  summarize(n_items = n(),
            n_names = n_distinct(name),
            n_category = n_distinct(category)) |>
  arrange(desc(n_names))
```
```{r chunk_10}
library(broom)

ikea_volume |>
  mutate(category = fct_relevel(category, "Tables & desks")) |>
  lm(log2(price_usd) ~ log2(volume_m3) + category + other_colors, data = .) |>
  tidy(conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(term = ifelse(term == "log2(volume_m3)", "Item volume (doubling)", term),
         term = str_remove(term, "^category")) |>
  mutate(term = fct_reorder(term, estimate)) |>
  ggplot(aes(estimate, term)) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = .1) +
  geom_vline(xintercept = 0, color = "red", lty = 2) +
  labs(x = "Impact on price (relative to Tables & desks)",
       y = "",
       title = "What objects are unusually expensive/inexpensive relative to volume?")
```

---

**→ [Hour 2: Modeling](h2_modeling_ikea_furniture.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
