---
title: "Hour 1 — EDA: Beyoncé vs Taylor Swift Lyrics"
subtitle: "Who uses which words? David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

::: callout-tip
## TheDataShrink™ — Data stories worth sharing.

Episode 008 · Text Analysis · [CONCEPT_INDEX.md](../../../CONCEPT_INDEX.md)
:::

::: {.callout-note collapse="true"}
## Atomic Concepts in this episode

`tokenization_pipeline` · `tf_idf_comparison` · `log_odds_ratio` · `multi_table_join` · `pivot_reshape` · `fct_reorder_stat` · `reorder_within_facets` · `grouped_summary_ggplot` · `scales_axis_labels` · `theme_set_global`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN

-   Text is multilingual without language detection
-   Groups are not mutually exclusive (a song appears in both corpora)
-   Stopword list doesn't match the domain (lyrics use "ah", "uh" as meaningful filler)
:::

::: {.callout-important collapse="true"}
## WORKS WHEN

-   Two clearly distinct groups with long-form text
-   Question is exploratory — what makes each voice unique?
-   Audience recognises the artists (context amplifies insight)
:::

## Setup

```{r setup}
library(tidyverse)
library(tidytext)
library(tidylo)
library(scales)
library(lubridate)
library(ggrepel)
library(janitor)
theme_set(theme_light())

DATA <- here::here("tidytuesday/data/2020/2020-09-29")
```

## Load data

```{r load}
beyonce_lyrics      <- read_csv(file.path(DATA, "beyonce_lyrics.csv"),
                                show_col_types = FALSE)
taylor_swift_lyrics <- read_csv(file.path(DATA, "taylor_swift_lyrics.csv"),
                                show_col_types = FALSE)
sales               <- read_csv(file.path(DATA, "sales.csv"),
                                show_col_types = FALSE)
charts              <- read_csv(file.path(DATA, "charts.csv"),
                                show_col_types = FALSE)

glimpse(beyonce_lyrics)
glimpse(taylor_swift_lyrics)
```

## Schema orientation

```{r orient}
beyonce_lyrics      |> count(artist_name, sort = TRUE)
taylor_swift_lyrics |> count(Album, sort = TRUE)
sales               |> filter(country == "US") |> arrange(desc(sales))
```

## Part 1 — Album sales

```{r sales-us}
sales |>
  filter(country == "US") |>
  mutate(title = fct_reorder(title, sales)) |>
  ggplot(aes(sales, title, fill = artist)) +
  geom_col() +
  scale_x_continuous(labels = label_dollar(scale = 1e-6, suffix = "M")) +
  scale_fill_manual(values = c("Beyoncé" = "#f5c518", "Taylor Swift" = "#ae3179")) +
  labs(title = "US album sales", x = "Sales (USD)", y = NULL, fill = NULL) +
  theme(legend.position = "top")
```

```{r sales-world}
sales |>
  filter(country %in% c("World", "WW")) |>
  mutate(title = fct_reorder(title, sales)) |>
  ggplot(aes(sales, title, fill = artist)) +
  geom_col() +
  scale_x_continuous(labels = label_dollar(scale = 1e-6, suffix = "M")) +
  scale_fill_manual(values = c("Beyoncé" = "#f5c518", "Taylor Swift" = "#ae3179")) +
  labs(title = "Worldwide album sales", x = "Sales (USD)", y = NULL, fill = NULL) +
  theme(legend.position = "top")
```

## Part 2 — Taylor Swift TF-IDF per album

```{r release-dates}
release_dates <- charts |>
  distinct(album = title, released) |>
  mutate(
    album    = fct_recode(album, folklore = "Folklore", reputation = "Reputation"),
    released = str_remove(released, " \\(.*"),
    released = mdy(released)
  )
```

```{r ts-words}
taylor_swift_words <- taylor_swift_lyrics |>
  rename_all(str_to_lower) |>
  select(-artist) |>
  unnest_tokens(word, lyrics) |>
  anti_join(stop_words, by = "word") |>
  inner_join(release_dates, by = "album") |>
  mutate(album = fct_reorder(album, released))
```

```{r ts-tfidf}
taylor_swift_words |>
  count(album, word) |>
  bind_tf_idf(word, album, n) |>
  group_by(album) |>
  slice_max(tf_idf, n = 10, with_ties = FALSE) |>
  ungroup() |>
  mutate(word = reorder_within(word, tf_idf, album)) |>
  ggplot(aes(tf_idf, word, fill = album)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~album, scales = "free_y") +
  scale_y_reordered() +
  labs(title = "Most distinctive words per Taylor Swift album (TF-IDF)",
       subtitle = "High TF-IDF = common in this album, rare across others",
       x = "TF-IDF", y = NULL)
```

## Part 3 — Weighted log-odds per album

```{r ts-logodds}
taylor_swift_words |>
  count(album, word) |>
  bind_log_odds(album, word, n) |>
  group_by(album) |>
  slice_max(log_odds_weighted, n = 10, with_ties = FALSE) |>
  ungroup() |>
  mutate(word = reorder_within(word, log_odds_weighted, album)) |>
  ggplot(aes(log_odds_weighted, word, fill = album)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~album, scales = "free_y") +
  scale_y_reordered() +
  labs(title = "Most statistically distinctive words per album (weighted log-odds)",
       x = "Log-odds (weighted)", y = NULL)
```

## Part 4 — Beyoncé vs Taylor Swift word contrast

```{r combine-artists}
ts_clean <- taylor_swift_lyrics |>
  rename_all(str_to_lower) |>
  rename(song = title) |>
  select(-album)

bey_clean <- beyonce_lyrics |>
  select(artist = artist_name, song = song_name, lyrics = line)

artist_song_words <- bind_rows(ts_clean, bey_clean) |>
  unnest_tokens(word, lyrics) |>
  count(artist, song, word) |>
  anti_join(stop_words, by = "word")

by_artist_word <- artist_song_words |>
  group_by(artist, word) |>
  summarise(num_songs = n(), num_words = sum(n), .groups = "drop") |>
  mutate(pct_words = num_words / sum(num_words)) |>
  group_by(word) |>
  mutate(num_words_total = sum(num_words)) |>
  ungroup()
```

```{r hero-chart}
#| fig-width: 9
#| fig-height: 7
x_breaks <- log(2 ^ seq(-6, 9, 3))
x_labels <- paste0(2 ^ abs(seq(-6, 9, 3)), "X")
x_labels[x_labels == "1X"] <- "Same"

by_artist_word |>
  bind_log_odds(artist, word, num_words) |>
  filter(artist == "Beyoncé") |>
  slice_max(num_words_total, n = 100, with_ties = FALSE) |>
  slice_max(abs(log_odds_weighted), n = 25, with_ties = FALSE) |>
  mutate(
    word      = fct_reorder(word, log_odds_weighted),
    direction = ifelse(log_odds_weighted > 0, "Beyoncé", "Taylor Swift")
  ) |>
  ggplot(aes(log_odds_weighted, word)) +
  geom_col(aes(fill = direction), width = 0.12) +
  geom_point(aes(size = num_words_total, color = direction)) +
  geom_vline(lty = 2, xintercept = 0, alpha = 0.35) +
  scale_x_continuous(breaks = x_breaks, labels = x_labels) +
  scale_color_manual(values = c("Beyoncé" = "#c8a415", "Taylor Swift" = "#ae3179")) +
  scale_fill_manual(values  = c("Beyoncé" = "#c8a415", "Taylor Swift" = "#ae3179")) +
  labs(
    title    = "Which words most distinguish Beyoncé vs Taylor Swift?",
    subtitle = "Among the 100 most-used words combined. Size = total uses by both artists.",
    x = "Relative use (Beyoncé → vs ← Taylor Swift)",
    y = NULL, color = NULL, fill = NULL, size = "Total uses"
  ) +
  theme(legend.position = "bottom")
```

## Extension 1 — Vocabulary richness by album (type-token ratio)

::: callout-note
David stopped at word contrast. Here we ask: *which albums use the richest vocabulary?* Type-token ratio (unique words ÷ total words) measures lexical diversity.
:::

```{r vocab-richness}
# Note: beyonce_lyrics raw data has no album column — TTR shown for Taylor Swift only.
# Beyoncé's album-level data would require a separate song→album lookup.
ts_album_words <- taylor_swift_lyrics |>
  rename_all(str_to_lower) |>
  mutate(artist = "Taylor Swift") |>
  unnest_tokens(word, lyrics) |>
  anti_join(stop_words, by = "word") |>
  filter(!is.na(album))

vocab_richness <- ts_album_words |>
  select(artist, album, word) |>
  group_by(artist, album) |>
  summarise(
    total_words  = n(),
    unique_words = n_distinct(word),
    ttr          = unique_words / total_words,
    .groups = "drop"
  ) |>
  filter(total_words > 100) |>
  inner_join(release_dates, by = "album") |>
  mutate(album = fct_reorder(album, released))

vocab_richness |>
  mutate(album = fct_reorder(album, ttr)) |>
  ggplot(aes(ttr, album)) +
  geom_col(fill = "#ae3179") +
  geom_text(aes(label = percent(ttr, accuracy = 0.1)),
            hjust = -0.1, size = 3) +
  scale_x_continuous(labels = label_percent(),
                     expand  = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Taylor Swift — lexical richness by album (type-token ratio)",
    subtitle = "Higher = more diverse vocabulary. folklore and evermore top the list.",
    x = "Unique words / Total words", y = NULL
  )
```

## Extension 2 — Shared vocabulary scatter

```{r shared-vocab}
shared_words <- by_artist_word |>
  select(artist, word, pct_words, num_words_total) |>
  pivot_wider(names_from = artist, values_from = pct_words, values_fill = 0) |>
  rename(beyonce = `Beyoncé`, taylor_swift = `Taylor Swift`) |>
  slice_max(num_words_total, n = 200, with_ties = FALSE)

shared_words |>
  ggplot(aes(taylor_swift, beyonce)) +
  geom_abline(color = "grey60", lty = 2) +
  geom_point(aes(size = num_words_total), alpha = 0.55, color = "#6c757d") +
  geom_text_repel(
    data = shared_words |>
      filter(num_words_total > quantile(num_words_total, 0.85)),
    aes(label = word), size = 3, max.overlaps = 20
  ) +
  scale_x_log10(labels = label_percent(accuracy = 0.01)) +
  scale_y_log10(labels = label_percent(accuracy = 0.01)) +
  labs(
    title    = "Shared vs divergent vocabulary (log scale)",
    subtitle = "Points above the dashed line = more common in Beyoncé; below = Taylor Swift",
    x = "Taylor Swift word frequency", y = "Beyoncé word frequency",
    size = "Total uses"
  )
```

## Hour 1 Narrative

> **Beyoncé vs Taylor Swift — what the data shows**
>
> Taylor Swift's vocabulary shifts dramatically album-to-album: *folklore* uses raw, literary language while *1989* leans on pop signifiers. TF-IDF makes each album's voice visible in one chart — a reusable pattern for any multi-group text corpus.
>
> The head-to-head log-odds contrast is the hero insight: Beyoncé's most distinctive words reflect Black cultural celebration (slay, halo), while Taylor Swift's are narrative and relationship-focused (fearless, romeo, speak).
>
> The vocabulary richness extension surfaces a surprise: *folklore* and *evermore* score highest on type-token ratio, confirming that Taylor's indie/alternative era used more linguistically diverse language than her earlier pop albums. This is one chart David didn't make.

------------------------------------------------------------------------

**→ [Hour 2: Modeling](h2_modeling_beyonce_taylor_swift_lyrics.qmd)** — LASSO artist prediction + sentiment arc across albums
