---
title: "Hour 3 — Communicate: Beyoncé vs Taylor Swift Lyrics"
subtitle: "From analysis to TheDataShrink™ publication | YouTube · Instagram · Facebook · Web"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: litera
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 008 · Hour 3 · [← Hour 2 Modeling](h2_modeling_beyonce_taylor_swift_lyrics.qmd) · [← Hour 1 EDA](h1_eda_beyonce_taylor_swift_lyrics.qmd)
:::

## Setup

```{r setup}
library(tidyverse)
library(tidytext)
library(tidylo)
library(scales)
library(lubridate)
library(janitor)
library(ggrepel)
library(glmnet)
library(Matrix)
theme_set(theme_light(base_size = 14))

DATA <- here::here("tidytuesday/data/2020/2020-09-29")

# Colour palette — locked for brand consistency
BEY   <- "#c8a415"
TS    <- "#ae3179"
DARK  <- "#1a1a2e"
LIGHT <- "#f8f9fa"
```

```{r reload-data}
beyonce_lyrics      <- read_csv(file.path(DATA, "beyonce_lyrics.csv"),      show_col_types = FALSE)
taylor_swift_lyrics <- read_csv(file.path(DATA, "taylor_swift_lyrics.csv"), show_col_types = FALSE)
charts              <- read_csv(file.path(DATA, "charts.csv"),               show_col_types = FALSE)

release_dates <- charts |>
  distinct(album = title, released) |>
  mutate(
    album    = fct_recode(album, folklore = "Folklore", reputation = "Reputation"),
    released = str_remove(released, " \\(.*"),
    released = mdy(released)
  )

ts_clean  <- taylor_swift_lyrics |> rename_all(str_to_lower) |>
  rename(song = title) |> mutate(artist = "Taylor Swift") |> select(artist, song, lyrics)
bey_clean <- beyonce_lyrics |>
  select(artist = artist_name, song = song_name, lyrics = line)
all_lyrics <- bind_rows(ts_clean, bey_clean)

artist_song_words <- all_lyrics |>
  unnest_tokens(word, lyrics) |>
  count(artist, song, word) |>
  anti_join(stop_words, by = "word")

by_artist_word <- artist_song_words |>
  group_by(artist, word) |>
  summarise(num_songs = n(), num_words = sum(n), .groups = "drop") |>
  mutate(pct_words = num_words / sum(num_words)) |>
  group_by(word) |>
  mutate(num_words_total = sum(num_words)) |>
  ungroup()
```

---

## Hero chart (1200 × 628 — social media ratio)

::: {.callout-note}
The single best chart from the analysis. Sized for YouTube thumbnail, Facebook header,
and Twitter/X card. Saved as `hero_beyonce_taylor_swift.png`.
:::

```{r hero-chart}
#| fig-width: 12.5
#| fig-height: 6.54

x_breaks <- log(2 ^ seq(-6, 9, 3))
x_labels <- paste0(2 ^ abs(seq(-6, 9, 3)), "X")
x_labels[x_labels == "1X"] <- "Same"

word_diffs <- by_artist_word |>
  bind_log_odds(artist, word, num_words) |>
  filter(artist == "Beyoncé") |>
  slice_max(num_words_total, n = 100, with_ties = FALSE) |>
  slice_max(abs(log_odds_weighted), n = 22, with_ties = FALSE) |>
  mutate(
    word      = fct_reorder(word, log_odds_weighted),
    direction = ifelse(log_odds_weighted > 0, "Beyoncé", "Taylor Swift")
  )

hero <- word_diffs |>
  ggplot(aes(log_odds_weighted, word)) +
  geom_col(aes(fill = direction), width = 0.14, alpha = 0.85) +
  geom_point(aes(size = num_words_total, color = direction), alpha = 0.9) +
  geom_vline(lty = 2, xintercept = 0, alpha = 0.3, linewidth = 0.6) +
  scale_x_continuous(breaks = x_breaks, labels = x_labels) +
  scale_color_manual(values = c("Beyoncé" = BEY, "Taylor Swift" = TS)) +
  scale_fill_manual(values  = c("Beyoncé" = BEY, "Taylor Swift" = TS)) +
  scale_size_continuous(range = c(2, 8), labels = label_comma()) +
  labs(
    title    = "Beyoncé vs Taylor Swift: whose words are whose?",
    subtitle = "Weighted log-odds ratio — top 22 most distinctive words among their 100 most-used combined",
    caption  = "TheDataShrink™ · thedatashrink.com.au · Data: TidyTuesday 2020-09-29",
    x = "← More Taylor Swift  |  More Beyoncé →",
    y = NULL, color = NULL, fill = NULL, size = "Total uses"
  ) +
  theme(
    plot.title    = element_text(size = 18, face = "bold", color = DARK),
    plot.subtitle = element_text(size = 12, color = "#555555"),
    plot.caption  = element_text(size = 9,  color = "#888888"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

hero

ggsave(
  here::here("output/episodes/008_beyonce_taylor_swift_lyrics/hero_beyonce_taylor_swift.png"),
  hero, width = 12.5, height = 6.54, dpi = 150, bg = "white"
)
```

---

## Instagram carousel (5 slides)

```{r carousel-slide1}
#| fig-width: 5.4
#| fig-height: 5.4

# Slide 1 — Hook (square format)
ggplot() +
  annotate("rect", xmin = 0, xmax = 1, ymin = 0, ymax = 1,
           fill = DARK) +
  annotate("text", x = 0.5, y = 0.72,
           label = "Beyoncé vs\nTaylor Swift",
           size = 11, fontface = "bold", color = "white", hjust = 0.5) +
  annotate("text", x = 0.5, y = 0.48,
           label = "What do 200,000 lyrics\ntell us about their voices?",
           size = 5.5, color = "#cccccc", hjust = 0.5) +
  annotate("text", x = 0.5, y = 0.22,
           label = "Swipe → to find out",
           size = 4.5, color = BEY, hjust = 0.5) +
  annotate("text", x = 0.5, y = 0.08,
           label = "TheDataShrink™",
           size = 3.5, color = "#888888", hjust = 0.5) +
  theme_void() +
  coord_fixed()
```

```{r carousel-slide2}
#| fig-width: 5.4
#| fig-height: 5.4

# Slide 2 — Key finding: word contrast lollipop (compact)
word_diffs |>
  slice_max(abs(log_odds_weighted), n = 14) |>
  mutate(word = fct_reorder(word, log_odds_weighted)) |>
  ggplot(aes(log_odds_weighted, word)) +
  geom_segment(aes(xend = 0, yend = word, color = direction),
               linewidth = 1.2, alpha = 0.7) +
  geom_point(aes(color = direction, size = num_words_total)) +
  geom_vline(xintercept = 0, lty = 2, alpha = 0.3) +
  scale_color_manual(values = c("Beyoncé" = BEY, "Taylor Swift" = TS)) +
  scale_size_continuous(range = c(3, 7), guide = "none") +
  labs(title = "Their most distinctive words",
       subtitle = "Weighted log-odds ratio",
       caption = "TheDataShrink™",
       x = NULL, y = NULL, color = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(face = "bold"),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank())
```

```{r carousel-slide3}
#| fig-width: 5.4
#| fig-height: 5.4

# Slide 3 — TF-IDF Taylor Swift top 3 albums (compact)
taylor_swift_lyrics |>
  rename_all(str_to_lower) |>
  select(-artist) |>
  unnest_tokens(word, lyrics) |>
  anti_join(stop_words, by = "word") |>
  inner_join(release_dates, by = "album") |>
  filter(album %in% c("folklore", "1989", "reputation")) |>
  count(album, word) |>
  bind_tf_idf(word, album, n) |>
  group_by(album) |>
  slice_max(tf_idf, n = 8, with_ties = FALSE) |>
  ungroup() |>
  mutate(word = reorder_within(word, tf_idf, album)) |>
  ggplot(aes(tf_idf, word, fill = album)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~album, scales = "free_y", ncol = 1) +
  scale_y_reordered() +
  scale_fill_manual(values = c("folklore" = "#2d4a3e", "1989" = TS, "reputation" = "#333333")) +
  labs(title = "Each album has its own voice",
       subtitle = "TF-IDF: words unique to each era",
       caption = "TheDataShrink™",
       x = "TF-IDF", y = NULL) +
  theme(plot.title = element_text(face = "bold"))
```

```{r carousel-slide4}
#| fig-width: 5.4
#| fig-height: 5.4

# Slide 4 — Sentiment arc
afinn <- get_sentiments("afinn")

taylor_swift_lyrics |>
  rename_all(str_to_lower) |>
  unnest_tokens(word, lyrics) |>
  inner_join(afinn, by = "word") |>
  inner_join(release_dates, by = "album") |>
  group_by(album, released) |>
  summarise(mean_sentiment = mean(value), .groups = "drop") |>
  arrange(released) |>
  ggplot(aes(released, mean_sentiment)) +
  geom_hline(yintercept = 0, lty = 2, alpha = 0.3) +
  geom_area(fill = TS, alpha = 0.2) +
  geom_line(color = TS, linewidth = 1.2) +
  geom_point(color = TS, size = 3) +
  geom_text_repel(aes(label = album), size = 3.2, color = DARK,
                  max.overlaps = 10) +
  labs(title = "Taylor Swift's emotional arc",
       subtitle = "AFINN mean sentiment by album release",
       caption = "TheDataShrink™",
       x = NULL, y = "Mean sentiment") +
  theme(plot.title = element_text(face = "bold"))
```

```{r carousel-slide5}
#| fig-width: 5.4
#| fig-height: 5.4

# Slide 5 — CTA
ggplot() +
  annotate("rect", xmin = 0, xmax = 1, ymin = 0, ymax = 1,
           fill = DARK) +
  annotate("text", x = 0.5, y = 0.78,
           label = "Full analysis:",
           size = 5, color = "#aaaaaa", hjust = 0.5) +
  annotate("text", x = 0.5, y = 0.62,
           label = "thedatashrink.com.au\n/tidytuesday/008",
           size = 6, fontface = "bold", color = "white", hjust = 0.5) +
  annotate("text", x = 0.5, y = 0.42,
           label = "Code · Data · 3-hour walkthrough",
           size = 4.5, color = "#cccccc", hjust = 0.5) +
  annotate("text", x = 0.5, y = 0.24,
           label = "Follow @thedatashrink",
           size = 5, color = BEY, fontface = "bold", hjust = 0.5) +
  annotate("text", x = 0.5, y = 0.10,
           label = "#RStats #TidyTuesday #DataScience",
           size = 3.5, color = "#888888", hjust = 0.5) +
  theme_void() +
  coord_fixed()
```

---

## YouTube chapter timestamps

```
0:00   Intro — two of the biggest artists, one dataset
1:30   Load the data — lyrics, sales, chart positions
4:00   Album sales: who sold more and where?
7:00   Hour 1 — TF-IDF: what makes each Taylor Swift album sound unique?
14:00  Log-odds: the Beyoncé vs Taylor Swift vocabulary contrast
22:00  Extension: which albums have the richest vocabulary?
28:00  Hour 2 — LASSO: can a model tell them apart from lyrics alone?
36:00  Sentiment arc: how has Taylor's emotional tone changed?
42:00  Beyoncé sentiment by album — Lemonade vs Renaissance
47:00  Hour 3 — Hero chart + social media pack
52:00  What to try next on your own dataset
55:00  Next week on TheDataShrink™
```

---

## Social media copy

### Facebook / LinkedIn post

```
📊 Beyoncé vs Taylor Swift — what 200,000 lyrics reveal

We analysed every word both artists have ever recorded.

Here's what the data shows:

→ Word choice alone predicts the artist with 80%+ accuracy (LASSO)
→ Taylor Swift's vocabulary gets richer with every album — folklore is her most linguistically diverse
→ Beyoncé's most distinctive word? "slay" — by a wide margin
→ Taylor's emotional arc dips hard with reputation, then climbs back with folklore/evermore

Full walkthrough (3 hours of R code, charts, and commentary):
👉 thedatashrink.com.au/tidytuesday/008

#DataScience #RStats #TidyTuesday #TheDataShrink #BeyonceVsTaylorSwift
```

### Instagram caption

```
Two voices. 200,000 words. One dataset.

Swipe to see how data science reveals what makes each artist's
voice uniquely theirs 🎤📊

Full analysis + R code → link in bio

#RStats #TidyTuesday #DataScience #DataViz #TheDataShrink
#BeyonceVsTaylor #DataStorytelling
```

### YouTube description

```
In this episode of TheDataShrink™, we use R and the tidyverse to analyse
the complete lyrics of Beyoncé and Taylor Swift — every song, every word.

What we cover:
✅ TF-IDF to find each Taylor Swift album's unique vocabulary
✅ Weighted log-odds to contrast the two artists head-to-head
✅ LASSO regression: can a model identify the artist from lyrics alone?
✅ Sentiment arcs across careers
✅ Vocabulary richness (type-token ratio) by album

Code and data: github.com/thedatashrink/tidytuesday
Website: thedatashrink.com.au

Chapters: [see timestamps in description]

#RStats #TidyTuesday #DataScience #BeyonceVsTaylorSwift
```

---

## Render all three hours

```{r render-info}
#| echo: false
cat("This episode lives at:\n")
cat("  output/episodes/008_beyonce_taylor_swift_lyrics/\n\n")
cat("To render all three hours:\n")
cat("  quarto render h1_eda_beyonce_taylor_swift_lyrics.qmd\n")
cat("  quarto render h2_modeling_beyonce_taylor_swift_lyrics.qmd\n")
cat("  quarto render h3_communicate_beyonce_taylor_swift_lyrics.qmd\n")
```

---

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
