---
title: "Hour 1 — EDA: Analyzing NYC restaurant inspections with R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 047 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `boxplot_jitter` · `coord_flip_horizontal` · `error_bar_coef_plot` · `fct_reorder_stat` · `grouped_summary_ggplot` · `janitor_clean` · `linear_model_broom` · `log_scale` · `lubridate_dates` · `pairwise_token_correlation` · `purrr_map_iterate`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(lubridate)
library(broom)
library(widyr)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
library(lubridate)

theme_set(theme_light())

# You can use this url to download the data directly into R (will take a few seconds)
restaurant_inspections_raw <- read_csv("https://data.cityofnewyork.us/api/views/43nn-pn8j/rows.csv")

restaurant_inspections <- restaurant_inspections_raw |> 
  janitor::clean_names() |>
  select(-phone, -grade_date, -record_date, -building, -street) |>
  mutate(inspection_date = mdy(inspection_date)) |>
  separate(inspection_type, c("inspection_program", "inspection_type"), sep = " / ")
```
```{r chunk_02}
restaurant_inspections |>
  count(dba, camis, sort = TRUE)

restaurant_inspections |>
  count(year = year(inspection_date))

restaurant_inspections |>
  count(grade, sort = TRUE)

restaurant_inspections |>
  count(violation_code, violation_description, sort = TRUE)

restaurant_inspections |>
  filter(camis == 41297769, inspection_date == "2018-09-25") |>
  count(camis, dba, inspection_date, sort = TRUE)

restaurant_inspections |>
  count(cuisine_description, sort = TRUE)

restaurant_inspections |>
  filter(action == "No violations were recorded at the time of this inspection.") |>
  count(critical_flag)

inspections <- restaurant_inspections |>
  group_by(camis,
           dba,
           boro,
           zipcode,
           cuisine_description,
           inspection_date,
           action,
           score,
           grade,
           inspection_type,
           inspection_program) |>
  summarize(critical_violations = sum(critical_flag == "Critical", na.rm = TRUE),
            non_critical_violations = sum(critical_flag == "Not Critical", na.rm = TRUE)) |>
  ungroup()

most_recent_cycle_inspection <- inspections |>
  filter(inspection_program == "Cycle Inspection",
         inspection_type == "Initial Inspection") |>
  arrange(desc(inspection_date)) |>
  distinct(camis, .keep_all = TRUE)
```
```{r chunk_03}
by_dba <- most_recent_cycle_inspection |>
  group_by(dba, cuisine = cuisine_description) |>
  summarize(locations = n(),
            avg_score = mean(score),
            median_score = median(score)) |>
  ungroup() |>
  arrange(desc(locations))

by_dba |>
  mutate(locations_bin = cut(locations, c(0, 1, 3, 10, Inf), labels = c("1", "2-3", "3-10", ">10"))) |>
  ggplot(aes(locations_bin, avg_score + 1)) +
  geom_boxplot() +
  scale_y_log10()
```
```{r chunk_04}
by_cuisine <- by_dba |>
  group_by(cuisine) |>
  summarize(avg_score = mean(avg_score),
            median_score = median(avg_score),
            restaurants = n()) |>
  arrange(desc(restaurants))

library(broom)

cuisine_conf_ints <- by_dba |>
  add_count(cuisine) |>
  filter(n > 100) |>
  nest(-cuisine) |>
  mutate(model = map(data, ~ t.test(.$avg_score))) |>
  unnest(map(model, tidy))

cuisine_conf_ints |>
  mutate(cuisine = str_remove(cuisine, " \\(.*"),
         cuisine = fct_reorder(cuisine, estimate)) |>
  ggplot(aes(estimate, cuisine)) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf.low,
                     xmax = conf.high)) +
  labs(x = "Average inspection score (higher means more violations)",
       y = "Type of cuisine",
       title = "Average inspection score by type of cuisine in NYC",
       subtitle = "Each restaurant chain was counted once based on its average score")
```
```{r chunk_05}
violation_cuisine_counts <- restaurant_inspections |>
  semi_join(most_recent_cycle_inspection, by = c("camis", "inspection_date")) |>
  count(critical_flag, violation_code, violation_description, cuisine = cuisine_description) |>
  group_by(violation_code) |>
  mutate(violation_total = sum(n)) |>
  group_by(cuisine) |>
  mutate(cuisine_total = sum(n)) |>
  ungroup() |>
  filter(violation_total >= 1000,
         cuisine_total >= 2000) |>
  group_by(violation_description) |>
  mutate(fraction = n / cuisine_total,
         avg_fraction = mean(fraction)) |>
  ungroup()

violation_cuisine_counts |>
  mutate(relative_risk = fraction / avg_fraction) |>
  arrange(desc(relative_risk)) |>
  filter(str_detect(violation_description, "mice"))
```

## What violations tend to occur together?
```{r chunk_06}
library(widyr)

violations <- restaurant_inspections |>
  semi_join(most_recent_cycle_inspection, by = c("camis", "inspection_date")) |>
  filter(!is.na(violation_description))

violations |>
  pairwise_cor(violation_description, camis, sort = TRUE)

principal_components <- violations |>
  mutate(value = 1) |>
  widely_svd(violation_description, camis, value, nv = 6)

principal_components |>
  filter(dimension == 2) |>
  top_n(10, abs(value)) |>
  mutate(violation_description = str_sub(violation_description, 1, 60),
         violation_description = fct_reorder(violation_description, value)) |>
  ggplot(aes(violation_description, value)) +
  geom_col() +
  coord_flip()
```

---

**→ [Hour 2: Modeling](h2_modeling_nyc_restaurant_inspections.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
