---
title: "Hour 1 — EDA: Analyzing exploring US beer production in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 027 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`animated_transitions` · `choropleth_sf` · `coord_flip_horizontal` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `grouped_summary_ggplot` · `lubridate_dates` · `multi_table_join` · `parse_number_dirty` · `purrr_map_iterate` · `rolling_window_metrics`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- region name spelling differs between data and shapefile — silent dropped rows
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- geographic column can be matched to a standard ISO/FIPS shapefile without fuzzy joins
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(lubridate)
library(tidymetrics)
library(shinymetrics)
library(maps)
library(sf)
library(gganimate)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
library(lubridate)
theme_set(theme_light())

brewing_materials <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-31/brewing_materials.csv') |>
  mutate(date = ymd(paste(year, month, 1))) |>
  filter(year < 2016)

beer_taxed <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-31/beer_taxed.csv')
brewer_size <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-31/brewer_size.csv')
beer_states <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-31/beer_states.csv')
```
```{r chunk_02}
brewing_materials |>
  filter(date == max(date)) |>
  filter(!str_detect(material_type, "Total")) |>
  mutate(type = fct_reorder(type, month_current)) |>
  ggplot(aes(type, month_current, fill = material_type)) +
  geom_col() +
  coord_flip()

library(lubridate)

brewing_materials |>
  filter(!str_detect(material_type, "Total")) |>
  mutate(type = fct_reorder(type, month_current, sum)) |>
  ggplot(aes(date, month_current, fill = type)) +
  geom_col() +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Time",
       y = "Pounds used in beer production",
       fill = "Material")

brewing_materials |>
  filter(str_detect(material_type, "Total.*products")) |>
  mutate(type = fct_reorder(type, month_current, sum)) |>
  ggplot(aes(date, month_current, fill = type)) +
  geom_col() +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Time",
       y = "Pounds used in beer production",
       fill = "Material")
```
```{r chunk_03}
# devtools::install_github("ramnathv/tidymetrics")
library(tidymetrics)

brewing_summarized <- brewing_materials |>
  rename(material = type) |>
  filter(!str_detect(material_type, "Total")) |>
  cross_by_dimensions(material, material_type) |>
  cross_by_periods(c("month", "quarter", "year")) |>
  summarize(nb_pounds = sum(month_current)) |>
  ungroup()

brewing_summarized |>
  filter(material_type == "All", material != "All", period == "quarter") |>
  ggplot(aes(date, nb_pounds, fill = material)) +
  geom_col()
```
```{r chunk_04}
brewing_metrics <- create_metrics(brewing_summarized)

library(shinymetrics)

preview_metric(brewing_metrics$beer_production_nb_pounds)
```

## Brewer size distribution
```{r chunk_05}
brewer_size |>
  filter(brewer_size != "Total", !is.na(total_barrels)) |>
  mutate(brewer_size = fct_lump(brewer_size, 5, w = total_barrels),
         barrel_number = coalesce(parse_number(as.character(brewer_size)), 1),
         brewer_size = fct_reorder(brewer_size, barrel_number)) |>
  ggplot(aes(year, total_barrels, fill = brewer_size)) +
  geom_col()
```
```{r chunk_06}
brewing_materials |>
  filter(material_type == "Total Used") |>
  ggplot(aes(month, month_current, color = factor(year))) +
  geom_line() +
  expand_limits(y = 0)
```
```{r chunk_07}
beer_states |>
  count(type, sort = TRUE, wt = barrels)

beer_states |>
  group_by(year) |>
  summarize(barrels = sum(barrels, na.rm = TRUE))

# Who consumes beer on premises
state_percents_2019 <- beer_states |>
  filter(year == max(year), state != "total") |>
  group_by(state, year) |>
  mutate(percent = barrels / sum(barrels)) |>
  ungroup()

state_percents_2019 |>
  filter(type == "On Premises") |>
  arrange(desc(percent))
```
```{r chunk_08}
library(maps)
library(sf)

states <- st_as_sf(map("state", plot = FALSE, fill = TRUE))

states_joined <- state_percents_2019 |>
  mutate(ID = str_to_lower(state.name[match(state, state.abb)])) |>
  inner_join(states, by = "ID")

states_joined |>
  filter(type == "On Premises") |>
  ggplot(aes(geometry = geom, fill = percent)) +
  geom_sf() +
  scale_fill_gradient2(low = "blue", high = "orange", midpoint = .5,
                       labels = scales::percent) +
  ggthemes::theme_map() +
  labs(title = "In which states is a lot of beer produced consumed on premises?",
       fill = "% consumed on prem.")

states_joined |>
  ggplot(aes(geometry = geom, fill = percent)) +
  geom_sf() +
  facet_wrap(~ type, nrow = 2) +
  scale_fill_gradient2(low = "blue", high = "orange", midpoint = .5,
                       labels = scales::percent) +
  ggthemes::theme_map() +
  theme(legend.position = "right") +
  labs(title = "How is beer consumed within each state?",
       fill = "% consumed")
```
```{r chunk_09}
library(gganimate)

beer_states |>
  filter(state != "total") |>
  group_by(state, year) |>
  mutate(percent = barrels / sum(barrels)) |>
  ungroup() |>
  filter(type == "On Premises") |>
  mutate(ID = str_to_lower(state.name[match(state, state.abb)])) |>
  inner_join(states, by = "ID") |>
  ggplot(aes(geometry = geom, fill = percent)) +
  geom_sf() +
  transition_time(year) +
  facet_wrap(~ type, nrow = 2) +
  scale_fill_gradient2(low = "blue", high = "orange", midpoint = .5,
                       labels = scales::percent) +
  ggthemes::theme_map() +
  theme(legend.position = "right") +
  labs(title = "How is beer consumed within each state? ({ as.integer(frame_time) })",
       fill = "% consumed")
```

---

**→ [Hour 2: Modeling](h2_modeling_exploring_beer_production.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
