---
title: "Hour 1 — EDA: Analyzing Ninja Warrior in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 045 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `boxplot_jitter` · `faceted_comparison` · `fct_lump_infrequent` · `fct_reorder_stat` · `geom_histogram_density` · `glue_strings` · `grouped_summary_ggplot` · `heatmap_tile` · `log_odds_ratio` · `purrr_map_iterate` · `reorder_within_facets`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- text is multilingual without language detection
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- text is long-form English and benefits from word-level decomposition
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(tidylo)
library(tidytext)
library(glue)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
ninja_warrior <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2020/2020-12-15/ninja_warrior.csv", show_col_types = FALSE)
glimpse(ninja_warrior)
```

## Load the weekly Data
```{r chunk_01}
ninja_warrior |>
  count(location, sort = TRUE)

ninja_warrior |>
  count(round_stage, obstacle_name, sort = TRUE)

ninja_warrior |>
  filter(location == "Venice") |>

ninja_warrior |>
  count(location, round_stage, sort = T) |>
  view()

ninja_warrior |>
  filter(!str_detect(round_stage, "National")) |>
  count(season, location, round_stage) |>
  ggplot(aes(n, fill = round_stage)) +
  geom_histogram() +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "# of obstacles")

ninja_warrior |>
  count(round_stage, sort = TRUE)
```
```{r chunk_02}
library(tidylo)

ninja_warrior |>
  filter(round_stage %in% c("Qualifying", "Finals")) |>
  count(round_stage, obstacle_name, sort = TRUE) |>
  bind_log_odds(round_stage, obstacle_name, n) |>
  arrange(desc(log_odds_weighted)) |>
  filter(round_stage == "Finals") |>
  top_n(16, abs(log_odds_weighted)) |>
  mutate(obstacle_name = fct_reorder(obstacle_name, log_odds_weighted)) |>
  ggplot(aes(log_odds_weighted, obstacle_name)) +
  geom_col() +
  labs(x = "More / less likely in finals")
```
```{r chunk_03}
total_rounds <- ninja_warrior |>
  filter(round_stage == "Qualifying") |>
  distinct(season, location) |>
  nrow()
  
library(tidytext)

ninja_warrior |>
  filter(round_stage %in% c("Qualifying", "Finals")) |>
  unite(season_location, season, location, remove = FALSE) |>
  group_by(round_stage) |>
  mutate(total_rounds = n_distinct(season_location)) |>
  group_by(round_stage, obstacle_name) |>
  summarize(avg_position = mean(obstacle_order),
            n_rounds = n(),
            pct_rounds = n_rounds / first(total_rounds)) |>
  arrange(desc(n_rounds)) |>
  top_n(10, n_rounds) |>
  ungroup() |>
  mutate(obstacle_name = reorder_within(obstacle_name, avg_position, round_stage)) |>
  ggplot(aes(avg_position, obstacle_name, size = pct_rounds)) +
  geom_point() +
  facet_wrap(~ round_stage, nrow = 2, scales = "free_y") +
  scale_x_continuous(breaks = 1:10) +
  scale_y_reordered() +
  scale_size_continuous(labels = percent) +
  labs(x = "Average position within the obstacle course",
       y = "",
       size = "% of courses")
```
```{r chunk_04}
visualize_steps <- function(tbl) {
  tbl |>
    add_count(obstacle_order, round_stage, name = "round_stage_total") |>
    filter(round_stage_total >= 10) |>
    mutate(obstacle_name = fct_lump(obstacle_name, 10)) |>
    mutate(obstacle_name = fct_reorder(obstacle_name, obstacle_order)) |>
    count(round_stage_total, obstacle_name, obstacle_order) |>
    ggplot(aes(obstacle_order, n / round_stage_total, fill = obstacle_name)) +
    geom_col(width = 1) +
    scale_x_continuous(breaks = 1:10) +
    scale_y_continuous(labels = percent) +
    labs(x = "Step",
         y = "% of courses",
         fill = "Obstacle")
}

ninja_warrior |>
  filter(round_stage == "Qualifying") |>
  visualize_steps() +
  labs(title = "What does a typical Qualifying course look like?")

ninja_warrior |>
  filter(round_stage == "Finals") |>
  visualize_steps() +
  labs(title = "What does a typical Finals course look like?")

ninja_warrior |>
  filter(round_stage == "Finals") |>
  visualize_steps() +
  labs(title = "What does a typical Finals course look like?") +
  facet_wrap(~ obstacle_name) +
  theme(legend.position = "none")
```
```{r chunk_05}
library(glue)

ninja_warrior |>
  filter(round_stage == "Qualifying") |>
  add_count(obstacle_order, round_stage, name = "round_stage_total") |>
  filter(round_stage_total >= 10) |>
  add_count(obstacle_name, name = "obstacle_total") |>
  mutate(obstacle_name = glue("{ obstacle_name } ({ obstacle_total })")) |>
  mutate(obstacle_name = fct_lump(obstacle_name, 10)) |>
  mutate(obstacle_name = fct_reorder(obstacle_name, obstacle_order)) |>
  ggplot(aes(obstacle_order, obstacle_name)) +
  geom_boxplot() +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "Step",
       y = "% of courses",
       fill = "Obstacle")
```
```{r chunk_06}
ninja_warrior |>
  filter(round_stage == "Qualifying",
         obstacle_order <= 6) |>
  mutate(lumped = fct_lump(obstacle_name, 12),
         lumped = fct_reorder(lumped, obstacle_order + season * .01)) |>
  unite(season_location, season, location, sep = " - ", remove = FALSE) |>
  mutate(season_location = fct_rev(fct_reorder(season_location, season))) |>
  ggplot(aes(obstacle_order, season_location, fill = lumped)) +
  geom_tile() +
  geom_text(aes(label = obstacle_name), size = 3) +
  theme(legend.position = "none") +
  scale_x_continuous(breaks = 1:6) +
  labs(x = "Step",
       y = "")

ninja_warrior |>
  filter(round_stage == "Finals",
         obstacle_order <= 10) |>
  mutate(obstacle_name = str_trunc(obstacle_name, 15),
         lumped = fct_lump(obstacle_name, 12),
         lumped = fct_reorder(lumped, obstacle_order + season * .01)) |>
  unite(season_location, season, location, sep = " - ", remove = FALSE) |>
  mutate(season_location = fct_rev(fct_reorder(season_location, season))) |>
  ggplot(aes(obstacle_order, season_location, fill = lumped)) +
  geom_tile() +
  geom_text(aes(label = obstacle_name), size = 3) +
  theme(legend.position = "none") +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "Step",
       y = "")
```
```{r chunk_07}
ninja_warrior |>
  filter(round_stage == "Qualifying") |>
  mutate(obstacle_name = fct_lump(obstacle_name, 15)) |>
  mutate(obstacle_name = fct_reorder(obstacle_name, season)) |>
  ggplot(aes(season, obstacle_name)) +
  geom_boxplot() +
  scale_x_continuous(breaks = 1:10)

ninja_warrior |>
  filter(round_stage == "Qualifying") |>
  mutate(obstacle_name = fct_lump(obstacle_name, 8)) |>
  mutate(obstacle_name = fct_rev(fct_reorder(obstacle_name, season))) |>
  count(obstacle_name, season) |>
  group_by(season) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(season, pct, fill = obstacle_name)) +
  geom_col() +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(labels = percent) +
  labs(y = "% of obstacles")
```
```{r chunk_08}
ninja_warrior |>
  filter()
```

## Readme
```{r Readme}
tt
```

## Glimpse Data
```{r Glimpse}
tt |> 
  map(glimpse)
```

## Save Image
```{r chunk_09}
# This will save your most recent plot
ggsave(
  filename = "My TidyTuesday Plot.png",
  device = "png")
```

---

**→ [Hour 2: Modeling](h2_modeling_ninja_warrior.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
