---
title: "Hour 1 — EDA: Analyzing tennis tournaments in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 059 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `boxplot_jitter` · `case_when_conditional_col` · `coord_flip_horizontal` · `cumsum_running_total` · `fct_relevel_manual` · `fct_reorder_stat` · `geom_histogram_density` · `grouped_summary_ggplot` · `lubridate_dates` · `multi_table_join` · `pivot_reshape`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(lubridate)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
library(lubridate)
theme_set(theme_light())

player_dob <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-04-09/player_dob.csv")

# Removing some players in 1977 who were duplicated
grand_slams <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-04-09/grand_slams.csv") |>
  arrange(year, grand_slam, name, gender) |>
  distinct(year, grand_slam, name, .keep_all = TRUE) |>
  mutate(grand_slam = str_replace(str_to_title(str_replace(grand_slam, "_", " ")), "Us", "US"))

grand_slam_timeline <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-04-09/grand_slam_timeline.csv")
```
```{r chunk_02}
grand_slam_timeline |>
  count(year, tournament, gender) |>
  arrange(year) |>

grand_slam_timeline |>
  filter(year == 2018, tournament == "US Open") |>
  count(outcome, sort = TRUE)
```

## Exploration of the winners
```{r chunk_03}
grand_slams |>
  count(name, grand_slam, sort = TRUE) |>
  add_count(name, wt = n) |>
  filter(nn >= 8) |>
  mutate(name = fct_reorder(name, n, sum)) |>
  ggplot(aes(name, n, fill = grand_slam)) +
  geom_col() +
  coord_flip() +
  labs(x = "",
       y = "# of Grand Slam tournaments won",
       title = "Tennis players with the most Grand Slam tournament wins",
       subtitle = "1968-Present",
       fill = "Grand Slam")
```
```{r chunk_04}
grand_slams_age <- player_dob |>
  select(name, date_of_birth) |>
  inner_join(grand_slams, by = "name") |>
  mutate(age = as.numeric(difftime(tournament_date, date_of_birth, unit = "days")) / 365.25)

grand_slams_age |>
  ggplot(aes(age)) +
  geom_histogram()
```
```{r chunk_05}
grand_slams_age |>
  ggplot(aes(age, fill = gender)) +
  geom_histogram(position = "identity", alpha = .75)

grand_slams_age |>
  mutate(decade = 10 * (year(tournament_date) %/% 10)) |>
  ggplot(aes(decade, age, fill = gender, group = interaction(gender, decade))) +
  geom_boxplot()
```
```{r age_over_time}
by_decade_gender <- grand_slams_age |>
  mutate(decade = 10 * (year(tournament_date) %/% 10)) |>
  filter(decade >= 1970) |>
  group_by(decade, gender, name) |>
  summarize(age = mean(age),
            wins = n()) |>
  summarize(age = mean(age),
            players = n())
```
```{r chunk_06}
by_decade_gender |>
  ggplot(aes(decade, age, color = gender)) +
  geom_line() +
  labs(title = "Average age of Grand Slam winners over time",
       subtitle = "Each player was counted only once per decade",
       x = "Decade",
       y = "Average age",
       color = "")
```

## Predicting the winner of a Grand Slam tournament
```{r chunk_07}
dob <- player_dob |>
  select(player = name, date_of_birth)

tournaments <- grand_slams |>
  select(year, tournament = grand_slam, gender, tournament_date)

timeline_processed <- grand_slam_timeline |>
  inner_join(tournaments, by = c("year", "tournament", "gender")) |>
  arrange(player, tournament_date) |>
  filter(outcome != "Absent",
         !str_detect(outcome, "Qualif")) |>
  group_by(player) |>
  mutate(rolling_play_count = row_number() - 1,
         rolling_won_count = lag(cumsum(outcome == "Won"), default = 0),
         rolling_finals_count = lag(cumsum(outcome %in% c("Won", "Finalist")), default = 0)) |>
  ungroup() |>
  filter(!(year == 1977 & tournament == "Australian Open")) |>
  mutate(won = outcome == "Won")

timeline_processed |>
  filter(outcome %in% c("Finalist", "Won")) |>
  arrange(tournament_date) |>
  group_by(rolling_won_count = pmin(rolling_won_count, 10)) |>
  summarize(pct_won = mean(won),
            observations = n()) |>
  ggplot(aes(rolling_won_count, pct_won)) +
  geom_line() +
  expand_limits(y = 0)

timeline_processed |>
  filter(outcome %in% c("Finalist", "Won")) |>
  select(year, tournament, gender, outcome, rolling_finals_count) |>
  spread(outcome, rolling_finals_count) |>
  count(result = case_when(
    Won > Finalist ~ "Won > Finalist",
    Won == Finalist ~ "Won == Finalist",
    TRUE ~ "Won < Finalist"
  )) |>
  mutate(n / sum(n))
```
```{r chunk_08}
outcome_rankings <- c("1st Round", "2nd Round", "3rd Round", "4th Round",
                      "Quarterfinalist", "Semi-finalist", "Finalist", "Won")

tournament_scores <- timeline_processed |>
  filter(outcome %in% outcome_rankings) |>
  mutate(score_contribution = match(outcome, outcome_rankings)) |>
  group_by(player) |>
  mutate(previous_average = lag(cummean(score_contribution), default = 1)) |>
  ungroup() |>
  mutate(previous_performance = outcome_rankings[round(previous_average)],
         previous_performance = fct_relevel(previous_performance, outcome_rankings))

tournament_scores |>
  group_by(previous_performance) |>
  summarize(observations = n(),
            probability_win = mean(won)) |>
  ggplot(aes(previous_performance, probability_win, group = 1)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Player's average previous performance",
       y = "Probability of winning tournament",
       title = "Does past performance in Grand Slams predict future success?",
       subtitle = "Treating rounds as if they can be averaged linearly")

tournament_scores |>
  group_by(outcome) |>
  summarize(avg_score = mean(running_score)) |>
  arrange(avg_score)
  select(year, tournament, gender, outcome, rolling_finals_count) |>
  spread(outcome, rolling_finals_count) |>
  count(result = case_when(
    Won > Finalist ~ "Won > Finalist",
    Won == Finalist ~ "Won == Finalist",
    TRUE ~ "Won < Finalist"
  )) |>
  mutate(n / sum(n))
```

---

**→ [Hour 2: Modeling](h2_modeling_tennis_tournaments.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
