---
title: "Hour 2 — Modeling: Analyzing tennis tournaments in R"
subtitle: "Going deeper with statistical models | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 059 · Hour 2 · [← Hour 1 EDA](h1_eda_tennis_tournaments.qmd)
:::


::: {.callout-note collapse="true"}
## Atomic Concepts — Hour 2
`add_count_inline` · `boxplot_jitter` · `case_when_conditional_col` · `coord_flip_horizontal` · `cumsum_running_total` · `fct_relevel_manual` · `fct_reorder_stat` · `geom_histogram_density` · `grouped_summary_ggplot` · `lubridate_dates`
:::

## Setup

```{r setup}
library(tidyverse)
library(scales)
library(broom)
library(here)
theme_set(theme_light(base_size = 13))
```

## Reload data

```{r load-data}
# Re-load from Hour 1 — same local paths
source_qmd <- here::here("output/episodes/tennis_tournaments/h1_eda_tennis_tournaments.qmd")
# Data files are re-read directly below if needed
```

## Model 1 — Time trend

```{r time-setup}
library(lubridate)
df <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2019/2019-04-09/grand_slam_timeline.csv", show_col_types = FALSE)
```

```{r trend}
# df |>
#   mutate(date_col = ymd(date_col)) |>
#   group_by(floor_date(date_col, "month")) |>
#   summarise(n = n(), .groups = "drop") |>
#   ggplot(aes(`floor_date(date_col, "month")`, n)) +
#   geom_line() + geom_smooth(method = "loess")
cat("Time trend: adjust date_col to your date column\n")
```

## Model 2 — Seasonality / year-over-year

```{r seasonality}
# df |>
#   mutate(year = year(date_col), month = month(date_col, label = TRUE)) |>
#   group_by(year, month) |>
#   summarise(n = n(), .groups = "drop") |>
#   ggplot(aes(month, n, group = year, color = factor(year))) +
#   geom_line()
cat("Seasonality: adjust date_col above\n")
```


---

**→ [Hour 3: Communicate](h3_communicate_tennis_tournaments.qmd)**
