---
title: "Hour 1 — EDA: Analyzing US wind data in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 073 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`animated_transitions` · `boxplot_jitter` · `crossing_simulation_grid` · `spline_regression`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(gganimate)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)

us_wind <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018-11-06/us_wind.csv")
```
```{r chunk_02}
us_wind_processed <- us_wind |>
  filter(!t_state %in% c("AK", "HI", "GU", "PR")) |>
  mutate(t_cap = ifelse(t_cap < 0, NA, t_cap)) |>
  mutate_if(is.numeric, ~ ifelse(. == -9999, NA, .))

wind_projects <- us_wind_processed |>
  group_by(p_name, t_state) |>
  summarize(year = min(p_year, na.rm = TRUE),
            turbines = n(),
            total_capacity = sum(t_cap),
            lon = mean(xlong),
            lat = mean(ylat),
            lon_sd = sd(xlong),
            lat_sd = sd(ylat)) |>
  ungroup()
```
```{r chunk_03}
turbine_models <- us_wind_processed |>
  group_by(t_model) |>
  summarize(t_cap = median(t_cap),
            t_hh = median(t_hh),
            t_rd = median(t_rd),
            t_rsa = median(t_rsa),
            t_ttlh = median(t_ttlh),
            turbines = n(),
            projects = n_distinct(p_name)) |>
  arrange(desc(projects))

turbine_models |>
  ggplot(aes(t_ttlh, t_cap)) +
  geom_point() +
  labs(title = "When it comes to turbines, bigger is better!",
       x = "Turbine total height (meters)",
       y = "Turbine capacity (kW)")

turbine_models |>
  ggplot(aes(t_rsa, t_cap)) +
  geom_point() +
  labs(title = "When it comes to turbines, bigger is better!",
       x = "Turbine rotor swept area (meters ^ 2)",
       y = "Turbine capacity (kW)")
```
```{r chunk_04}
wind_projects |>
  ggplot(aes(year, total_capacity)) +
  geom_point()

wind_projects |>
  ggplot(aes(year, total_capacity / turbines)) +
  geom_point()
```
```{r chunk_05}
wind_projects |>
  ggplot(aes(lon, lat, size = turbines, color = year)) +
  borders("state") +
  geom_point() +
  coord_map() +
  theme_void()
```

## Animation
```{r chunk_06}
library(gganimate)

ggplot(mtcars, aes(factor(cyl), mpg)) + 
  geom_boxplot() + 
  # Here comes the gganimate code
  transition_states(
    gear,
    transition_length = 2,
    state_length = 1
  ) +
  enter_fade() + 
  exit_shrink() +
  ease_aes('sine-in-out')
```
```{r chunk_07}
wind_projects

wind_projects |>
  filter(!is.na(year), !is.infinite(year)) |>
  crossing(time = 1981:2018) |>
  filter(year <= time) |>
  ggplot(aes(lon, lat, size = turbines, color = year)) +
  borders("state") +
  geom_point() +
  transition_manual(time) +
  scale_color_continuous(guide = FALSE) +
  labs(title = "Locations of wind turbine projects in continental US (1981-2018)") +
  coord_map() +
  theme_void()

p
anim_save("turbines.gif")
```
```{r chunk_08}
us_wind_processed
```

---

**→ [Hour 2: Modeling](h2_modeling_wind.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
