---
title: "Hour 1 — EDA: Analyzing predicting wine ratings in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 053 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `boxplot_jitter` · `broom_model_output` · `coord_flip_horizontal` · `error_bar_coef_plot` · `faceted_comparison` · `fct_lump_infrequent` · `fct_relevel_manual` · `fct_reorder_stat` · `geom_histogram_density` · `geom_smooth_loess` · `grouped_summary_ggplot`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- text is multilingual without language detection
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- text is long-form English and benefits from word-level decomposition
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(broom)
library(tidytext)
library(widyr)
library(Matrix)
library(glmnet)
library(doMC)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
theme_set(theme_light())

wine_ratings <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-28/winemag-data-130k-v2.csv") |>
  select(-X1) |>
  extract(title, "year", "(20\\d\\d)", convert = TRUE, remove = FALSE) |>
  mutate(year = ifelse(year < 1900, NA, year)) |>
  filter(!is.na(price))
```
```{r chunk_02}
wine_ratings |>
  count(country, sort = T)

wine_ratings |>
  count(designation, sort = T)

wine_ratings |>
  count(country, region_1, sort = TRUE)

wine_ratings |>
  count(taster_name, sort = TRUE)

wine_ratings |>
  filter(!is.na(designation)) |>
  count(variety, designation, sort = TRUE)

wine_ratings |>
  ggplot(aes(year)) +
  geom_histogram()

wine_ratings |>
  ggplot(aes(points)) +
  geom_histogram(binwidth = 1)

wine_ratings |>
  ggplot(aes(price)) +
  geom_histogram() +
  scale_x_log10()
```
```{r chunk_03}
ggplot(wine_ratings, aes(price, points)) +
  geom_point(alpha = .1) +
  geom_smooth(method = "lm") +
  scale_x_log10()

summary(lm(points ~ log2(price), wine_ratings))
```
```{r chunk_04}
library(broom)

model <- wine_ratings |>
  replace_na(list(taster_name = "Missing", country = "Missing")) |>
  mutate(country = fct_relevel(fct_lump(country, 7), "US"),
         taster_name = fct_relevel(fct_lump(taster_name, 6), "Missing")) |>
  lm(points ~ log2(price) + country + year + taster_name, data = .)

model |>
  tidy(conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(term = str_replace(term, "country", "Country: "),
         term = str_replace(term, "taster_name", "Taster: "),
         term = fct_reorder(term, estimate)) |>
  ggplot(aes(estimate, term)) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high))
```
```{r chunk_05}
model |>
  augment(data = wine_ratings) |>
  ggplot(aes(.fitted, points)) +
  geom_point(alpha = .1)

tidy(anova(model)) |>
  mutate(sumsq / sum(sumsq))
```

## Lasso regression on words in description
```{r chunk_06}
library(tidytext)

wine_rating_words <- wine_ratings |>
  mutate(wine_id = row_number()) |>
  unnest_tokens(word, description) |>
  anti_join(stop_words, by = "word") |>
  filter(!word %in% c("wine", "drink"),
         str_detect(word, "[a-z]"))

wine_rating_words |>
  count(word, sort = TRUE) |>
  head(20) |>
  mutate(word = fct_reorder(word, n)) |>
  ggplot(aes(word, n)) +
  geom_col() +
  coord_flip()
```
```{r chunk_07}
library(widyr)

wine_words_filtered <- wine_rating_words |>
  distinct(wine_id, word) |>
  add_count(word) |>
  filter(n >= 100)

wine_words_filtered |>
  pairwise_cor(word, wine_id, sort = TRUE)
```
```{r chunk_08}
library(Matrix)

wine_word_matrix <- wine_words_filtered |>
  cast_sparse(wine_id, word)

wine_ids <- as.integer(rownames(wine_word_matrix))
scores <- wine_ratings$points[wine_ids]

library(glmnet)

wine_word_matrix_extra <- cbind(wine_word_matrix, log_price = log2(wine_ratings$price[wine_ids]))

library(doMC)
registerDoMC(cores = 4)

cv_glmnet_model <- cv.glmnet(wine_word_matrix_extra, scores, parallel = TRUE)

plot(cv_glmnet_model)
```
```{r chunk_09}
lexicon <- cv_glmnet_model$glmnet.fit |>
  tidy() |>
  filter(lambda == cv_glmnet_model$lambda.1se,
         term != "(Intercept)",
         term != "log_price") |>
  select(word = term, coefficient = estimate)

lexicon |>
  arrange(coefficient) |>
  group_by(direction = ifelse(coefficient < 0, "Negative", "Positive")) |>
  top_n(16, abs(coefficient)) |>
  ungroup() |>
  mutate(word = fct_reorder(word, coefficient)) |>
  ggplot(aes(word, coefficient, fill = direction)) +
  geom_col() +
  coord_flip() +
  labs(x = "",
       y = "Estimated effect of the word on the score",
       title = "What words are predictive of a wine's score?")
```
```{r chunk_10}
wine_ratings |>
  mutate(wine_id = row_number()) |>
  arrange(points) |>
  head(1) |>
  select(wine_id, description) |>
  pull(description)

wine_rating_words |>
  filter(wine_id %in% sample(unique(wine_id), 6)) |>
  distinct(word, title, points) |>
  mutate(wine = paste0(str_trunc(title, 40), " (", points, ")")) |>
  inner_join(lexicon, by = "word") |>
  mutate(word = fct_reorder(word, coefficient)) |>
  ggplot(aes(word, coefficient, fill = coefficient > 0)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ wine, scales = "free_y") +
  labs(title = "How a lasso regression would predict each wine's score",
       subtitle = "Using a lasso regression with an extra term for price",
       x = "",
       y = "Effect on score")
```

## What is glmnet?
```{r chunk_11}
cv_glmnet_model$glmnet.fit |>
  tidy() |>
  filter(term %in% c("rich", "black", "simple", "complex", "vineyard", "concentrated")) |>
  ggplot(aes(lambda, estimate, color = term)) +
  geom_line() +
  scale_x_log10() +
  geom_hline(lty = 2, yintercept = 0)

cv_glmnet_model$glmnet.fit |>
  tidy() |>
  count(lambda) |>
  ggplot(aes(lambda, n)) +
  geom_line() +
  scale_x_log10()
```
```{r chunk_12}
wine_ratings |>
  mutate(country = fct_relevel(fct_lump(country, 7), "US")) |>
  mutate(country = fct_reorder(country, points)) |>
  ggplot(aes(country, points)) +
  geom_boxplot() +
  coord_flip()

wine_ratings |>
  group_by(year) |>
  summarize(average_points = mean(points), n())
  
wine_ratings |>
  mutate(reviewer = fct_reorder(fct_lump(taster_name, 10), points)) |>
  ggplot(aes(reviewer, points)) +
  geom_boxplot() +
  coord_flip()
```

---

**→ [Hour 2: Modeling](h2_modeling_predicting_wine_ratings.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
