---
title: "Hour 1 — EDA: Analyzing tidytuesday and rstats tweets in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 067 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`coord_flip_horizontal` · `geom_histogram_density` · `grouped_summary_ggplot` · `label_repulsion_ggrepel` · `log_scale` · `lubridate_dates` · `multi_table_join` · `spline_regression` · `str_detect_filter` · `tf_idf_comparison` · `theme_set_global` · `time_series_trend`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- source HTML structure changes — selector-based scraping breaks silently
- text is multilingual without language detection
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- supplemental context is unavailable in the flat file and worth the scraping overhead
- text is long-form English and benefits from word-level decomposition
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(lubridate)
library(tidytext)
library(rvest)
library(ggrepel)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)

theme_set(theme_light())

tidytuesday_tweets <- read_rds(url("https://github.com/rfordatascience/tidytuesday/blob/master/data/2019/2019-01-01/tidytuesday_tweets.rds?raw=true")) |>
  mutate(week = as.Date(floor_date(created_at, "week", week_start = 1)))
```
```{r chunk_02}
library(lubridate)

week_summary <- tidytuesday_tweets |>
  group_by(week) |>
  summarize(tweets = n(),
            avg_retweets = exp(mean(log(retweet_count + 1))) - 1)

week_summary |>
  ggplot(aes(week, tweets)) +
  geom_line() +
  expand_limits(y = 0) +
  labs(x = "Time",
       y = "# of #tidytuesday tweets each week")

week_summary |>
  ggplot(aes(week, avg_retweets)) +
  geom_line() +
  expand_limits(y = 0) +
  labs(x = "Time",
       y = "Average (geometric mean) retweets each week")
```
```{r chunk_03}
tidytuesday_tweets |>
  count(screen_name, sort = TRUE) |>
  head(12) |>
  mutate(screen_name = reorder(screen_name, n)) |>
  ggplot(aes(screen_name, n)) +
  geom_col() +
  coord_flip()
```
```{r chunk_04}
tidytuesday_tweets |>
  arrange(desc(retweet_count)) |>
  select(screen_name, text, retweet_count)

tidytuesday_tweets |>
  group_by(screen_name) |>
  summarize(tweets = n(),
            retweets = sum(retweet_count)) |>
  arrange(desc(tweets)) |>
  arrange(desc(retweets))
```
```{r chunk_05}
tidytuesday_tweets |>
  select(screen_name, text, retweet_count, favorite_count) |>
  mutate(ratio = (favorite_count + 1) / (retweet_count + 1)) |>
  arrange(desc(ratio))
```
```{r chunk_06}
library(tidytext)

tweet_words <- tidytuesday_tweets |>
  select(screen_name, text, retweet_count, favorite_count, created_at, status_id,
         week) |>
  unnest_tokens(word, text, token = "tweets") |>
  anti_join(stop_words, by = "word") |>
  filter(!word %in% c("#tidytuesday", "de", "|"),
         str_detect(word, "[a-z]"))

tweet_words |>
  count(word, sort = TRUE) |>
  head(16) |>
  mutate(word = reorder(word, n)) |>
  ggplot(aes(word, n)) +
  geom_col() +
  coord_flip() +
  labs(title = "Most common words in #tidytuesday tweets",
       y = "Frequency of words")
```
```{r chunk_07}
tidytuesday_tweets |>
  ggplot(aes(favorite_count + 1)) +
  geom_histogram() +
  scale_x_log10()

word_summary <- tweet_words |>
  filter(!screen_name %in% c("thomas_mock", "R4DScommunity")) |>
  group_by(word) |>
  summarize(n = n(),
            avg_retweets = exp(mean(log(retweet_count + 1))) - 1,
            avg_favorites = exp(mean(log(favorite_count + 1))) - 1) |>
  filter(n >= 30) |>
  arrange(desc(avg_retweets))

tweet_words |>
  filter(word == "https://t.co/selb4fcv3u") |>
```
```{r chunk_08}
tidytuesday_tweets |>
  mutate(week = floor_date(created_at, "week", week_start = 1))

  ggplot(aes(date, n)) +
  geom_line() +
  expand_limits(y = 0) +
  labs(x = "Time",
       y = "# of #tidytuesday tweets")
```

## What topic is each week about?
```{r chunk_09}
top_word <- tweet_words |>
  count(word, week) |>
  bind_tf_idf(word, week, n) |>
  arrange(desc(tf_idf)) |>
  distinct(week, .keep_all = TRUE) |>
  arrange(week)

week_summary |>
  inner_join(top_word, by = c("week")) |>
  arrange(desc(avg_retweets)) |>
```
```{r chunk_10}
library(rvest)

week_titles <- read_html("https://github.com/rfordatascience/tidytuesday/tree/master/data/2018") |>
  html_node(".entry-content") |>
  html_node("table") |>
  html_table() |>
  tbl_df() |>
  transmute(week = floor_date(as.Date(Date), "week", week_start = 1),
            title = Data)
```
```{r chunk_11}
library(ggrepel)

week_summary |>
  inner_join(top_word, by = c("week")) |>
  inner_join(week_titles, by = "week") |>
  ggplot(aes(week, tweets)) +
  geom_line(color = "lightblue", size = 2) +
  geom_text(aes(label = title), check_overlap = TRUE) +
  expand_limits(y = 0) +
  labs(x = "Time",
       y = "# of #tidytuesday tweets each week",
       title = "# of tweets about each week's #tidytuesday",
       subtitle = "Shown is the word most specific to each week (based on TF-IDF)")
```

## #rstats hashtag
```{r chunk_12}
rstats_tweets <- read_rds(url("https://github.com/rfordatascience/tidytuesday/blob/master/data/2019/2019-01-01/rstats_tweets.rds?raw=true")) |>
  mutate(week = as.Date(floor_date(created_at, "week", week_start = 1)))
```
```{r chunk_13}
rstats_week_summary <- rstats_tweets |>
  group_by(week) |>
  summarize(tweets = n(),
            avg_retweets = exp(mean(log(retweet_count + 1))) - 1)

rstats_week_summary |>
  ggplot(aes(week, tweets)) +
  geom_line() +
  expand_limits(y = 0) +
  labs(x = "Time",
       y = "# of #tidytuesday tweets each week")

rstats_week_summary |>
  ggplot(aes(week, avg_retweets)) +
  geom_line() +
  expand_limits(y = 0) +
  labs(x = "Time",
       y = "Average retweets each week")

rstats_words <- rstats_tweets |>
  mutate(hashtags = str_count(text, "#[a-zA-Z]"), sort = TRUE) |>
  select(screen_name, text, retweet_count, favorite_count, created_at, status_id,
         week, hashtags) |>
  unnest_tokens(word, text, token = "tweets") |>
  anti_join(stop_words, by = "word") |>
  filter(!word %in% c("de", "|"),
         str_detect(word, "[a-z]"))
```
```{r chunk_14}
rstats_word_summary <- rstats_words |>
  filter(created_at >= "2018-01-01",
         hashtags < 6) |>
  group_by(word) |>
  summarize(n = n(),
            avg_retweets = exp(mean(log(retweet_count + 1))) - 1,
            avg_favorites = exp(mean(log(favorite_count + 1))) - 1)

rstats_word_summary |>
  filter(n >= 100,
         !str_detect(word, "https")) |>
  arrange(desc(avg_retweets)) |>
```
```{r chunk_15}
rstats_word_summary |>
  filter(n >= 100,
         !str_detect(word, "https")) |>
  ggplot(aes(n, avg_retweets)) +
  geom_point() +
  geom_text(aes(label = word), check_overlap = TRUE) +
  scale_x_log10() +
  scale_y_log10()

rstats_word_summary |>
  filter(n >= 100,
         !str_detect(word, "https")) |>
  arrange(desc(avg_retweets)) |>
  head(16) |>
  mutate(word = reorder(word, avg_retweets)) |>
  ggplot(aes(word, avg_retweets)) +
  geom_col() +
  coord_flip() +
  labs(title = "Which words get the most retweets in #rstats?",
       subtitle = "Only words appearing in at least 100 tweets in 2018",
       y = "Geometric mean of # of retweets")
```
```{r chunk_16}
rstats_tweets |>
  mutate(hashtags = str_count(text, "#[a-zA-Z]"), sort = TRUE) |>
  filter(hashtags < 6) |>
  filter(created_at >= "2018-01-01") |>
  group_by(screen_name) |>
  summarize(tweets = n(),
            avg_retweets = exp(mean(log(retweet_count + 1))) - 1) |>
  filter(tweets >= 30) |>
  arrange(desc(avg_retweets)) |>
```

---

**→ [Hour 2: Modeling](h2_modeling_tidytuesday_rstats_tweets.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
