---
title: "Hour 1 — EDA: Analyzing the Bechdel test in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 061 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `boxplot_jitter` · `faceted_comparison` · `fct_lump_infrequent` · `fct_relevel_manual` · `fct_reorder_stat` · `geom_histogram_density` · `grouped_summary_ggplot` · `label_repulsion_ggrepel` · `lasso_glmnet` · `linear_model_broom` · `log_scale`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- outcome column is missing or near-zero variance
- training set is too small for the number of predictors
- text is multilingual without language detection
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
- irregular time intervals inflate or deflate apparent trends
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- outcome column is clearly defined and has sufficient variance
- features can be engineered from existing columns without external joins
- text is long-form English and benefits from word-level decomposition
- time dimension is regular and spans multiple periods with meaningful change
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(ggrepel)
library(tidytext)
library(glmnet)
library(broom)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```
## Load data
```{r load-data}
movies <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-03-09/movies.csv", show_col_types = FALSE)
raw_bechdel <- read_csv("/Users/stoiana/sa-tidy-tuesday/tidytuesday/data/2021/2021-03-09/raw_bechdel.csv", show_col_types = FALSE)
glimpse(movies)
glimpse(raw_bechdel)
```

## Load the weekly Data
```{r Readme}
movies |>
  count(year, binary) |>
  ggplot(aes(year, n, fill = binary)) +
  geom_col()

summarize_bechdel <- function(tbl) {
  tbl |>
    summarize(n_movies = n(),
              pct_pass = mean(binary == "PASS"),
              median_intgross_2013 = median(intgross_2013, na.rm = TRUE)) |>
    arrange(desc(n_movies))
}

movies |>
  group_by(decade = 10 * (year %/% 10)) |>
  summarize_bechdel() |>
  ggplot(aes(decade, pct_pass)) +
  geom_line() +
  geom_point(aes(size = n_movies)) +
  expand_limits(y = 0)

movies |>
  arrange(desc(intgross_2013)) |>
  head(25) |>
  mutate(title = fct_reorder(title, intgross_2013)) |>
  ggplot(aes(intgross_2013, title, fill = binary)) +
  geom_col()

movies |>
  ggplot(aes(intgross_2013, binary)) +
  geom_boxplot() +
  scale_x_log10()

movies |>
  filter(year >= 1990) |>
  ggplot(aes(intgross_2013, fill = binary)) +
  geom_density(alpha = .5) +
  scale_x_log10()

lm(log2(intgross_2013) ~ binary, data = movies) |>
  summary()
```
```{r chunk_01}
library(ggrepel)

movie_genres <- movies |>
  separate_rows(genre, sep = ", ") |>
  filter(year >= 1990, !is.na(genre))

movie_genres |>
  group_by(genre) |>
  summarize_bechdel() |>
  head(12) |>
  ggplot(aes(median_intgross_2013, pct_pass)) +
  geom_point(aes(size = n_movies)) +
  geom_text_repel(aes(label = genre)) +
  expand_limits(x = 0, y = 0) +
  scale_y_continuous(labels = percent) +
  labs(x = "Median International Gross (2013 dollars)",
       y = "% that pass the Bechdel test")

movie_genres |>
  filter(fct_lump(genre, 9) != "Other") |>
  ggplot(aes(intgross_2013, fill = binary)) +
  geom_density(alpha = .5) +
  facet_wrap(~ genre) +
  scale_x_log10(labels = dollar) +
  labs(x = "International Gross (2013 dollars)")

movies |>
  filter(fct_lump(genre, 9) != "Other") |>
  ggplot(aes(intgross_2013, genre, fill = binary)) +
  geom_boxplot() +
  scale_x_log10(labels = dollar) +
  labs(x = "International Gross (2013 dollars)")

movies

movies |>
  filter(year >= 1990) |>
  separate_rows(genre, sep = ", ") |>
  mutate(value = 1) |>
  spread(genre, value, fill = 0) |>
  lm(log2(intgross_2013) ~ Action + Adventure + Crime +
       Thriller + `Sci-Fi` + Fantasy + Comedy +
       Drama + Romance + Horror +
       binary, data = .) |>
  summary()
```
```{r chunk_02}
genre_reasons <- movie_genres |>
  count(genre = fct_lump(genre, 8),
        clean_test,
        sort = TRUE) |>
  group_by(genre) |>
  mutate(pct = n / sum(n)) |>
  ungroup()

genre_reasons |>
  ggplot(aes(pct, clean_test)) +
  geom_col() +
  facet_wrap(~ genre) +
  scale_x_continuous(labels = percent) +
  labs(title = "Does the reason a movie fails the Bechdel test differ between genres?",
       y = "Test result",
       x = "% of movies in this genre")

library(tidytext)

genre_reasons |>
  filter(clean_test != "dubious") |>
  mutate(clean_test = fct_relevel(clean_test, "nowomen", "notalk", "men"),
         clean_test = fct_recode(clean_test,
                                 "Passes!" = "ok",
                                 "Has two women..." = "nowomen",
                                 "...who talk to each other..." = "notalk",
                                 "...about something other than a man" = "men")) |>
  mutate(genre = reorder_within(genre, pct, clean_test)) |>
  ggplot(aes(pct, genre)) +
  geom_col() +
  facet_wrap(~ clean_test, scales = "free_y") +
  scale_x_continuous(labels = percent) +
  scale_y_reordered() +
  labs(title = "Which step do movies fail on?",
       y = "",
       x = "% of movies in this genre that fail at this step?")
```
```{r chunk_03}
movies |>
  filter(!is.na(director)) |>
  group_by(director = fct_lump(director, 12)) |>
  summarize_bechdel() |>
  arrange(desc(pct_pass))

movies |>
  filter(director == "Steven Spielberg") |>
  arrange(desc(year)) |>

movies |>
  filter(!is.na(writer)) |>
  separate_rows(writer, sep = ", ") |>
  mutate(writer = str_remove(writer, " \\(.*")) |>
  distinct(title, writer, .keep_all = TRUE) |>
  group_by(writer = fct_lump(writer, 10)) |>
  summarize_bechdel() |>
```
```{r chunk_04}
movie_words <- movies |>
  filter(!is.na(plot)) |>
  unnest_tokens(word, plot) |>
  anti_join(stop_words, by = "word")

movie_words |>
  distinct(word, title, .keep_all = TRUE) |>
  group_by(word = fct_lump(word, 50)) |>
  summarize_bechdel() |>
  arrange(desc(pct_pass)) |>
  view()
```
```{r chunk_05}
library(glmnet)

movie_features <- bind_rows(
  movie_words |> distinct(imdb, title, binary, feature = word),
  movie_genres |> distinct(imdb, title, binary, feature = genre)
) |>
  add_count(feature) |>
  filter(n >= 10)

movie_feature_matrix <- movie_features |>
  cast_sparse(imdb, feature)

pass <- movies$binary[match(rownames(movie_feature_matrix), movies$imdb)] == "PASS"

mod <- cv.glmnet(movie_feature_matrix, pass, family = "binomial")

plot(mod)

library(broom)

tidy(mod$glmnet.fit) |>
  filter(lambda == mod$lambda.1se,
         term != "(Intercept)") |>
  top_n(40, abs(estimate)) |>
  mutate(term = fct_reorder(term, estimate)) |>
  ggplot(aes(estimate, term)) +
  geom_col() +
  labs(y = "Word/Genre",
       x = "Coefficient towards passing Bechdel test")
```

## Glimpse Data
```{r Glimpse}
tt |> 
  map(glimpse)
```

## Save Image
```{r chunk_06}
# This will save your most recent plot
ggsave(
  filename = "My TidyTuesday Plot.png",
  device = "png")
```

---

**→ [Hour 2: Modeling](h2_modeling_bechdel_test.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
