---
title: "Hour 1 — EDA: Analyzing Simpsons guest stars and dialogue in R"
subtitle: "David Robinson's analysis, extended | TheDataShrink™"
author: "TheDataShrink™"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: visual
---


::: {.callout-tip}
## TheDataShrink™ — Data stories worth sharing.
Episode 055 · Hour 1
:::


::: {.callout-note collapse="true"}
## Atomic Concepts in this episode
`add_count_inline` · `coord_flip_horizontal` · `faceted_comparison` · `fct_reorder_stat` · `geom_histogram_density` · `grouped_summary_ggplot` · `multi_table_join` · `parse_number_dirty` · `spline_regression` · `str_detect_filter` · `tf_idf_comparison` · `theme_set_global`
:::

::: {.callout-warning collapse="true"}
## FAILS WHEN
- dataset has no natural grouping variable — comparisons become meaningless
- column names are unstable across TidyTuesday editions (re-check schema each year)
- text is multilingual without language detection
- join key has trailing whitespace or case inconsistency — produces unexpected NAs
:::

::: {.callout-important collapse="true"}
## WORKS WHEN
- dataset has clear categorical + numeric columns for grouped comparisons
- question is exploratory — no strong prior hypothesis
- text is long-form English and benefits from word-level decomposition
:::

```{r setup}
library(here)
library(scales)
library(tidyverse)
library(tidytext)

# TheDataShrink™ brand colours
TDS_BLUE  <- "#1b6ca8"
TDS_GOLD  <- "#c8a415"
TDS_DARK  <- "#1a1a2e"
TDS_LIGHT <- "#f8f9fa"
theme_set(theme_light(base_size = 13))

```

## Explore
```{r chunk_01}
library(tidyverse)
theme_set(theme_light())

simpsons <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-08-27/simpsons-guests.csv") |>
  mutate(self = str_detect(role, "self|selves"),
         season = parse_number(season))
```
```{r chunk_02}
simpsons |>
  filter(self) |>
  count(guest_star, sort = TRUE) |>
  filter(n > 1) |>
  mutate(guest_star = fct_reorder(guest_star, n)) |>
  ggplot(aes(guest_star, n)) +
  geom_col() +
  coord_flip() +
  labs(title = "Who has played themselves in multiple Simpsons episodes?")
```
```{r chunk_03}
simpsons |>
  separate_rows(role, sep = ";\\s+") |>
  add_count(role) |>
  filter(n >= 8) |>
  count(season, role) |>
  mutate(role = fct_reorder(role, -n, sum)) |>
  ggplot(aes(season, n)) +
  geom_col() +
  facet_wrap(~ role)
```
```{r chunk_04}
simpsons
```

## Bringing in Simpsons dialogue
```{r chunk_05}
dialogue <- read_csv("~/Downloads/simpsons_dataset.csv") |>
  select(role = raw_character_text, line = spoken_words)

guests_processed <- simpsons |>
  separate_rows(role, sep = ";\\s+") |>
  mutate(role = ifelse(self, guest_star, role),
         role = ifelse(role == "Edna Krabappel", "Edna Krabappel-Flanders", role))

guests_summarized <- guests_processed |>
  filter(season <= 27) |>
  group_by(guest_star, role, self) |>
  summarize(nb_episodes = n(),
            first_season = min(season),
            last_season = max(season)) |>
  arrange(desc(nb_episodes)) |>
  group_by(role) |>
  filter(n() == 1) |>
  ungroup() |>
  filter(!is.na(role))

dialogue_summarized <- dialogue |>
  group_by(role) |>
  summarize(nb_lines = n(),
            random_line = sample(line, 1)) |>
  arrange(desc(nb_lines))

guest_roles <- guests_summarized |>
  inner_join(dialogue_summarized, by = "role") |>
  mutate(lines_per_episode = nb_lines / nb_episodes)

guest_roles |>
  mutate(self = ifelse(self, "Playing Themselves", "Playing a Character")) |>
  ggplot(aes(lines_per_episode)) +
  geom_histogram(binwidth = 2, center = 1) +
  facet_wrap(~ self, ncol = 1) +
  labs(x = "Average # of lines per episode",
       title = "Most guest stars, especially those playing themselves, have relatively few lines per episode")

guest_roles |>
  arrange(desc(lines_per_episode))
```
```{r chunk_06}
library(tidytext)

role_words <- dialogue |>
  filter(!is.na(line), !is.na(role)) |>
  mutate(line_number = row_number()) |>
  unnest_tokens(word, line) |>
  anti_join(stop_words, by = "word") |>
  distinct(role, line_number, word) |>
  count(role, word, sort = TRUE)

role_word_tf_idf <- role_words |>
  group_by(role) |>
  mutate(total_words = sum(n)) |>
  ungroup() |>
  bind_tf_idf(word, role, n) |>
  arrange(desc(tf_idf))

role_word_tf_idf |>
  filter(total_words >= 500) |>
  distinct(role, .keep_all = TRUE) |>
  mutate(role_word = paste0(role, ": ", word)) |>
  head(20) |>
  mutate(role_word = fct_reorder(role_word, tf_idf)) |>
  ggplot(aes(role_word, tf_idf)) +
  geom_col() +
  coord_flip() +
  labs(title = "Using TF-IDF as a Simpsons catchphrase detector",
       subtitle = "Only the 53 characters that speak at least 500 words in 27 seasons",
       x = "",
       y = "TF-IDF")
```
```{r chunk_07}
guests_summarized |>
  filter(nb_episodes > 1) |>
  inner_join(role_word_tf_idf, by = "role") |>
  filter(total_words >= 100) |>
  arrange(desc(tf_idf)) |>
  distinct(role, .keep_all = TRUE) |>
  select(guest_star, role, word, tf_idf)
```

---

**→ [Hour 2: Modeling](h2_modeling_simpsons_guest_stars_dialogue.qmd)**

*TheDataShrink™ — thedatashrink.com.au — Weekly data stories in R*
